### 2020-05-20 docs/disaggregation.tex
disaggregate FIES for survey countries

\documentclass[10pt]{article}         %% What type of document you're writing.

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

\begin{document}

\section*{This describes how to disaggregate national FIES rates by urban and rural areas}

For a given country, we have the following variables:

\begin{itemize}
 \item $\omega$: The percentage of households in urban areas in a country with food insecurity (UNKNOWN)

 \item $\rho$: The percentage of households in rural areas in a country with food insecurity (UNKNOWN)

 \item $t$: The overall percentage of households in a country with food insecurity (KNOWN)

 \item $RUR$: The total population in rural areas (KNOWN)

 \item $URB$: The total population in urban areas (KNOWN)

 \item $TOT$: The total population (KNOWN)

 \item $ratio$: The ratio of the rates of urban food insecurity to rural food insecurity.  This is known at a regional level, and it is estimated that the same ratio holds at the individual country level.
\end{itemize}

Lets start from the following two assumptions/equations:

\begin{equation} 
	\omega / \rho = ratio 
	\label{eq1}
\end{equation}

\begin{equation} 
	\omega * URB + \rho * RUR = t * TOT
	\label{eq2}
\end{equation}
\newline

Now we can derive the value of $\omega$ from Equation \ref{eq1}.

\begin{gather*} 
\omega = \rho * ratio                   
\end{gather*}

Substituting that value into Equation \ref{eq2} yields:

\begin{gather*} 
\rho * ratio * URB + \rho * RUR = t * TOT                
\end{gather*}

Solving for $\rho$:

\begin{align*}
\rho * ratio * URB + \rho * RUR &= t * TOT     \\
\rho * (ratio * URB + RUR) &= t * TOT \\
\rho &= \frac{t * TOT}{ratio * URB + RUR}
\end{align*}

Then, given $\rho$, we can solve for $\omega$ with Equation \ref{eq2}.

\end{document}

### 2020-05-20 lib/country_map.R

@@ -1,0 +1,31 @@
+country_map <- function(dat, iso3_col, var_col){
+  #Make a quick map
+  # data: A dataframe with the data
+  # iso3_col: the column with the country codes, in iso3 format
+  # var_col: a continuous variable to be mapped
+  
+  cty <- ne_countries(scale = 50, returnclass = 'sf')
+  
+  dat$iso_a3 <- dat[ , iso3_col]
+  dat$var <- dat[ , var_col]
+
+  if (sum(duplicated(dat$iso_a3)) > 0){
+    dat <- dat %>%
+      group_by(iso_a3) %>%
+      summarize(var = mean(var, na.rm=T))
+  }
+  
+  cty <- merge(cty, dat, all.x=T)
+  
+  plt <- ggplot(cty) + 
+    geom_sf(aes(fill=var)) + 
+    labs(fill=var_col) + 
+    scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", "#e4ea9a", "#fbf8c0", 
+                                   "#fce08a", "#faae61", "#f36c44", "#a01c44")) + 
+    coord_sf(crs='+proj=robin') + 
+    theme_void() + 
+    theme(legend.position = 'bottom',
+          plot.title = element_text(hjust = 0.5))
+  
+  return(plt)
+}

### 2020-05-20 munge/01-Combine Surveys.R

@@ -1,0 +1,26 @@
+fs <- list.files('data/fies_surveys', full.names = T)
+
+fies_raw <- data.frame()
+for (f in fs){
+  load(f)
+  data$iso3 <- substr(f, 19, 21)
+  data$year <- as.numeric(substr(f, 23, 26))
+
+  fies_raw <- bind_rows(fies_raw, data)  
+}
+
+fies_raw <- fies_raw %>%
+  #Convert data to numeric
+  mutate(WORRIED = as.numeric(WORRIED),
+         HEALTHY = as.numeric(HEALTHY),
+         FEWFOOD = as.numeric(FEWFOOD),
+         SKIPPED = as.numeric(SKIPPED),
+         ATELESS = as.numeric(ATELESS),
+         RUNOUT = as.numeric(RUNOUT),
+         HUNGRY = as.numeric(HUNGRY),
+         WHLDAY = as.numeric(WHLDAY),
+         #Looks like cambodia was mis-coded?
+         iso3 = case_when(iso3=='KMH' ~ 'KHM',
+                             TRUE ~ iso3))
+
+cache('fies_raw')

### 2020-05-20 munge/02-Derive FIES.R

@@ -1,0 +1,42 @@
+#Load comparison data from FAO
+compare <- FAOSTAT.data.4.14.2020 %>%
+  select(Area, Year, Value, Item) %>%
+  transmute(country = Area,
+            year = as.numeric(substr(Year, 1, 4)) + 1,
+            value = as.numeric(Value)/100,
+            item = case_when(grepl('moderate', Item) ~ 'fies.mod.3yr.fao',
+                             TRUE ~ 'fies.sev.3yr.fao')) %>%
+  spread(item, value) %>%
+  mutate(iso3 = countrycode(country, 'country.name', 'iso3c')) %>%
+  select(-country) %>%
+  na.omit
+
+#Aggregate raw data to country level 
+fies_cty <- fies_raw %>% 
+  group_by(iso3, year) %>% 
+  summarize(fies.mod = weighted.mean(Prob_Mod_Sev > 0.5, w = wt, na.rm=T),
+            fies.sev = weighted.mean(Prob_sev > 0.5, w = wt, na.rm=T)) %>%
+  mutate(fies.mod.3yr = rollapply(fies.mod, width=3, FUN=mean, na.pad=T),
+         fies.mod.3yr = rollapply(fies.sev, width=3, FUN=mean, na.pad=T)) %>%
+  merge(compare, all.x=T, all.y=F)
+
+#Disaggregate by urban-rural
+fies_ur <- fies_raw %>% 
+  merge(regions, all.x=T, all.y=F) %>%
+  filter(!Area %in% c('(Refused)', 'Dont_know'),
+         !is.na(Area)) %>%
+  group_by(region, Area) %>% 
+  summarize(fies.reg.mod = weighted.mean(Prob_Mod_Sev > 0.5, w = wt, na.rm=T),
+            fies.reg.sev = weighted.mean(Prob_sev > 0.5, w = wt, na.rm=T)) %>%
+  mutate(Area = case_when(grepl('Urban', Area) ~ '.urb',
+                          TRUE ~ '.rur')) %>%
+  gather(key, value, -region, -Area) %>%
+  mutate(key=paste0(key, Area),
+         Area=NULL) %>%
+  spread(key, value) %>%
+  mutate(ur.ratio.sev = fies.reg.sev.urb/fies.reg.sev.rur,
+         ur.ratio.mod = fies.reg.mod.urb/fies.reg.mod.rur)
+
+cache('fies_cty')
+cache('fies_ur')
+cache('regions')

### 2020-05-20 viz/explore_country_trends.R

@@ -1,0 +1,17 @@
+fies_trends <- merge(
+  fies_cty %>%
+    group_by(iso3) %>%
+    do(tidy(lm(fies.mod ~ year, .))) %>% 
+    filter(!is.nan(estimate),
+           term == 'year') %>%
+    select(iso3, mod.coef=estimate, mod.pval=p.value),
+  fies_cty %>%
+    group_by(iso3) %>%
+    do(tidy(lm(fies.sev ~ year, .))) %>% 
+    filter(!is.nan(estimate),
+           term == 'year') %>%
+    select(iso3, sev.coef=estimate, sev.pval=p.value)
+  )
+
+country_map(fies_trends, 'iso3', 'mod.coef')
+country_map(fies_trends, 'iso3', 'sev.coef')

### 2020-05-20 viz/map_regions.R

@@ -1,0 +1,15 @@
+cty <- ne_countries(scale = 50, returnclass = 'sf')
+
+regions$iso_a3 <- regions$iso3
+
+cty <- merge(cty, regions, all.x=T)
+
+plt <- ggplot(cty) + 
+  geom_sf(aes(fill=region)) + 
+  coord_sf(crs='+proj=robin') + 
+  theme_void() + 
+  theme(legend.position = 'bottom',
+        plot.title = element_text(hjust = 0.5))
+plt
+
+ggsave(plot = plt, filename = 'figures/regions.png', width=12, height=7)

### 2020-05-20 viz/subindicator_maps.R

@@ -1,0 +1,34 @@
+library(ProjectTemplate)
+setwd('C://Users/matt/wdl-fies')
+load.project()
+
+subindicators <- 
+  data.frame(
+    matrix(
+      c('WORRIED', 'During the last 12 MONTHS, was there a time when\nyou were worried you would not have enough food to eat\nbecause of a lack of money or other resources?', 
+        'HEALTHY', 'Still thinking about the last 12 MONTHS, was there\na time when you were unable to eat healthy and nutritious\nfood because of a lack of money or other resources?', 
+        'FEWFOOD', 'Was there a time when you ate only a few kinds of\nfoods because of a lack of money or other resources?', 
+        'SKIPPED', 'Was there a time when you had to skip a meal because\nthere was not enough money or other resources to get food?', 
+        'ATELESS', 'Still thinking about the last 12 MONTHS, was there a\ntime when you ate less than you thought you should\nbecause of a lack of money or other resources?', 
+        'RUNOUT', 'Was there a time when your household ran out of food\nbecause of a lack of money or other resources?', 
+        'HUNGRY', 'Was there a time when you were hungry but did not eat\nbecause there was not enough money or other\nresources for food?',
+        'WHLDAY', 'During the last 12 MONTHS, was there a time when you\nwent without eating for a whole day because of\na lack of money or other resources?'
+      ),
+      byrow=T,
+      ncol = 2
+    )
+  )
+
+names(subindicators) <- c('code', 'question')
+
+for (i in 1:nrow(subindicators)){
+  res <- country_map(fies_raw, 'country', subindicators$code[i])
+  
+  res <- res + 
+    labs(title = subindicators$question[i])
+
+  ggsave(plot=res, filename=paste0('figures/subindicator - ', subindicators$code[i], '.png'), width = 12, height=7)  
+}
+
+
+

### 2020-05-20 viz/subnational_maps.R

@@ -1,0 +1,30 @@
+gdl_fies <- merge(gdl,
+                  fies_subnat %>%
+                    group_by(GDLCODE) %>%
+                    summarize_if(is.numeric, mean))
+
+ggplot() + 
+  geom_sf(data=gdl, fill = '#999999', size=0.1) + 
+  geom_sf(data=gdl_fies, aes(fill=fies.mod), size=0.1) +
+  coord_sf(crs='+proj=robin') + 
+  theme_void() + 
+  scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", "#e4ea9a", "#fbf8c0", 
+                                 "#fce08a", "#faae61", "#f36c44", "#a01c44")) + 
+  theme(legend.position = 'bottom',
+        plot.title = element_text(hjust = 0.5)) + 
+  labs(fill='',
+       title='Subnational Rate of Moderate Food Insecurity\n(Averaged over all years)')
+ggsave('figures/subnational-moderate.png', width=12, height=7)
+
+ggplot() + 
+  geom_sf(data=gdl, fill = '#999999', size=0.1) + 
+  geom_sf(data=gdl_fies, aes(fill=fies.sev), size=0.1) +
+  coord_sf(crs='+proj=robin') + 
+  theme_void() + 
+  scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", "#e4ea9a", "#fbf8c0", 
+                                 "#fce08a", "#faae61", "#f36c44", "#a01c44")) + 
+  theme(legend.position = 'bottom',
+        plot.title = element_text(hjust = 0.5)) + 
+  labs(fill='',
+       title='Subnational Rate of Severe Food Insecurity\n(Averaged over all years)')
+ggsave('figures/subnational-severe.png', width=12, height=7)

### 2020-05-20 munge/02-Derive FIES.R
compare rates with FAO

@@ -17,7 +17,7 @@ fies_cty <- fies_raw %>%
   summarize(fies.mod = weighted.mean(Prob_Mod_Sev > 0.5, w = wt, na.rm=T),
             fies.sev = weighted.mean(Prob_sev > 0.5, w = wt, na.rm=T)) %>%
   mutate(fies.mod.3yr = rollapply(fies.mod, width=3, FUN=mean, na.pad=T),
-         fies.mod.3yr = rollapply(fies.sev, width=3, FUN=mean, na.pad=T)) %>%
+         fies.sev.3yr = rollapply(fies.sev, width=3, FUN=mean, na.pad=T)) %>%
   merge(compare, all.x=T, all.y=F)
 
 #Disaggregate by urban-rural

### 2020-05-20 viz/compare_fao_rates.R

@@ -1,0 +1,13 @@
+ggplot(fies_cty) + 
+  geom_point(aes(x=fies.sev.3yr.fao,
+                 y=fies.sev.3yr)) + 
+  labs(x='Published Estimates of Severe Food Insecurity',
+       y='Manually Calculated Estimates of Severe Food Insecurity')
+ggsave('figures/FAO Compare - Severe.png')
+
+ggplot(fies_cty) + 
+  geom_point(aes(x=fies.mod.3yr.fao,
+                 y=fies.mod.3yr)) + 
+  labs(x='Published Estimates of Moderate Food Insecurity',
+       y='Manually Calculated Estimates of Moderate Food Insecurity')
+ggsave('figures/FAO Compare - Moderate.png')

### 2020-06-03 wdl-fies_load.project.R
BMS

@@ -9,5 +9,5 @@ setwd("C:/Users/bmuel/Desktop/WDL/GitHub/wdl-fies")
 
 
 #migrate.project()
-load.project(list(data_loading=TRUE, munging=TRUE))
-#load.project()
+#load.project(list(data_loading=TRUE, munging=TRUE))
+load.project()

### 2020-06-04 viz/regions_bar.R
make urban-rural bar chart

@@ -1,0 +1,15 @@
+dat <- fies_ur %>%
+	select(region, fies.reg.mod.rur, fies.reg.mod.urb) %>%
+	gather(key, val, -region) %>%
+	mutate(key=ifelse(grepl('rur', key), 'Rural', 'Urban'))
+
+ggplot(dat) +
+	geom_bar(aes(x=region, y=val, fill=key), stat='identity', position='dodge') + 
+	labs(y='Rate of Food Insecurity',
+			 x='',
+			 fill='') + 
+	coord_flip() + 
+	theme_bw() + 
+	theme(legend.position=c(0.75, 0.25))
+
+ggsave('figures/Urban_Rural.png', height=4, width=5)

### 2020-06-04 models/model_allcovars.R
remove near-idenitcal predictors

@@ -16,9 +16,13 @@ preddat <- covar_ext %>%
 	data.frame
 
 # Set up Model
-vars <- names(moddat)[!names(moddat) %in% c('iso3', 'GDLCODE', 'fies.mod.rur', 'fies.sev.rur', 
-																						'fies.mod.urb', 'fies.sev.urb', 'Urban', 'Rural', 'fies.mod', 
-																						'fies.sev', 'rural', 'urban', 'year')]
+vars <- names(moddat)[!names(moddat) %in% c('iso3', 'GDLCODE', 'fies.mod.rur', 
+																						'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb', 
+																						'Urban', 'Rural', 'fies.mod', 
+																						'fies.sev', 'rural', 'urban', 'year',
+																						#Exclue a few more that are collinear
+																						'nutritiondiversity_s', 'nutritiondiversity_h',
+																						'irrig_aei', 'high_settle', 'low_settle')]
 
 x <- model.matrix(as.formula(paste0('fies.mod ~ ', paste0(vars, collapse=' + '))), data=moddat)
 

### 2020-06-04 models/model_SSP.R
cleaned up model outputs

@@ -67,9 +67,13 @@ ggsave('figures/SSP2_LASSO.png', width=7, height=12)
 
 # Assess residuals
 mae <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred))
+r2 <- cor(moddat$fies.mod, moddat$fies.mod.pred)
 ggplot(moddat) + 
 	geom_point(aes(x=fies.mod, y=fies.mod.pred)) + 
-	labs(caption=paste0('MAE: ',  mae)) 
+	labs(caption=paste0('Mean Absolute Error: ',  round(mae, 4),
+											'\nR-squared: ', round(r2, 4)),
+			x='Observed Rates of Food Insecurity',
+			y='Modeled Rate of Food Insecurity')	
 ggsave('figures/SSP2_Residuals.png', width=5, height=5)
 
 # Plot scaled covariates
@@ -86,7 +90,8 @@ df$term <- factor(df$term, levels=df$term[order(df$scaled)], ordered=TRUE)
 ggplot(df %>% filter(term != '(Intercept)')) + 
 	  geom_bar(aes(x=term, y=scaled), stat='identity') + 
 		coord_flip() + 
-		labs(title='Change in Rate of Food Insecurity\nWith increase of 1 SD in Var\nFor SSP2 LASSO Regression Model') + 
+		labs(title='Change in Rate of Food Insecurity\nWith increase of 1 SD in Var\nFor SSP2 LASSO Regression Model',
+				 x="", y="") + 
 		theme_minimal()
 
 ggsave('figures/SSP2_Coefs.png', width=5, height=5)

### 2020-06-04 models/model_allcovars.R

@@ -71,9 +71,13 @@ ggsave('figures/Pred2020_LASSO.png', width=7, height=5)
 
 # Assess residuals
 mae <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred))
+r2 <- cor(moddat$fies.mod, moddat$fies.mod.pred)
 ggplot(moddat) + 
 	geom_point(aes(x=fies.mod, y=fies.mod.pred)) + 
-	labs(caption=paste0('MAE: ',  mae)) 
+	labs(caption=paste0('Mean Absolute Error: ',  round(mae, 4), 
+											'\nR-squared: ', round(r2, 4)),
+			 x='Observed Rates of Food Insecurity',
+			 y='Modeled Rate of Food Insecurity') 
 ggsave('figures/Pred2020_Residuals.png', width=5, height=5)
 
 # Plot scaled covariates
@@ -90,7 +94,9 @@ df$term <- factor(df$term, levels=df$term[order(df$scaled)], ordered=TRUE)
 ggplot(df %>% filter(term != '(Intercept)')) + 
 	  geom_bar(aes(x=term, y=scaled), stat='identity') + 
 		coord_flip() + 
-		labs(title='Change in Rate of Food Insecurity\nWith increase of 1 SD in Variable\nFor 2020 LASSO Regression Model') + 
+		labs(title='Change in Rate of Food Insecurity\nWith increase of 1 SD in Variable\nFor 2020 LASSO Regression Model', 
+				 x="", 
+				 y="") + 
 		theme_minimal()
 
 ggsave('figures/Pred2020_Coefs.png', width=5, height=5)

### 2020-06-04 docs/modeling_fies.tex
First draft of manuscript

\documentclass{article}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[a4paper, margin=1.25in]{geometry}
\usepackage{breakcites}
\usepackage{subcaption}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{longtable}

\begin{document}

\title{Modeling the Food Insecurity Experience Scale (FIES) in the Present and Future}

\author[1]{Matthew Cooper}
\author[2]{Benjamin Müller}

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{TU Wien}

\maketitle
\begin{abstract}
The Food Insecurity Experience Scale (FIES) measures the experiential aspects of food insecurity, and has been selected as a target for the second Sustainable Development Goal (SDG) of Zero Hunger.  Here, we present the preliminary results of our work modeling this indicator globally for the years 2020-2030.

\end{abstract}

\section{Introduction}

\subsection{Measuring Food Security and the Genesis of the FIES}
Food security is a key part of human development, but has traditionally been difficult to measure.  Metrics of macro-health, such as anthropometry and mortality rates are correlated broadly with food insecurity \cite{Puffer1973, Habicht1974}, but are confounded with other determinants of health such as infectious disease.  Other proxies for food security, such as food availability estimated from crop yields \cite{Maxwell1992}, are inadequate because they do not indicate how accessible food is to the general population, and food insecurity can certainly occur in the absence of food availability decline \cite{Sen1983}.

As researchers began to focus on food insecurity at the individual and household level, household microdata collecting information on household finances and consumption became a common proxy for food security \cite{Haddad1994}.  However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \cite{Maxwell1996}.  This led to the emergence of several indicators designed to be rapidly deployable, culturally cross-comparable, and based on the lived experience of food security.  These metrics include the Household Food Insecurity and Access Scale (HFIAS) \cite{Coates2007}; the Coping Strategies Index (CSI) \cite{Maxwell1999}; the Household Hunger Scale (HHS) \cite{Ballard2011}; the Food Consumption Score (FCS); and the Household Dietary Diversity Scale (HDDS) \cite{Kennedy2010}.

Drawing on the insights derived in designing and implementing these food security metrics, the Food Insecurity Experience Scale (FIES) was developed by the Food and Agricultural Organization (FAO) of the UN.

\subsection{The FIES}
The FIES asks participants eight questions about their experience of food insecurity during the previous year \cite{Cafiero2018}, collected in countries around the world both in person and over the telephone with the assistance of Gallup World Poll.  These questions are:

\begin{enumerate}
	\item During the last 12 MONTHS, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
	\item Still thinking about the last 12 MONTHS, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
	\item Was there a time when your household ran out of food because of a lack of money or other resources?
	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
\end{enumerate}

Based on the responses to these questions, a Rasch model is used to generate the score for the population \cite{Engelhard2013}.

\section{Data}
For our analysis, we draw on several datasets.  First, we use the individual responses from 335,139 individuals from 331 surveys across 77 countries, downloaded from the FAO microdata catalogue.  Additionally, we use country-level FIES scores downloaded from FAOSTAT.

To disaggregate the FIES by urban-rural areas, we used data on urbanization from \cite{Jiang2017}.

\subsection{Covariates Available in the Present}
We use a large stack of covariates available for the present or the near-present, reflective of factors related to many aspects of food systems.  Datasets that were originally gridded were aggregated to the Admin-1 level taking the mean value.  A full table of datasets is given in Table \ref{tab:covars1}.

\begin{table}[H]
	\begin{tabular}{llll}
		\toprule
		Name & Availability & Scale & Source \\
		\midrule
		Agriculture As A Percentage Of GDP & Annual & National & \cite{TheWorldBank2016} \\
		Net Oda Per Capita & Annual & National & \cite{TheWorldBank2016} \\
		Percent of Area Bare Ground & Annual & Admin-1 & \cite{Song2018} \\
		Percent Of Area With Built Up Land Cover & 2000 \& 2014 & Admin-1 & \cite{Pesaresi2015} \\
		Staple Crop Proction Per Capita & Annual & National & \cite{FAOSTAT2018} \\
		Elevation & Static & Admin-1 & \cite{USGS1996} \\
		Net School Enrollment & Annual & National & \cite{TheWorldBank2016} \\
		Forest Cover & Annual & Admin-1 & \cite{Song2018} \\
		Government Effectiveness Indicator & Annual & National & \cite{Kaufmann2011} \\
		GDP PPP Per Capita & Annual & Admin-1 & \cite{Kummu2018} \\
		HDI & Annual & Admin-1 & \cite{Kummu2018} \\
		Value Of All Imports Per Capita & Annual & National & \cite{TheWorldBank2016} \\
		Irrigated Area & 2000 & Admin-1 & \cite{SiebertS.DollP.FeickS.FrenkenK.Hoogeveen2013} \\
		Time To Travel To A City  & 2015 & Admin-1 & \cite{Uchida2008} \\
		Mean Annual Precipitation & Static & Admin-1 & \cite{Funk2015} \\
		NDVIi & Annual & Admin-1 & \cite{Song2018} \\
		People Per Pixel & 2010, 2015, 2020 & Admin-1 & \cite{Doxsey-Whitfield2015} \\
		Topographic Roughness Index & Static & Admin-1 & \cite{USGS1996, Riley1999} \\
		Political Stability Index & Annual & Admin-1 & \cite{Kaufmann2011} \\
		Number Of Conflict Deaths Within 50Km & Annual & Admin-1 & \cite{Eriksson2015} \\
		Nutritional Diversity Of Agriculture & 2015 & Admin-1 & \cite{Herrero2017a} \\
		Average Maximum Temperature & Roughly Static & Admin-1 & \cite{Sheffield2006} \\
		Number Of Buffaloes Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
		Number Of Cattle Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
		Number Of Chickens Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
		Number Of Ducks Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
		Number Of Horses Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
		Number Of Pigs Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
		Number Of Sheep Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
		Incidence \textit{P. falciparum} Malaria & Annual & Admin-1 & \cite{Weiss2019} \\
		Incidence \textit{P. vivax} Malaria & Annual & Admin-1 & \cite{Weiss2019} \\
		\bottomrule
	\end{tabular}
	\caption{Co-Variates Included in Present-Day Analysis}
	\label{tab:covars1}
\end{table}

\subsection{Covariates Available in the Present and Future}
To model the FIES in the future, we use projections derived in line with the Shared Socio-Economic Pathways (SSPs).  The SSPs are a framework to estimate what the future might look like in terms of human development and climate-change mitigation, based on a number of scenarios.  We use projections of GDP \cite{Dellink2017}, education \cite{KC2017}, and urbanization \cite{Jiang2017} for SSP2, the middle-of the road pathway.

\section{Methods \& Results}
\subsection{Disaggregating the FIES Subnationally}
We begin by using the probability an individual has moderate-to-severe food insecurity for the individual-level survey data derive the FIES at the country level.  Using the sampling weights reported by the FAO, we found the percentage of people in each country with a greater than 50\% chance of being moderate or severely food insecure. We then checked our own results against those reported by FAOSTAT and found similar, but not identical, estimates (See Figure \ref{fig:comparison}).

\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\linewidth]{../figures/"FAO Compare - Moderate".png}
	\caption{Comparison of derived and FAO reported estimates of food insecurity by FIES Score}
	\label{fig:comparison}
\end{figure}

The FIES surveys include information on whether respondents were located in urban or rural areas.  However, due to the small numbers of individuals in urban areas within each country (about 200 observations), we group countries into regions and estimate regional prevalences of food insecurity for both rural and urban areas.  In our grouping, we use the following regions: Eastern Europe, Latin American and the Caribbean (LAC), Southeast Asia, West Africa, Southern Africa, the Mideast and Central Asia, and High-Income Countries (HICs) (See Figure \ref{fig:regions}).

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{../figures/regions.png}
	\caption{Grouping used to estimate regional rates of food insecurity in urban and rural areas.}
	\label{fig:regions}
\end{figure}

\begin{figure}[h]
	\centering
	\includegraphics[width=0.75\linewidth]{../figures/Urban_Rural.png}
	\caption{Rates of food insecurity in urban and rural areas by world region.}
	\label{fig:regionrate}
\end{figure}

Examining rates of food insecurity by both urban and rural areas shows that, in nearly every world region, food insecurity is worse in rural areas, with the largest gap in Southern Africa.  High-income countries area the only exception, and food insecurity is actually worse in urban parts of these countries than in rural areas.

Given the overall rate of food insecurity in each country, the disparities in rates of urban and rural food insecurity in each world region, and the percentage of the population and urban and rural areas within each admin-1 area of each country, it is possible to derive a rate of food insecurity specific to each admin-1 area.

Formally, given the following two assumptions:

\begin{equation} 
\omega / \rho = ratio 
\label{eq1}
\end{equation}

\begin{equation} 
\omega * URB + \rho * RUR = t * TOT
\label{eq2}
\end{equation}
\newline

Where:

\begin{itemize}
	\item $\omega$: The percentage of households in urban areas in a country with food insecurity (UNKNOWN)

	\item $\rho$: The percentage of households in rural areas in a country with food insecurity (UNKNOWN)

	\item $t$: The overall percentage of households in a country with food insecurity (KNOWN)

	\item $RUR$: The total population in rural areas (KNOWN)

	\item $URB$: The total population in urban areas (KNOWN)

	\item $TOT$: The total population (KNOWN)

	\item $ratio$: The ratio of the rates of urban food insecurity to rural food insecurity.  This is known at a regional level, and it is estimated that the same ratio holds at the individual country level.
\end{itemize}

We can solve for the two unknown parameters, $\omega$ and $\rho$ to get country-level estimates of urban and rural rates of food insecurity and derive subnational prevalences.

Thus, we derive the following map of disaggregated rates of food insecurity in countries where we have direct FIES data (See Figure \ref{fig:disaggregated}).

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{../figures/subnational-moderate.png}
	\caption{Rates of food insecurity in urban and rural areas by world region, averaged across all years for which survey data is available}
	\label{fig:disaggregated}
\end{figure}

\subsection{Modeling the FIES in the Present}
We then used these Admin-1 estimates, in combination with a large stack of covariates in the present, to model the prevalence of food insecurity globally for the year 2020, based on covariates for as close to the year 2020 as possible.  We used a LASSO regression approach, where the size of the regression coefficients are penalized, which is most appropriate for making predictions based on a large number of covariates \cite{Tibshirani2011}.

For our in-sample data, the LASSO performed well (see Figure \ref{fig:lasso_residuals}).  The coefficients for each variable are given in Figure \ref{fig:lasso_coefs}.  Note that some coefficients are penalized to 0, which effectively removes them from the model.  The map is the results is given in Figure \ref{fig:lasso_map}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\linewidth]{../figures/Pred2020_Residuals.png}
	\caption{Correlation between observed disaggregated rates of food insecurity and modeled rates of food insecurity.}
	\label{fig:lasso_residuals}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\linewidth]{../figures/Pred2020_Coefs.png}
	\caption{Change in rate of food insecurity with a 1 standard deviation increase in each variable.  NOTE: IN A FUTURE EDITION THE LABELS WILL BE BETTER.}
	\label{fig:lasso_coefs}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{../figures/Pred2020_LASSO.png}
	\caption{Estimated rates of food insecurity in 2020. NOTE: SOME COVARIATES ARE MISSING AT HIGH LATITUDES, THIS WILL BE FIXED IN A FUTURE EDITION.}
	\label{fig:lasso_map}
\end{figure}

\subsection{Modeling the FIES in the Future}
In addition to modeling the FIES in the present, we also modeled the FIES for the years 2020, 2025, and 2030, based covariates that were already extrapolated to the year 2030 in peer-reviewed literature.  NOTE: MANY MORE COVARIATES ARE BEING ADDED.  Similar to the previous model, we below show the correlation between the observed and predicted values (Figure \ref{fig:ssp_residuals}), the estimated coefficients (Figure \ref{fig:ssp_coefs}), and the mapped results (Figure \ref{fig:ssp_map}).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\linewidth]{../figures/SSP2_Residuals.png}
	\caption{Correlation between observed disaggregated rates of food insecurity and modeled rates of food insecurity.}
	\label{fig:ssp_residuals}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\linewidth]{../figures/SSP2_Coefs.png}
	\caption{Change in rate of food insecurity with a 1 standard deviation increase in each variable.  NOTE: IN A FUTURE EDITION THE LABELS WILL BE BETTER AND MORE COVARIATES WILL BE INCLUDED.}
	\label{fig:ssp_coefs}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{../figures/SSP2_LASSO.png}
	\caption{Estimated rates of food insecurity in 2020, 2025, and 2030.}
	\label{fig:ssp_map}
\end{figure}

\section{Discussion}
There is still a wide gap between the predictions of the 2020 model based on the large stack of covariates and the 2020 model based on just the SSP covariates.

\section{Conclusion}

\clearpage
\bibliographystyle{apalike}

%Determine local environment
\bibliography{C://Users/matt/library}

\section*{Supplemental Info}

\end{document}

### 2020-09-02 models/model_bayes_ssvs.R
(prelim) final things

@@ -1,5 +1,5 @@
 #######################################################
-# Use BMA with SSP covariates
+# Use Bayes SSVS with SSP covariates
 # to estimate food insecurity, based on data for
 # modeled for 2020, 2025, and 2030
 #######################################################
@@ -39,11 +39,7 @@ exc <- c()
 #exc <- c('mal_falciparum', 'mal_vivax', 'wasting', 'crop_prod')
 exc <- c('crops_prod', 'forest', 'builtup', 'livestock',
          'pasture', 'crops_prod', 'cropland',
-<<<<<<< HEAD
          'mal_vivax', 'stunting', 'mal_falciparum')
-=======
-         'mal_falciparum', 'mal_vivax', 'stunting')
->>>>>>> 4bd1c7a1266f2df0de493859838def5bb1e8699d
 
 # Set up Model
 vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
@@ -669,7 +665,7 @@ ggplot(sdf %>% filter(term != '(Intercept)')) +
 ggsave('figures/bayes_SSVS/SSP2_Sev_Coefs_bayes_ssvs_wt.png', width=5, height=5)
 
 
-<<<<<<< HEAD
+
 ############################
 # IFAD call
 ############################

### 2020-09-02 models/model_randomforest.R

@@ -10,7 +10,7 @@ covars <- covars %>% merge(ne_countries()@data %>%
                            select(ISO3=iso_a3, region=region_wb))
 
 ne_countries()@data %>%
-  select(ISO3=iso_a3, region=wb_regions)
+  select(ISO3=iso_a3, region=region_wb)
 
 #Just to fix Siberia
 covars$wci_index[covars$GDLCODE=='RUSr108'] <- 1
@@ -35,11 +35,12 @@ moddat <- merge(fies_subnat,
 preddat <- covars %>%
   data.frame
 
+preddat <- na.omit(preddat); summary(preddat)
+
 exc <- c()
 #exc <- c('mal_falciparum', 'mal_vivax', 'wasting', 'crop_prod')
-exc <- c('crops_prod', 'forest', 'builtup', 'livestock',
-         'pasture', 'crops_prod', 'cropland',
-         'mal_vivax', 'stunting', 'mal_falciparum')
+exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
+         'pasture', 'cropland', 'mal_vivax')
 
 # Set up Model
 vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
@@ -53,23 +54,73 @@ moddat <- moddat %>%
   mutate(fies.mod = logit(fies.mod),
          fies.sev = logit(fies.sev))
 
-rf.mod <- randomForest(fies.mod ~ wasting + school_mean + ruggedness + gdp_percap + 
-                       gini + mean_annual_precip + hci + tavg + urban_perc + 
-                       wci_index, data=moddat)
-rf.sev <- randomForest(fies.sev ~ wasting + school_mean + ruggedness + gdp_percap + 
-                       gini + mean_annual_precip + hci + tavg + urban_perc + 
-                       wci_index, data=moddat)
+## Mod+Sev
+avg.moddat.mod <- list()
+avg.preddat.mod <- list()
+for(i in 1:100) {
+  
+  # Print iteration
+  print(paste0("-------------------------------------------------------------------------| ", i, "/100 |"))
+  
+  if(i%%10 == 0 | i == 1) {
+    # Tune parameters
+    rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod", paste(vars, collapse = "+"), sep= "~")),
+                              data = moddat,
[…]

### 2020-09-02 models/model_randomforest_tune.R

@@ -1,0 +1,388 @@
+#######################################################
+# Tune Random Forest model, Find the best model
+#######################################################
+
+covars <- covars %>% merge(ne_countries()@data %>%
+                           select(ISO3=iso_a3, region=region_wb))
+
+ne_countries()@data %>%
+  select(ISO3=iso_a3, region=region_wb)
+
+#Just to fix Siberia
+covars$wci_index[covars$GDLCODE=='RUSr108'] <- 1
+
+covars$gdp_percap <- log(covars$gdp_percap)
+
+for (i in unique(covars$region)){
+  if (i == 'HIC'){
+    next
+  }
+  covars[ , paste0('region', i)] <- as.numeric(covars$region == i)
+}
+
+
+# Play around with model variables
+exc <- list(
+  c(), #1
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #2
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting', 'mal_falciparum'),
+  
+  c('crops_prod', 'builtup', 'livestock', #3
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting'),
+  
+  c('crops_prod', 'builtup', 'livestock', #4
+    'pasture', 'crops_prod', 'cropland',
+    'wasting'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #5
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting', 'mal_falciparum',
+    'school_mean', 'mean_annual_precip',
+    'tavg', 'wci_index', 'ruggedness', 'urban_perc'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #6
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting', 'mal_falciparum',
+    'wci_index'),
+  
+  c('crops_prod', 'mal_vivax', 'wasting', #7
+    'mal_falciparum', 'wci_index'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #8
+    'pasture', 'cropland', 'mal_vivax'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #9
+    'pasture', 'cropland', 'mal_vivax', 'wasting')
+)
+
+tsplit <- c(0.25) #split training and test, tsplit = % of test sample
+ntree <- c(500) #number of trees
[…]

### 2020-09-02 docs/modeling_fies_v2.tex
Updates from Overleaf

--- 2020-06-04 docs/modeling_fies.tex
@@
 \affil[1]{T.H. Chan School of Public Health, Harvard University}
-\affil[2]{TU Wien}
+\affil[2]{Department of Economics, Vienna University of Economics and Business}
 
@@
 
+\begin{table}[H]
+	\begin{tabular}{llll}
+		\toprule
+		Name & Availability & Scale & Source \\
+		\midrule
+        HDI & Annual & Subnational & GDL \\
+        Life Expectancy & Annual & Subnational & GDL \\
+        GNI Per Capita & Annual & Subnational & GDL \\
+        Expected Years School & Annual & Subnational & GDL \\
+        Mean Years School & Annual & Subnational & GDL \\
+        Country Programmable Assistance (CPA) & Annual & National & OECD Database \\
+        AG PCT GD & Annual & National & World Bank \\
+        Import Value Per Capita & Annual & National & World Bank \\
+        Undernourishment & Annual & National & World Bank, FAO \\
+        government_effectiveness & Annual & National & World Bank WGIs \\
+        control_corruption & Annual & National & World Bank WGIs \\
+        regulatory_quality & Annual & National & World Bank WGIs \\
+        rule_of_law & Annual & National & World Bank WGIs \\
+        stability_noviolence & Annual & National & World Bank WGIs \\
+        voice_accountability & Annual & National & World Bank WGIs \\
+        mean annual precip & static & subnational & WorldClim \\
+        mean average temp & static & subnational & WorldClim \\
+        forest cover & 2016 & subnational & Song et al \\
+        elevation & static & subnational & ETOPO5 \\
+        ruggedness & static & subnational & ETOPO5 \\
+
+        %TODO:
+        %Nutritional Diversity
+        %Malaria
+        %Stunting
+        %Livestock
+        %Buitup
+        %Irrigation
+        %TTC
+        %Sum of conflict deaths
+        %Urban-Rural pop
+
+		\bottomrule
+	\end{tabular}
+	\caption{Co-Variates Included in Present-Day Analysis NEW}
+	\label{tab:covars2}
+\end{table}
+
 \subsection{Covariates Available in the Present and Future}
-To model the FIES in the future, we use projections derived in line with the Shared Socio-Economic Pathways (SSPs).
+To model the FIES in the future, we use projections from multiple sources.
+Most covariates are derived in line with the Shared Socio-Economic Pathways (SSPs).
 The SSPs are a framework to estimate what the future might look like in terms of human development and climate-change mitigation, based on a number of scenarios.
-We use projections of GDP \cite{Dellink2017}, education \cite{KC2017}, and urbanization \cite{Jiang2017} for SSP2, the middle-of the road pathway.
+We use projections for SSP2, the middle-of the road pathway, from \cite{Crespo2017}, \cite{Dellink2017}, \cite{Rao2019a}, \cite{KC2017}, \cite{Jiang2017} and \cite{SSPData2017}.
+Additionally, we use projections for poverty and water scarcity from \cite{WDL2020PC} and \cite{WDL2020WSC} respectively and estimate our own forecasts (with elasticities to GDP) for ``Imports per capita'' and ``Agriculture as a Percent of GDP'' (See Table \ref{tab:covars3}.
+
+\begin{table}[H]
+	\begin{tabular}{llll}
+		\toprule
+		Name & Availability & Scale & Source \\
+		\midrule
+		GDP PPP per capita & Annual & National & \cite{Crespo2017} \\
+		- & - & - & \cite{Dellink2017} \\
+		Gini index & Annual & National & \cite{Rao2019a} \\
+		Share primary education & Annual & National & \cite{KC2017} \\
+		Share secondary education & Annual & National & \cite{KC2017} \\
+		%Share tertiary education & Annual & National & \cite{KC2017} \\
+		Share people in urban areas & Annual & Subnational & \cite{Jiang2017} \\
+        Crops production & Annual & Regional & \cite{Jiang2017} \\
+        Cropland cover & Annual & Regional & \cite{SSPData2017} \\
+        Forest land cover & Annual & Regional & \cite{SSPData2017} \\
+        Built up land cover & Annual & Regional & \cite{SSPData2017} \\
+        Livestock production & Annual & Regional & \cite{SSPData2017} \\
+        Poverty headcount index & Annual & National & \cite{WDL2020PC} \\
+        %Water stress index & Annual & Subnational & \cite{WDL2020WSC} \\
+        Water scarcity index & Annual & Subnational & \cite{WDL2020WSC} \\
+        %Absolute water scarcity index & Annual & Subnational & \cite{WDL2020WSC} \\
+        Imports per capita & Annual & National & \cite{TheWorldBank2016} \\
+        Agriculture percent of GDP & Annual & National & \cite{TheWorldBank2016} \\
+        \bottomrule
+	\end{tabular}
+	\caption{Co-Variates Included in Forecast Analysis}
+	\label{tab:covars3}
+\end{table}
 
@@
 %Determine local environment
-\bibliography{C://Users/matt/library}
+\bibliography{references}
 

### 2020-09-02 docs/modeling_fies.tex
Notes of Covariates

@@
 \author[2]{Benjamin Müller}
+\author[3]{Juan Carlos Laso}
+\author[4]{Homi Kharas}
 
 \affil[1]{T.H. Chan School of Public Health, Harvard University}
-\affil[2]{TU Wien}
+\affil[2]{Department of Economics, Vienna University of Economics and Business}
 
@@
 To disaggregate the FIES by urban-rural areas, we used data on urbanization from \cite{Jiang2017}.
+%once implemented add here more to disaggregation
 
@@
 
+%NOTE: WHEN JUSTIFYING COVARIATES, SAY THAT WE USE MANY FROM DEVELOPING COUNTRIES BECAUSE THE CAUSES OF FOOD INSECURITY ARE MORE DIVERSE THERE: ALMOST A KARENINA PRINCIPLE
+
+%update/create this table once we have our final model
 \begin{table}[H]
@@
 
+\begin{table}[H]
+	\begin{tabular}{llll}
+		\toprule
+		Name & Availability & Scale & Source \\
+		\midrule
+        HDI & Annual & Subnational & GDL \\
+        Life Expectancy & Annual & Subnational & GDL \\
+        GNI Per Capita & Annual & Subnational & GDL \\
+        Expected Years School & Annual & Subnational & GDL \\
+        Mean Years School & Annual & Subnational & GDL \\
+        Country Programmable Assistance (CPA) & Annual & National & OECD Database \\
+        AG PCT GD & Annual & National & World Bank \\
+        Import Value Per Capita & Annual & National & World Bank \\
+        Undernourishment & Annual & National & World Bank, FAO \\
+        government_effectiveness & Annual & National & World Bank WGIs \\
+        control_corruption & Annual & National & World Bank WGIs \\
+        regulatory_quality & Annual & National & World Bank WGIs \\
+        rule_of_law & Annual & National & World Bank WGIs \\
+        stability_noviolence & Annual & National & World Bank WGIs \\
+        voice_accountability & Annual & National & World Bank WGIs \\
+        mean annual precip & static & subnational & WorldClim \\
+        mean average temp & static & subnational & WorldClim \\
+        forest cover & 2016 & subnational & Song et al \\
+        elevation & static & subnational & ETOPO5 \\
+        ruggedness & static & subnational & ETOPO5 \\
+
+        %TODO:
+        %Nutritional Diversity
+        %Malaria
+        %Stunting
+        %Livestock
+        %Buitup
+        %Irrigation
+        %TTC
+        %Sum of conflict deaths
+        %Urban-Rural pop
+
+		\bottomrule
+	\end{tabular}
+	\caption{Co-Variates Included in Present-Day Analysis NEW}
+	\label{tab:covars2}
+\end{table}
+
 \subsection{Covariates Available in the Present and Future}
@@
 We use projections of GDP \cite{Dellink2017}, education \cite{KC2017}, and urbanization \cite{Jiang2017} for SSP2, the middle-of the road pathway.
+%in general we use middle of the road but maybe also other scenarios to account for uncertainty
 

### 2020-09-03 data/covars/scripts/anthropometry.R
final model outputs, cleaned and updated covars

@@ -76,4 +76,13 @@ ep <- em %>%
   mutate(stunting=modelFuture(stunting),
          wasting=modelFuture(wasting))
 
+#For countries without data, set to 0
+ep$ISO3 <- substr(ep$GDLCODE, 1, 3)
+
+na_countries <- c('USA', 'CAN', 'AUS', 'CHL', 'ARG', 'URY', 'SAU', 'OMN', 
+                  'RUS', 'KAZ', 'ARM', 'AZE')
+
+ep$wasting[ep$ISO3 %in% na_countries] <- 0
+ep$stunting[ep$ISO3 %in% na_countries] <- 0
+
 write.csv(ep, 'data/covars/results/anthro_vars.csv', row.names=F)

### 2020-09-03 data/covars/scripts/malaria.R

@@ -6,7 +6,7 @@ library(tidyverse)
 
 sp <- readOGR('data/GDL Shapefiles V4 0.005', 'GDL Shapefiles V4 Edit')
 
-fs <- list.files('data/covars/rawdata/malaria', pattern='^2019_Global_P._Incidence_20...tif$', full.names = T, recursive=T)
+fs <- list.files('data/covars/rawdata/malaria', pattern='^2019_Global_P._Mortality_Rate_20...tif$', full.names = T, recursive=T)
 
 s <- stack(fs)
 
@@ -14,7 +14,7 @@ e <- raster::extract(s, sp)
 
 getColSumsOrNull <- function(x){
 	if (is.null(dim(x))){
-		return(data.frame(X2019_Global_Pv_Incidence_2012=NA))
+		return(data.frame(X2019_Global_Pf_Mortality_Rate_2017=NA))
 	} else{
 		return(colSums(x, na.rm=T))
 	}
@@ -25,15 +25,12 @@ e2 <- lapply(e, getColSumsOrNull) %>%
 
 e3 <- e2 %>%
 	mutate(GDLCODE=sp$GDLcode) %>%
-  gather(timetype, incidence, -GDLCODE) %>%
-  mutate(YEAR=as.numeric(substr(timetype, nchar(timetype) - 3, nchar(timetype))),
-         type=ifelse(grepl('Pf', timetype), 'mal_falciparum', 'mal_vivax')) %>%
-  select(-timetype) %>%
-  spread(type, incidence)
+  gather(timetype, mal_falciparum, -GDLCODE) %>%
+  mutate(YEAR=as.numeric(substr(timetype, nchar(timetype) - 3, nchar(timetype)))) %>%
+  select(-timetype)
 
 #Missing data at northern latitudes
-e3$falciparum[is.na(e3$falciparum)] <- 0
-e3$vivax[is.na(e3$vivax)] <- 0
+e3$mal_falciparum[is.na(e3$mal_falciparum)] <- 0
 
 ##################################################
 # Now project to 2030 using methodlogy
@@ -81,8 +78,6 @@ e4 <- e3 %>%
   merge(expand.grid(list(GDLCODE=unique(e3$GDLCODE), YEAR=2000:2030)), all.y=T) %>%
   group_by(GDLCODE) %>%
   arrange(YEAR) %>%
-  mutate(mal_falciparum=modelFuture(mal_falciparum/100000),
[…]

### 2020-09-03 data/covars/scripts/tavg.R

@@ -4,6 +4,24 @@ library(raster)
 library(rgdal)
 library(tidyverse)
 
+mod1_2010s <- stack('data/covars/rawdata/tavg/tas_day_GFDL-ESM2M_rcp60_r1i1p1_EWEMBI_landonly_20110101-20201231.nc4')
+indices <- as.numeric(format(as.Date(names(mod1_2010s), format = "X%Y.%m.%d"), format = "%Y"))
+mod1_2010s_year <- stackApply(mod1_2010s, indices, fun=mean)
+
+mod1_2020s <- stack('data/covars/rawdata/tavg/tas_day_GFDL-ESM2M_rcp60_r1i1p1_EWEMBI_landonly_20210101-20301231.nc4')
+indices <- as.numeric(format(as.Date(names(mod1_2020s), format = "X%Y.%m.%d"), format = "%Y"))
+mod1_2020s_year <- stackApply(mod1_2020s, indices, fun=mean)
+
+mod2_2010s <- stack('data/covars/rawdata/tavg/tas_day_HadGEM2-ES_rcp60_r1i1p1_EWEMBI_landonly_20110101-20201231.nc4')
+indices <- as.numeric(format(as.Date(names(mod2_2010s), format = "X%Y.%m.%d"), format = "%Y"))
+mod2_2010s_year <- stackApply(mod2_2010s, indices, fun=mean)
+
+mod2_2020s <- stack('data/covars/rawdata/tavg/tas_day_HadGEM2-ES_rcp60_r1i1p1_EWEMBI_landonly_20210101-20301231.nc4')
+indices <- as.numeric(format(as.Date(names(mod2_2020s), format = "X%Y.%m.%d"), format = "%Y"))
+mod2_2020s_year <- stackApply(mod2_2020s, indices, fun=mean)
+
+
+
 sp <- readOGR('data/GDL Shapefiles V4 0.005', 'GDL Shapefiles V4 Edit')
 
 #FROM https://www.worldclim.org/data/worldclim21.html

### 2020-09-03 data/covars/scripts/under5pop.R

@@ -2,60 +2,13 @@ setwd('~/wdl-fies')
 
 library(tidyverse)
 library(zoo)
-library(rgdal)
+library(sf)
 
-logit <- function(x){
-  log(x/(1-x))
-}
+sp <- read_sf('data/GDL Shapefiles V4 0.005', 'GDL Shapefiles V4 Edit')
 
-inv.logit <- function(x){
-  1/(1 + exp(-x))
-}
+pop <- read.csv('data/covars/rawdata/ssp/SspDb_country_data_2013-06-12.csv')
 
-modelFuture <- function(var){
-  #Model Future using methodology from Osgood-Zimmerman
-  #Get year on year Average Rate of Change (AROC)
-  #Get weighted average rate of change,
-  #Then apply it to future
-
-  #Var must be in order sequentially,
-  # with only non-NAs preceeding NAs
-  # must be a fraction (rate, or incidence)
-
-  #Logit fails if any value is 1 or 0.
-  #For my purposes, values very close to 1 and 0 will work
-  #So set 1 = 0.99999
-  
-  if (any(var > 1, na.rm=T) | any(var < 0, na.rm=T)){
-    stop("Value should be a proportion, cant be > 1 or < 0")
-  }
-
-  var[var == 1] <- 0.99999
-  var[var == 0] <- 0.00001
-
-  dat <- var[!is.na(var)]
-  nas <- sum(is.na(var))
-
-  roc <- mapply(function(y1, y2) logit(y2) - logit(y1),
-         dat[1:(length(dat) -1)],
-         dat[2:length(dat)])
-
-  w <- (2:length(dat))/sum(2:length(dat)) 
-  
-  aroc <- sum(w*roc)
-
-  pred <- inv.logit(logit(dat[length(dat)]) + aroc*(1:nas))
-  
-  pred[is.nan(pred)] <- 0
-
-  res <- c(dat, pred)
-
-  return(res)
-}
-
-sp <- readOGR('data/GDL Shapefiles V4 0.005', 'GDL Shapefiles V4 Edit')
-
-u5 <- read.csv('data/covars/rawdata/ssp/SspDb_country_data_2013-06-12.csv') %>%
+u5 <- pop %>%
   filter(grepl('Aged0-4$', VARIABLE), 
          MODEL=='IIASA-WiC POP', 
          SCENARIO=='SSP2_v9_130115') %>%
[…]

### 2020-09-03 models/model_randomforest.R

@@ -1,52 +1,29 @@
 #######################################################
-# Use Random Forest with SSP covariates
+# use random forest with ssp covariates
 # to estimate food insecurity, based on data for
 # modeled for 2020, 2025, and 2030
 #######################################################
 
-# setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
+# setwd('~/wdl-fies'); library(projecttemplate); load.project()
 
-covars <- covars %>% merge(ne_countries()@data %>%
-                           select(ISO3=iso_a3, region=region_wb))
-
-ne_countries()@data %>%
-  select(ISO3=iso_a3, region=region_wb)
-
-#Just to fix Siberia
-covars$wci_index[covars$GDLCODE=='RUSr108'] <- 1
-
-covars$gdp_percap <- log(covars$gdp_percap)
-
-for (i in unique(covars$region)){
-  if (i == 'HIC'){
-    next
-  }
-  covars[ , paste0('region', i)] <- as.numeric(covars$region == i)
-}
-
-
-# Read in data
+# read in data
 moddat <- merge(fies_subnat, 
                 covars, 
-                all.x=T, all.y=F) %>%
+                all.x=t, all.y=f) %>%
   na.omit %>%
   data.frame
 
 preddat <- covars %>%
   data.frame
 
-preddat <- na.omit(preddat); summary(preddat)
-
-exc <- c()
-#exc <- c('mal_falciparum', 'mal_vivax', 'wasting', 'crop_prod')
 exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
-         'pasture', 'cropland', 'mal_vivax')
+         'pasture', 'cropland')
 
-# Set up Model
-vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
+# set up model
+vars <- names(moddat)[!names(moddat) %in% c('iso3', 'gdlcode', 'fies.mod.rur',
                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
-                                            'Urban', 'Rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region',
+                                            'urban', 'rural', 'fies.sev', 'fies.mod',
+                                            'population', 'year', 'rural_perc', 'region',
[…]

### 2020-09-03 munge/03-Covariates.R

@@ -11,13 +11,16 @@ for (f in fs){
 #Select ony vars that have future data
 covars <- covars %>%
   select(YEAR, ISO3, GDLCODE, stunting, wasting, school_mean, ruggedness,
-         gdp_percap, gini, mal_falciparum, mal_vivax, mean_annual_precip, hci, tavg, 
+         gdp_percap, gini, mal_falciparum, mean_annual_precip, hci, tavg, 
          population, rural_perc, urban_perc, wci_index, builtup, cropland, crops_prod,
-         forest, livestock, pasture)
+         forest, livestock, pasture) %>%
+  merge(regions)
 
-##############################
-#Post-hoc data matching
-################################
+#########################################################################
+# Missing entire variable at the country level.  
+# Fill a missing countries with neighbors
+# (Mostly Syria, South Sudan, North Korea, Kosovo)
+#########################################################################
 
 ########  Syria  #############
 #HCI is the same as Yemen
@@ -41,7 +44,6 @@ covars$gini[covars$ISO3=='SSD'] <- rep(caf_gini, each=n_admin1)
 #WCI, School mean, GDP Percap, same as Tajikstan
 #Malaria is 0
 covars$mal_falciparum[covars$ISO3=='PRK'] <- 0
-covars$mal_vivax[covars$ISO3=='PRK'] <- 0
 
 tjk <- covars %>%
   filter(ISO3=='TJK') %>%
@@ -55,5 +57,56 @@ covars$school_mean[covars$ISO3=='PRK'] <- tjk$school_mean
 covars$gdp_percap[covars$ISO3=='PRK'] <- tjk$gdp_percap
 covars$gini[covars$ISO3=='PRK'] <- tjk$gini
 covars$wci_index[covars$ISO3=='PRK'] <- tjk$wci_index
+
+###########################################################################
+# Missing data at the region level
+# Fill with mean of other regions
+# (Mostly regions missing from GDL ontology - Jersey, Puerto Rico, Taiwan)
+############################################################################
+#Use country average for missing values
+impute.mean <- function(x){replace(x, is.na(x), mean(x, na.rm=T))}
+
+covars <- covars %>%
+  group_by(ISO3, YEAR) %>%
+  mutate_if(is.numeric, impute.mean)
+
+##################################################################
+# Missing dat chronologically
[…]

### 2020-09-03 viz/covar_maps.R

@@ -1,18 +1,24 @@
 #Add 2018 data to shapefile
+# setwd('~/wdl-fies');library(ProjectTemplate);load.project()
 
-covar_map <- merge(gdl, covar_now %>% filter(YEAR==2018))
+covar_map <- merge(gdl, covars %>% filter(YEAR %in% c(2015, 2020, 2025, 2030)))
 
-v <- names(covar_map)[!names(covar_map) %in% c('GDLCODE', 'constant', 'iso_code', 'country',
-                                               'region', 'YEAR', 'ISO3')]
+exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
+         'pasture', 'cropland')
 
-countries <- ne_countries(returnclass='sf')
+# Set up Model
+v <- names(covar_map)[!names(covar_map) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
+                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
+                                            'Urban', 'Rural', 'fies.sev', 'fies.mod',
+                                            'population', 'YEAR', 'rural_perc', 'region',
+                                            names(covar_map)[grepl('region', names(covar_map))],
+                                            exc)]
 
 for (n in v){
   covar_map$mapvar <- covar_map[ , n, drop=TRUE]
 
   ggplot() + 
-    geom_sf(data=covar_map, aes(fill=mapvar)) +
-    geom_sf(data=countries, fill='transparent') + 
+    geom_sf(data=covar_map, aes(fill=mapvar), size=0.05) +
     labs(fill=n,
          title=n) + 
     scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", 
@@ -21,7 +27,8 @@ for (n in v){
     coord_sf(crs='+proj=robin') +
     theme_void() +
     theme(legend.position = 'bottom',
-          plot.title = element_text(hjust = 0.5))
+          plot.title = element_text(hjust = 0.5)) + 
+    facet_wrap(~ YEAR, nrow=2, ncol=2)
 
-  ggsave(paste0('figures/covar_maps/', n, '.png'), width=10, height=8)
+  ggsave(paste0('figures/covar_maps/', n, '.png'))
 }

### 2020-09-03 viz/covar_maps.R
covariate maps

@@ -10,7 +10,8 @@ exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
 v <- names(covar_map)[!names(covar_map) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
                                             'Urban', 'Rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region',
+                                            'population', 'YEAR', 'rural_perc', 'region', 
+                                            'geometry',
                                             names(covar_map)[grepl('region', names(covar_map))],
                                             exc)]
 
@@ -30,5 +31,5 @@ for (n in v){
           plot.title = element_text(hjust = 0.5)) + 
     facet_wrap(~ YEAR, nrow=2, ncol=2)
 
-  ggsave(paste0('figures/covar_maps/', n, '.png'))
+  ggsave(paste0('figures/covar_maps/', n, '.png'), width=18, height=9.5)
 }

### 2020-09-07 viz/Viz_Mod_Results.R
New Visualizations for paper

@@ -2,108 +2,116 @@
 
 #Save outputs to wdl-fies-tex for manuscript
 setwd('~/wdl-fies-tex/img')
+library(cowplot)
+options(scipen=100)
 
 ############################
 # visualize results
 ############################
+
+########################
+# Make Map
+##########################
+
+mapdat <- merge(gdl,
+                preddat %>%
+                  filter(YEAR %in% c(2010, 2020, 2030)) %>%
+                  select(YEAR, GDLCODE, fies.mod.pred, fies.sev.pred) %>%
+                  gather(outcome, value, -YEAR, -GDLCODE) %>%
+                  mutate(FIES=ifelse(grepl('sev', outcome), 'Severe\n', 'Moderate-to-Severe\n')))
+
+ggplot(mapdat) +
+  geom_sf(aes(fill=value), color=NA) + 
+  scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", "#e4ea9a", "#fbf8c0", 
+                                 "#fce08a", "#faae61", "#f36c44", "#a01c44"), 
+                       breaks=seq(0, 1, by=0.2),
+                       limits=c(0, max(mapdat$value))) + 
+  geom_sf(data=cty, color='#000000', fill=NA, size=0.15) + 
+  coord_sf(crs='+proj=robin', expand=FALSE) + 
+  theme_void() + 
+  theme(legend.position = 'bottom',
+        plot.title = element_text(hjust = 0.5),
+        panel.spacing.x = unit(-0.5, "in")) + 
+  labs(fill='') + 
+  facet_grid(YEAR ~ FIES, switch='y')
+ggsave('FullMap.pdf', width=7, dpi=750)
+system('pdfcrop FullMap.pdf FullMap.pdf')
+
+
+###########################
+# Setup Reference Map
+############################
+region_cols <- c("South Asia" = "#e41a1c",
+                 "Sub-Saharan Africa" = "#377eb8",
+                 "Europe & Central Asia" = "#4daf4a",
+                 "Middle East & North Africa" = "#f781bf",
+                 "Latin America & Caribbean" = "#ff7f00",
+                 "East Asia & Pacific" = "#a65628",
+                 "North America" = "#984ea3")
+
+cty <- ne_countries(returnclass='sf') %>%
+  filter(region_wb != 'Antarctica')
+
+reg <- ggplot(cty) + 
[…]

### 2020-09-07 docs/modeling_fies.tex
fix visualizations path

--- 2020-06-04 docs/modeling_fies.tex
@@
 \usepackage{longtable}
+\usepackage{lineno}
+\usepackage[
+  style=numeric,
+  citestyle=numeric-comp,
+  backend=biber,
+  doi=true,
+  natbib=true,
+  sorting=none
+]{biblatex}
+
+\addbibresource{library.bib}
 
@@
 
-\title{Modeling the Food Insecurity Experience Scale (FIES) in the Present and Future}
+\title{A Global Picture of Food Insecurity and Hunger}
 
-\author[1]{Matthew Cooper}
-\author[2]{Benjamin Müller}
+\author[1,2,*]{Matthew Cooper}
+\author[2,3]{Benjamin Müller}
+\author[4]{Carlo Cafiero}
+\author[2,5]{Juan Carlos Laso}
+\author[2,6]{Homi Kharas}
 
 \affil[1]{T.H. Chan School of Public Health, Harvard University}
-\affil[2]{TU Wien}
+\affil[2]{The World Data Lab, Vienna, Austria}
+\affil[3]{Department of Economics, Vienna University of Economics and Business}
+\affil[4]{The Food and Agricultural Organization of the United Nations}
+\affil[5]{The International Institute for Applied Systems Analysis}
+\affil[6]{The Brookings Institute}
+\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}
 
@@
 \begin{abstract}
-The Food Insecurity Experience Scale (FIES) measures the experiential aspects of food insecurity, and has been selected as a target for the second Sustainable Development Goal (SDG) of Zero Hunger.
-Here, we present the preliminary results of our work modeling this indicator globally for the years 2020-2030.
-
+We modeled levels of of food security at a subnational level from 2010 to 2030 using microdata from 77 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project.
+We find significant heterogeneity in levels of food security around the world, ranging from parts of the global north with less than 1\% of the population food insecure to parts of the global south where as much as 3 out of 4 people are severely food insecure.
+Examining global temporal trends, we find that food insecurity has been increasing over the past several years, but under middle-of-the-road assumptions for development and population change, food insecurity will decline throughout the 2020s.
+This global decrease is largely driven by trends in South and East Asia, and some parts of the world, particularly sub-Saharan Africa, are projected to show increases in the prevalence of food insecurity throughout the next decade.
+This is the first global picture of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
 \end{abstract}
@@
 \section{Introduction}
+Being food secure is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG2) is ``Zero Hunger".
+While a number of indicators are used to track progress towards this goal, the indicator that is most indicative of the human experience of food insecurity is the Food Insecurity Experience Scale, or FIES.
+The FIES was developed due to a recognized need to measure actual food insecurity based on personal exposure to hunger, rather than proxies for food insecurity, such as estimates of calories per capita, or health consequences of food insecurity, such as child stunting.
+However, as a relatively new metric, little work has been done so far to model food security at a global scale or estimate current trends in the number of food insecure people using the FIES.
 
-\subsection{Measuring Food Security and the Genesis of the FIES}
-Food security is a key part of human development, but has traditionally been difficult to measure.
-Metrics of macro-health, such as anthropometry and mortality rates are correlated broadly with food insecurity \cite{Puffer1973, Habicht1974}, but are confounded with other determinants of health such as infectious disease.
-Other proxies for food security, such as food availability estimated from crop yields \cite{Maxwell1992}, are inadequate because they do not indicate how accessible food is to the general population, and food insecurity can certainly occur in the absence of food availability decline \cite{Sen1983}.
+Food security has traditionally been difficult to measure, and this has led to incomplete or inaccurate pictures of global hunger.
+Metrics of macro-health, such as anthropometry and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.
+However, these metrics are confounded with other determinants of health such as infectious disease and are not meaningful at the scale of individuals or households \citep{Perumal2018}.
+Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they do not indicate how accessible food is to the general population, and food insecurity can certainly occur in the absence of food availability decline \citep{Sen1983}.
+Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.
+Thus, global estimates of hunger and food insecurity based on these metrics carry forward similar flaws.
 
-As researchers began to focus on food insecurity at the individual and household level, household microdata collecting information on household finances and consumption became a common proxy for food security \cite{Haddad1994}.
-However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \cite{Maxwell1996}.
-This led to the emergence of several indicators designed to be rapidly deployable, culturally cross-comparable, and based on the lived experience of food security.
-These metrics include the Household Food Insecurity and Access Scale (HFIAS) \cite{Coates2007}; the Coping Strategies Index (CSI) \cite{Maxwell1999}; the Household Hunger Scale (HHS) \cite{Ballard2011}; the Food Consumption Score (FCS); and the Household Dietary Diversity Scale (HDDS) \cite{Kennedy2010}.
+As researchers began to focus on food insecurity at the individual and household level, household microdata collecting information on household finances and consumption became a common proxy for food security \citep{Haddad1994}.
+However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \citep{Maxwell1996}.
+This led to the emergence of several indicators designed to be rapidly deployable, culturally cross-comparable, and based on the lived experience of food security \citep{Jones2013}.
+These metrics include the Household Food Insecurity and Access Scale (HFIAS) \citep{Coates2007}; the Coping Strategies Index (CSI) \citep{Maxwell1999}; and the Household Hunger Scale (HHS) \citep{Ballard2011}.
 
-Drawing on the insights derived in designing and implementing these food security metrics, the Food Insecurity Experience Scale (FIES) was developed by the Food and Agricultural Organization (FAO) of the UN.
+Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013}.
+The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
+Using a Rash model \citep{Cafiero2018}, each individual in the survey is given a severity score, and thresholds for moderate and severe food insecurity are set at the behaviors of eating less than the respondent thought they should, and not eating for a while day, respectively.
+Thus, people in moderate-to-severe food insecurity are estimated to have not eaten as much as they should have at least once in the previous year, while people in severe food insecurity are estimated to have gone an entire day without eating in the previous year.
 
-\subsection{The FIES}
-The FIES asks participants eight questions about their experience of food insecurity during the previous year \cite{Cafiero2018}, collected in countries around the world both in person and over the telephone with the assistance of Gallup World Poll.
-These questions are:
+Since 2014, in partnership with Gallup World Poll, the FAO conducted surveys around the world in 77 countries and nearly 200 languages to estimate the prevalence of food insecurity using the FIES.
+Using this data, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
+Our analysis highlights several important global and regional trends in food security.
 
+\section{Results}
+\subsection{Spatial Distribution of Food Insecurity}
+We found significant heterogeneities in the global distribution of severe food insecurity, estimated to be people who had to go an entire day without eating at least once in the previous year (See Fig.
+\ref{fig:map}).
+For the year 2020, mainland sub-Saharan Africa is the continent with with the highest rates of severe food insecurity, with at least 15\% of people in severe food insecurity in at last one region in every country except Gabon and Equatorial Guinea.
+Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.
+In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10-15\% in areas like northern Brazil, many central Asian and middle-eastern countries, and India and Indonesia.
+
+The experience of at least moderate food insecurity, or eating less than one would prefer, is quite common in many parts of the world.
+In 2020, it is quite common in Africa, south and southeast Asia, and parts of Latin America, where in many places over 50\% of the population experiences this degree of food insecurity.
+Even in countries widely considered to be developed, such as Australia, the United States, and parts of eastern and southern Europe, moderate-to-severe food insecurity is as high as 10-15\%.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/FullMap.pdf}
+  \caption{Modeled rates of severe and moderate-to-severe food insecurity for the years 2010, 2020, and 2030}
+  \label{fig:map}
+\end{figure}
+
+Globally, in the year 2020, we estimate that 2.33 billion people experience at least moderate food insecurity, and 829 million people experience severe food insecurity (See Fig.
+\ref{fig:timeseries}).
+While the total number of people in moderate and severe food security held roughly constant for the 2010s and even increased in latter part of that decade, we predict that it will decline throughout the 2020s.
+However, this overall decline will not be uniform across the world.
+Our model expects that, in sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau.
+Food Security will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+
+\section{Discussion}
+Compare with SOFI reports, other metrics of hunger.
+
+\section{Methods}
+
+\subsection{Disaggregation}
+Assumptions
+
+\subsection{Covariates}
+
+Selection and Processing
+
+(Maybe put some in Appendix)
+
+Cite determinants of FIES form \citep{smith2017world}.
+
+\subsection{Modeling}
+
+Random Forest
+
+Hyperparameter selection
+
+\section{Conclusion}
+
+\printbibliography
+
+\section*{Supplemental Info}
+\setcounter{table}{0}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\subsection{Component Questions of the FIES}
 \begin{enumerate}
@@
 
-Based on the responses to these questions, a Rasch model is used to generate the score for the population \cite{Engelhard2013}.
-
-\section{Data}
-For our analysis, we draw on several datasets.
-First, we use the individual responses from 335,139 individuals from 331 surveys across 77 countries, downloaded from the FAO microdata catalogue.
-Additionally, we use country-level FIES scores downloaded from FAOSTAT.
-
-To disaggregate the FIES by urban-rural areas, we used data on urbanization from \cite{Jiang2017}.
-
-\subsection{Covariates Available in the Present}
-We use a large stack of covariates available for the present or the near-present, reflective of factors related to many aspects of food systems.
-Datasets that were originally gridded were aggregated to the Admin-1 level taking the mean value.
-A full table of datasets is given in Table \ref{tab:covars1}.
-
-\begin{table}[H]
-	\begin{tabular}{llll}
-		\toprule
-		Name & Availability & Scale & Source \\
-		\midrule
-		Agriculture As A Percentage Of GDP & Annual & National & \cite{TheWorldBank2016} \\
-		Net Oda Per Capita & Annual & National & \cite{TheWorldBank2016} \\
-		Percent of Area Bare Ground & Annual & Admin-1 & \cite{Song2018} \\
-		Percent Of Area With Built Up Land Cover & 2000 \& 2014 & Admin-1 & \cite{Pesaresi2015} \\
-		Staple Crop Proction Per Capita & Annual & National & \cite{FAOSTAT2018} \\
-		Elevation & Static & Admin-1 & \cite{USGS1996} \\
-		Net School Enrollment & Annual & National & \cite{TheWorldBank2016} \\
-		Forest Cover & Annual & Admin-1 & \cite{Song2018} \\
-		Government Effectiveness Indicator & Annual & National & \cite{Kaufmann2011} \\
-		GDP PPP Per Capita & Annual & Admin-1 & \cite{Kummu2018} \\
-		HDI & Annual & Admin-1 & \cite{Kummu2018} \\
-		Value Of All Imports Per Capita & Annual & National & \cite{TheWorldBank2016} \\
-		Irrigated Area & 2000 & Admin-1 & \cite{SiebertS.DollP.FeickS.FrenkenK.Hoogeveen2013} \\
-		Time To Travel To A City  & 2015 & Admin-1 & \cite{Uchida2008} \\
-		Mean Annual Precipitation & Static & Admin-1 & \cite{Funk2015} \\
-		NDVIi & Annual & Admin-1 & \cite{Song2018} \\
-		People Per Pixel & 2010, 2015, 2020 & Admin-1 & \cite{Doxsey-Whitfield2015} \\
-		Topographic Roughness Index & Static & Admin-1 & \cite{USGS1996, Riley1999} \\
-		Political Stability Index & Annual & Admin-1 & \cite{Kaufmann2011} \\
-		Number Of Conflict Deaths Within 50Km & Annual & Admin-1 & \cite{Eriksson2015} \\
-		Nutritional Diversity Of Agriculture & 2015 & Admin-1 & \cite{Herrero2017a} \\
-		Average Maximum Temperature & Roughly Static & Admin-1 & \cite{Sheffield2006} \\
-		Number Of Buffaloes Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
-		Number Of Cattle Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
-		Number Of Chickens Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
-		Number Of Ducks Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
-		Number Of Horses Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
-		Number Of Pigs Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
-		Number Of Sheep Per Square Km & 2010 & Admin-1 & \cite{Robinson2014} \\
-		Incidence \textit{P. falciparum} Malaria & Annual & Admin-1 & \cite{Weiss2019} \\
-		Incidence \textit{P. vivax} Malaria & Annual & Admin-1 & \cite{Weiss2019} \\
-		\bottomrule
-	\end{tabular}
-	\caption{Co-Variates Included in Present-Day Analysis}
-	\label{tab:covars1}
-\end{table}
-
-\subsection{Covariates Available in the Present and Future}
-To model the FIES in the future, we use projections derived in line with the Shared Socio-Economic Pathways (SSPs).
-The SSPs are a framework to estimate what the future might look like in terms of human development and climate-change mitigation, based on a number of scenarios.
-We use projections of GDP \cite{Dellink2017}, education \cite{KC2017}, and urbanization \cite{Jiang2017} for SSP2, the middle-of the road pathway.
-
-\section{Methods \& Results}
-\subsection{Disaggregating the FIES Subnationally}
-We begin by using the probability an individual has moderate-to-severe food insecurity for the individual-level survey data derive the FIES at the country level.
-Using the sampling weights reported by the FAO, we found the percentage of people in each country with a greater than 50\% chance of being moderate or severely food insecure.
-We then checked our own results against those reported by FAOSTAT and found similar, but not identical, estimates (See Figure \ref{fig:comparison}).
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.6\linewidth]{../figures/"FAO Compare - Moderate".png}
-	\caption{Comparison of derived and FAO reported estimates of food insecurity by FIES Score}
-	\label{fig:comparison}
-\end{figure}
-
-The FIES surveys include information on whether respondents were located in urban or rural areas.
-However, due to the small numbers of individuals in urban areas within each country (about 200 observations), we group countries into regions and estimate regional prevalences of food insecurity for both rural and urban areas.
-In our grouping, we use the following regions: Eastern Europe, Latin American and the Caribbean (LAC), Southeast Asia, West Africa, Southern Africa, the Mideast and Central Asia, and High-Income Countries (HICs) (See Figure \ref{fig:regions}).
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=\linewidth]{../figures/regions.png}
-	\caption{Grouping used to estimate regional rates of food insecurity in urban and rural areas.}
-	\label{fig:regions}
-\end{figure}
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.75\linewidth]{../figures/Urban_Rural.png}
-	\caption{Rates of food insecurity in urban and rural areas by world region.}
-	\label{fig:regionrate}
-\end{figure}
-
-Examining rates of food insecurity by both urban and rural areas shows that, in nearly every world region, food insecurity is worse in rural areas, with the largest gap in Southern Africa.
-High-income countries area the only exception, and food insecurity is actually worse in urban parts of these countries than in rural areas.
-
-Given the overall rate of food insecurity in each country, the disparities in rates of urban and rural food insecurity in each world region, and the percentage of the population and urban and rural areas within each admin-1 area of each country, it is possible to derive a rate of food insecurity specific to each admin-1 area.
-
-Formally, given the following two assumptions:
-
-\begin{equation} 
-\omega / \rho = ratio 
-\label{eq1}
-\end{equation}
-
-\begin{equation} 
-\omega * URB + \rho * RUR = t * TOT
-\label{eq2}
-\end{equation}
-\newline
-
-Where:
-
-\begin{itemize}
-	\item $\omega$: The percentage of households in urban areas in a country with food insecurity (UNKNOWN)
-
-	\item $\rho$: The percentage of households in rural areas in a country with food insecurity (UNKNOWN)
-
-	\item $t$: The overall percentage of households in a country with food insecurity (KNOWN)
-
-	\item $RUR$: The total population in rural areas (KNOWN)
-
-	\item $URB$: The total population in urban areas (KNOWN)
-
-	\item $TOT$: The total population (KNOWN)
-
-	\item $ratio$: The ratio of the rates of urban food insecurity to rural food insecurity.
-This is known at a regional level, and it is estimated that the same ratio holds at the individual country level.
-\end{itemize}
-
-We can solve for the two unknown parameters, $\omega$ and $\rho$ to get country-level estimates of urban and rural rates of food insecurity and derive subnational prevalences.
-
-Thus, we derive the following map of disaggregated rates of food insecurity in countries where we have direct FIES data (See Figure \ref{fig:disaggregated}).
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=\linewidth]{../figures/subnational-moderate.png}
-	\caption{Rates of food insecurity in urban and rural areas by world region, averaged across all years for which survey data is available}
-	\label{fig:disaggregated}
-\end{figure}
-
-\subsection{Modeling the FIES in the Present}
-We then used these Admin-1 estimates, in combination with a large stack of covariates in the present, to model the prevalence of food insecurity globally for the year 2020, based on covariates for as close to the year 2020 as possible.
-We used a LASSO regression approach, where the size of the regression coefficients are penalized, which is most appropriate for making predictions based on a large number of covariates \cite{Tibshirani2011}.
-
-For our in-sample data, the LASSO performed well (see Figure \ref{fig:lasso_residuals}).
-The coefficients for each variable are given in Figure \ref{fig:lasso_coefs}.
-Note that some coefficients are penalized to 0, which effectively removes them from the model.
-The map is the results is given in Figure \ref{fig:lasso_map}.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\linewidth]{../figures/Pred2020_Residuals.png}
-	\caption{Correlation between observed disaggregated rates of food insecurity and modeled rates of food insecurity.}
-	\label{fig:lasso_residuals}
-\end{figure}
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\linewidth]{../figures/Pred2020_Coefs.png}
-	\caption{Change in rate of food insecurity with a 1 standard deviation increase in each variable.  NOTE: IN A FUTURE EDITION THE LABELS WILL BE BETTER.}
-	\label{fig:lasso_coefs}
-\end{figure}
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{../figures/Pred2020_LASSO.png}
-	\caption{Estimated rates of food insecurity in 2020. NOTE: SOME COVARIATES ARE MISSING AT HIGH LATITUDES, THIS WILL BE FIXED IN A FUTURE EDITION.}
-	\label{fig:lasso_map}
-\end{figure}
-
-\subsection{Modeling the FIES in the Future}
-In addition to modeling the FIES in the present, we also modeled the FIES for the years 2020, 2025, and 2030, based covariates that were already extrapolated to the year 2030 in peer-reviewed literature.
-NOTE: MANY MORE COVARIATES ARE BEING ADDED.
-Similar to the previous model, we below show the correlation between the observed and predicted values (Figure \ref{fig:ssp_residuals}), the estimated coefficients (Figure \ref{fig:ssp_coefs}), and the mapped results (Figure \ref{fig:ssp_map}).
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\linewidth]{../figures/SSP2_Residuals.png}
-	\caption{Correlation between observed disaggregated rates of food insecurity and modeled rates of food insecurity.}
-	\label{fig:ssp_residuals}
-\end{figure}
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\linewidth]{../figures/SSP2_Coefs.png}
-	\caption{Change in rate of food insecurity with a 1 standard deviation increase in each variable.  NOTE: IN A FUTURE EDITION THE LABELS WILL BE BETTER AND MORE COVARIATES WILL BE INCLUDED.}
-	\label{fig:ssp_coefs}
-\end{figure}
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{../figures/SSP2_LASSO.png}
-	\caption{Estimated rates of food insecurity in 2020, 2025, and 2030.}
-	\label{fig:ssp_map}
-\end{figure}
-
-\section{Discussion}
-There is still a wide gap between the predictions of the 2020 model based on the large stack of covariates and the 2020 model based on just the SSP covariates.
-
-\section{Conclusion}
-
-\clearpage
-\bibliographystyle{apalike}
-
-%Determine local environment
-\bibliography{C://Users/matt/library}
-
-\section*{Supplemental Info}
-
 \end{document}

### 2020-09-07 viz/Viz_Mod_Results.R

@@ -1,7 +1,7 @@
 # setwd('~/wdl-fies'); library(ProjectTemplate); load.project(list(data_loading=T))
 
 #Save outputs to wdl-fies-tex for manuscript
-setwd('~/wdl-fies-tex/img')
+setwd('~/wdl-fies/docs/img')
 library(cowplot)
 options(scipen=100)
 

### 2020-09-07 docs/modeling_fies.tex
Note about improvements in some African countries

@@
 However, this overall decline will not be uniform across the world.
-Our model expects that, in sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau.
+Our model expects that, in sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau, although it will improve in some countires, such as Kenya, Senegal, and Ethiopia.
 Food Security will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.

### 2020-09-08 models/model_SSP.R
thesis folder and plots of all models

@@ -1,291 +1,0 @@
-#######################################################
-# Use LASSO regression with SSP covariates
-# to estimate food insecurity, based on data for
-# modeled for 2020, 2025, and 2030
-#######################################################
-
-#Just to fix Siberia
-covars$wci_index[covars$GDLCODE=='RUSr108'] <- 1
-
-covars$gdp_percap <- log(covars$gdp_percap)
-
-for (i in unique(covars$region)){
-  if (i == 'HIC'){
-    next
-  }
-  covars[ , paste0('region', i)] <- as.numeric(covars$region == i)
-}
-
-
-# Read in data
-moddat <- merge(fies_subnat, 
-                covars, 
-                all.x=T, all.y=F) %>%
-  na.omit %>%
-  data.frame
-
-preddat <- covars %>%
-  data.frame
-
-# Set up Model
-vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur', 
-                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',                                             'Urban', 'Rural', 'fies.mod', 'fies.sev', 
-                                            'population', 'YEAR', 'rural_perc',
-                                            'crops_prod', 'forest', 'builtup', 'livestock',
-                                            'pasture', 'crops_prod', 'cropland', 
-                                            'mal_vivax', 'stunting', 'region',
-                                            names(moddat)[grepl('region', names(moddat))])]
-####################################
-# Model moderate food insecurity
-##################################
-x <- model.matrix(as.formula(paste0('fies.mod ~ ', paste0(vars, collapse=' + '))), 
-                  data=moddat)
-
-mod <- cv.glmnet(x, logit(moddat$fies.mod), alpha=1)
-mod$lambda.min
-
-tmp_coeffs <- coef(mod, s = "lambda.min")
-mdf <- data.frame(term = tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1], estimate =tmp_coeffs@x,
-                  stringsAsFactors=F)
-
-# Get model predictions
[…]

### 2020-09-08 models/model_bayes_ssvs.R

@@ -1,1029 +1,0 @@
-#######################################################
-# Use Bayes SSVS with SSP covariates
-# to estimate food insecurity, based on data for
-# modeled for 2020, 2025, and 2030
-#######################################################
-
-# setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
-
-covars <- covars %>% merge(ne_countries()@data %>%
-                           select(ISO3=iso_a3, region=region_wb))
-
-ne_countries()@data %>%
-  select(ISO3=iso_a3, region=wb_regions)
-
-#Just to fix Siberia
-covars$wci_index[covars$GDLCODE=='RUSr108'] <- 1
-
-covars$gdp_percap <- log(covars$gdp_percap)
-
-for (i in unique(covars$region)){
-  if (i == 'HIC'){
-    next
-  }
-  covars[ , paste0('region', i)] <- as.numeric(covars$region == i)
-}
-
-
-# Read in data
-moddat <- merge(fies_subnat, 
-                covars, 
-                all.x=T, all.y=F) %>%
-  na.omit %>%
-  data.frame
-
-preddat <- covars %>%
-  data.frame
-
-exc <- c()
-#exc <- c('mal_falciparum', 'mal_vivax', 'wasting', 'crop_prod')
-exc <- c('crops_prod', 'forest', 'builtup', 'livestock',
-         'pasture', 'crops_prod', 'cropland',
-         'mal_vivax', 'stunting', 'mal_falciparum')
-
-# Set up Model
-vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
-                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
-                                            'Urban', 'Rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region',
-                                            names(moddat)[grepl('region', names(moddat))],
-                                            exc)]
-
-# vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur', 
-#                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',                                             
-#                                             'Urban', 'Rural', 'fies.mod', 'fies.sev', 
[…]

### 2020-09-08 thesis/model_lasso.R

@@ -1,0 +1,154 @@
+#######################################################
+# Use LASSO regression with SSP covariates
+# to estimate food insecurity, based on data for
+# modeled for 2020, 2025, and 2030
+#######################################################
+
+# read in data
+moddat <- merge(fies_subnat, 
+                covars, 
+                all.x=T, all.y=F) %>%
+  na.omit %>%
+  data.frame
+
+preddat <- covars %>%
+  data.frame
+
+exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
+         'pasture', 'cropland')
+
+# set up model
+vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
+                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
+                                            'urban', 'rural', 'fies.sev', 'fies.mod',
+                                            'population', 'YEAR', 'rural_perc', 'region',
+                                            names(moddat)[grepl('region', names(moddat))],
+                                            exc)]
+
+moddat <- moddat %>%
+  mutate(fies.mod = logit(fies.mod),
+         fies.sev = logit(fies.sev))
+
+###################################
+# Model moderate food insecurity
+##################################
+
+x <- model.matrix(as.formula(paste0('fies.mod ~ ', paste0(vars, collapse=' + '))), 
+                  data=moddat)
+
+mod <- cv.glmnet(x, moddat$fies.mod, alpha=1)
+mod$lambda.min
+
+tmp_coeffs <- coef(mod, s = "lambda.min")
+mdf <- data.frame(term = tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1], estimate =tmp_coeffs@x,
+                  stringsAsFactors=F)
+
+# Get model predictions
+preddat$fies.mod.pred <- mdf$estimate[mdf$term == '(Intercept)']
+moddat$fies.mod.pred <- mdf$estimate[mdf$term == '(Intercept)']
+for (i in 2:nrow(mdf)){
[…]

### 2020-09-08 thesis/model_randomforest_tune.R

@@ -1,0 +1,384 @@
+#######################################################
+# Tune Random Forest model, Find the best model
+#######################################################
+
+covars <- covars %>% merge(ne_countries()@data %>%
+                             select(ISO3=iso_a3, region=region_wb))
+
+ne_countries()@data %>%
+  select(ISO3=iso_a3, region=region_wb)
+
+#Just to fix Siberia
+covars$wci_index[covars$GDLCODE=='RUSr108'] <- 1
+
+covars$gdp_percap <- log(covars$gdp_percap)
+
+for (i in unique(covars$region)){
+  if (i == 'HIC'){
+    next
+  }
+  covars[ , paste0('region', i)] <- as.numeric(covars$region == i)
+}
+
+
+# Play around with model variables
+exc <- list(
+  c(), #1
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #2
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting', 'mal_falciparum'),
+  
+  c('crops_prod', 'builtup', 'livestock', #3
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting'),
+  
+  c('crops_prod', 'builtup', 'livestock', #4
+    'pasture', 'crops_prod', 'cropland',
+    'wasting'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #5
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting', 'mal_falciparum',
+    'school_mean', 'mean_annual_precip',
+    'tavg', 'wci_index', 'ruggedness', 'urban_perc'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #6
+    'pasture', 'crops_prod', 'cropland',
+    'mal_vivax', 'wasting', 'mal_falciparum',
+    'wci_index'),
+  
+  c('crops_prod', 'mal_vivax', 'wasting', #7
+    'mal_falciparum', 'wci_index'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #8
+    'pasture', 'cropland', 'mal_vivax'),
+  
+  c('crops_prod', 'forest', 'builtup', 'livestock', #9
+    'pasture', 'cropland', 'mal_vivax', 'wasting')
+)
+
+tsplit <- c(0.25) #split training and test, tsplit = % of test sample
+ntree <- c(500) #number of trees
[…]

### 2020-09-09 docs/modeling_fies.tex
Updates from Overleaf

--- 2020-09-07 docs/modeling_fies.tex
@@
 However, this overall decline will not be uniform across the world.
-Our model expects that, in sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau.
+Our model expects that, in sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau, although it will improve in some countries, such as Kenya, Senegal, and Ethiopia.
 Food Security will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.
@@
 Compare with SOFI reports, other metrics of hunger.
+
+Dynamics of population growth and development
+
+Deal with scale - why modeling at admin1 is a different exercise than local level modeling
+
+What is driving the increases?
+What will drive the decreases? (caveats of RF method)
 
@@
 
-Random Forest
+- tried different modelling approaches: lasso, bayes ssvs and random forest
 
-Hyperparameter selection
+- random forest performed best in terms of r2/mse and predictions
+
+- random forest used because it captures interactions, machine learning approach
+
+- tuned hyperparameters based on OOB error rate for different combinations, mtry and nodesize
+
+- estimated 5000 trees
+
+- awesome in-sample fit, overfitting?
+
+- out-of-sample fit also very good, r2 around 0.99
+
+- variable importance and coefficient table?
 

### 2020-09-10 data/covars/scripts/tavg_precip.R
begin processing temps

@@ -1,0 +1,142 @@
+library(tidyverse)
+library(raster)
+library(rgdal)
+library(gdalUtils)
+library(foreach)
+library(doParallel)
+
+summarizeYear <- function(path){
+  s <- stack(path)
+  indices <- as.numeric(format(as.Date(names(s), format = "X%Y.%m.%d"), format = "%Y"))
+  year_sum <- stackApply(s, indices, fun=mean, na.rm=T)
+  year_sum
+}
+
+writeStack <- function(x, path){
+  for (n in names(x)){
+    s <- x[[n]]
+    writeRaster(s, paste0('proc/',
+                          path, '_', n, '.tif'),
+                format='GTiff')
+
+  }
+}
+
+start <- Sys.time()
+extract(s, matrix(c(15, 15, 16, 16), ncol=2))
+end <- Sys.time()
+
+start - end
+
+#######################################
+# Process future data from Ensembles
+######################################
+setwd('~/mortalityblob/scenarios/rcp60')
+
+mod1_2020s <- summarizeYear('GFDL-ESM2M/tas_day_GFDL-ESM2M_rcp60_r1i1p1_EWEMBI_landonly_20210101-20301231.nc4')
+writeStack(mod1_2020s, 'mod1_tave')
+
+mod2_2020s <- summarizeYear('HadGEM2-ES/tas_day_HadGEM2-ES_rcp60_r1i1p1_EWEMBI_landonly_20210101-20301231.nc4')
+writeStack(mod2_2020s, 'mod2_tave')
+
+mod3_2020s <- summarizeYear('IPSL-CM5A-LR/tas_day_IPSL-CM5A-LR_rcp60_r1i1p1_EWEMBI_landonly_20210101-20301231.nc4')
+writeStack(mod3_2020s, 'mod3_tave')
+
+mod4_2020s <- summarizeYear('MIROC5/tas_day_MIROC5_rcp60_r1i1p1_EWEMBI_landonly_20210101-20301231.nc4')
+writeStack(mod4_2020s, 'mod4_tave')
+
+#######################################
+# Process future data from Ensembles
+######################################
+mod1_2020s <- summarizeYear('GFDL-ESM2M/pr_day_GFDL-ESM2M_rcp60_r1i1p1_EWEMBI_landonly_20210101-20301231.nc4')
+writeStack(mod1_2020s, 'mod1_precip')
+
[…]

### 2020-09-11 docs/modeling_fies.tex
Account for Covid

--- 2020-09-07 docs/modeling_fies.tex
@@
 \subsection{Disaggregation}
+The data on the FIES collected by Gallup
+
 Assumptions
@@
 	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
-	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
-	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
-	\item Was there a time when your household ran out of food because of a lack of money or other resources?
-	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
-	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
-\end{enumerate}
+- out-of-sample fit also very good, r2 around 0.99
 
-\end{document}
+- variable importance and coefficient table?
+
+\section{Conclusion}
+
+\printbibliography
+
+\section*{Supplemental Info}

### 2020-09-11 data/covars/scripts/=e gdl.R

@@ -1,0 +1,92 @@
+setwd('~/wdl-fies')
+
+library(raster)
+library(rgdal)
+library(tidyverse)
+
+sp <- readOGR('data/GDL Shapefiles V4 0.005', 'GDL Shapefiles V4 Edit')
+
+stunting <- stack(list.files('data/covars/rawdata/stunting', full.names=T))
+wasting <- stack(list.files('data/covars/rawdata/wasting', full.names=T))
+      
+es <- raster::extract(stunting, sp, method='simple', fun=mean, na.rm=T,
+             sp=TRUE, df=TRUE)
+
+ew <- raster::extract(wasting, sp, method='simple', fun=mean, na.rm=T,
+             sp=TRUE, df=TRUE)
+
+
+em <- merge(es@data, ew@data) %>%
+  dplyr::select(GDLCODE=GDLcode, matches("IHME")) %>%
+  gather(var, val, -GDLCODE) %>%
+  mutate(YEAR = as.numeric(str_match(var, "MEAN_(\\d+)")[ , 2]),
+         var = ifelse(grepl('STUNT', var), 'stunting', 'wasting')) %>%
+  spread(var, val)
+
+#Set developing countries to 0
+em$stunting[is.na(em$stunting)] <- 0
+em$wasting[is.na(em$wasting)] <- 0
+
+##################################################
+# Now project to 2030 using methodlogy
+##################################################
+
+logit <- function(x){
+  log(x/(1-x))
+}
+
+inv.logit <- function(x){
+  1/(1 + exp(-x))
+}
+
+modelFuture <- function(var){
+  #Model Future using methodology from Osgood-Zimmerman
+  #Get year on year Average Rate of Change (AROC)
+  #Get weighted average rate of change,
+  #Then apply it to future
+
+  #Var must be in order sequentially,
+  # with only non-NAs preceeding NAs
+  # must be a fraction (rate, or incidence)
+
+  dat <- var[!is.na(var)]
+  nas <- sum(is.na(var))
+
+  roc <- mapply(function(y1, y2) logit(y2) - logit(y1),
+         dat[1:(length(dat) -1)],
+         dat[2:length(dat)])
+
+  w <- (2:length(dat))/sum(2:length(dat)) 
+  
+  aroc <- sum(w*roc)
+
[…]

### 2020-09-11 data/covars/scripts/gdp.R

@@ -187,8 +187,39 @@ comb <- merge(ssp, gdp_proc, all=T) %>%
   mutate(gdp = GDP*admin_gdp_share) %>%
   merge(pop %>%
           select(GDLCODE, ISO3, YEAR, population)) %>%
-  mutate(gdp_percap = gdp/population,
-         gdp_percap = ifelse(is.infinite(gdp_percap), NA, gdp_percap)) %>%
-  select(ISO3, YEAR, GDLCODE, gdp_percap)
+  select(ISO3, YEAR, GDLCODE, gdp, population)
 
-write.csv(comb, 'data/covars/results/gdp.csv', row.names=F)
+############################################
+# Impact of COVID!
+#############################################
+get_roc <- function(x){
+  c(NA, diff(x))/c(NA, x[-length(x)])
+}
+
+apply_roc <- function(val, roc){
+  newvals <- c(val[1])
+  for (i in 2:length(val)){
+    newvals[i] <- newvals[i-1]*(1 + roc[i])
+  }
+  newvals
+}
+
+covid <- read.csv('data/covars/rawdata/gdp/covid_impact.csv') %>%
+  gather(YEAR, roc_pred, -ISO3) %>%
+  mutate(YEAR = as.numeric(substr(YEAR, 2, 5)),
+         roc_pred = roc_pred/100) %>%
+  filter(YEAR %in% c(2020, 2021))
+
+#Get yoy percentage changes
+final <- comb %>%
+  arrange(GDLCODE, YEAR) %>%
+  group_by(GDLCODE) %>%
+  mutate(roc = get_roc(gdp)) %>%
+  merge(covid, all.x=T, all.y=F) %>%
+  group_by(GDLCODE) %>%
+  mutate(roc = ifelse(is.na(roc_pred), roc, roc_pred),
+         gdp = apply_roc(gdp, roc),
+         gdp_percap = gdp/population) %>%
+  select(GDLCODE, YEAR, gdp_percap)
+
+write.csv(final, 'data/covars/results/gdp.csv', row.names=F)

### 2020-09-11 data/covars/scripts/prep_gdp.R

@@ -1,0 +1,88 @@
+library(rnaturalearth)
+library(tidyverse)
+library(sf)
+library(countrycode)
+library(readxl)
+
+sp <- read_sf('~/wdl-fies/data/GDL Shapefiles V4 0.005/', 'GDL Shapefiles V4 Edit')
+
+d <- ne_countries(returnclass='sf') %>% 
+  select(ISO3=iso_a3, region_wb) %>% 
+  st_drop_geometry
+
+gdp <- read_xlsx('~/wdl-fies/data/covars/rawdata/gdp/GlobalEconomicProspectsJune2020GDPgrowthdata.xlsx',
+                 sheet='Statistical Appendix', skip=4) 
+names(gdp)[1:9] <- c('bb', 'region', 'country', 'x', 'r2017', 'r2018', 'r2019', 'r2020', 'r2021')
+
+#########################
+# Get Country-level Data
+#########################
+cty <- gdp %>%
+  select(country, matches('r20')) %>%
+  filter(!is.na(country)) %>%
+  mutate(ISO3 = countrycode(country, 'country.name', 'iso3c'))
+
+cty$ISO3[cty$country == 'Kosovo'] <- 'KSV'
+
+euro <- c("AUT", "BEL", "BGR", "HRV", "CYP", "CZE", "DNK", "EST", "FIN", "FRA", "DEU", "GRC", "HUN", "IRL", "ITA", "LVA", "LTU", "LUX", "MLT", "NLD", "POL", "PRT", "ROU", "SVK", "SVN", "ESP", "SWE", "GBR")
+euro <- euro[!euro %in% cty$ISO3]
+
+eur <- data.frame(cty)[rep(which(cty$country == 'Euro Area'), 28), ]
+eur$ISO3 <- euro
+
+cty <- bind_rows(cty %>% filter(country != 'Euro Area'),
+                 eur)
+#################################
+# Ger Region-level Dta
+#################################
+reg <- gdp %>%
+  select(region, matches('r20')) %>%
+  filter(!is.na(region))
+
+eap_cty <- d$ISO3[d$region_wb == 'East Asia & Pacific' & !d$ISO3 %in% cty$ISO3]
+#Drop AUS and NZL
+eap_cty <- eap_cty[!eap_cty %in% c('AUS', 'NZL')]
[…]

### 2020-09-11 models/model_randomforest.R

@@ -92,7 +92,7 @@ write.csv(moddat, 'data/moddat.csv', row.names=F)
 sel <- preddat %>%
   select(ISO3, YEAR, GDLCODE, stunting, urban_perc, fies.mod.pred, population) %>%
   filter(YEAR %in% c(2020, 2025, 2030)) %>%
-  merge(u5.population) %>%
+  merge(u5.population, all.x=T, all.y=F) %>%
   mutate(u5pop.urban = urban_perc*u5_frac*population,
          u5pop.rural = (1 - urban_perc)*u5_frac*population,
          stunting.urban = stunting*u5pop.urban,

### 2020-09-11 docs/modeling_fies.tex
Updates from Overleaf

@@
 
+Dynamics of population growth and development
+
+Deal with scale - why modeling at admin1 is a different exercise than local level modeling
+
+What is driving the increases?
+What will drive the decreases? (caveats of RF method)
+
+Next decade improvements in food insecurity will only fall like 10\%, not major in overall terms and nowhere near ambitious SDG target.
+
+Cliamte change is not yet a major threat on a global scale, although there will be regional impacts
+
 \section{Methods}

### 2020-09-11 docs/modeling_fies.tex
more manuscript updates

@@
 
+NOTE: SHOULD CITE \citep{smith2017world}, a previous study that used the FIES to estimate correlates of food insecurity.
+
 \section{Results}
@@
 \subsection{Disaggregation}
-The data on the FIES collected by Gallup
-
-Assumptions
+The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
+Using national data on these variables, a standard weighting schema is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
+We used a similar methodology, except at a subnational level.
+We used year-specific subnational estimates of population percentages by gender, age, urbanization, and, where available, wealth, and created a separate set of weights for each admin area across all individuals.
+For age and gender, we used gridded data from WorldPop \citep{Tatem2017}, for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data from Demographic and Health Surveys \citep{dhsall}.
+This methodology rests on the mild assumption that subnational differences in rates of food insecurity are more due to factors related to demographics, urbanization, and wealth than other unobservable factors.
 
 \subsection{Covariates}
+We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, and the environment (See Table \ref{tab:covars}.
+We aimed for covariates that had subnational spatial resolution or peer-reviewed projections available through the year 2030.
+For many variables, we combined historic subnational data with national projections through the 2020s to derive simple subnational estimates for the next decade.
+For projections based on the Shared Socioeconomic Pathways (SSP) framework, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway 6.
+Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020june}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index (CITATION).
+For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.
 
-Selection and Processing
-
-(Maybe put some in Appendix)
-
-Cite determinants of FIES form \citep{smith2017world}.
+\begin{table}[H]
+	\begin{tabular}{llll}
+		\toprule
+		Name & Source \\
+		\midrule
+		Population & \citep{Smits2019, KC2017} \\
+		Urban Percentage & \citep{Jonesi2016} \\
+		Stunting & \citep{Local2020} \\
+		Wasting & \citep{Local2020} \\
+		Mean Years of Schooling & \citep{Smits2019, KC2017} \\
+		GDP Per Capita & \citep{Smits2019, Dellink2017} \\
+		Gini Coefficient & \citep{Rao2019a} \\
+		Poverty Headcount Index & (CITATION) \\
+		Water Scarcity & (CITATION) \\
+		Mean Annual Precipitation &  \cite{abatzoglou2018terraclimate, warszawski2014inter} \\
+		Topographic Ruggedness &  \cite{USGS1996, Riley1999} \\
+		Mean Temperature &  \cite{abatzoglou2018terraclimate, warszawski2014inter} \\
+		Malaria (\textit{P. falciparum}) Mortality Rate &  \cite{Weiss2019} \\
+		\bottomrule
+	\end{tabular}
+	\caption{Covariates Included in Present-Day Analysis}
+	\label{tab:covars1}
+\end{table}
 
@@
 \setcounter{figure}{0}
+\setcounter{section}{0}
 \renewcommand{\thetable}{S\arabic{table}}
 \renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thesection}{S\arabic{section}}
 
-\subsection{Component Questions of the FIES}
+\section{Component Questions of the FIES}
 \begin{enumerate}
@@
 	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
-- out-of-sample fit also very good, r2 around 0.99
+	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
+	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
+	\item Was there a time when your household ran out of food because of a lack of money or other resources?
+	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
+	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
+\end{enumerate}
 
-- variable importance and coefficient table?
+\section{Covariates}
+\subsection{Population}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ \citep{Smits2019, KC2017}
 
-\section{Conclusion}
+\subsection{Urban Percentage}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ \citep{Jonesi2016}
 
-\printbibliography
+\subsection{Stunting}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ \citep{Local2020}
 
-\section*{Supplemental Info}
+\subsection{Wasting}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ \citep{Local2020}
+
+\subsection{Mean Years of Schooling}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ \citep{Smits2019, KC2017}
+
+\subsection{GDP Per Capita}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ \citep{Smits2019, Dellink2017}
+
+\subsection{Gini Coefficient}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ \citep{Rao2019a}
+
+\subsection{Poverty Headcount Index}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ (CITATION)
+
+\subsection{Water Scarcity}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+ (CITATION)
+
+\subsection{Mean Annual Precipitation}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+  \cite{abatzoglou2018terraclimate, warszawski2014inter}
+
+\subsection{Topographic Ruggedness}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+  \cite{USGS1996, Riley1999}
+
+\subsection{Mean Temperature}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+  \cite{abatzoglou2018terraclimate, warszawski2014inter}
+
+\subsection{Malaria (\textit{P. falciparum}) Mortality Rate}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
+Panel (B) shows regional totals compared against each other over time.}
+  \label{fig:timeseries}
+\end{figure}
+  \cite{Weiss2019}
+
+\end{document}

### 2020-09-11 data/covars/scripts/=e gdl.R

@@ -1,92 +1,0 @@
-setwd('~/wdl-fies')
-
-library(raster)
-library(rgdal)
-library(tidyverse)
-
-sp <- readOGR('data/GDL Shapefiles V4 0.005', 'GDL Shapefiles V4 Edit')
-
-stunting <- stack(list.files('data/covars/rawdata/stunting', full.names=T))
-wasting <- stack(list.files('data/covars/rawdata/wasting', full.names=T))
-      
-es <- raster::extract(stunting, sp, method='simple', fun=mean, na.rm=T,
-             sp=TRUE, df=TRUE)
-
-ew <- raster::extract(wasting, sp, method='simple', fun=mean, na.rm=T,
-             sp=TRUE, df=TRUE)
-
-
-em <- merge(es@data, ew@data) %>%
-  dplyr::select(GDLCODE=GDLcode, matches("IHME")) %>%
-  gather(var, val, -GDLCODE) %>%
-  mutate(YEAR = as.numeric(str_match(var, "MEAN_(\\d+)")[ , 2]),
-         var = ifelse(grepl('STUNT', var), 'stunting', 'wasting')) %>%
-  spread(var, val)
-
-#Set developing countries to 0
-em$stunting[is.na(em$stunting)] <- 0
-em$wasting[is.na(em$wasting)] <- 0
-
-##################################################
-# Now project to 2030 using methodlogy
-##################################################
-
-logit <- function(x){
-  log(x/(1-x))
-}
-
-inv.logit <- function(x){
-  1/(1 + exp(-x))
-}
-
-modelFuture <- function(var){
-  #Model Future using methodology from Osgood-Zimmerman
-  #Get year on year Average Rate of Change (AROC)
-  #Get weighted average rate of change,
-  #Then apply it to future
-
-  #Var must be in order sequentially,
-  # with only non-NAs preceeding NAs
-  # must be a fraction (rate, or incidence)
-
-  dat <- var[!is.na(var)]
-  nas <- sum(is.na(var))
-
-  roc <- mapply(function(y1, y2) logit(y2) - logit(y1),
-         dat[1:(length(dat) -1)],
-         dat[2:length(dat)])
-
-  w <- (2:length(dat))/sum(2:length(dat)) 
-  
-  aroc <- sum(w*roc)
-
[…]

### 2020-09-11 viz/Viz_Mod_Results.R

@@ -74,8 +74,8 @@ totals <- preddat %>%
             sev.total=sum(fies.sev.pred * (population), na.rm=t)) %>%
   gather(var, value, -YEAR, -region) %>%
   mutate(var = ifelse(grepl('mod', var), "Moderate-to-Severe", "Severe")) %>%
-  group_by(region, var)# %>%
-  #mutate(value = rollapply(value, width=3, FUN=mean, align='center', partial=TRUE))
+  group_by(region, var) %>%
+  mutate(value = rollapply(value, width=3, FUN=mean, align='center', partial=TRUE))
 
 
 stack <- ggplot(totals) +
@@ -113,22 +113,22 @@ plot_grid(plot_grid(stack, lines, align='v', nrow=2, labels='AUTO'),
           reg, ncol=1, rel_heights=c(5, 1))
 ggsave('TimeSeries.pdf', width=7, height=7)
 
-# Figure out whassup with south asia in 2017
-sa <- preddat %>%
-  filter(region == 'South Asia') %>%
-  select(stunting, wasting, school_mean, hci, gdp_percap, gini, YEAR, population) %>% 
-  group_by(YEAR) %>%
-  summarize(stunting=weighted.mean(stunting, w=population),
-            wasting=weighted.mean(wasting, w=population),
-            school_mean=weighted.mean(school_mean, w=population),
-            hci=weighted.mean(hci, w=population),
-            gdp_percap=weighted.mean(gdp_percap, w=population),
-            gini=weighted.mean(gini, w=population)) %>%
-  mutate_at(vars(-group_cols()),function(x){x/max(x)}) %>%
-  gather(key, value, -YEAR)
-
-ggplot(sa) +
-  geom_line(aes(x=YEAR*2030, y=value, color=key))
+# # Figure out whassup with south asia in 2017
+# sa <- preddat %>%
+#   filter(region == 'South Asia') %>%
+#   select(stunting, wasting, school_mean, hci, gdp_percap, gini, YEAR, population) %>% 
+#   group_by(YEAR) %>%
+#   summarize(stunting=weighted.mean(stunting, w=population),
+#             wasting=weighted.mean(wasting, w=population),
+#             school_mean=weighted.mean(school_mean, w=population),
+#             hci=weighted.mean(hci, w=population),
+#             gdp_percap=weighted.mean(gdp_percap, w=population),
+#             gini=weighted.mean(gini, w=population)) %>%
[…]

### 2020-09-11 viz/covar_maps.R

@@ -20,8 +20,8 @@ for (n in v){
 
   ggplot() + 
     geom_sf(data=covar_map, aes(fill=mapvar), size=0.05) +
-    labs(fill=n,
-         title=n) + 
+    labs(fill='',
+         title='') + 
     scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", 
                                    "#e4ea9a", "#fbf8c0", "#fce08a", "#faae61", 
                                    "#f36c44", "#a01c44")) +
@@ -31,5 +31,5 @@ for (n in v){
           plot.title = element_text(hjust = 0.5)) + 
     facet_wrap(~ YEAR, nrow=2, ncol=2)
 
-  ggsave(paste0('figures/covar_maps/', n, '.png'), width=18, height=9.5)
+  ggsave(paste0('docs/img/covars/', n, '.png'), width=7, height=5)
 }

### 2020-09-15 docs/modeling_fies.tex
write discussion and appendix

@@
 \usepackage{longtable}
+\usepackage{pdflscape}
 \usepackage{lineno}
@@
 We find significant heterogeneity in levels of food security around the world, ranging from parts of the global north with less than 1\% of the population food insecure to parts of the global south where as much as 3 out of 4 people are severely food insecure.
-Examining global temporal trends, we find that food insecurity has been increasing over the past several years, but under middle-of-the-road assumptions for development and population change, food insecurity will decline throughout the 2020s.
+Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that food insecurity has been increasing over the past several years, but under middle-of-the-road assumptions for development and population change, food insecurity will decline throughout the 2020s.
 This global decrease is largely driven by trends in South and East Asia, and some parts of the world, particularly sub-Saharan Africa, are projected to show increases in the prevalence of food insecurity throughout the next decade.
@@
 \section{Introduction}
-Being food secure is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG2) is ``Zero Hunger".
+Being food secure is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is ``Zero Hunger".
 While a number of indicators are used to track progress towards this goal, the indicator that is most indicative of the human experience of food insecurity is the Food Insecurity Experience Scale, or FIES.
@@
 
-Since 2014, in partnership with Gallup World Poll, the FAO conducted surveys around the world in 77 countries and nearly 200 languages to estimate the prevalence of food insecurity using the FIES.
-Using this data, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
+Since 2014, in partnership with Gallup World Poll, the FAO has published surveys from around the world in 77 countries to estimate the prevalence of food insecurity using the FIES.
+With this data, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
 Our analysis highlights several important global and regional trends in food security.
@@
 
-\begin{figure}[H]
+\begin{landscape}
+\begin{figure}[h]
   \centering
-  \includegraphics[width=\linewidth]{img/FullMap.pdf}
+  \includegraphics{img/FullMap.pdf}
   \caption{Modeled rates of severe and moderate-to-severe food insecurity for the years 2010, 2020, and 2030}
@@
 \end{figure}
+\end{landscape}
 
@@
 However, this overall decline will not be uniform across the world.
-Our model expects that, in sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau.
+Our model expects that, in most of sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau, although some countries, such as Kenya, Ethiopia, and Ghana, will see improvements.
 Food Security will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.
@@
 
+This results of this analysis can be explored in more detail on the \href{worldhunger.io}{World Hunger Clock}, including statistics for each subnational administrative area.
+
 \section{Discussion}
-Compare with SOFI reports, other metrics of hunger.
+This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts out to the year 2030.
+We found that, while food insecurity has been increasing and has worsened in 2020 with the novel coronavirus, our models predict that number of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in south and east Asia.
+Nevertheless, many regions, most notably sub-Saharan Africa, are expected to see overall increases in the number of food insecure people.
 
-Dynamics of population growth and development
+While this is the first comprehensive projection of indicator 2.1.2 for SDG 2, projections for other indicators for that goal have been made.
+The most recent State of Food Insecurity (SOFI) report by the FAO projected indicator 2.1.1, the prevalence of undernourishment, and found that that, much like food insecurity, undernourishment had been increasing throughout the 2010s \citep{sofi2020}.
+The SOFI report also modeled the prevalence of undernourishment in 2030 based on recent trends, giving more weight to recent years, and projected that undernourishment will increase by 2030 \citep[Part 1, Page 11]{sofi2020}.
+Using a similar method for modeling the prevalence of stunting (indicator 2.2.1), the SOFI reported that the rate of stunting would decrease by 2030 \citep[Part 1, Page 30]{sofi2020}.
 
-Deal with scale - why modeling at admin1 is a different exercise than local level modeling
+The fact that the SOFI expects increases in undernourishment while we predict decreases in food insecurity is likely due to our differing modeling approaches.
+While the SOFI report extrapolated recent trends in undernourishment, we modeled food insecurity based on projections of factors like GDP per capita, population, and the prevalence of stunting.
+We found that, in a random forest model, these predictors were very successful at predicting recent rates of both moderate-to-severe and severe food insecurity.
 
-What is driving the increases?
-What will drive the decreases? (caveats of RF method)
-
-Next decade improvements in food insecurity will only fall like 10\%, not major in overall terms and nowhere near ambitious SDG target.
-
-Cliamte change is not yet a major threat on a global scale, although there will be regional impacts
+While we found that food insecurity will decrease overall by 2030, our middle-of-the-road scenario only expects the number of people experiencing moderate or severe food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.
+This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, far short of the SDG 2 goal of ``zero hunger."  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.
+Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done in pursuit of SDG 2.
 
@@
 Using national data on these variables, a standard weighting schema is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
-We used a similar methodology, except at a subnational level.
+We used a similar methodology, except we calculated post-stratification weights at a subnational level.
 We used year-specific subnational estimates of population percentages by gender, age, urbanization, and, where available, wealth, and created a separate set of weights for each admin area across all individuals.
@@
 For projections based on the Shared Socioeconomic Pathways (SSP) framework, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway 6.
-Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020june}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index (CITATION).
+Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index (CITATION).
 For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.
@@
 \begin{table}[H]
-	\begin{tabular}{llll}
+	\begin{tabular}{ll}
 		\toprule
@@
 		\midrule
-		Population & \citep{Smits2019, KC2017} \\
-		Urban Percentage & \citep{Jonesi2016} \\
+		Urban Percentage & \citep{Jones2016} \\
 		Stunting & \citep{Local2020} \\
@@
 		Gini Coefficient & \citep{Rao2019a} \\
-		Poverty Headcount Index & (CITATION) \\
+    Poverty Headcount Index & \citep{Cuaresma2018} \\
 		Water Scarcity & (CITATION) \\
@@
 	\caption{Covariates Included in Present-Day Analysis}
-	\label{tab:covars1}
+	\label{tab:covars}
 \end{table}
@@
 Hyperparameter selection
+
+Population \citep{Smits2019, KC2017}
 
@@
 
-\section{Covariates}
+\section{Modeling Covariates Into the Future}
+\subsection{Annualized Rate of Change}
+For many of our covariates, to extrapolate recent trends to the future, it is necessary to draw on annualized rates of change, which are then used to anticipate the rate of future change.
+This method can be used for any variable that is a fraction or a rate.
+We use this method to estimate future rates of stunting, wasting, and malaria mortality, as well as to model future admin shares of GDP and population.
+Thus, we give an overview of the methodology here.
+
+First, for each admin area, $a$, the rate of change (ROC) is calculated between each pair of adjacent years, $y$, based on a value, $p$, such that $0 < p < 1$, available for each year.
+
+\begin{equation}
+  \text{ROC}_{a,y} = \text{logit} \left( \frac{p_{a,y}}{p_{a,y-1}} \right)
+  \label{eqn:a}
+\end{equation}
+
+Then, each ROC is weighted to give more weight to recent years, such that weight $W$ at year $y$ is.  For a dataset with observations for 2010 to 2017, that would be:
+
+\begin{equation}
+  W_y = \frac{y - 2010}{\sum_{y=2010}^{2017} y - 2010}
+  \label{eqn:b}
+\end{equation}
+
+Then, the annualized rate of change (AROC) is calculated as:
+
+\begin{equation}
+  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2017} W_y \times \text{AROC}_{a, y} \right)
+  \label{eqn:c}
+\end{equation}
+
+Finally, the projections, Proj, for each year from 2018 to 2030 are given as:
+
+\begin{equation}
+  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2017} ) + AROC_{a} \times ( y - 2017 ) )
+  \label{eqn:d}
+\end{equation}
+
+For cases where we are modeling shares of a whole, for example when modeling subnational shares of population or GDP, we then re-scale the shares to ensure they total to 1.
+
 \subsection{Population}
+Although population is not a covariate in our random forest model, it is an important precursor to other covariates, such as GDP per capita, and future estimates of population are also necessary to convert modeled future rates of food insecurity into total headcounts of food insecurity.
+Thus, in addition to our other covariates, we also modeled population into the future.
+
+We do this by combining historic subnational data from the Subnational Development Database \citep{Smits2019} with national level population estimates for the 21st century \citep{KC2017}.
+To account for within-country trends and changes in population distribution, we use the AROC method outlined in equations \ref{eqn:a} - \ref{eqn:d} to project each admin areas share of national population totals.
+We then disaggregate the national totals given by KC et al.
+to estimate future subnational populations.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/population.png}
+  \caption{Population}
 \end{figure}
- \citep{Smits2019, KC2017}
 
 \subsection{Urban Percentage}
+
+We drew our historic and future estimates of urbanization entirely from Jones and O'Neill, published in \textit{Environmental Research Letters} \citep{Jones2016}, who modeled spatially explicit scenarios of urbanization consistent with the Shared Socioeconomic Pathways.
+As with our other covariates, we used estimates for SSP2, the middle-of-the-road scenario.
+Because definitions of ``urban" and ``rural" vary significantly across datasets, using one dataset that is entirely consistent within itself was an important consideration.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/urban_perc.png}
+  \caption{Percentage of people living in urban areas.}
 \end{figure}
- \citep{Jonesi2016}
+
+\subsection{Wasting}
+To estimate the prevalence of wasting for each administrative area in our dataset, we used data from the Local Burden of Disease project published in \textit{Nature} \citep{Local2020}.
+For the years 2010 to 2017, we simply took the mean rate of wasting in each administrative area from the dataset.
+Higher-income countries that were not included in the dataset were modeled as having a rate of 0 wasting.
+We then modeled wasting for the years 2018-2030 using the AROC method, which the Local Burden of Disease group similarly used to estimate wasting for the year 2025 \citep{Local2020}.
+
+To account for the effects of the novel coronavirus on global rates of wasting, we assumed that long-term trends and rates of wasting would hold steady, but that they increase globally by 14.3\% in 2020, based on estimates published in \textit{The Lancet} \citep{headey2020impacts}.
+We model this impact as uniform across all countries where wasting occurs.
+Then we model the rate of wasting in each administrative area in 2021 as being the mean of the rate in 2020 that accounts for the shock and the previously modeled rate for 2022.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/covars/wasting.png}
+  \caption{Prevalence of wasting.}
+\end{figure}
 
 \subsection{Stunting}
+
+We model stunting using the same methodology we use to summarize and model wasting, because the Local Burden of Disease dataset that we draw on for wasting also includes stunting.
+We also included the 14.3\% increase in prevalence for the year 2020 that was estimated for wasting \citep{headey2020impacts}, even though stunting is a long-term consequence of malnutrition and will likely not be as readily observable in a population as wasting would be.
+However, we use stunting in our model not as a cause of food insecurity, but rather as a proxy for chronic conditions of hunger, which are exacerbated by the coronavirus pandemic.
+Thus, estimating an impact of the coronavirus shock on stunting is appropriate for our model.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/stunting.png}
+  \caption{Prevalence of stunting.}
 \end{figure}
- \citep{Local2020}
 
-\subsection{Wasting}
+\subsection{GDP Per Capita}
+We estimate global subnational GDP per capita using historic subnational estimates of Gross National Income (GNI) from the Subnational Human Development Database \cite{Smits2019} and forecasts of national GDP within the SSP framework from Dellink et al.
+\cite{Dellink2017}.
+We first use data from the World Bank to estimate the mean country-level ratio of GNI to GDP and adjust the subnational estimates of GNI according to that country-specific ratio.
+We then calculate each admin area's share of country GDP, and then use the AROC method to estimate each admin area's share of total country GDP.
+Based on these trajectories in each admin areas share of total country GDP, we estimate what each admin area's share of GDP will be for each year from 2019 to 2030.
+We divide that share of GDP by the population of the administrative area, which we had previously estimated.
+Thus, national GDP totals are consistent with the projections of Dellink et al., but are disaggregated according to the historic patters of GDP share in the Subnational Human Development Database.
+
+Finally, we used estimates of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the coronavirus \citep{prospects2020}.
+We then model GDP as resuming its previous growth trajectory based on previously-estimated growth rates from 2022-2030, but growing from an estimate for 2021 that includes the coronavirus shock.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/gdp_percap.png}
+  \caption{GDP per capita.}
 \end{figure}
- \citep{Local2020}
 
 \subsection{Mean Years of Schooling}
+We estimate global subnational mean years of schooling using historic subnational estimates from the Subnational Human Development Database \cite{Smits2019} and forecasts of national mean years of schooling from KC et al.
+\cite{KC2017}.
+Using the historic subnational data, we estimate each subnational administrative area's difference in years of schooling from the national mean, and use this value to disaggregate the future projections to a subnational level.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/school_mean.png}
+  \caption{Mean years of schooling.}
 \end{figure}
- \citep{Smits2019, KC2017}
 
-\subsection{GDP Per Capita}
+\subsection{Gini Coefficient}
+We use estimates of national income inequality from Rao et al. \citep{Rao2019a}.  These estimates include all years in our analysis, and are not disaggregated to a subnational level.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/gini.png}
+  \caption{Gini Coefficient}
 \end{figure}
- \citep{Smits2019, Dellink2017}
 
-\subsection{Gini Coefficient}
+\subsection{Poverty Headcount Index}
+We used estimates of the Poverty Headcount Index for the recent past and future from Cuaresma et al.
+\citep{Cuaresma2018}, updated to include the impact of the coronavirus.
+We include the Poverty Headcount Index in our model at the national level.
+SOMEBODY FAMILIAR WITH THE WORLD POVERTY CLOCK FEEL FREE TO ADD MORE.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/hci.png}
+  \caption{Poverty Headcount Index}
 \end{figure}
- \citep{Rao2019a}
 
-\subsection{Poverty Headcount Index}
+\subsection{Water Scarcity}
+SOMEBODY FAMILIAR WITH THE WATER CLOCK METHODOLOGY, FEEL FREE TO ADD MORE.
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/wci_index.png}
+  \caption{Water Scarcity Index}
 \end{figure}
- (CITATION)
 
-\subsection{Water Scarcity}
+\subsection{Mean Annual Precipitation}
+For mean annual precipitation, we use data for the years 2010 to 2019 from the TerraClimate dataset \citep{abatzoglou2018terraclimate}, summarized to each admin area.
+For data for the years 2020-2030, we use models from the Inter-Sectoral Model Intercomparison Project \citep{warszawski2014inter}.
+Specifically, we use an ensemble of bias-corrected projections under representative concentration pathway (RCP) 6.0, taking the mean of the ensemble.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/precip.png}
+  \caption{Mean annual precipitation}
 \end{figure}
- (CITATION)
 
-\subsection{Mean Annual Precipitation}
+\subsection{Mean Temperature}
+For mean temperature, we used the same datasets that we'd used for precipitation, as each of those datasets included both temperature in addition to precipitation.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/tave.png}
+  \caption{Mean temperature (Celsius)}
 \end{figure}
-  \cite{abatzoglou2018terraclimate, warszawski2014inter}
 
 \subsection{Topographic Ruggedness}
+As a measure of accessibility, we include the mean topographic ruggedness of each admin area.
+We first used a gridded dataset of elevation from the USGS \cite{USGS1996} and calculated the index at each grid cell using the methodology from Riley et al.
+\citep{Riley1999}.
+We then summarized the values of all the grid cells within each administrative area.
+Because this variable does not change over time, it is constant across all years in the model.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/ruggedness.png}
+  \caption{Topographic ruggedness}
 \end{figure}
-  \cite{USGS1996, Riley1999}
 
-\subsection{Mean Temperature}
+\subsection{Malaria (\textit{P. falciparum}) Mortality Rate}
+We used data from \textit{The Lancet} to estimate the mortality (deaths per 100,000 population per year) due to Malaria \citep{Weiss2019}, taking the average pixel values within each subnational administrative area.
+We then estimated future malaria mortality based on AROC method.
+
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
+  \includegraphics[width=\linewidth]{img/covars/mal_falciparum.png}
+  \caption{Rate of mortality due to \textit{P. falciparum} Malaria}
 \end{figure}
-  \cite{abatzoglou2018terraclimate, warszawski2014inter}
-
-\subsection{Malaria (\textit{P. falciparum}) Mortality Rate}
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
-Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
-Panel (B) shows regional totals compared against each other over time.}
-  \label{fig:timeseries}
-\end{figure}
-  \cite{Weiss2019}
 

### 2020-09-15 viz/compare_Undernourishment.R

@@ -1,0 +1,47 @@
+library(tidyverse)
+
+setwd('~/wdl-fies')
+
+ours <- read.csv('data/preddat.csv') %>%
+  group_by(YEAR) %>%
+  summarize(Mod.Number = sum(fies.mod.pred*population, na.rm=T),
+            Sev.Number = sum(fies.sev.pred*population, na.rm=T),
+            Mod.Prevalence = weighted.mean(fies.mod.pred, w=population),
+            Sev.Prevalence = weighted.mean(fies.sev.pred, w=population)) %>%
+  gather(Var, Value, -YEAR) %>%
+  separate(Var, c("Level", "Var"), '\\.') %>%
+  mutate(source="Ours")
+
+fao <- data.frame(YEAR = c(2014, 2015, 2016, 2017, 2018, 2019),
+                  Mod.Prevalence = c(0.224, 0.225, 0.232, 0.248, 0.258, 0.259),
+                  Sev.Prevalence = c(0.083, 0.079, 0.081, 0.086, 0.094, 0.097),
+                  Mod.Number = c(1633.5, 1649.5, 1735.2, 1874.5, 1969.6, 2001.1)*1e6,
+                  Sev.Number = c(602, 586, 605.5, 646.4, 717.5, 746)*1e6) %>%
+  gather(Var, Value, -YEAR) %>%
+  separate(Var, c("Level", "Var"), '\\.') %>%
+  mutate(source="FAO")
+
+fies <- bind_rows(ours, fao) %>%
+  filter(YEAR < 2020, YEAR > 2013)
+
+ggplot(fies) + 
+  geom_line(aes(x=YEAR, y=Value, color=Level, linetype=source)) + 
+  facet_wrap(. ~ Var, scales="free_y")
+
+und <- data.frame(YEAR = c(2005, 2010, 2015, 2016, 2017, 2018, 2019, 2030),
+                  Number = c(825.6, 668.2, 653.3, 657.6, 653.2, 678.1, 687.8, 841.4)*1e6,
+                  Prevalence = c(0.126, 0.096, 0.089, 0.088, 0.087, 0.089, 0.089, 0.098)) %>%
+  gather(Var, Value, -YEAR) %>%
+  mutate(source = 'Undernourishment')
+
+sdg2.1 <- bind_rows(ours %>% filter(Level == "Sev"), und)
+
+ggplot(sdg2.1) + 
[…]

### 2020-09-15 viz/covar_maps.R

@@ -10,10 +10,12 @@ exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
 v <- names(covar_map)[!names(covar_map) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
                                             'Urban', 'Rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region', 
+                                            'YEAR', 'rural_perc', 'region', 
                                             'geometry',
                                             names(covar_map)[grepl('region', names(covar_map))],
                                             exc)]
+
+covar_map$mal_falciparum <- covar_map$mal_falciparum*1000000
 
 for (n in v){
   covar_map$mapvar <- covar_map[ , n, drop=TRUE]
@@ -27,9 +29,25 @@ for (n in v){
                                    "#f36c44", "#a01c44")) +
     coord_sf(crs='+proj=robin') +
     theme_void() +
-    theme(legend.position = 'bottom',
-          plot.title = element_text(hjust = 0.5)) + 
+    theme(plot.title = element_text(hjust = 0.5)) + 
     facet_wrap(~ YEAR, nrow=2, ncol=2)
 
   ggsave(paste0('docs/img/covars/', n, '.png'), width=7, height=5)
 }
+
+#Overwrite topographic ruggedness with just one year
+n <- 'ruggedness'
+covar_map$mapvar <- covar_map[ , n, drop=TRUE]
+covar_map <- covar_map %>% filter(YEAR == 2015)
+ggplot() + 
+  geom_sf(data=covar_map, aes(fill=mapvar), size=0.05) +
+  labs(fill='',
+       title='') + 
+  scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", 
+                                 "#e4ea9a", "#fbf8c0", "#fce08a", "#faae61", 
+                                 "#f36c44", "#a01c44")) +
+  coord_sf(crs='+proj=robin') +
+  theme_void() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+ggsave(paste0('docs/img/covars/', n, '.png'), width=7, height=5)

### 2020-09-15 viz/Viz_Mod_Results.R
try sideways maps for more detail

@@ -36,8 +36,8 @@ ggplot(mapdat) +
         plot.title = element_text(hjust = 0.5),
         panel.spacing.x = unit(-0.5, "in")) + 
   labs(fill='') + 
-  facet_grid(YEAR ~ FIES, switch='y')
-ggsave('FullMap.pdf', width=7, dpi=750)
+  facet_grid(FIES ~ YEAR, switch='y')
+ggsave('FullMap.pdf', width=10, dpi=750)
 system('pdfcrop FullMap.pdf FullMap.pdf')
 
 

### 2020-09-15 models/model_randomforest.R
change samplesize of tune, updated figures

@@ -39,7 +39,7 @@ rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod",
                           mtrystart = max(1, floor(sqrt(length(vars)))), 
                           nodesizetry = c(1:3),
                           ntreetry = 5000,
-                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(4/5))),
+                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
                           trace = T,
                           dobest = T); rf.tune.mod$optimal
 
@@ -63,7 +63,7 @@ rf.tune.sev <- tune.rfsrc(formula = as.formula(paste("fies.sev",
                           mtrystart = max(1, floor(sqrt(length(vars)))), 
                           nodesizetry = c(1:3),
                           ntreetry = 5000,
-                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(4/5))),
+                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
                           trace = T,
                           dobest = T); rf.tune.sev$optimal
 # model
@@ -114,7 +114,7 @@ write.csv(sel, 'figures/fies.mod.results_randomforest.csv', row.names=F)
 
 ##################################
 #### plot scaled covariates
-########################################
+##################################
 
 
 # rf.var.sel <- var.select.rfsrc(formula = as.formula(paste("fies.mod", paste(vars, collapse = "+"), sep= "~")),
@@ -122,23 +122,51 @@ write.csv(sel, 'figures/fies.mod.results_randomforest.csv', row.names=F)
 #                                method = "md",
 #                                ntree = 500)
 
-setwd('~/wdl-fies/docs/img')
+#setwd('~/wdl-fies/docs/img')
+setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
 
 # variable importance
-png('mod_vimp.png', width = 1000, height = 500, units="px")
+png('model/mod_vimp.png', width = 1000, height = 500, units="px")
 plot(vimp(rf.mod))
 dev.off()
 
-png('sev_vimp.png', width = 1000, height = 500, units="px")
[…]

### 2020-09-15 thesis/model_randomforest.R

@@ -40,7 +40,7 @@ rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod",
                           mtrystart = max(1, floor(sqrt(length(vars)))), 
                           nodesizetry = c(1:3),
                           ntreetry = 5000,
-                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(4/5))),
+                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
                           trace = T,
                           dobest = T); rf.tune.mod$optimal
 
@@ -50,7 +50,7 @@ rf.tune.sev <- tune.rfsrc(formula = as.formula(paste("fies.sev",
                           mtrystart = max(1, floor(sqrt(length(vars)))), 
                           nodesizetry = c(1:3),
                           ntreetry = 5000,
-                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(4/5))),
+                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
                           trace = T,
                           dobest = T); rf.tune.sev$optimal
 

### 2020-09-15 viz/Viz_Mod_Results.R

@@ -1,7 +1,9 @@
 # setwd('~/wdl-fies'); library(ProjectTemplate); load.project(list(data_loading=T))
 
 #Save outputs to wdl-fies-tex for manuscript
-setwd('~/wdl-fies/docs/img')
+#setwd('~/wdl-fies/docs/img')
+setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
+
 library(cowplot)
 options(scipen=100)
 
@@ -129,32 +131,6 @@ ggsave('TimeSeries.pdf', width=7, height=7)
 # 
 # ggplot(sa) +
 #   geom_line(aes(x=YEAR*2030, y=value, color=key))
-
-
-##################################
-# assess residuals
-##################################
-mae <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred))
-r2 <- cor(moddat$fies.mod, moddat$fies.mod.pred)
-ggplot(moddat) + 
-  geom_point(aes(x=fies.mod, y=fies.mod.pred)) + 
-  labs(title='model performance (random forest)',
-       caption=paste0('mean absolute error: ',  round(mae, 4),
-                      '\nr-squared: ', round(r2, 4)),
-       x='observed rates of mod+sev food insecurity',
-       y='modeled rate of mod+sev food insecurity')	
-ggsave('mod_residuals.png', width=5, height=5)
-
-mae <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred))
-r2 <- cor(moddat$fies.sev, moddat$fies.sev.pred)
-ggplot(moddat) + 
-  geom_point(aes(x=fies.sev, y=fies.sev.pred)) + 
-  labs(title='model performance (random forest)',
-       caption=paste0('mean absolute error: ',  round(mae, 4),
-                      '\nr-squared: ', round(r2, 4)),
-       x='observed rates of sev food insecurity',
-       y='modeled rate of sev food insecurity')	
-ggsave('sev_residuals.png', width=5, height=5)
 
 
 ############################

### 2020-09-15 docs/modeling_fies.tex
added modeling in tex

@@
 \subsection{Covariates}
-We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, and the environment (See Table \ref{tab:covars}.
+We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, and the environment (See Table \ref{tab:covars}).
 We aimed for covariates that had subnational spatial resolution or peer-reviewed projections available through the year 2030.
@@
 
-Random Forest
+Given the subnational FIES data resulting from the disaggregation combined with the covariates stated above (See Table \ref{tab:covars}), we modelled subnational rates of severe and moderate-to-severe food insecurity for the whole world in the period of 2010 to 2030.
+We conduced suitable statistical methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression.
+Based on various model validation metrics, Random Forest (RF) regression performs best which can be seen as a contribution to the increasing importance of such methods in the field (CITATION: Master Thesis M\"uller?).
 
-Hyperparameter selection
+By using the most popular model validation metrics, we observed a in-sample mean squared error (MSE) of 0.00??
+with a R2 of 0.999?
+for the RF regression model.
+Since these models are prone to overfitting (CITATION) it is crucial to check the out-of-sample prediction performance.
+Hence, we trained the RF regression model with 75\% of the data and predicted rates of severe and moderate-to-severe food insecurity for the remaining 25\%.
+We got an out-of-sample MSE of 0.00??
+with a R2 of 0.99??.
+This excellent out-of-sample performance makes the RF regression model particularly suitable for our application.
+See the Supplement Info for a detailed description and validation of all models.
+
+In RF regression models two main hyperparameters can be selected.
+One parameter which describes the number of variables randomly selected as candidates for splitting a node and another which describes the forest average number of unique cases (data points) in a terminal node, “mtry” and “nodesize” respectively (CITATION, manual r package randomforestSRC).
+We calculated the Out-Of-Bag (OOB) error for different combinations of the hyperparameters.
+The OOB error is the mean prediction error of the observation not included in the bootstrap aggregating or bagging (CITATION).
+By choosing the combination of hyperparameters with the lowest OOB error we found the optimal values which we used in our RF regression model to predict subnational rates of severe and moderate-to-severe food insecurity.
 
@@
 
-\section{Modeling Covariates Into the Future}
+\section{Predicting Covariates Into the Future}
 \subsection{Annualized Rate of Change}
@@
 
+\section{Description and Validation of Models}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/residuals.png}
+  \caption{In-Sample Residuals of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+\end{figure}
+
 \end{document}

### 2020-09-15 models/model_randomforest.R

@@ -149,7 +149,7 @@ dev.off()
 ##################################
 mse <- mean(sqrt((moddat$fies.mod - moddat$fies.mod.pred)^2))
 r2 <- cor(moddat$fies.mod, moddat$fies.mod.pred)
-ggplot(moddat) + 
+res.mod <- ggplot(moddat) + 
   geom_point(aes(x=fies.mod, y=fies.mod.pred), alpha = 0.3) + 
   geom_abline(intercept = 0, slope = 1, color = "red", size = 0.5) +
   xlim(0, 1) + ylim(0, 1) +
@@ -157,11 +157,11 @@ ggplot(moddat) +
                       '\nR-Squared: ', round(r2, 4)),
        x='Observed Rates Of Moderat-to-Severe Food Insecurity',
        y='Modeled Rates Of Moderat-to-Severe Food Insecurity')
-ggsave('model/mod_residuals.png', width=5, height=5)
+res.mod; ggsave('model/mod_residuals.png', width=5, height=5)
 
 mse <- mean(sqrt((moddat$fies.sev - moddat$fies.sev.pred)^2))
 r2 <- cor(moddat$fies.sev, moddat$fies.sev.pred)
-ggplot(moddat) + 
+res.sev <- ggplot(moddat) + 
   geom_point(aes(x=fies.sev, y=fies.sev.pred), alpha = 0.3) +
   geom_abline(intercept = 0, slope = 1, color = "red", size = 0.5) +
   xlim(0, 1) + ylim(0, 1) +
@@ -169,4 +169,8 @@ ggplot(moddat) +
                       '\nR-Squared: ', round(r2, 4)),
        x='Observed Rates Of Severe Food Insecurity',
        y='Modeled Rates Of Severe Food Insecurity')	
-ggsave('model/sev_residuals.png', width=5, height=5)
+res.sev; ggsave('model/sev_residuals.png', width=5, height=5)
+
+plot_grid(res.mod, res.sev, align='h',labels='AUTO')
+ggsave('model/residuals.png', width=14, height=7)
+

### 2020-09-16 docs/modeling_fies.tex
Updates from Overleaf
model plots for paper, out-of-sample script

@@
 We conduced suitable statistical methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression.
-Based on various model validation metrics, Random Forest (RF) regression performs best which can be seen as a contribution to the increasing importance of such methods in the field (CITATION: Master Thesis M\"uller?).
+Based on various model validation metrics, Random Forest (RF) regression performs best which can be seen as a contribution to the increasing importance of such methods in the field.
+For a detailed description and comparison of all models check M\"uller (CITATION: Master Thesis M\"uller 2020).
 
-By using the most popular model validation metrics, we observed a in-sample mean squared error (MSE) of 0.00??
-with a R2 of 0.999?
-for the RF regression model.
-Since these models are prone to overfitting (CITATION) it is crucial to check the out-of-sample prediction performance.
+By using the most popular model validation metrics, we observed for the severe as well as for the moderate-to-severe RF regression model a in-sample mean squared error (MSE) of around zero with a R-squared (${R}^2$) of 0.99 (See Fig.
+\ref{fig:rf_in-sample}).
+Although RF regression models avoid overfitting by randomly selecting features at every node split (CITATION: Breiman 2001), it is still crucial to check the out-of-sample prediction performance.
 Hence, we trained the RF regression model with 75\% of the data and predicted rates of severe and moderate-to-severe food insecurity for the remaining 25\%.
-We got an out-of-sample MSE of 0.00??
-with a R2 of 0.99??.
+Similarly to the in-sample MSE and ${R}^2$, the out-of-sample MSE is close to zero with a ${R}^2$ of around 0.99 for both models (See Fig.
+\ref{fig:rf_out-sample}).
 This excellent out-of-sample performance makes the RF regression model particularly suitable for our application.
-See the Supplement Info for a detailed description and validation of all models.
+%See the Supplement Info for a detailed description and validation of all models.
 
 In RF regression models two main hyperparameters can be selected.
-One parameter which describes the number of variables randomly selected as candidates for splitting a node and another which describes the forest average number of unique cases (data points) in a terminal node, “mtry” and “nodesize” respectively (CITATION, manual r package randomforestSRC).
-We calculated the Out-Of-Bag (OOB) error for different combinations of the hyperparameters.
-The OOB error is the mean prediction error of the observation not included in the bootstrap aggregating or bagging (CITATION).
-By choosing the combination of hyperparameters with the lowest OOB error we found the optimal values which we used in our RF regression model to predict subnational rates of severe and moderate-to-severe food insecurity.
+One parameter which describes the number of variables randomly selected as candidates for splitting a node and another which describes the forest average number of unique cases (data points) in a terminal node, “mtry” and “nodesize” respectively (CITATION: Manual R-package randomforestSRC).
+We calculated the Out-Of-Bag (OOB) error for different combinations of the hyperparameters for the severe and moderate-to-severe model.
+The OOB error is the mean prediction error of the observations not included in the bootstrap aggregating or bagging (CITATION: Some fundamental work on Random Forest).
+By choosing the combination of hyperparameters with the lowest OOB error we found the optimal values for both model respectively.
+
+Apart from tuning the hyperparameters, we examined the importance of the individual covariates (See Table \ref{tab:covars}) in explaining rates of severe and moderate-to-severe food insecurity.
+This is entirely based on the prediction error.
+By training the RF regression models with all covariates and predicting an outcome with and without a specific variables we calculated the average prediction error when leaving out that variable.
+According to the highest average prediction error we rank the importance of the models.
+We found that for both, the severe and moderate-to-severe model, the Poverty Headcount Index, Stunting, GDP Per Capita and the Gini Coefficient are the most important variables in explaining rates of severe and moderate-to-severe food insecurity (See Fig.
+\ref{fig:rf_vimp}).
 
@@
   \centering
-  \includegraphics[width=\linewidth]{img/model/residuals.png}
-  \caption{In-Sample Residuals of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
+  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_in-sample}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
+  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_out-sample}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/vimp_rf.png}
+  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_vimp}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/error_rf.png}
+  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_error}
 \end{figure}

### 2020-09-16 models/model_randomforest.R

@@ -6,15 +6,18 @@
 
 # setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
 
-# read in data
-moddat <- merge(fies_subnat, 
-                covars, 
-                all.x=T, all.y=F) %>%
-  na.omit %>%
-  data.frame
+# # read in data
+# moddat <- merge(fies_subnat, 
+#                 covars, 
+#                 all.x=T, all.y=F) %>%
+#   na.omit %>%
+#   data.frame
+# 
+# preddat <- covars %>%
+#   data.frame
 
-preddat <- covars %>%
-  data.frame
+moddat <- read.csv('data/moddat.csv')
+preddat <- read.csv('data/preddat.csv')
 
 exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
          'pasture', 'cropland')
@@ -24,6 +27,7 @@ vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
                                             'urban', 'rural', 'fies.sev', 'fies.mod',
                                             'population', 'YEAR', 'rural_perc', 'region',
+                                            "fies.mod.pred", "fies.sev.pred", #only because cache is not updated
                                             names(moddat)[grepl('region', names(moddat))],
                                             exc)]
 
@@ -82,66 +86,95 @@ moddat <- moddat %>%
   mutate(fies.mod = inv.logit(fies.mod),
          fies.sev = inv.logit(fies.sev))
 
-write.csv(preddat, 'data/preddat.csv', row.names=F)
-write.csv(moddat, 'data/moddat.csv', row.names=F)
+# write.csv(preddat, 'data/preddat.csv', row.names=F)
+# write.csv(moddat, 'data/moddat.csv', row.names=F)
 
 ######################
 # save output for poli
 ########################
 
-sel <- preddat %>%
-  select(ISO3, YEAR, GDLCODE, stunting, urban_perc, fies.mod.pred, population) %>%
-  filter(YEAR %in% c(2020, 2025, 2030)) %>%
-  merge(u5.population, all.x=T, all.y=F) %>%
-  mutate(u5pop.urban = urban_perc*u5_frac*population,
[…]

### 2020-09-16 thesis/model_randomforest_outsamp.R

@@ -1,0 +1,132 @@
+#######################################################
+# Tune Random Forest model, Find the best model
+#######################################################
+
+# # read in data
+# moddat <- merge(fies_subnat, 
+#                 covars, 
+#                 all.x=T, all.y=F) %>%
+#   na.omit %>%
+#   data.frame
+# 
+# preddat <- covars %>%
+#   data.frame
+
+moddat <- read.csv('data/moddat.csv')
+preddat <- read.csv('data/preddat.csv')
+
+exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
+         'pasture', 'cropland')
+
+# set up model
+vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
+                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
+                                            'urban', 'rural', 'fies.sev', 'fies.mod',
+                                            'population', 'YEAR', 'rural_perc', 'region',
+                                            "fies.mod.pred", "fies.sev.pred", #only because cache is not updated
+                                            names(moddat)[grepl('region', names(moddat))],
+                                            exc)]
+
+moddat <- moddat %>%
+  mutate(fies.mod = logit(fies.mod),
+         fies.sev = logit(fies.sev))
+
+
+#take 25% of obersations as test sample
+moddat_test <- sample_frac(moddat, size = 0.25, replace = F)
+moddat_train <- setdiff(moddat, moddat_test)
+moddat_test$i <- 1; moddat_train$i <- 0
+
+
+## mod+sev
+# tune parameters
+rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod", 
+                                                     paste(vars, collapse = "+"), sep= "~")),
+                          data = moddat_train,
+                          mtrystart = max(1, floor(sqrt(length(vars)))), 
+                          nodesizetry = c(1:3),
+                          ntreetry = 5000,
+                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
+                          trace = T,
+                          dobest = T); rf.tune.mod$optimal
+
+# model
[…]

### 2020-09-16 viz/Viz_Mod_Results.R

@@ -4,7 +4,7 @@
 #setwd('~/wdl-fies/docs/img')
 setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
 
-library(cowplot)
+#library(cowplot)
 options(scipen=100)
 
 ############################

### 2020-09-16 docs/modeling_fies.tex
model: main done, supplement still work in progress

--- 2020-09-15 docs/modeling_fies.tex
@@
 
-Random Forest
+Given the subnational FIES data resulting from the disaggregation combined with the covariates stated above (See Table \ref{tab:covars}) we model subnational rates of moderate-to-severe and severe food insecurity for the whole world in the period of 2010 to 2030.
+We conduced suitable statistical methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression.
 
-Hyperparameter selection
+Based on various model validation metrics, Random Forest (RF) regression performs best which can be seen as a contribution to the increasing importance of such methods in the field.
+The RF regression is based on the creation of a large number of decision trees and is known to perform well in the presence of nonlinearities and possible important interaction terms of individual covariates.
+Since this is likely to be the case in our application, this method is particularly suitable for modeling rates of moderate-to-severe and severe food insecurity.
 
-Population \citep{Smits2019, KC2017}
+By using the most popular model validation metrics, we observe for the moderate-to-sever as well as the severe RF regression model a in-sample mean squared error (MSE) of around zero with a R-squared (${R}^2$) value of 0.99 (See Fig.
+\ref{fig:rf_in-sample}).
+Although RF regression models tackle the issue of overfitting by applying various ensemble methods (CITATION: Breiman 2001), it is crucial to validate the out-of-sample prediction performance.
+Hence, we trained the RF regression model with 75\% of the available observations and predicted rates of moderate-to-severe and severe food insecurity for the remaining 25\%.
+Similarly to the in-sample model validation metrics, the out-of-sample MSE is close to zero with a ${R}^2$ of around 0.99 for both models (See Fig.
+\ref{fig:rf_out-sample}).
+
+To get the best possible model performance, it is necessary to tune the hyperparamter.
+In RF regression models two main hyperparameters can be specified.
+One parameter which describes the number of variables randomly selected as candidates for splitting a node and another which describes the forest average number of observations in a leaf node.
+We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe model.
+By choosing the combination of hyperparameters with the smallest OOB error we found the optimal combination for both models respectively.
+Additionally to these two hyperparamters a third parameter describes the number of trees to be created.
+We found that 5000 trees are more than sufficient for the error rates to converge (See Fig.
+\ref{fig:rf_error}).
+
+Apart from tuning the hyperparameters, we examined the importance of the individual covariates in explaining rates of moderate-to-severe and severe food insecurity.
+This is based on the prediction error with and without a specific variable.
+According to the highest average prediction error we ranked the importance of the variables and found that for both, the moderate-to-severe and severe model, the Poverty Headcount Index, Stunting, GDP Per Capita and the Gini Coefficient are the most important variables for explaining differences in rates of moderate-to-severe and severe food insecurity (See Fig.
+\ref{fig:rf_vimp}).
+
+For a detailed description and validation of the RF regression models see the Supplemental Info and for a comparison of all models check M\"uller (CITATION: Master Thesis M\"uller 2020).
 
@@
 \subsection{Poverty Headcount Index}
-We used estimates of the Poverty Headcount Index for the recent past and future from Cuaresma et al.
-\citep{Cuaresma2018}, updated to include the impact of the coronavirus.
-We include the Poverty Headcount Index in our model at the national level.
-SOMEBODY FAMILIAR WITH THE WORLD POVERTY CLOCK FEEL FREE TO ADD MORE.
+We used estimates of the Poverty Headcount Index for the recent past and future from the World Poverty Clock by the World Data Lab (CITATION: World Data Lab), updated to include the impact of the coronavirus.
+In our model we include national level poverty rates at a threshold of \$1.90 which is defined as \textit{extreme} poverty.
+By using the concept of parametrized Lorenz curves in combination with mean income/consumption and population projections future poverty rates are calculated.
+For more details on the methodology and information on global and country level poverty check the website \textit{worldpoverty.io}.
 
@@
 
+\section{Description and Validation of Random Forest regression models}
+
+%RF regression models are based on the creation of a large number of decision trees.
+To boost the performance of such models different ensemble methods are used, where two of them stand out.
+The bagging and
+
+%To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions.
+
+Without providing a detailed theoretical description of the RF regression model, we focus on the implementation and specification of the model.
+For implementation we use the R-package \textit{randomForestSRC} by Ishwaran and Kogalur (CITATION: R-package randomforestSRC).
+Specifically, we apply the functions \textit{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \textit{rfsrc()} to perform the RF regression models.
+After predicting rates of moderate-to-severe and severe food insecurity we plot the modeled vs.
+observed rates and calculated the MSE and ${R}^2$ (See Fig.
+\ref{fig:rf_in-sample} and Fig.
+\ref{fig:rf_out-sample}) for both models.
+Additionally, we use the function \textit{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig.
+\ref{fig:rf_error}) and the importance of individual variables (See Fig.
+\ref{fig:rf_vimp}).
+
+\subsection{Tuning the hyperparamters}
+The optimal hyperparamters \textit{mtry} and \textit{nodesize} are found by calculating the OOB error and choosing the combination with the smallest OOB.
+
+We find the optimal hyperparamters by choosing the combination with the smallest OOB error.
+The \textit{tune.rfsrc()} function does exactly that and returns the optimal values of \textit{mtry} and \textit{nodesize}.
+For the moderate-to-severe as well as for the severe model we specify a number of function arguments.
+INCLUDE ALL FUNCTION ARGUMENTS HERE AND DESCRIBE WHY WE CHOOSE THEM
+
+Since RF regression models use bootstrap aggregation (also called bagging) some observations are not used when building a specific tree.
+These observations are used to calulate the predictions error of a specific tree.
+We get the mean OOB by
+
+Based on these observations
+The OOB error is the mean prediction error of the observations not included in the bootstrap aggregating or bagging (CITATION: Some fundamental work on Random Forest).
+
+\subsection{Variable Importance}
+By training the RF regression models with all covariates and predicting rates with and without a specific variable we calculated the average prediction error when leaving out the specific variable
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
+  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_in-sample}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
+  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_out-sample}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/vimp_rf.png}
+  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_vimp}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/error_rf.png}
+  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_error}
+\end{figure}
+
 \end{document}

### 2020-09-17 docs/modeling_fies.tex
added some stuff in supplement models

@@
 Apart from tuning the hyperparameters, we examined the importance of the individual covariates in explaining rates of moderate-to-severe and severe food insecurity.
-This is based on the prediction error with and without a specific variable.
-According to the highest average prediction error we ranked the importance of the variables and found that for both, the moderate-to-severe and severe model, the Poverty Headcount Index, Stunting, GDP Per Capita and the Gini Coefficient are the most important variables for explaining differences in rates of moderate-to-severe and severe food insecurity (See Fig.
+By training the models with all covariates and predicting rates with and without a individual variable we calculated the average prediction error of omitting that variable.
+Using the highest average prediction error we ranked the importance of the variables and found that for both, the moderate-to-severe and severe model, the Poverty Headcount Index, Stunting, GDP Per Capita and the Gini Coefficient are the most important variables for explaining differences in rates of moderate-to-severe and severe food insecurity (See Fig.
 \ref{fig:rf_vimp}).
 
-For a detailed description and validation of the RF regression models see the Supplemental Info and for a comparison of all models check M\"uller (CITATION: Master Thesis M\"uller 2020).
+For a detailed description, implementation and validation of the RF regression models see the Supplemental Info and for a comparison of all models check M\"uller (CITATION: Master Thesis M\"uller 2020).
 
@@
 
-\section{Description and Validation of Random Forest regression models}
+\section{Description, Implementation and Validation of the RF regression model}
 
-%RF regression models are based on the creation of a large number of decision trees.
-To boost the performance of such models different ensemble methods are used, where two of them stand out.
-The bagging and
+WE NEED HERE A SHORT PARAGRAPH ON HOW RF MODELS WORK, THIS WILL HELP: https://www.kdnuggets.com/2017/10/random-forests-explained.html
 
-%To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions.
+The RF regression is a machine learning method based on the creation of a large number of decision trees.
+Such models are used for prediction purposes and do not describe any relationships in the training data...
 
-Without providing a detailed theoretical description of the RF regression model, we focus on the implementation and specification of the model.
 For implementation we use the R-package \textit{randomForestSRC} by Ishwaran and Kogalur (CITATION: R-package randomforestSRC).
@@
 
-\subsection{Tuning the hyperparamters}
-The optimal hyperparamters \textit{mtry} and \textit{nodesize} are found by calculating the OOB error and choosing the combination with the smallest OOB.
+To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions.
 
-We find the optimal hyperparamters by choosing the combination with the smallest OOB error.
-The \textit{tune.rfsrc()} function does exactly that and returns the optimal values of \textit{mtry} and \textit{nodesize}.
-For the moderate-to-severe as well as for the severe model we specify a number of function arguments.
-INCLUDE ALL FUNCTION ARGUMENTS HERE AND DESCRIBE WHY WE CHOOSE THEM
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/model/error_rf.png}
+  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \label{fig:rf_error}
+\end{figure}
 
-Since RF regression models use bootstrap aggregation (also called bagging) some observations are not used when building a specific tree.
-These observations are used to calulate the predictions error of a specific tree.
-We get the mean OOB by
+HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH
 
-Based on these observations
-The OOB error is the mean prediction error of the observations not included in the bootstrap aggregating or bagging (CITATION: Some fundamental work on Random Forest).
+In addition to the parameter that describes the number of trees to be created, two hyperparameters must be tuned.
+In the function \textit{tune.rfsrc()} these paramters are called \textit{mtry} and \textit{nodesize}.
+\textit{Mtry} describes the number of variables randomly selected as candidates for splitting a node and \textit{nodesize} describes the forest average number of observations in a leaf node.
+The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest OOB error.
 
-\subsection{Variable Importance}
-By training the RF regression models with all covariates and predicting rates with and without a specific variable we calculated the average prediction error when leaving out the specific variable
+EXPLAIN THE OOB ERROR HERE?
+
+The function \textit{tune.rfsrc()} does exactly that and returns optimal values for \textit{mtry} and \textit{nodesize}.
+We applied this function on the both, the moderate-to-severe and the severe model.
+For the modarate-to-severe model the optimal combination is \textit{mtry} = 12 and \textit{nodesize} = 1 and for the severe model we found \textit{mtry} = 10 and \textit{nodesize} = 2 to be optimal.
+
+STATE FUNCTION ARGUMENTS HERE?
 
@@
 
+HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
+For modeling the rates of moderate-to-severe and severe food insecurity we use the function \textit{rfsrc()}. We specify
+
 \begin{figure}[H]
@@
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/model/error_rf.png}
-  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_error}
-\end{figure}
+Lastly we asses the importance of individual variables.
+By training the RF regression models with all covariates and predicting rates with and without a specific variable we calculated the average prediction error when leaving out the specific variable
+
+MORE INFORMATION HERE
 

### 2020-09-17 models/model_randomforest.R

@@ -40,10 +40,11 @@ moddat <- moddat %>%
 rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod", 
                                                      paste(vars, collapse = "+"), sep= "~")),
                           data = moddat,
-                          mtrystart = max(1, floor(sqrt(length(vars)))), 
-                          nodesizetry = c(1:3),
-                          ntreetry = 5000,
+                          mtrystart = ncol(moddat)/2, 
+                          nodesizetry = c(1:5),
+                          ntree = 5000,
                           sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
+                          improve = 1e-3,
                           trace = T,
                           dobest = T); rf.tune.mod$optimal
 
@@ -64,10 +65,10 @@ preddat$fies.mod.pred <- inv.logit(predict(rf.mod, preddat)$predicted)
 rf.tune.sev <- tune.rfsrc(formula = as.formula(paste("fies.sev", 
                                                      paste(vars, collapse = "+"), sep= "~")),
                           data = moddat,
-                          mtrystart = max(1, floor(sqrt(length(vars)))), 
-                          nodesizetry = c(1:3),
-                          ntreetry = 5000,
-                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
+                          #mtrystart = max(1, floor(sqrt(length(vars)))), 
+                          #nodesizetry = c(1:3),
+                          ntreetry = 10000,
+                          #sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
                           trace = T,
                           dobest = T); rf.tune.sev$optimal
 # model

### 2020-09-17 docs/modeling_fies.tex
some comments on the supp info in model section

@@
 WE NEED HERE A SHORT PARAGRAPH ON HOW RF MODELS WORK, THIS WILL HELP: https://www.kdnuggets.com/2017/10/random-forests-explained.html
+INCLUDE BAGGING AND RANDOM FEATURE SELECTION, NECESSARY TO EXPLAIN OOB ERROR
 
@@
 
+STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
+
 To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions. 
+MAYBE STATE THIS SOMEWHERE ELSE, NOT SURE WHERE
 

### 2020-09-18 docs/modeling_fies.tex
final draft of paper for coauthors

--- 2020-09-17 docs/modeling_fies.tex
@@
 \author[2,5]{Juan Carlos Laso}
-\author[2,6]{Homi Kharas}
+\author[2,6]{Wolfgang Fengler}
+\author[2,7]{Homi Kharas}
 
@@
 \affil[5]{The International Institute for Applied Systems Analysis}
-\affil[6]{The Brookings Institute}
+\affil[6]{The World Bank}
+\affil[7]{The Brookings Institute}
 \affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}
@@
 This global decrease is largely driven by trends in South and East Asia, and some parts of the world, particularly sub-Saharan Africa, are projected to show increases in the prevalence of food insecurity throughout the next decade.
-This is the first global picture of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
+This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
 \end{abstract}
 
-\section{Introduction}
-Being food secure is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is ``Zero Hunger".
-While a number of indicators are used to track progress towards this goal, the indicator that is most indicative of the human experience of food insecurity is the Food Insecurity Experience Scale, or FIES.
+\section{Main}
+Being food secure is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is ``Zero Hunger."  While a number of indicators are used to track progress towards this goal, the indicator that is most indicative of the human experience of food insecurity is the Food Insecurity Experience Scale, or FIES.
 The FIES was developed due to a recognized need to measure actual food insecurity based on personal exposure to hunger, rather than proxies for food insecurity, such as estimates of calories per capita, or health consequences of food insecurity, such as child stunting.
-However, as a relatively new metric, little work has been done so far to model food security at a global scale or estimate current trends in the number of food insecure people using the FIES.
+However, as a relatively new metric, little work has been done so far to model food security or estimate current trends using the FIES at more local spatial scales.
 
@@
 However, these metrics are confounded with other determinants of health such as infectious disease and are not meaningful at the scale of individuals or households \citep{Perumal2018}.
-Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they do not indicate how accessible food is to the general population, and food insecurity can certainly occur in the absence of food availability decline \citep{Sen1983}.
+Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and food insecurity can certainly occur in the absence of food availability decline \citep{Sen1983}.
 Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.
@@
 
-Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013}.
+Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013} and quickly became a useful tool for understanding individual and household-level food insecurity \citep{smith2017world}.
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
@@
 
-Since 2014, in partnership with Gallup World Poll, the FAO has published surveys from around the world in 77 countries to estimate the prevalence of food insecurity using the FIES.
-With this data, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
+Since 2014, in partnership with Gallup World Poll, the FAO has published surveys from around the world to estimate the prevalence of food insecurity using the FIES.
+Drawing on surveys from 77 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
 Our analysis highlights several important global and regional trends in food security.
-
-NOTE: SHOULD CITE \citep{smith2017world}, a previous study that used the FIES to estimate correlates of food insecurity.
 
@@
 For age and gender, we used gridded data from WorldPop \citep{Tatem2017}, for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data from Demographic and Health Surveys \citep{dhsall}.
-This methodology rests on the mild assumption that subnational differences in rates of food insecurity are more due to factors related to demographics, urbanization, and wealth than other unobservable factors.
+This methodology rests on the mild assumption that subnational differences in rates of food insecurity are more due to factors related to demographics, urbanization, and wealth than to other unobservable factors.
 
@@
 We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, and the environment (See Table \ref{tab:covars}).
-We aimed for covariates that had subnational spatial resolution or peer-reviewed projections available through the year 2030.
+We aimed for covariates that had subnational spatial resolution and peer-reviewed projections available through the year 2030.
 For many variables, we combined historic subnational data with national projections through the 2020s to derive simple subnational estimates for the next decade.
-For projections based on the Shared Socioeconomic Pathways (SSP) framework, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway 6.
-Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index (CITATION).
+For projections based on the Shared Socioeconomic Pathways (SSP) framework, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway (RCP) 6.0.
+Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index \cite{cuaresma2018will}.
 For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.
@@
 \begin{table}[H]
+  \centering
 	\begin{tabular}{ll}
@@
 
-Given the subnational FIES data resulting from the disaggregation combined with the covariates stated above (See Table \ref{tab:covars}) we model subnational rates of moderate-to-severe and severe food insecurity for the whole world in the period of 2010 to 2030.
-We conduced suitable statistical methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression.
+We modeled subnational rates of moderate-to-severe and severe food insecurity globally at the subnational level in the period of 2010 to 2030 using the disaggregated FIES data combined with the covariates.
+Using the log-transformed rate of food insecurity as our outcome variable, we explored several modeling approaches, including linear regression methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression.
 
-Based on various model validation metrics, Random Forest (RF) regression performs best which can be seen as a contribution to the increasing importance of such methods in the field.
-The RF regression is based on the creation of a large number of decision trees and is known to perform well in the presence of nonlinearities and possible important interaction terms of individual covariates.
+Based on various model validation metrics, Random Forest (RF) regression performed the best.
+The RF regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.
+Random forest models are known to perform well in the presence of non-linearities and important interactions among predictor variables \citep{breiman2001random}.
 Since this is likely to be the case in our application, this method is particularly suitable for modeling rates of moderate-to-severe and severe food insecurity.
 
-By using the most popular model validation metrics, we observe for the moderate-to-sever as well as the severe RF regression model a in-sample mean squared error (MSE) of around zero with a R-squared (${R}^2$) value of 0.99 (See Fig.
+We observed for the moderate-to-severe as well as the severe RF regression model an in-sample mean squared error (MSE) between 0.002-0.005, with a R-squared (${R}^2$) value of 0.999 (See Fig.
 \ref{fig:rf_in-sample}).
-Although RF regression models tackle the issue of overfitting by applying various ensemble methods (CITATION: Breiman 2001), it is crucial to validate the out-of-sample prediction performance.
+Although RF regression models tackle the issue of overfitting by applying various ensemble methods \citep{breiman2001random}, it is crucial to validate the out-of-sample prediction performance.
 Hence, we trained the RF regression model with 75\% of the available observations and predicted rates of moderate-to-severe and severe food insecurity for the remaining 25\%.
-Similarly to the in-sample model validation metrics, the out-of-sample MSE is close to zero with a ${R}^2$ of around 0.99 for both models (See Fig.
+Similarly to the in-sample model validation metrics, the out-of-sample MSE is between 0.008-0.017, with a ${R}^2$ of around 0.98-0.99 for both models (See Fig.
 \ref{fig:rf_out-sample}).
 
-To get the best possible model performance, it is necessary to tune the hyperparamter.
-In RF regression models two main hyperparameters can be specified.
-One parameter which describes the number of variables randomly selected as candidates for splitting a node and another which describes the forest average number of observations in a leaf node.
-We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe model.
+To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.
+Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.
+We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe models.
 By choosing the combination of hyperparameters with the smallest OOB error we found the optimal combination for both models respectively.
-Additionally to these two hyperparamters a third parameter describes the number of trees to be created.
 We found that 5000 trees are more than sufficient for the error rates to converge (See Fig.
@@
 
-For a detailed description, implementation and validation of the RF regression models see the Supplemental Info and for a comparison of all models check M\"uller (CITATION: Master Thesis M\"uller 2020).
+For a detailed description, implementation and validation of the RF regression models see the Supplemental Info.
 
@@
 \subsection{Poverty Headcount Index}
-We used estimates of the Poverty Headcount Index for the recent past and future from the World Poverty Clock by the World Data Lab (CITATION: World Data Lab), updated to include the impact of the coronavirus.
+We used estimates of the Poverty Headcount Index for the recent past and future from the World Poverty Clock by the World Data Lab \citep{cuaresma2018will}, updated to include the impact of the coronavirus.
 In our model we include national level poverty rates at a threshold of \$1.90 which is defined as \textit{extreme} poverty.
-By using the concept of parametrized Lorenz curves in combination with mean income/consumption and population projections future poverty rates are calculated.
-For more details on the methodology and information on global and country level poverty check the website \textit{worldpoverty.io}.
+By using the concept of parameterized Lorenz curves in combination with mean income/consumption and population projections future poverty rates are calculated.
+For more details on the methodology and information on global and country level poverty check the website \url{worldpoverty.io}.
 
@@
 \subsection{Water Scarcity}
-SOMEBODY FAMILIAR WITH THE WATER CLOCK METHODOLOGY, FEEL FREE TO ADD MORE.
 \begin{figure}[H]
@@
 
-WE NEED HERE A SHORT PARAGRAPH ON HOW RF MODELS WORK, THIS WILL HELP: https://www.kdnuggets.com/2017/10/random-forests-explained.html
+%WE NEED HERE A SHORT PARAGRAPH ON HOW RF MODELS WORK, THIS WILL HELP: https://www.kdnuggets.com/2017/10/random-forests-explained.html
+%INCLUDE BAGGING AND RANDOM FEATURE SELECTION, NECESSARY TO EXPLAIN OOB ERROR
 
 The RF regression is a machine learning method based on the creation of a large number of decision trees.
-Such models are used for prediction purposes and do not describe any relationships in the training data...
+By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to single decision trees \citep{friedman2001elements}.
+For each of the decision trees in a random forest model, observations are selected at random with replacement, a method known as bootstrap aggregating, or bagging, and features are area also selected at random.
 
-For implementation we use the R-package \textit{randomForestSRC} by Ishwaran and Kogalur (CITATION: R-package randomforestSRC).
-Specifically, we apply the functions \textit{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \textit{rfsrc()} to perform the RF regression models.
+For implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}.
+Specifically, we apply the functions \texttt{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \texttt{rfsrc()} to perform the RF regression models.
 After predicting rates of moderate-to-severe and severe food insecurity we plot the modeled vs.
@@
 \ref{fig:rf_out-sample}) for both models.
-Additionally, we use the function \textit{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig.
+Additionally, we use the function \texttt{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig.
 \ref{fig:rf_error}) and the importance of individual variables (See Fig.
@@
 
+%STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
+
 To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions. 
+%MAYBE STATE THIS SOMEWHERE ELSE, NOT SURE WHERE
 
@@
 
-HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH
+%HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH
 
-In addition to the parameter that describes the number of trees to be created, two hyperparameters must be tuned.
-In the function \textit{tune.rfsrc()} these paramters are called \textit{mtry} and \textit{nodesize}.
-\textit{Mtry} describes the number of variables randomly selected as candidates for splitting a node and \textit{nodesize} describes the forest average number of observations in a leaf node.
-The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest OOB error.
+In addition to the parameter that describes the number of trees to be created, two other important hyperparameters must be tuned.
+In the function \texttt{tune.rfsrc()} these parameters are called \texttt{mtry} and \texttt{nodesize}.
+\texttt{mtry} describes the number of variables randomly selected as candidates for splitting a node and \texttt{nodesize} describes the average number of observations in a leaf node.
+The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest out-of-bag (OOB) error.
 
-EXPLAIN THE OOB ERROR HERE?
+%EXPLAIN THE OOB ERROR HERE?
 
-The function \textit{tune.rfsrc()} does exactly that and returns optimal values for \textit{mtry} and \textit{nodesize}.
+The function \texttt{tune.rfsrc()} does exactly that and returns optimal values for \texttt{mtry} and \texttt{nodesize}.
 We applied this function on the both, the moderate-to-severe and the severe model.
-For the modarate-to-severe model the optimal combination is \textit{mtry} = 12 and \textit{nodesize} = 1 and for the severe model we found \textit{mtry} = 10 and \textit{nodesize} = 2 to be optimal.
+For the moderate-to-severe model the optimal combination is \texttt{mtry} = 12 and \texttt{nodesize} = 1 and for the severe model we found \texttt{mtry} = 10 and \texttt{nodesize} = 2 to be optimal.
 
-STATE FUNCTION ARGUMENTS HERE?
+%STATE FUNCTION ARGUMENTS HERE?
 
@@
 
-HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
-For modeling the rates of moderate-to-severe and severe food insecurity we use the function \textit{rfsrc()}. We specify
+%HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
+%For modeling the rates of moderate-to-severe and severe food insecurity we use the function \texttt{rfsrc()}. We specify
 
@@
 
+%https://kogalur.github.io/randomForestSRC/theory.html#section4.2
 Lastly we asses the importance of individual variables.
-By training the RF regression models with all covariates and predicting rates with and without a specific variable we calculated the average prediction error when leaving out the specific variable
-
-MORE INFORMATION HERE
+This is based on the Breiman-Culter permutation variable importance information \citep{breiman2001random}.
+This involved comparing the prediction error on the OOB data to the prediction error of OOB cases where the x-variable $x$ is randomly permutated.
+The importance of an individual variable is thus the mean difference between the perturbed and unperturbed error rate, across all trees.
+%MORE INFORMATION HERE
 

### 2020-09-18 models/model_randomforest.R

@@ -138,6 +138,12 @@ rf.mod$xvar.names <- vars_full
 rf.sev$yvar.names <- c("Severe Food Insecurity")
 rf.sev$xvar.names <- vars_full
 
+# hv.mod <- holdout.vimp(formula = as.formula(paste("fies.mod", paste(vars, collapse = "+"), sep= "~")),
+#                 data = moddat,
+#                 ntree = 5000, 
+#                 mtry = rf.tune.mod$optimal[[2]],
+#                 nodesize = rf.tune.mod$optimal[[1]],
+#                 verbose = TRUE)
 
 # error rate and variable importance
 plot(vimp(rf.mod), plots.one.page = F, sorted = T, verbose = T)

### 2020-09-19 docs/modeling_fies.tex
final final lol

@@
 For projections based on the Shared Socioeconomic Pathways (SSP) framework, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway (RCP) 6.0.
-Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index \cite{cuaresma2018will}.
+Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index \cite{Cuaresma2018}.
 For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.
@@
 \subsection{Poverty Headcount Index}
-We used estimates of the Poverty Headcount Index for the recent past and future from the World Poverty Clock by the World Data Lab \citep{cuaresma2018will}, updated to include the impact of the coronavirus.
+We used estimates of the Poverty Headcount Index for the recent past and future from the World Poverty Clock by the World Data Lab \citep{Cuaresma2018}, updated to include the impact of the coronavirus.
 In our model we include national level poverty rates at a threshold of \$1.90 which is defined as \textit{extreme} poverty.

### 2020-10-07 models/model_randomforest.R
updated model and all plots with water scarcity share

@@ -4,116 +4,114 @@
 # modeled for 2020, 2025, and 2030
 #######################################################
 
-# setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
-
+# # setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
+# 
 # # read in data
-# moddat <- merge(fies_subnat, 
-#                 covars, 
+# moddat <- merge(fies_subnat,
+#                 covars,
 #                 all.x=T, all.y=F) %>%
 #   na.omit %>%
 #   data.frame
 # 
 # preddat <- covars %>%
 #   data.frame
-
-moddat <- read.csv('data/moddat.csv')
-preddat <- read.csv('data/preddat.csv')
-
-exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
-         'pasture', 'cropland')
-
-# set up model
-vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
-                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
-                                            'urban', 'rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region',
-                                            "fies.mod.pred", "fies.sev.pred", #only because cache is not updated
-                                            names(moddat)[grepl('region', names(moddat))],
-                                            exc)]
-
-moddat <- moddat %>%
-  mutate(fies.mod = logit(fies.mod),
-         fies.sev = logit(fies.sev))
-
-## mod+sev
-# tune parameters
-rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod", 
-                                                     paste(vars, collapse = "+"), sep= "~")),
-                          data = moddat,
-                          mtrystart = ncol(moddat)/2, 
-                          nodesizetry = c(1:5),
-                          ntree = 5000,
-                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
-                          improve = 1e-3,
-                          trace = T,
-                          dobest = T); rf.tune.mod$optimal
-
-# model
[…]

### 2020-10-07 munge/03-Covariates.R

@@ -12,7 +12,7 @@ for (f in fs){
 covars <- covars %>%
   select(YEAR, ISO3, GDLCODE, stunting, wasting, school_mean, ruggedness,
          gdp_percap, gini, mal_falciparum, precip, hci, tave, 
-         population, rural_perc, urban_perc, wci_index, builtup, cropland, crops_prod,
+         population, rural_perc, urban_perc, ws_share, builtup, cropland, crops_prod,
          forest, livestock, pasture) %>%
   merge(regions)
 
@@ -41,7 +41,7 @@ covars$gdp_percap[covars$ISO3=='SSD'] <- rep(caf_gdp_percap, each=n_admin1)
 covars$gini[covars$ISO3=='SSD'] <- rep(caf_gini, each=n_admin1)
 
 ######## North Korea #############
-#WCI, School mean, GDP Percap, same as Tajikstan
+#Water Scarcity Share, School mean, GDP Percap, same as Tajikstan
 #Malaria is 0
 covars$mal_falciparum[covars$ISO3=='PRK'] <- 0
 
@@ -51,12 +51,12 @@ tjk <- covars %>%
   summarize(school_mean = weighted.mean(school_mean, w=population),
             gdp_percap = weighted.mean(gdp_percap, w=population),
             gini = weighted.mean(gini, w=population),
-            wci_index = weighted.mean(wci_index, w=population))
+            ws_share = weighted.mean(ws_share, w=population))
 
 covars$school_mean[covars$ISO3=='PRK'] <- tjk$school_mean
 covars$gdp_percap[covars$ISO3=='PRK'] <- tjk$gdp_percap
 covars$gini[covars$ISO3=='PRK'] <- tjk$gini
-covars$wci_index[covars$ISO3=='PRK'] <- tjk$wci_index
+covars$ws_share[covars$ISO3=='PRK'] <- tjk$ws_share
 
 ###########################################################################
 # Missing data at the region level
@@ -110,3 +110,4 @@ covars <- covars %>%
 covars <- na.omit(covars)
 
 cache('covars')
+

### 2020-10-07 viz/Viz_Mod_Results.R
new images

@@ -1,8 +1,8 @@
 # setwd('~/wdl-fies'); library(ProjectTemplate); load.project(list(data_loading=T))
 
 #Save outputs to wdl-fies-tex for manuscript
-#setwd('~/wdl-fies/docs/img')
-setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
+setwd('~/wdl-fies/docs/img')
+#setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
 
 #library(cowplot)
 options(scipen=100)
@@ -30,7 +30,8 @@ ggplot(mapdat) +
   scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", "#e4ea9a", "#fbf8c0", 
                                  "#fce08a", "#faae61", "#f36c44", "#a01c44"), 
                        breaks=seq(0, 1, by=0.2),
-                       limits=c(0, max(mapdat$value))) + 
+                       limits=c(0, max(mapdat$value)),
+                       labels=function(x){paste0(x*100, '%')}) + 
   geom_sf(data=cty, color='#000000', fill=NA, size=0.15) + 
   coord_sf(crs='+proj=robin', expand=FALSE) + 
   theme_void() + 
@@ -39,7 +40,7 @@ ggplot(mapdat) +
         panel.spacing.x = unit(-0.5, "in")) + 
   labs(fill='') + 
   facet_grid(FIES ~ YEAR, switch='y')
-ggsave('FullMap.pdf', width=10, dpi=750)
+ggsave('FullMap.pdf', width=10)
 system('pdfcrop FullMap.pdf FullMap.pdf')
 
 

### 2020-10-13 docs/modeling_fies.tex

@@
 The RF regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.
-Random forest models are known to perform well in the presence of non-linearities and important interactions among predictor variables \citep{breiman2001random}.
+Random forest models are known to perform well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}.
 Since this is likely to be the case in our application, this method is particularly suitable for modeling rates of moderate-to-severe and severe food insecurity.
@@
 \ref{fig:rf_in-sample}).
-Although RF regression models tackle the issue of overfitting by applying various ensemble methods \citep{breiman2001random}, it is crucial to validate the out-of-sample prediction performance.
-Hence, we trained the RF regression model with 75\% of the available observations and predicted rates of moderate-to-severe and severe food insecurity for the remaining 25\%.
-Similarly to the in-sample model validation metrics, the out-of-sample MSE is between 0.008-0.017, with a ${R}^2$ of around 0.98-0.99 for both models (See Fig.
+To test the out-of-sample prediction performance we used ten-fold cross-validation.
+First, we partitioned the data into ten equal sized subsamples.
+Once partitioned, we trained the RF regression models on nine subsamples and tested the models on the remaining one.
+This process gets repeated ten times, with each subsample being used once for testing.
+We then calculated the average MSE and ${R}^2$ over all ten out-of-sample predictions.
+Similarly to the in-sample model validation metrics, the average out-of-sample MSE is between 0.008-0.016, with an average ${R}^2$ of around 0.99 for both models (See Fig.
 \ref{fig:rf_out-sample}).
+Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries.
+We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.
 
@@
 
-\subsection{Water Scarcity}
+\subsection{Water Scarcity Share}
 \begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{img/covars/wci_index.png}
-  \caption{Water Scarcity Index}
+  \includegraphics[width=\linewidth]{img/covars/ws_share.png}
+  \caption{Water Scarcity Share}
 \end{figure}

### 2020-10-21 viz/Blog_Graph.R
final blog and start of methodology for tool

@@ -1,6 +1,7 @@
 # setwd('~/wdl-fies'); library(ProjectTemplate); load.project(list(data_loading=T))
 
 setwd('~/wdl-fies/docs/img')
+setwd('./docs/img')
 
 options(scipen=100)
 
@@ -20,15 +21,15 @@ totals <- preddat %>%
   mutate(value = rollapply(value, width=3, FUN=mean, align='center', partial=TRUE))
 
 ggplot(totals) + 
-  geom_line(aes(x=YEAR, y=value, color=var), size=2) + 
-  scale_color_manual(values=c("Moderate And Severe"='#ff7f00',
-                              "Severe"='#e41a1c')) + 
+  geom_line(aes(x=YEAR, y=value, linetype=var), size=1.2) + 
+  #scale_color_manual(values=c("Moderate And Severe"='#ff7f00',
+  #                            "Severe"='#e41a1c')) + 
   scale_x_continuous(expand=c(0,0)) +
   scale_y_continuous(expand=expansion(mult=c(0,0.05), add=0), 
                      labels=function(x){prettyNum(x, big.mark=',')},
                      limits=c(0, max(totals$value))) +
-  theme_bw() + 
+  theme(legend.title=element_blank()) + 
   labs(y="Number of People in Food Insecurity", title="", 
        x='Year', color='')
 
-ggsave('Blog_Graph.png')
+ggsave('Blog_Graph.png', width=7, height=4)

### 2020-10-21 models/model_randomforest.R
write methods for dashboard and move model viz to viz folder

@@ -4,211 +4,83 @@
 # modeled for 2020, 2025, and 2030
 #######################################################
 
-# # setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
-# 
-# # read in data
-# moddat <- merge(fies_subnat,
-#                 covars,
-#                 all.x=T, all.y=F) %>%
-#   na.omit %>%
-#   data.frame
-# 
-# preddat <- covars %>%
-#   data.frame
-# 
-# exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
-#          'pasture', 'cropland')
-# 
-# # set up model
-# vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
-#                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
-#                                             'urban', 'rural', 'fies.sev', 'fies.mod',
-#                                             'population', 'YEAR', 'rural_perc', 'region',
-#                                             names(moddat)[grepl('region', names(moddat))],
-#                                             exc)]
-# 
-# moddat <- moddat %>%
-#   mutate(fies.mod = logit(fies.mod),
-#          fies.sev = logit(fies.sev))
-# 
-# ## mod+sev
-# # tune parameters
-# rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod", 
-#                                                      paste(vars, collapse = "+"), sep= "~")),
-#                           data = moddat,
-#                           mtryStart = ncol(moddat)/2, 
-#                           nodesizeTry = c(1:3),
-#                           ntree = 5000,
-#                           sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
-#                           trace = T,
-#                           dobest = T); rf.tune.mod$optimal
-# 
-# # model
-# rf.mod <- rfsrc(formula = as.formula(paste("fies.mod", paste(vars, collapse = "+"), sep= "~")),
-#                 data = moddat,
-#                 ntree = 5000, 
-#                 mtry = rf.tune.mod$optimal[[2]],
-#                 nodesize = rf.tune.mod$optimal[[1]],
-#                 do.trace = TRUE)
-# 
[…]

### 2020-10-21 viz/Error_and_VIP.R

@@ -1,0 +1,129 @@
+
+
+preddat <- read.csv('data/preddat.csv')
+moddat <- read.csv('data/moddat.csv')
+
+######################
+# save output for poli
+########################
+
+sel <- preddat %>%
+  select(ISO3, YEAR, GDLCODE, stunting, urban_perc, fies.mod.pred, population) %>%
+  filter(YEAR %in% c(2020, 2025, 2030)) %>%
+  merge(u5.population, all.x=T, all.y=F) %>%
+  mutate(u5pop.urban = urban_perc*u5_frac*population,
+         u5pop.rural = (1 - urban_perc)*u5_frac*population,
+         stunting.urban = stunting*u5pop.urban,
+         stunting.rural = stunting*u5pop.rural,
+         population.urban = population*urban_perc,
+         population.rural = population*(1 - urban_perc),
+         fies.urban = fies.mod.pred*population.urban,
+         fies.rural = fies.mod.pred*population.rural) %>%
+  select(-stunting, -urban_perc, -fies.mod.pred, -population, -u5_frac) %>%
+  gather(var, value, -ISO3, -YEAR, -GDLCODE) %>%
+  mutate(geo_area=ifelse(grepl('urban', var), 'Urban', 'Rural'),
+         var=gsub('.rural|.urban', '', var),
+         value = round(value)) %>%
+  spread(var, value)
+
+write.csv(sel, 'figures/HC_2020-10-07.csv', row.names=F) #update date
+
+
+
+##########################################
+#### plot error and variable importance
+##########################################
+
+#setwd('~/wdl-fies/docs/img')
+setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
+
+library(cowplot)
+library(gridGraphics)
+options(scipen=100)
+
+# change varaibles names
+vars_full <- c("Stunting", "Wasting", "Mean Years of Schooling", "Topographic Ruggedness",
+              "GDP Per Capita", "Gini Coefficient", "Malaria Mortality Rate", "Mean Annual Precipitation",
+              "Poverty Headcount Index", "Mean Temperature", "Urban Percentage", "Water Scarcity") #has to be in the same order as "vars"
+
[…]

### 2020-10-22 docs/modeling_fies.tex
Updates from Overleaf

@@
 
+%do we need to add a section where we state which fies data we use? like countries and years?
+
 \subsection{Disaggregation}
@@
 \ref{fig:rf_in-sample}).
-To test the out-of-sample prediction performance we used ten-fold cross-validation.
+To test the out-of-sample prediction performance we performed ten-fold cross-validation.
 First, we partitioned the data into ten equal sized subsamples.
@@
 \ref{fig:rf_out-sample}).
-Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries.
-We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.
+Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED?) and performed ten-fold cross-validation for this limited set of countries.
+We found that excluding HIC does not change the out-of-sample performance of the models.
+The average out-of-sample MSE is between 0.013-0.023 with an average ${R}^2$ of around 0.98 for both models.
 
@@
 Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.
-We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe models.
+For the latter two we calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe models.
 By choosing the combination of hyperparameters with the smallest OOB error we found the optimal combination for both models respectively.
-We found that 5000 trees are more than sufficient for the error rates to converge (See Fig.
+For the number of decision trees we found that 5000 trees are more than sufficient for the error rates to converge (See Fig.
 \ref{fig:rf_error}).

### 2020-10-27 docs/modeling_fies.tex
update manuscript; add new regions; work on graphs

--- 2020-10-13 docs/modeling_fies.tex
@@
 \documentclass{article}
+
 \usepackage{graphicx}
@@
 \addbibresource{library.bib}
+
+% TODO:
+% - re-run model
+%   - Use validation with FAO data, if possible
+% - discuss validation more
+%   - various exclusions: HICs
+%   - Compare to country reports
+% - get new results:
+%   - new regions
+%   - Get rates over time
+% - describing FIES
+%   - Go over different data sources
+% - add some kind of framework for selecting covars
+% - make appendix separate
+% - add higher level overview of modeling the covars
 
@@
 \section{Main}
-Being food secure is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is ``Zero Hunger."  While a number of indicators are used to track progress towards this goal, the indicator that is most indicative of the human experience of food insecurity is the Food Insecurity Experience Scale, or FIES.
-The FIES was developed due to a recognized need to measure actual food insecurity based on personal exposure to hunger, rather than proxies for food insecurity, such as estimates of calories per capita, or health consequences of food insecurity, such as child stunting.
-However, as a relatively new metric, little work has been done so far to model food security or estimate current trends using the FIES at more local spatial scales.
+Food security is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is “Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
+The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience it.
+Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories consumed per capita in a population.
+These do not capture the actual lived exposure of individuals to food shortages.
+
+Because the FIES is based on responses to polls, the data is limited to 77 countries where the polls have been conducted and microdata has been vetted and released by the FAO.
+The current practice is to assign the regional average prevalence rate for countries without survey data in order to compile regional and global aggregates.
+This is not very satisfactory.
+In addition, because the FIES is a relatively new metric, it does not have a sufficiently long time series to estimate trends with any degree of precision.
+To understand whether SDG 2 is being met, one needs to estimate food insecurity outcomes in 2030.
+
+The contribution of this paper is to fill these two gaps of coverage and forecasts by developing a model of FIES outcomes that is based on covariates that are available for places where actual polling data is not available, and that can be forecast to 2030 to get a sense of what food insecurity at that time might look like.
+We also take advantage of the within-country spatial distribution of FIES reports to push the analysis beyond country level to a sub-national level, that is administratively once removed from the national level (Admin-1).
+To our knowledge, this is the first effort to model the FIES outcomes at that scale.
 
@@
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
-Using a Rash model \citep{Cafiero2018}, each individual in the survey is given a severity score, and thresholds for moderate and severe food insecurity are set at the behaviors of eating less than the respondent thought they should, and not eating for a while day, respectively.
-Thus, people in moderate-to-severe food insecurity are estimated to have not eaten as much as they should have at least once in the previous year, while people in severe food insecurity are estimated to have gone an entire day without eating in the previous year.
+Using a Rash model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing he responses given to the 8 questions asked in the survey.
+Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalences can be computed.
+A moderately food insecure adult has compromised on food quality and variety, as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
+A severely food insecure adult has run out of food and has gone an entire day without eating at times during the year.
 
@@
 Drawing on surveys from 77 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
-Our analysis highlights several important global and regional trends in food security.
+The model is then used to infer levels of food insecurity around the world at the subnational level and to forecast food insecurity to the year 2030.
 
 \section{Results}
-\subsection{Spatial Distribution of Food Insecurity}
-We found significant heterogeneities in the global distribution of severe food insecurity, estimated to be people who had to go an entire day without eating at least once in the previous year (See Fig.
-\ref{fig:map}).
-For the year 2020, mainland sub-Saharan Africa is the continent with with the highest rates of severe food insecurity, with at least 15\% of people in severe food insecurity in at last one region in every country except Gabon and Equatorial Guinea.
-Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.
-In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10-15\% in areas like northern Brazil, many central Asian and middle-eastern countries, and India and Indonesia.
+% Maybe pull out individual interesting scenarios?
+Globally, in the year 2020, we estimate that 829 million people are above the threshold for severe food insecurity, and an additional 1.5 billion people are above the threshold for moderate food insecurity (See Fig.
+\ref{fig:timeseries}).
+These numbers are higher than the estimates in the 2019 FAO State of Food Insecurity report, which estimated about 700 million severely food insecure people and an additional 1.31 billion people experiencing at least moderate food insecurity in 2018.
 
-The experience of at least moderate food insecurity, or eating less than one would prefer, is quite common in many parts of the world.
-In 2020, it is quite common in Africa, south and southeast Asia, and parts of Latin America, where in many places over 50\% of the population experiences this degree of food insecurity.
-Even in countries widely considered to be developed, such as Australia, the United States, and parts of eastern and southern Europe, moderate-to-severe food insecurity is as high as 10-15\%.
-
-\begin{landscape}
-\begin{figure}[h]
-  \centering
-  \includegraphics{img/FullMap.pdf}
-  \caption{Modeled rates of severe and moderate-to-severe food insecurity for the years 2010, 2020, and 2030}
-  \label{fig:map}
-\end{figure}
-\end{landscape}
-
-Globally, in the year 2020, we estimate that 2.33 billion people experience at least moderate food insecurity, and 829 million people experience severe food insecurity (See Fig.
-\ref{fig:timeseries}).
-While the total number of people in moderate and severe food security held roughly constant for the 2010s and even increased in latter part of that decade, we predict that it will decline throughout the 2020s.
+While the total number of people experiencing moderate and severe food insecurity held roughly constant for the 2010s and even increased in latter part of that decade, we predict that it will decline throughout the 2020s.
 However, this overall decline will not be uniform across the world.
 Our model expects that, in most of sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau, although some countries, such as Kenya, Ethiopia, and Ghana, will see improvements.
-Food Security will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.
+Food insecurity will fall in most other world regions, particularly South Asia and East Asia & the Pacific.
 
@@
   \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people in moderate-to-severe and severe food insecurity, by World Bank regions over time, with a 3-year smoothing.
+  \caption{Number of people over the thresholds for moderate and severe food insecurity, by UN SDG regions over time, with a 3-year smoothing.
 Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
@@
 
-This results of this analysis can be explored in more detail on the \href{worldhunger.io}{World Hunger Clock}, including statistics for each subnational administrative area.
+We found substantial heterogeneity in the global distribution of severe food insecurity (See Fig.
+\ref{fig:map}).
+For the year 2020, mainland sub-Saharan Africa is the continent with with the highest rates of severe food insecurity, with at least 15\% of people over the threshold for severe food insecurity in at last one subnational area in every country except Gabon and Equatorial Guinea.
+Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.
+In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10-15\% in areas like northern Brazil, many central Asian and middle-eastern countries, and India and Indonesia.
+
+The experience of at least moderate food insecurity is quite common in many parts of the world.
+In 2020, much of Africa, south and southeast Asia, and parts of Latin America had over 50\% of the population living with moderate food insecurity.
+Even in countries widely considered to be developed, such as Australia, the United States, and parts of eastern and southern Europe, over 10\% of the population is above the threshold for moderate food insecurity.
+
+\begin{landscape}
+\begin{figure}[h]
+  \centering
+  \includegraphics{img/FullMap.png}
+  \caption{Modeled spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
+  \label{fig:map}
+\end{figure}
+\end{landscape}
+
+The results of this analysis can be further explored in more detail on the \href{worldhunger.io}{World Hunger Clock}, including statistics for each subnational administrative area.
 
@@
 This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts out to the year 2030.
-We found that, while food insecurity has been increasing and has worsened in 2020 with the novel coronavirus, our models predict that number of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in south and east Asia.
+We found that, while food insecurity has been increasing and has likely worsened in 2020 with economic downturn associated with the novel coronavirus, our models predict that number of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in south and east Asia.
 Nevertheless, many regions, most notably sub-Saharan Africa, are expected to see overall increases in the number of food insecure people.
+%Add here comparison with SOFI, but emphasize that discrepancies may also come from measuring different things
 
+%Homi says delete this para
 While this is the first comprehensive projection of indicator 2.1.2 for SDG 2, projections for other indicators for that goal have been made.
@@
 
-While we found that food insecurity will decrease overall by 2030, our middle-of-the-road scenario only expects the number of people experiencing moderate or severe food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.
-This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, far short of the SDG 2 goal of ``zero hunger."  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.
+%Add section about the drivers of food insecurity, as well as how they largely account for the outcomes observed.
+
+While we found that food insecurity will decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, only expects the number of people experiencing moderate food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.
+This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, a number that falls far short of the SDG 2 goal of ``zero hunger."  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.
 Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done in pursuit of SDG 2.
@@
 The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
-Using national data on these variables, a standard weighting schema is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
+Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
 We used a similar methodology, except we calculated post-stratification weights at a subnational level.
-We used year-specific subnational estimates of population percentages by gender, age, urbanization, and, where available, wealth, and created a separate set of weights for each admin area across all individuals.
-For age and gender, we used gridded data from WorldPop \citep{Tatem2017}, for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data from Demographic and Health Surveys \citep{dhsall}.
+We used year-specific subnational estimates of population percentages by gender, age, urbanization, and, where available, wealth, and created a separate set of weights for each subnational area across all individuals.
+For age and gender, we used gridded data from WorldPop \citep{Tatem2017}, for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}, where available.
 This methodology rests on the mild assumption that subnational differences in rates of food insecurity are more due to factors related to demographics, urbanization, and wealth than to other unobservable factors.
@@
 \subsection{Covariates}
-We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, and the environment (See Table \ref{tab:covars}).
+We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).
 We aimed for covariates that had subnational spatial resolution and peer-reviewed projections available through the year 2030.
-For many variables, we combined historic subnational data with national projections through the 2020s to derive simple subnational estimates for the next decade.
 For projections based on the Shared Socioeconomic Pathways (SSP) framework, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway (RCP) 6.0.
+
+Many of the variables required harmonizing subnational historic data with projections at the national level.
+This included mean years of schooling, GDP per capita, as well as for projections of population.
+We used observed trajectories in the historical distribution of variables among subnational areas within a country to estimate the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.
+
+In cases of health variables that have shown long-term trends and represented a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these variables for 2030.
+The involves taking the rate of change between each pair of years in the dataset, and then taking the mean rate of change over the period for which data is available, giving greater weight to more recent years.
+This method has been used to estimate rates of stunting and wasting in 2025 \citep{Local2020}; we simply use the same method to make forecasts to the year 2030.
+
+For the climate variables temperature and precipitation, we combined historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
+
 Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index \cite{Cuaresma2018}.
@@
   \centering
-	\begin{tabular}{ll}
+	\begin{tabular}{lll}
 		\toprule
-		Name & Source \\
+    Name & Source & Scale \\
 		\midrule
-		Urban Percentage & \citep{Jones2016} \\
-		Stunting & \citep{Local2020} \\
-		Wasting & \citep{Local2020} \\
-		Mean Years of Schooling & \citep{Smits2019, KC2017} \\
-		GDP Per Capita & \citep{Smits2019, Dellink2017} \\
-		Gini Coefficient & \citep{Rao2019a} \\
-    Poverty Headcount Index & \citep{Cuaresma2018} \\
-		Water Scarcity & (CITATION) \\
-		Mean Annual Precipitation &  \cite{abatzoglou2018terraclimate, warszawski2014inter} \\
-		Topographic Ruggedness &  \cite{USGS1996, Riley1999} \\
-		Mean Temperature &  \cite{abatzoglou2018terraclimate, warszawski2014inter} \\
-		Malaria (\textit{P. falciparum}) Mortality Rate &  \cite{Weiss2019} \\
+    Urban Percentage & \citep{Jones2016} & Subnational \\
+    Stunting & \citep{Local2020} & Subnational \\
+    Wasting & \citep{Local2020} & Subnational \\
+    Mean Years of Schooling & \citep{Smits2019, KC2017} & Subnational\\
+    GDP Per Capita & \citep{Smits2019, Dellink2017} & Subnational \\
+    Gini Coefficient & \citep{Rao2019a} & National\\
+    Poverty Headcount Index & \citep{Cuaresma2018} & National \\
+    Water Scarcity & \citep{greve2018global} & Subnational \\
+    Mean Annual Precipitation &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
+    Topographic Ruggedness &  \cite{USGS1996, Riley1999} & Subnational \\
+    Mean Temperature &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
+    Malaria (\textit{P. falciparum}) Mortality Rate &  \cite{Weiss2019} & Subnational \\
 		\bottomrule
@@
 \subsection{Modeling}
+We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
+We used random forest regressions, as this approach performs well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}, while providing more model interpretability than other machine learning approaches \citep{ishwaran2007variable}.
+A random forest regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.
 
-We modeled subnational rates of moderate-to-severe and severe food insecurity globally at the subnational level in the period of 2010 to 2030 using the disaggregated FIES data combined with the covariates.
-Using the log-transformed rate of food insecurity as our outcome variable, we explored several modeling approaches, including linear regression methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression.
+Because the outcome variables in our models, the rate of moderate or severe food insecurity, is a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.
+Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.
 
-Based on various model validation metrics, Random Forest (RF) regression performed the best.
-The RF regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.
-Random forest models are known to perform well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}.
-Since this is likely to be the case in our application, this method is particularly suitable for modeling rates of moderate-to-severe and severe food insecurity.
+To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.
+Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.
+We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe models.
+By choosing the combinations of hyperparameters with the smallest OOB error we found the optimal combination for both models.
+We found that 5000 trees are more than sufficient for the error rates to converge (See Fig.
+\ref{fig:rf_error}).
 
-We observed for the moderate-to-severe as well as the severe RF regression model an in-sample mean squared error (MSE) between 0.002-0.005, with a R-squared (${R}^2$) value of 0.999 (See Fig.
+%Talk more about validation: excluding HICs, how many runs, etc
+We observed for the moderate as well as the severe model an in-sample mean squared error (MSE) between 0.002-0.005, with a R-squared (${R}^2$) value of 0.999 (See Fig.
 \ref{fig:rf_in-sample}).
@@
 
-To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.
-Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.
-We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe models.
-By choosing the combination of hyperparameters with the smallest OOB error we found the optimal combination for both models respectively.
-We found that 5000 trees are more than sufficient for the error rates to converge (See Fig.
-\ref{fig:rf_error}).
-
-Apart from tuning the hyperparameters, we examined the importance of the individual covariates in explaining rates of moderate-to-severe and severe food insecurity.
+After fitting the models, we examined the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
 By training the models with all covariates and predicting rates with and without a individual variable we calculated the average prediction error of omitting that variable.
@@
 
-For a detailed description, implementation and validation of the RF regression models see the Supplemental Info.
+For a detailed description, implementation and validation of the random forest models see the Supplemental Info.
 
@@
 This method can be used for any variable that is a fraction or a rate.
-We use this method to estimate future rates of stunting, wasting, and malaria mortality, as well as to model future admin shares of GDP and population.
+We use this method to estimate future rates of stunting, wasting, and malaria mortality, as well as to model future subnational shares of GDP and population.
 Thus, we give an overview of the methodology here.
 
-First, for each admin area, $a$, the rate of change (ROC) is calculated between each pair of adjacent years, $y$, based on a value, $p$, such that $0 < p < 1$, available for each year.
+First, for each subnational area, $a$, the rate of change (ROC) is calculated between each pair of adjacent years, $y$, based on a value, $p$, such that $0 < p < 1$, available for each year.
 
@@
 We do this by combining historic subnational data from the Subnational Development Database \citep{Smits2019} with national level population estimates for the 21st century \citep{KC2017}.
-To account for within-country trends and changes in population distribution, we use the AROC method outlined in equations \ref{eqn:a} - \ref{eqn:d} to project each admin areas share of national population totals.
+To account for within-country trends and changes in population distribution, we use the AROC method outlined in equations \ref{eqn:a} - \ref{eqn:d} to project each subnational areas share of national population totals.
 We then disaggregate the national totals given by KC et al.
@@
 We first use data from the World Bank to estimate the mean country-level ratio of GNI to GDP and adjust the subnational estimates of GNI according to that country-specific ratio.
-We then calculate each admin area's share of country GDP, and then use the AROC method to estimate each admin area's share of total country GDP.
-Based on these trajectories in each admin areas share of total country GDP, we estimate what each admin area's share of GDP will be for each year from 2019 to 2030.
+We then calculate each subnational area's share of country GDP, and then use the AROC method to estimate each subnational area's share of total country GDP.
+Based on these trajectories in each subnational areas share of total country GDP, we estimate what each subnational area's share of GDP will be for each year from 2019 to 2030.
 We divide that share of GDP by the population of the administrative area, which we had previously estimated.
@@
 \subsection{Water Scarcity Share}
+We used the water scarcity index (WSI) developed by \citep{greve2018global}.
+The index represents the ratio of the decadal averages in water demand to water supply and it assesses changes in water scarcity on a 0.5 by 0.5 degree global grid.
+Based on this index, we estimate the share of people living in areas with less than 1000m\textsuperscript{3} of water available per capita per year.
+The WSI has been used already to build the water scarcity clock.
+For more information visit \href{https://worldwater.io/}{worldwater.io}.
+
 \begin{figure}[H]
@@
 \subsection{Mean Annual Precipitation}
-For mean annual precipitation, we use data for the years 2010 to 2019 from the TerraClimate dataset \citep{abatzoglou2018terraclimate}, summarized to each admin area.
+For mean annual precipitation, we use data for the years 2010 to 2019 from the TerraClimate dataset \citep{abatzoglou2018terraclimate}, summarized to each subnational area.
 For data for the years 2020-2030, we use models from the Inter-Sectoral Model Intercomparison Project \citep{warszawski2014inter}.
@@
 \subsection{Topographic Ruggedness}
-As a measure of accessibility, we include the mean topographic ruggedness of each admin area.
+As a measure of accessibility, we include the mean topographic ruggedness of each subnational area.
 We first used a gridded dataset of elevation from the USGS \cite{USGS1996} and calculated the index at each grid cell using the methodology from Riley et al.

### 2020-10-27 data/covars/scripts/water_clock.R

@@ -1,6 +1,7 @@
 setwd('~/wdl-fies')
 
 library(tidyverse)
+library(zoo)
 
 #water scarcity
 fs <- list.files('data/covars/rawdata/wc', pattern='csv$', recursive = T, full.names = T)

### 2020-10-27 models/model_randomforest.R

@@ -16,16 +16,11 @@ moddat <- merge(fies_subnat,
 preddat <- covars %>%
   data.frame
 
-exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
-         'pasture', 'cropland')
-
 # set up model
 vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
                                             'urban', 'rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region',
-                                            names(moddat)[grepl('region', names(moddat))],
-                                            exc)]
+                                            'population', 'YEAR', 'rural_perc', 'region')]
 
 moddat <- moddat %>%
   mutate(fies.mod = logit(fies.mod),
@@ -84,3 +79,4 @@ moddat <- moddat %>%
 
 write.csv(preddat, 'data/preddat.csv', row.names=F)
 write.csv(moddat, 'data/moddat.csv', row.names=F)
+save(list=c('rf.mod', 'rf.sev', 'rf.tune.mod', 'rf.tune.sev'), file='data/models.RData')

### 2020-10-27 models/model_randomforest_1yearlag.R

@@ -1,153 +1,0 @@
-#######################################################
-# use random forest with ssp covariates
-# to estimate food insecurity, based on data for
-# modeled for 2020, 2025, and 2030
-#######################################################
-
-# setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
-
-#Add 1-year lags as features
-prev_year <- covars %>%
-  mutate(YEAR = YEAR - 1)
-
-ix <- !names(prev_year) %in% c('GDLCODE', 'YEAR', 'ISO3')
-names(prev_year)[ix] <- paste0('prev_', names(prev_year)[ix])
-
-covars <- merge(covars, prev_year)
-
-# read in data
-moddat <- merge(fies_subnat, 
-                covars, 
-                all.x=T, all.y=F) %>%
-  na.omit %>%
-  data.frame
-
-preddat <- covars %>%
-  data.frame
-
-exc <- c('crops_prod', 'forest', 'builtup', 'livestock', #8
-         'pasture', 'cropland')
-
-# set up model
-vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
-                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
-                                            'urban', 'rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region',
-                                            names(moddat)[grepl('region', names(moddat))],
-                                            exc, paste0('prev_', exc))]
-
-moddat <- moddat %>%
-  mutate(fies.mod = logit(fies.mod),
-         fies.sev = logit(fies.sev))
-
-## mod+sev
-# tune parameters
-rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod", 
-                                                     paste(vars, collapse = "+"), sep= "~")),
-                          data = moddat,
-                          mtrystart = max(1, floor(sqrt(length(vars)))), 
-                          nodesizetry = c(1:3),
-                          ntreetry = 5000,
-                          sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
-                          trace = T,
-                          dobest = T); rf.tune.mod$optimal
-
-# model
[…]

### 2020-10-27 munge/03-Covariates.R

@@ -12,9 +12,9 @@ for (f in fs){
 covars <- covars %>%
   select(YEAR, ISO3, GDLCODE, stunting, wasting, school_mean, ruggedness,
          gdp_percap, gini, mal_falciparum, precip, hci, tave, 
-         population, rural_perc, urban_perc, ws_share, builtup, cropland, crops_prod,
-         forest, livestock, pasture) %>%
-  merge(regions)
+         population, rural_perc, urban_perc, ws_share) %>%
+  merge(regions %>%
+          select(region=region_sdg, ISO3))
 
 #########################################################################
 # Missing entire variable at the country level.  

### 2020-10-27 viz/Error_and_VIP.R

@@ -1,39 +1,4 @@
-
-
-preddat <- read.csv('data/preddat.csv')
-moddat <- read.csv('data/moddat.csv')
-
-######################
-# save output for poli
-########################
-
-sel <- preddat %>%
-  select(ISO3, YEAR, GDLCODE, stunting, urban_perc, fies.mod.pred, population) %>%
-  filter(YEAR %in% c(2020, 2025, 2030)) %>%
-  merge(u5.population, all.x=T, all.y=F) %>%
-  mutate(u5pop.urban = urban_perc*u5_frac*population,
-         u5pop.rural = (1 - urban_perc)*u5_frac*population,
-         stunting.urban = stunting*u5pop.urban,
-         stunting.rural = stunting*u5pop.rural,
-         population.urban = population*urban_perc,
-         population.rural = population*(1 - urban_perc),
-         fies.urban = fies.mod.pred*population.urban,
-         fies.rural = fies.mod.pred*population.rural) %>%
-  select(-stunting, -urban_perc, -fies.mod.pred, -population, -u5_frac) %>%
-  gather(var, value, -ISO3, -YEAR, -GDLCODE) %>%
-  mutate(geo_area=ifelse(grepl('urban', var), 'Urban', 'Rural'),
-         var=gsub('.rural|.urban', '', var),
-         value = round(value)) %>%
-  spread(var, value)
-
-write.csv(sel, 'figures/HC_2020-10-07.csv', row.names=F) #update date
-
-
-
-##########################################
-#### plot error and variable importance
-##########################################
-
+# setwd('~/wdl-fies');library(ProjectTemplate);load.project()
 #setwd('~/wdl-fies/docs/img')
 setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
 
@@ -41,39 +6,46 @@ library(cowplot)
 library(gridGraphics)
 options(scipen=100)
 
+vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
+                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
+                                            'fies.mod.pred', 'fies.sev.pred',
+                                            'urban', 'rural', 'fies.sev', 'fies.mod',
+                                            'population', 'YEAR', 'rural_perc', 'region')]
+
 # change varaibles names
[…]

### 2020-10-27 viz/Hunger_Clock_Data.R

@@ -1,0 +1,26 @@
+
+######################
+# save output for poli
+########################
+
+sel <- preddat %>%
+  select(ISO3, YEAR, GDLCODE, stunting, urban_perc, fies.mod.pred, population) %>%
+  filter(YEAR %in% c(2020, 2025, 2030)) %>%
+  merge(u5.population, all.x=T, all.y=F) %>%
+  mutate(u5pop.urban = urban_perc*u5_frac*population,
+         u5pop.rural = (1 - urban_perc)*u5_frac*population,
+         stunting.urban = stunting*u5pop.urban,
+         stunting.rural = stunting*u5pop.rural,
+         population.urban = population*urban_perc,
+         population.rural = population*(1 - urban_perc),
+         fies.urban = fies.mod.pred*population.urban,
+         fies.rural = fies.mod.pred*population.rural) %>%
+  select(-stunting, -urban_perc, -fies.mod.pred, -population, -u5_frac) %>%
+  gather(var, value, -ISO3, -YEAR, -GDLCODE) %>%
+  mutate(geo_area=ifelse(grepl('urban', var), 'Urban', 'Rural'),
+         var=gsub('.rural|.urban', '', var),
+         value = round(value)) %>%
+  spread(var, value)
+
+write.csv(sel, 'figures/HC_2020-10-07.csv', row.names=F) #update date
+

### 2020-10-27 viz/Viz_Mod_Results.R

@@ -4,7 +4,7 @@
 setwd('~/wdl-fies/docs/img')
 #setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
 
-#library(cowplot)
+library(cowplot)
 options(scipen=100)
 
 ############################
@@ -23,13 +23,13 @@ mapdat <- merge(gdl,
                   filter(YEAR %in% c(2010, 2020, 2030)) %>%
                   select(YEAR, GDLCODE, fies.mod.pred, fies.sev.pred) %>%
                   gather(outcome, value, -YEAR, -GDLCODE) %>%
-                  mutate(FIES=ifelse(grepl('sev', outcome), 'Severe\n', 'Moderate-to-Severe\n')))
+                  mutate(FIES=ifelse(grepl('sev', outcome), 'Severe\n', 'Moderate\n')))
 
 ggplot(mapdat) +
   geom_sf(aes(fill=value), color=NA) + 
   scale_fill_gradientn(colours=c("#5e51a2", "#2f89be", "#66c3a6", "#add8a4", "#e4ea9a", "#fbf8c0", 
                                  "#fce08a", "#faae61", "#f36c44", "#a01c44"), 
-                       breaks=seq(0, 1, by=0.2),
+                       breaks=seq(0, 1, by=0.25),
                        limits=c(0, max(mapdat$value)),
                        labels=function(x){paste0(x*100, '%')}) + 
   geom_sf(data=cty, color='#000000', fill=NA, size=0.15) + 
@@ -47,16 +47,21 @@ system('pdfcrop FullMap.pdf FullMap.pdf')
 ###########################
 # Setup Reference Map
 ############################
-region_cols <- c("South Asia" = "#e41a1c",
-                 "Sub-Saharan Africa" = "#377eb8",
-                 "Europe & Central Asia" = "#4daf4a",
-                 "Middle East & North Africa" = "#f781bf",
-                 "Latin America & Caribbean" = "#ff7f00",
-                 "East Asia & Pacific" = "#a65628",
-                 "North America" = "#984ea3")
+region_cols <- c("Central & Southern Asia"="#F36D25",
+                 "Sub-Saharan Africa"="#E11484",
+                 "Europe & Northern America"="#02558B",
+                  "Northern Africa & Western Asia"="#F99D26",
[…]

### 2020-10-27 docs/modeling_fies.tex

@@
 Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.
-We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate-to-severe and severe models.
+We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate and severe models.
 By choosing the combinations of hyperparameters with the smallest OOB error we found the optimal combination for both models.

### 2020-10-29 docs/modeling_fies.tex
add new VIMP scripts, continue work on manuscript

@@
 % TODO:
-% - re-run model
-%   - Use validation with FAO data, if possible
 % - discuss validation more
@@
 %   - Compare to country reports
-% - get new results:
-%   - new regions
-%   - Get rates over time
 % - describing FIES
@@
 % - make appendix separate
-% - add higher level overview of modeling the covars
+% - Go over the fies in more detail in appendix
 
@@
 \author[4]{Carlo Cafiero}
-\author[2,5]{Juan Carlos Laso}
-\author[2,6]{Wolfgang Fengler}
-\author[2,7]{Homi Kharas}
+\author[2,5]{Juan Carlos Laso Bayas}
+\author[2,6]{Homi Kharas}
 
@@
 \affil[5]{The International Institute for Applied Systems Analysis}
-\affil[6]{The World Bank}
-\affil[7]{The Brookings Institute}
+\affil[6]{The Brookings Institute}
 \affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}
@@
 Food security is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is “Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
-The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience it.
+The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger.
 Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories consumed per capita in a population.
@@
 However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \citep{Maxwell1996}.
-This led to the emergence of several indicators designed to be rapidly deployable, culturally cross-comparable, and based on the lived experience of food security \citep{Jones2013}.
-These metrics include the Household Food Insecurity and Access Scale (HFIAS) \citep{Coates2007}; the Coping Strategies Index (CSI) \citep{Maxwell1999}; and the Household Hunger Scale (HHS) \citep{Ballard2011}.
+This led to the emergence of several indicators designed to be rapidly deployable, and based on the lived experience of food security \citep{Jones2013}.
+These metrics include the Household Food Security Survey Module (HFSSM) \citep{kennedy2005keynote}, originally developed for use in the US; the Latin American and Caribbean Food Security Scale (ELCSA); and the Household Food Insecurity and Access Scale (HFIAS) \citep{Coates2007}.
 
-Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013} and quickly became a useful tool for understanding individual and household-level food insecurity \citep{smith2017world}.
+Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
@@
 Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalences can be computed.
-A moderately food insecure adult has compromised on food quality and variety, as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
-A severely food insecure adult has run out of food and has gone an entire day without eating at times during the year.
+A moderately food insecure adult is likely to have compromised on food quality and variety, as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
+A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.
 
-Since 2014, in partnership with Gallup World Poll, the FAO has published surveys from around the world to estimate the prevalence of food insecurity using the FIES.
-Drawing on surveys from 77 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
+Since 2014, in partnership with Gallup World Poll, the FAO administered the FIES in surveys around the world.
+The FAO uses this data to estimate the prevalence of food insecurity, and reports national-level estimates of the percentage of the population over the thresholds for moderate and severe food insecurity.
+Drawing on individual-level data from these microdatasets in 77 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
 The model is then used to infer levels of food insecurity around the world at the subnational level and to forecast food insecurity to the year 2030.
@@
 \section{Results}
-% Maybe pull out individual interesting scenarios?
+\subsection{Food Insecurity over Time}
 Globally, in the year 2020, we estimate that 829 million people are above the threshold for severe food insecurity, and an additional 1.5 billion people are above the threshold for moderate food insecurity (See Fig.
@@
 
-While the total number of people experiencing moderate and severe food insecurity held roughly constant for the 2010s and even increased in latter part of that decade, we predict that it will decline throughout the 2020s.
+While the total number of people experiencing moderate and severe food insecurity held roughly constant for the 2010s and even increased in latter part of that decade, our model predict that it will decline throughout the 2020s.
 However, this overall decline will not be uniform across the world.
 Our model expects that, in most of sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau, although some countries, such as Kenya, Ethiopia, and Ghana, will see improvements.
-Food insecurity will fall in most other world regions, particularly South Asia and East Asia & the Pacific.
+Food insecurity will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.
 
@@
 
+Examining the rate of food insecurity in a population, rather than the total number of food insecure people, our model finds the most progress has been made in East \& Southeast Asia, while Central \& Southern Asia is expected to see improvements in food insecurity throughout the 2020s.
+Other poor and middle income areas, including much of Africa, the Middle East and Latin America, have made only slight progress and will continue to make slight progress between 2010 and 2030.
+Finally, in high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/Rates.pdf}
+  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by UN SDG regions over time, with a 3-year smoothing.}
+  \label{fig:timeseries}
+\end{figure}
+
+\subsection{Food Insecurity over Space}
 We found substantial heterogeneity in the global distribution of severe food insecurity (See Fig.
@@
 
-The results of this analysis can be further explored in more detail on the \href{worldhunger.io}{World Hunger Clock}, including statistics for each subnational administrative area.
+\subsection{Important Predictors of Food Insecurity}
+Overall, our model was able to predict food insecurity with high accuracy in the countries that had training data ($R^2 = 0.999$ in sample, $R^2 = 0.99$ out of sample).
+Examining how permuting a variable affects the predictive accuracy of the model can give an idea of how important that variable is for predicting food insecurity.
+
+For both moderate and severe food insecurity, the most important variable for predicting food security levels was the Poverty Headcount Index, followed by the rate of Stunting.
+For predicting the rate of people with at least moderate food insecurity, GDP per Capita was quite important.
+However, for the rate of severe food insecurity, the country's Gini coefficient was very important, relative to moderate food insecurity.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{img/VIMP.pdf}
+  \caption{Importance of each variable in predicting the percent of a population over thresholds for moderate and severe food insecurity.
+This is determined by the increase in the error when the variable is permuted before fitting the model.}
+  \label{fig:map}
+\end{figure}
+
+The results of this analysis can be further explored in more detail on the \href{https://worldhunger.io}{World Hunger Clock}, including statistics for each subnational administrative area.
 
@@
 %Homi says delete this para
-While this is the first comprehensive projection of indicator 2.1.2 for SDG 2, projections for other indicators for that goal have been made.
+These results contrast with other projections that have been made for While this is the first comprehensive projection of indicator 2.1.2 for SDG 2, projections for other indicators for that goal have been made.
 The most recent State of Food Insecurity (SOFI) report by the FAO projected indicator 2.1.1, the prevalence of undernourishment, and found that that, much like food insecurity, undernourishment had been increasing throughout the 2010s \citep{sofi2020}.
@@
 We aimed for covariates that had subnational spatial resolution and peer-reviewed projections available through the year 2030.
-For projections based on the Shared Socioeconomic Pathways (SSP) framework, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway (RCP) 6.0.
+For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.
 
@@
 In cases of health variables that have shown long-term trends and represented a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these variables for 2030.
-The involves taking the rate of change between each pair of years in the dataset, and then taking the mean rate of change over the period for which data is available, giving greater weight to more recent years.
+The involves taking the rate of change between each pair of years in the dataset, and then taking the mean rate of change over the period for which data is available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.
 This method has been used to estimate rates of stunting and wasting in 2025 \citep{Local2020}; we simply use the same method to make forecasts to the year 2030.
@@
 \section{Component Questions of the FIES}
+%%% Discuss reference scale and calibration
 \begin{enumerate}
@@
 The function \texttt{tune.rfsrc()} does exactly that and returns optimal values for \texttt{mtry} and \texttt{nodesize}.
-We applied this function on the both, the moderate-to-severe and the severe model.
+We applied this function on both, the moderate-to-severe and the severe model.
 For the moderate-to-severe model the optimal combination is \texttt{mtry} = 12 and \texttt{nodesize} = 1 and for the severe model we found \texttt{mtry} = 10 and \texttt{nodesize} = 2 to be optimal.

### 2020-10-29 viz/Assess_Accuracy.R

@@ -1,0 +1,3 @@
+setwd('~/wdl-fies');library(ProjectTemplate);load.project()
+
+

### 2020-10-29 viz/Error_and_VIP.R
more local changes to Error and VIMP

@@ -17,35 +17,31 @@ vars_full <- c("Stunting", "Wasting", "Mean Years of Schooling", "Topographic Ru
               "GDP Per Capita", "Gini Coefficient", "Malaria Mortality Rate", "Mean Annual Precipitation",
               "Poverty Headcount Index", "Mean Temperature", "Urban Percentage", "Water Scarcity") #has to be in the same order as "vars"
 
+v.mod <- vimp(rf.mod, importance='permute')
+v.sev <- vimp(rf.sev, importance='permute')
 
-hv.mod <- holdout.vimp(formula = as.formula(paste("fies.mod", 
-                                                  paste(vars, collapse = "+"), sep= "~")),
-                 data = moddat,
-                 ntree = 5000, 
-                 mtry = rf.tune.mod$optimal[[2]],
-                 nodesize = rf.tune.mod$optimal[[1]],
-                 verbose = TRUE)
-
-hv.sev <- holdout.vimp(formula = as.formula(paste("fies.sev", 
-                                                  paste(vars, collapse = "+"), sep= "~")),
-                 data = moddat,
-                 ntree = 5000, 
-                 mtry = rf.tune.sev$optimal[[2]],
-                 nodesize = rf.tune.sev$optimal[[1]],
-                 verbose = TRUE)
-
-df <- data.frame(var=c(names(hv.mod), names(hv.sev)),
-                 val=c(hv.mod, hv.sev),
+df <- data.frame(var=c(names(v.mod$importance), names(v.sev$importance)),
+                 val=c(v.mod$importance, v.sev$importance),
                  mod=rep(c('Moderate', 'Severe'), each=12),
                  lab=rep(vars_full, 2))
 
-df$lab <- factor(df$lab, levels=vars_full[order(hv.mod)])
+df$lab <- factor(df$lab, levels=vars_full[order(v.sev$importance)])
+
+convertToOutcomeScale <- function(x){
+  #Determine how an error of x on a logit transformed scale
+  #affects the mean absolute error after being transformed back to [0,1]
+  mean(abs(moddat$fies.mod - inv.logit(logit(moddat$fies.mod) + x)))
+}
+
+df$val <- sapply(df$val, convertToOutcomeScale)
 
 ggplot(df) + 
   geom_bar(aes(x=lab, y=val), stat='identity', fill='#0F1290') +
   coord_flip() + 
   facet_grid(. ~ mod) + 
-  theme_bw()
[…]

### 2020-10-30 docs/modeling_fies.tex
almost done with paper; improve visualizations

@@
 % - Go over the fies in more detail in appendix
+
+% Ask Carlo - has the FAO made predictions of the FIES yet?
 
@@
 Other poor and middle income areas, including much of Africa, the Middle East and Latin America, have made only slight progress and will continue to make slight progress between 2010 and 2030.
-Finally, in high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.
+In high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.
+Overall, the world has made has made steady but slow progress on reducing the rate of food insecurity.
 
@@
 Nevertheless, many regions, most notably sub-Saharan Africa, are expected to see overall increases in the number of food insecure people.
-%Add here comparison with SOFI, but emphasize that discrepancies may also come from measuring different things
 
-%Homi says delete this para
-These results contrast with other projections that have been made for While this is the first comprehensive projection of indicator 2.1.2 for SDG 2, projections for other indicators for that goal have been made.
-The most recent State of Food Insecurity (SOFI) report by the FAO projected indicator 2.1.1, the prevalence of undernourishment, and found that that, much like food insecurity, undernourishment had been increasing throughout the 2010s \citep{sofi2020}.
-The SOFI report also modeled the prevalence of undernourishment in 2030 based on recent trends, giving more weight to recent years, and projected that undernourishment will increase by 2030 \citep[Part 1, Page 11]{sofi2020}.
-Using a similar method for modeling the prevalence of stunting (indicator 2.2.1), the SOFI reported that the rate of stunting would decrease by 2030 \citep[Part 1, Page 30]{sofi2020}.
+Our variable importance metrics show that economic conditions are the largest driver of global variation in rates of food insecurity.
+For both moderate and severe food insecurity, the poverty headcount index, the Gini coefficient, and GDP per capita were major predictors of the rate of food insecurity.
+The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger was also an important predictor of food insecurity.
+However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality were less important predictors of food insecurity.
+Geographic and climate factors, such as urban percentage, topographic ruggedness and mean annual temperature and precipitation were not important predictors of food insecurity at a global scale.
+The fact that our models had such high accuracy indicates that these variables alone are sufficient to predict food insecurity globally across countries, although other variables may be more important for explaining variation at more local scales.
 
-The fact that the SOFI expects increases in undernourishment while we predict decreases in food insecurity is likely due to our differing modeling approaches.
-While the SOFI report extrapolated recent trends in undernourishment, we modeled food insecurity based on projections of factors like GDP per capita, population, and the prevalence of stunting.
-We found that, in a random forest model, these predictors were very successful at predicting recent rates of both moderate-to-severe and severe food insecurity.
+These results contrast with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
+This is likely due in part to our differing modeling approaches.
+While the SOFI report extrapolated recent trends in undernourishment and stunting, we modeled food insecurity based on interactions between projections in factors like GDP per capita, population, and the prevalence of stunting.
+Additionally, our diverging predictions for different SDG 2 indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while food insecurity, as measured by the FIES, decreases.
+The discrepancies between our projections illustrate both how hunger is a complex phenomenon that is resistant to simple quantification, as well as the challenges in forecasting the future, where different modeling approaches will yield different predictions.
 
-%Add section about the drivers of food insecurity, as well as how they largely account for the outcomes observed.
+When making predictions in countries where we did not have primary data, we are assuming that the relationship between variables like poverty, stunting, inequality, and hunger are the same in all locations.
+We were able to test whether this was the case using data from countries that reported their aggregate rates of food insecurity but did not release microdata, and we found that our model largely proved well in those cases, with an accuracy of (ACCURACY HERE).
+Additionally, when making predictions into the future, we are assuming that long-term patterns of demographic change, urbanization, and development will maintain their long-held trajectories for the next decade.
+At a global scale, this is almost certainly the case: it is highly unlikely that rates of fertility or economic growth with shift suddenly and globally.
+Nevertheless, at more local scales, sudden crises can lead to severe and unforeseeable increases in food insecurity, as crises in the previous decade in Yemen, Syria and Venezuela have shown.
+Thus, while our model's predictions at a global scale are very likely to bear out, there will probably be individual countries and areas where our forecasts end up being off.
 
 While we found that food insecurity will decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, only expects the number of people experiencing moderate food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.
-This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, a number that falls far short of the SDG 2 goal of ``zero hunger."  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.
+This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, a number that falls far short of the SDG 2 goal of ending hunger.
+Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.
 Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done in pursuit of SDG 2.
@@
 We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
-We used random forest regressions, as this approach performs well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}, while providing more model interpretability than other machine learning approaches \citep{ishwaran2007variable}.
+We used random forest regressions, as this approach performs well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}.
 A random forest regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.
 
-Because the outcome variables in our models, the rate of moderate or severe food insecurity, is a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.
+Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.
 Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.
@@
 By choosing the combinations of hyperparameters with the smallest OOB error we found the optimal combination for both models.
-We found that 5000 trees are more than sufficient for the error rates to converge (See Fig.
-\ref{fig:rf_error}).
-
-%Talk more about validation: excluding HICs, how many runs, etc
-We observed for the moderate as well as the severe model an in-sample mean squared error (MSE) between 0.002-0.005, with a R-squared (${R}^2$) value of 0.999 (See Fig.
-\ref{fig:rf_in-sample}).
-To test the out-of-sample prediction performance we performed ten-fold cross-validation.
-First, we partitioned the data into ten equal sized subsamples.
-Once partitioned, we trained the RF regression models on nine subsamples and tested the models on the remaining one.
-This process gets repeated ten times, with each subsample being used once for testing.
-We then calculated the average MSE and ${R}^2$ over all ten out-of-sample predictions.
-Similarly to the in-sample model validation metrics, the average out-of-sample MSE is between 0.008-0.016, with an average ${R}^2$ of around 0.99 for both models (See Fig.
-\ref{fig:rf_out-sample}).
-Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries.
-We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.
+We found that 5000 trees are more than sufficient for the error rates to converge.
 
 After fitting the models, we examined the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
-By training the models with all covariates and predicting rates with and without a individual variable we calculated the average prediction error of omitting that variable.
-Using the highest average prediction error we ranked the importance of the variables and found that for both, the moderate-to-severe and severe model, the Poverty Headcount Index, Stunting, GDP Per Capita and the Gini Coefficient are the most important variables for explaining differences in rates of moderate-to-severe and severe food insecurity (See Fig.
-\ref{fig:rf_vimp}).
+We used a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable}.
+Similarly to model predictions, we transformed the error term with the inverse logit so that the errors would be on the scale of the outcome variable and therefore more interpretable.
+
+\subsection{Model Validation}
+We observed a MAE of 0.0029 for our model of severe food insecurity ($R^2$ of 0.9989) and an MAE 0.0054 for our model of moderate food insecurity ($R^2$ of 0.9990).
+This means that, on average, our predictions were only of by 0.29\% and 0.54\%, respectively, for the subnational areas in countries where we had microdata.
+
+To further validate the models, we aggregated our predictions to the country level to compare with national-level reports by the FAO.
+These included countries which reported the rate of people over the thresholds for moderate and severe food insecurity, but did not release raw microdata, thus providing an truly out-of-sample means of testing our model.
+Among these 21 countries, our model predicted the rate of at least moderate food insecurity with a MAE of 0.0961 ($R^2$ = 0.8094), and severe food insecurity with a MAE of 0.0718 ($R^2$ = 0.7782).
 
@@
 
+To test the out-of-sample prediction performance we performed ten-fold cross-validation.
+First, we partitioned the data into ten equal sized subsamples.
+Once partitioned, we trained the RF regression models on nine subsamples and tested the models on the remaining one.
+This process gets repeated ten times, with each subsample being used once for testing.
+We then calculated the average MSE and ${R}^2$ over all ten out-of-sample predictions.
+Similarly to the in-sample model validation metrics, the average out-of-sample MSE is between 0.008-0.016, with an average ${R}^2$ of around 0.99 for both models (See Fig.
+\ref{fig:rf_out-sample}).
+Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries.
+We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.
+
 \end{document}

### 2020-10-30 models/make_sample_data.R

@@ -1,52 +1,0 @@
-##############################################3
-# Make example CSV for Poli
-######################################################
-
-#Fields
-# -GDLCODE
-# -2020.fies.urb
-# -2020.fies.rur
-# -2025.fies.urb
-# -2025.fies.rur
-# -2030.fies.urb
-# -2030.fies.rur
-# -2020.pop.urb
-# -2020.pop.rur
-# -2025.pop.urb
-# -2025.pop.rur
-# -2030.pop.urb
-# -2030.pop.rur
-# -2020.stunt
-# -2025.stunt
-# -2030.stunt
-
-example <- data.frame(GDLCODE=unique(ssp_past$GDLCODE))
-
-cols <- c("2025.fies.urb",
-"2025.fies.rur",
-"2030.fies.urb",
-"2030.fies.rur",
-"2020.pop.urb",
-"2020.pop.rur",
-"2025.pop.urb",
-"2025.pop.rur",
-"2030.pop.urb",
-"2030.pop.rur",
-"2020.stunt",
-"2025.stunt",
-"2030.stunt")
-
-for (c in cols){
-	if (grepl('fies|stunt', c)){
-		d <- runif(nrow(example))
-	}
-	if (grepl('pop', c)){
-		d <- rpois(nrow(example), 100000)
-	}
-	
-	example[ , c] <- d	
-
-}
-
-write.csv(example, 'figures/example_hunger_data.csv', row.names=F)
-

### 2020-10-30 models/model_randomforest.R

@@ -1,10 +1,12 @@
-#######################################################
+######################################################
 # use random forest with ssp covariates
 # to estimate food insecurity, based on data for
 # modeled for 2020, 2025, and 2030
 #######################################################
 
-# setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
+set.seed(100)
+
+setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
 
 # read in data
 moddat <- merge(fies_subnat,
@@ -33,7 +35,7 @@ rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod",
                           data = moddat,
                           mtryStart = ncol(moddat)/2, 
                           nodesizeTry = c(1:3),
-                          ntree = 5000,
+                          ntree = 1000,
                           sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
                           trace = T,
                           dobest = T); rf.tune.mod$optimal
@@ -41,7 +43,7 @@ rf.tune.mod <- tune.rfsrc(formula = as.formula(paste("fies.mod",
 # model
 rf.mod <- rfsrc(formula = as.formula(paste("fies.mod", paste(vars, collapse = "+"), sep= "~")),
                 data = moddat,
-                ntree = 5000, 
+                ntree = 1000, 
                 mtry = rf.tune.mod$optimal[[2]],
                 nodesize = rf.tune.mod$optimal[[1]],
                 do.trace = TRUE)
@@ -57,14 +59,14 @@ rf.tune.sev <- tune.rfsrc(formula = as.formula(paste("fies.sev",
                           data = moddat,
                           mtryStart = ncol(moddat)/2, 
                           nodesizeTry = c(1:3),
-                          ntree = 5000,
+                          ntree = 1000,
                           sampsize = min(nrow(moddat)*.632, max(150, nrow(moddat)^(3/4))),
                           trace = T,
                           dobest = T); rf.tune.sev$optimal
 # model
 rf.sev <- rfsrc(formula = as.formula(paste("fies.sev", paste(vars, collapse = "+"), sep= "~")),
                 data = moddat,
-                ntree = 5000, 
+                ntree = 1000, 
[…]

### 2020-10-30 viz/Assess_Accuracy.R

@@ -1,3 +1,0 @@
-setwd('~/wdl-fies');library(ProjectTemplate);load.project()
-
-

### 2020-10-30 viz/Error_and_VIP.R

@@ -6,6 +6,10 @@ library(cowplot)
 library(gridGraphics)
 options(scipen=100)
 
+####################################
+# Variable Importance
+###################################
+
 vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
                                             'fies.mod.pred', 'fies.sev.pred',
@@ -43,33 +47,16 @@ ggplot(df) +
   labs(x='', y='Increase in Model Error When Variable is Permutated')
 ggsave('VIMP.pdf', width=7, height=3.75)
 
-# variable effect
-# png('model/mod.coefs_rf.png', width = 1000, height = 800, units="px")
-# plot.variable.rfsrc(rf.mod, sorted = T)
-# dev.off()
-# 
-# png('model/sev.coefs_rf.png', width = 1000, height = 800, units="px")
-# plot.variable.rfsrc(rf.sev, sorted = T)
-# dev.off()
-# 
-# plot.variable.rfsrc(rf.mod, sorted = T)
-# grid.echo()
-# mod.coef <- grid.grab()
-# 
-# plot.variable.rfsrc(rf.sev, sorted = T)
-# grid.echo()
-# sev.coef <- grid.grab()
-# 
-# plot_grid(plot.variable.rfsrc(rf.mod, sorted = T), plot.variable.rfsrc(rf.mod, sorted = T), align='h',labels='AUTO', scale = 0.8)
-# ggsave('model/coef_rf.png', width=18, height=8)
-
-
-
 ##################################
 # assess residuals
 ##################################
-mse <- mean(sqrt((moddat$fies.mod - moddat$fies.mod.pred)^2))
-r2 <- cor(moddat$fies.mod, moddat$fies.mod.pred)
+mae.mod <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred))
+mae.sev <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred))
+
+r2.mod <- cor(moddat$fies.mod, moddat$fies.mod.pred)
+r2.sev <- cor(moddat$fies.sev, moddat$fies.sev.pred)
+
+
 mod.res <- ggplot(moddat) + 
   geom_point(aes(x=fies.mod, y=fies.mod.pred), alpha = 0.3) + 
   geom_abline(intercept = 0, slope = 1, color = "red", size = 0.5) +

### 2020-10-30 viz/Viz_Mod_Results.R

@@ -127,15 +127,31 @@ rates <- preddat %>%
   filter(YEAR >= 2010) %>%
   group_by(YEAR, region) %>%
   summarize(mod.rate=weighted.mean(fies.mod.pred, w=population, na.rm=T),
-            sev.rate=weighted.mean(fies.sev.pred, w=population, na.rm=T)) %>%
-  gather(var, value, -YEAR, -region) %>%
+            sev.rate=weighted.mean(fies.sev.pred, w=population, na.rm=T),
+            population=sum(population)) %>%
+  gather(var, value, -YEAR, -region, -population) %>%
   mutate(var = ifelse(grepl('mod', var), "Moderate", "Severe")) %>%
   group_by(region, var) %>%
   mutate(value = rollapply(value, width=3, FUN=mean, align='center', partial=TRUE))
 
-ratelines <- ggplot(rates) +
-  geom_line(aes(x=YEAR, y=value, color=region), size=1) +
-  scale_color_manual(values=region_cols) + 
+world <- rates %>%
+  group_by(YEAR, var) %>%
+  summarize(value=weighted.mean(value, w=population, na.rm=T)) %>%
+  mutate(region='World')
+
+ratesb <- bind_rows(rates, world)
+
+region_cols2 <- c(region_cols, "World"="#000000")
+size <- c(rep(1, 8), 2)
+names(size) <- names(region_cols2)
+lty <- c(rep(1, 8), 2)
+names(lty) <- names(region_cols2)
+
+(ratelines <- ggplot() +
+  geom_line(data=ratesb, aes(x=YEAR, y=value, color=region, size=region, linetype=region), show.legend=F) +
+  scale_color_manual(values=region_cols2) + 
+  scale_size_manual(values=size) + 
+  scale_linetype_manual(values=lty) + 
   theme_bw() + 
   theme(plot.margin=unit(c(-1, 0.5, 0, 0), 'lines'),
         panel.spacing=unit(2, 'lines')) + 
@@ -144,8 +160,42 @@ ratelines <- ggplot(rates) +
                      labels=function(x){paste0(x*100, '%')}) +
   labs(x='', y="",
        title="") +
-  guides(color=FALSE) +           
-  facet_grid(. ~ var)
+  facet_grid(. ~ var))
 
-plot_grid(ratelines, reg, ncol=1, rel_heights=c(2.5, 1))
+
+(legendplot <- ggplot() + 
+  geom_line(aes(x=c(0,1), y=c(0, 1), color=c('World', 'World'), 
[…]

### 2020-10-30 docs/modeling_fies.tex
more updates to paper

@@
 
-% TODO:
-% - discuss validation more
-%   - various exclusions: HICs
-%   - Compare to country reports
-% - describing FIES
-%   - Go over different data sources
-% - add some kind of framework for selecting covars
-% - make appendix separate
-% - Go over the fies in more detail in appendix
-
-% Ask Carlo - has the FAO made predictions of the FIES yet?
-
 \begin{document}
@@
 \begin{abstract}
-We modeled levels of of food security at a subnational level from 2010 to 2030 using microdata from 77 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project.
+We modeled levels of of food security at a subnational level from 2010 to 2030 using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project.
 We find significant heterogeneity in levels of food security around the world, ranging from parts of the global north with less than 1\% of the population food insecure to parts of the global south where as much as 3 out of 4 people are severely food insecure.
-Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that food insecurity has been increasing over the past several years, but under middle-of-the-road assumptions for development and population change, food insecurity will decline throughout the 2020s.
+Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that food insecurity has increased slightly over the past several years, but under middle-of-the-road assumptions for development and population change, food insecurity will decline throughout the 2020s.
 This global decrease is largely driven by trends in South and East Asia, and some parts of the world, particularly sub-Saharan Africa, are projected to show increases in the prevalence of food insecurity throughout the next decade.
+Our model's forecast of decreasing food insecurity through the 2020s contrasts with predictions from the FAO that other metrics of hunger will increase.
 This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
@@
 
-Because the FIES is based on responses to polls, the data is limited to 77 countries where the polls have been conducted and microdata has been vetted and released by the FAO.
+Because the FIES is based on responses to polls, the data is limited to 75 countries where the polls have been conducted and microdata has been vetted and released by the FAO.
 The current practice is to assign the regional average prevalence rate for countries without survey data in order to compile regional and global aggregates.
@@
 A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.
+Because the model estimates the population over a threshold on a univariate food insecurity scale, the percentage of people over the threshold for moderate food insecurity includes people that are also over the threshold for severe food insecurity.
 
@@
 The FAO uses this data to estimate the prevalence of food insecurity, and reports national-level estimates of the percentage of the population over the thresholds for moderate and severe food insecurity.
-Drawing on individual-level data from these microdatasets in 77 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
+Drawing on individual-level data from these microdatasets in 75 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
 The model is then used to infer levels of food insecurity around the world at the subnational level and to forecast food insecurity to the year 2030.
@@
 
-Examining the rate of food insecurity in a population, rather than the total number of food insecure people, our model finds the most progress has been made in East \& Southeast Asia, while Central \& Southern Asia is expected to see improvements in food insecurity throughout the 2020s.
+Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig.
+\ref{fig:rates}, our model finds the most progress has been made in East \& Southeast Asia, while Central \& Southern Asia is expected to see improvements in food insecurity throughout the 2020s.
 Other poor and middle income areas, including much of Africa, the Middle East and Latin America, have made only slight progress and will continue to make slight progress between 2010 and 2030.
@@
   \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by UN SDG regions over time, with a 3-year smoothing.}
-  \label{fig:timeseries}
+  \label{fig:rates}
 \end{figure}
@@
 \subsection{Important Predictors of Food Insecurity}
-Overall, our model was able to predict food insecurity with high accuracy in the countries that had training data ($R^2 = 0.999$ in sample, $R^2 = 0.99$ out of sample).
+Overall, our model was able to predict food insecurity with high accuracy in the countries that had training data ($R^2 = 0.9981$ and $R^2 = 0.9976$ for moderate and severe food insecurity, respectively).
 Examining how permuting a variable affects the predictive accuracy of the model can give an idea of how important that variable is for predicting food insecurity.
 
-For both moderate and severe food insecurity, the most important variable for predicting food security levels was the Poverty Headcount Index, followed by the rate of Stunting.
-For predicting the rate of people with at least moderate food insecurity, GDP per Capita was quite important.
-However, for the rate of severe food insecurity, the country's Gini coefficient was very important, relative to moderate food insecurity.
+For both moderate and severe food insecurity, the most important variable for predicting food security levels was the Poverty Headcount Index, followed by the rate of stunting.
+For predicting the rate of people with at least moderate food insecurity, GDP per capita was quite important.
+However, for the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.
 
@@
 Geographic and climate factors, such as urban percentage, topographic ruggedness and mean annual temperature and precipitation were not important predictors of food insecurity at a global scale.
-The fact that our models had such high accuracy indicates that these variables alone are sufficient to predict food insecurity globally across countries, although other variables may be more important for explaining variation at more local scales.
+Finally, while economic conditions are widely associated with food insecurity at a global scale, the fact that Europe and North America is predicted to see little improvement in food insecurity outcomes between 2010 and 2030, in spite of economic growth and continued reductions in poverty, suggests that there is an upper limit to the extent to which increasing economic development can improve food security outcomes.
 
@@
 When making predictions in countries where we did not have primary data, we are assuming that the relationship between variables like poverty, stunting, inequality, and hunger are the same in all locations.
-We were able to test whether this was the case using data from countries that reported their aggregate rates of food insecurity but did not release microdata, and we found that our model largely proved well in those cases, with an accuracy of (ACCURACY HERE).
+We tested the accuracy of these cases by using ten-fold cross validation at the country level, and we found reasonable accuracy ($R^2 = 0.778$, MAE = 0.073 for moderate and $R^2 = 0.696$, MAE = 0.044 for severe food insecurity).
+This suggestions that our model explains between 70-78\% of the variation in food security outcomes and is on average, within 4-7\% in its estimates of rates of food insecurity, when applied to countries outside of the training set.
+
 Additionally, when making predictions into the future, we are assuming that long-term patterns of demographic change, urbanization, and development will maintain their long-held trajectories for the next decade.
@@
 \subsection{Model Validation}
-We observed a MAE of 0.0029 for our model of severe food insecurity ($R^2$ of 0.9989) and an MAE 0.0054 for our model of moderate food insecurity ($R^2$ of 0.9990).
-This means that, on average, our predictions were only of by 0.29\% and 0.54\%, respectively, for the subnational areas in countries where we had microdata.
+We observed a MAE of 0.0029 for our model of severe food insecurity ($R^2$ of 0.9978) and an MAE 0.0054 for our model of moderate food insecurity ($R^2$ of 0.9981).
+This means that, on average, our model predictions were only of by 0.29\% and 0.54\%, respectively, for the subnational areas in countries where we had microdata.
 
-To further validate the models, we aggregated our predictions to the country level to compare with national-level reports by the FAO.
-These included countries which reported the rate of people over the thresholds for moderate and severe food insecurity, but did not release raw microdata, thus providing an truly out-of-sample means of testing our model.
-Among these 21 countries, our model predicted the rate of at least moderate food insecurity with a MAE of 0.0961 ($R^2$ = 0.8094), and severe food insecurity with a MAE of 0.0718 ($R^2$ = 0.7782).
+To further validate the models, we ran a ten-fold cross validation at the country level.
+We randomly divided the countries into ten parts, and re-fit the model 10 times, with each part left out one.
+Then, we predicted the values for the ten percent that was left out using the model fit with 90\% of the countries.
+We randomly left out data at the country level because within each country we had values from different subnational areas and different years, and these values are likely to be very similar to each other.
+Because we are extending our models predictions to countries that were not found at all in our training data, a better test of our model is to assess its predictive accuracy on countries that were entirely absent from the training data.
+By this validation metric, we found that our model performed reasonably well, with an $R^2$ of 0.778 and an MAE of 0.073 for moderate food insecurity and an $R^2$ of 0.696 and a MAE of 0.044 for severe food insecurity.
 
@@
 
-\section{Component Questions of the FIES}
+\section{Overview of the FIES}
+We used raw microdata released by the FAO from 75 countries that vary by world region and income level (See Fig. \ref{fig:fies_countries}.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/FIES_Countries.pdf}
+  \caption{Countries that had microdata on the FIES and were used in our model, shown in blue.}
+  \label{fig:fies_countries}
+\end{figure}
+
+This microdata comes with the probability that an individual is over the threshold for moderate and severe food insecurity already calculated by the FAO, based on their responses to the 8 FIES questions (See List \ref{itm:fies}).
+Thus, we did not perform those calculations ourselves.
+Nevertheless, we give here an overview of the procedure.
+For more detail see Cafiero (2018) in \textit{Measurement} \citep{Cafiero2018}.
+
+The FIES is calculated using a Rash model, which was developed by the psychometrics literature.
+The Rash model assumes that each individual and their responses to the FIES questions can be placed on a one-dimensional scale of food insecurity, and that the log odds of a respondent \textit{r} answering yes to one of the FIES questions \textit{i} is a linear function of the difference between the severity of the food insecurity experienced by \textit{r} and the severity of item {i}.
+
+The severity of each item and each respondents level of food insecurity can be estimated with Maximum Likelihood.
+After fitting the model to each country, the FAO found that the assumptions of the Rash model held up well, as evidenced by model fit diagnostics.
+
+After separately estimating the Rash model for each country, the FAO then developed a global reference scale for the severity of each of the 8 questions, by iteratively harmonizing the severity values in each country.
+Then, these global reference points were calibrated against other surveys that had asked questions similar to the FIES, such as the HFSSM and the ELCSA.
+
+Based on the global reference scale, two thresholds are set based on questions 1 and 8: whether the responded was worried they would not have enough food to eat because of a lack of money or other resources, and whether there was a time when they went without eating for a whole day because of a lack of money or other resources.
+
 %%% Discuss reference scale and calibration
+%%% Map of countries with Training Data
 \begin{enumerate}
@@
 	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
+  \label{itm:fies}
 \end{enumerate}
@@
 
-%WE NEED HERE A SHORT PARAGRAPH ON HOW RF MODELS WORK, THIS WILL HELP: https://www.kdnuggets.com/2017/10/random-forests-explained.html
-%INCLUDE BAGGING AND RANDOM FEATURE SELECTION, NECESSARY TO EXPLAIN OOB ERROR
-
 The RF regression is a machine learning method based on the creation of a large number of decision trees.
@@
   \centering
-  \includegraphics[width=\linewidth]{img/model/vimp_rf.png}
+  \includegraphics[width=\linewidth]{img/VIMP.pdf}
   \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
@@
 %MORE INFORMATION HERE
-
-To test the out-of-sample prediction performance we performed ten-fold cross-validation.
-First, we partitioned the data into ten equal sized subsamples.
-Once partitioned, we trained the RF regression models on nine subsamples and tested the models on the remaining one.
-This process gets repeated ten times, with each subsample being used once for testing.
-We then calculated the average MSE and ${R}^2$ over all ten out-of-sample predictions.
-Similarly to the in-sample model validation metrics, the average out-of-sample MSE is between 0.008-0.016, with an average ${R}^2$ of around 0.99 for both models (See Fig.
-\ref{fig:rf_out-sample}).
-Besides using the full set of countries to validate the out-of-sample performance, we halved the sample size by excluding high-income countries (HIC) (according to the World Bank definition, REFERENCE NEEDED) and performed ten-fold cross-validation for this limited set of countries.
-We found that even when excluding HIC the average out-of-sample MSE is between 0.013-0.023, with an average ${R}^2$ of around 0.98 for both models.
-
 \end{document}

### 2020-10-30 models/model_randomforest.R

@@ -76,9 +76,10 @@ moddat$fies.sev.pred <- inv.logit(as.numeric(predict(rf.sev, moddat)$predicted))
 preddat$fies.sev.pred <- inv.logit(as.numeric(predict(rf.sev, preddat)$predicted))
 
 #Run model under k-fold cross validation
-samp <- sample(1:10, nrow(moddat), replace=T)
+iso3s <- unique(moddat$ISO3)
+samp <- sample(1:10, length(iso3s), replace=T)
 for (s in 1:10){
-  ix <- samp != s
+  ix <- moddat$ISO3 %in% iso3s[samp != s]
 
   rf.mod.cv <- rfsrc(formula = as.formula(paste("fies.mod", paste(vars, collapse = "+"), sep= "~")),
                   data = moddat[ix, ],

### 2020-10-30 viz/Error_and_VIP.R

@@ -1,6 +1,6 @@
 # setwd('~/wdl-fies');library(ProjectTemplate);load.project()
-#setwd('~/wdl-fies/docs/img')
-setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
+setwd('~/wdl-fies/docs/img')
+#setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
 
 library(cowplot)
 library(gridGraphics)
@@ -53,32 +53,59 @@ ggsave('VIMP.pdf', width=7, height=3.75)
 mae.mod <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred))
 mae.sev <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred))
 
-r2.mod <- cor(moddat$fies.mod, moddat$fies.mod.pred)
-r2.sev <- cor(moddat$fies.sev, moddat$fies.sev.pred)
-
+r2.mod <- cor(moddat$fies.mod, moddat$fies.mod.pred)^2
+r2.sev <- cor(moddat$fies.sev, moddat$fies.sev.pred)^2
 
 mod.res <- ggplot(moddat) + 
   geom_point(aes(x=fies.mod, y=fies.mod.pred), alpha = 0.3) + 
   geom_abline(intercept = 0, slope = 1, color = "red", size = 0.5) +
   xlim(0, 1) + ylim(0, 1) +
-  labs(caption=paste0('Mean Squared Error: ',  round(mse, 4),
-                      '\nR-Squared: ', round(r2, 4)),
+  labs(caption=paste0('Mean Average Error: ',  round(mae.mod, 4),
+                      '\nR-Squared: ', round(r2.mod, 4)),
        x='Observed Rates Of Moderat-to-Severe Food Insecurity',
        y='Modeled Rates Of Moderat-to-Severe Food Insecurity'); mod.res
-#ggsave('model/mod.residuals_rf.png', width=5, height=5)
 
-mse <- mean(sqrt((moddat$fies.sev - moddat$fies.sev.pred)^2))
-r2 <- cor(moddat$fies.sev, moddat$fies.sev.pred)
 sev.res <- ggplot(moddat) + 
   geom_point(aes(x=fies.sev, y=fies.sev.pred), alpha = 0.3) +
   geom_abline(intercept = 0, slope = 1, color = "red", size = 0.5) +
   xlim(0, 1) + ylim(0, 1) +
[…]

### 2020-10-30 viz/Map_FIES_Countries.R

@@ -1,0 +1,19 @@
+setwd('~/wdl-fies');library(ProjectTemplate);load.project()
+
+cty <- ne_countries(returnclass='sf')
+
+iso3 <- unique(moddat$ISO3)
+
+cty$FIES_DAT <- cty$iso_a3 %in% iso3
+cty <- cty %>%
+  filter(sovereignt != 'Antarctica')
+
+ggplot(cty) + 
+  geom_sf(aes(fill=FIES_DAT), show.legend=F, size=0.25) + 
+  scale_fill_manual(values=rev(c("#000245", "#C0C0C0"))) + 
+  coord_sf(crs='+proj=robin') +
+  labs(title='Countries That Have FIES Microdata') + 
+  theme_void() +
+  theme(plot.title = element_text(hjust = 0.5))
+
+ggsave('docs/img/FIES_Countries.pdf', width=6, height=4)

### 2020-10-30 viz/compare_predictions_FAO.R

@@ -25,6 +25,31 @@ preddat2 <- preddat %>%
 m <- merge(compare, preddat2, all.x=T, all.y=F) %>%
   na.omit
 
+fao <- unique(compare$ISO3)
+mod <- unique(moddat$ISO3)
+out <- fao[!fao %in% mod]
+
+#Remove china - they have estimates better than Norway??
+out <- out[out != 'CHN']
+
+s <- m %>%
+  filter(ISO3 %in% out)
+
+plot(s$fies.mod.3yr.fao, s$fies.mod.3yr.pred)
+cor(s$fies.mod.3yr.fao, s$fies.mod.3yr.pred)
+mean(abs(s$fies.mod.3yr.fao - s$fies.mod.3yr.pred))
+
+plot(s$fies.sev.3yr.fao, s$fies.sev.3yr.pred)
+cor(s$fies.sev.3yr.fao, s$fies.sev.3yr.pred)
+mean(abs(s$fies.sev.3yr.fao - s$fies.sev.3yr.pred))
+
+
+
+
+
+
+
+
 plot(m$fies.mod.3yr.fao, m$fies.mod.3yr.pred)
 cor(m$fies.mod.3yr.fao, m$fies.mod.3yr.pred)
 

### 2020-11-10 docs/modeling_fies.tex
some minor changes in the text of the manuscript

@@
 In high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.
-Overall, the world has made has made steady but slow progress on reducing the rate of food insecurity.
+Overall, the world has made steady but slow progress on reducing the rate of food insecurity.
 
@@
 \ref{fig:map}).
-For the year 2020, mainland sub-Saharan Africa is the continent with with the highest rates of severe food insecurity, with at least 15\% of people over the threshold for severe food insecurity in at last one subnational area in every country except Gabon and Equatorial Guinea.
+For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with at least 15\% of people over the threshold for severe food insecurity in at last one subnational area in every country except Gabon and Equatorial Guinea.
 Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.
@@
 Additionally, when making predictions into the future, we are assuming that long-term patterns of demographic change, urbanization, and development will maintain their long-held trajectories for the next decade.
-At a global scale, this is almost certainly the case: it is highly unlikely that rates of fertility or economic growth with shift suddenly and globally.
+At a global scale, this is almost certainly the case: it is highly unlikely that rates of fertility or economic growth will shift suddenly and globally.
 Nevertheless, at more local scales, sudden crises can lead to severe and unforeseeable increases in food insecurity, as crises in the previous decade in Yemen, Syria and Venezuela have shown.

### 2020-11-15 docs/modeling_fies.tex
new model outputs

@@
 
-\title{A Global Picture of Food Insecurity and Hunger}
+\title{A Global Picture of Hunger}
 
@@
 \author[2,5]{Juan Carlos Laso Bayas}
-\author[2,6]{Homi Kharas}
+\author[5,6]{Jesús Crespo Cuaresma}
+\author[2,7]{Homi Kharas}
 
@@
 \affil[5]{The International Institute for Applied Systems Analysis}
-\affil[6]{The Brookings Institute}
+\affil[6]{Wittgenstein Centre For Demography and Global Human Capital}
+\affil[7]{The Brookings Institute}
 \affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}
@@
 \begin{abstract}
-We modeled levels of of food security at a subnational level from 2010 to 2030 using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project.
-We find significant heterogeneity in levels of food security around the world, ranging from parts of the global north with less than 1\% of the population food insecure to parts of the global south where as much as 3 out of 4 people are severely food insecure.
-Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that food insecurity has increased slightly over the past several years, but under middle-of-the-road assumptions for development and population change, food insecurity will decline throughout the 2020s.
-This global decrease is largely driven by trends in South and East Asia, and some parts of the world, particularly sub-Saharan Africa, are projected to show increases in the prevalence of food insecurity throughout the next decade.
-Our model's forecast of decreasing food insecurity through the 2020s contrasts with predictions from the FAO that other metrics of hunger will increase.
+Using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the \href{https://worldhunger.io}{World Hunger Clock}.
 This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
+We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the developing world where over half the population is severely food insecure.
+Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been steadily declining, with large observed and forecasted gains in lower and lower-middle income countries, and decreases in total number of severely food insecure people.
+However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.
+Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.
 \end{abstract}
@@
 \section{Main}
-Food security is a critical component of human flourishing, which is why the second Sustainable Development Goal (SDG 2) is “Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
+Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
 The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger.
-Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories consumed per capita in a population.
-These do not capture the actual lived exposure of individuals to food shortages.
+Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population.
+These do not directly capture the actual lived exposure of individuals to food shortages.
 
-Because the FIES is based on responses to polls, the data is limited to 75 countries where the polls have been conducted and microdata has been vetted and released by the FAO.
-The current practice is to assign the regional average prevalence rate for countries without survey data in order to compile regional and global aggregates.
-This is not very satisfactory.
-In addition, because the FIES is a relatively new metric, it does not have a sufficiently long time series to estimate trends with any degree of precision.
-To understand whether SDG 2 is being met, one needs to estimate food insecurity outcomes in 2030.
+The FIES is based on responses to polls and the data are limited to 75 countries where these have been conducted and microdata has been vetted and released by the FAO.
+The current practice is to assign the average regional prevalence rate for countries without survey data in order to compile regional and global aggregates.
+In addition, since the FIES is a relatively new metric, it does not have a sufficiently long time series to assess time trends with a reasonable degree of precision.
+To understand whether SDG 2 is being met, estimates of food insecurity outcomes in 2030 are required.
 
-The contribution of this paper is to fill these two gaps of coverage and forecasts by developing a model of FIES outcomes that is based on covariates that are available for places where actual polling data is not available, and that can be forecast to 2030 to get a sense of what food insecurity at that time might look like.
-We also take advantage of the within-country spatial distribution of FIES reports to push the analysis beyond country level to a sub-national level, that is administratively once removed from the national level (Admin-1).
+This paper contributes to shifting the research frontier by filling these two gaps in terms of data coverage and forecasts by developing a statistical model of FIES outcomes based on covariates that are available for territories where actual polling data are not available.
+Such a model can be employed to obtain forecasts to 2030 and assess likely future trends in food insecurity.
+We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond country level to a sub-national level.
 To our knowledge, this is the first effort to model the FIES outcomes at that scale.
 
-Food security has traditionally been difficult to measure, and this has led to incomplete or inaccurate pictures of global hunger.
-Metrics of macro-health, such as anthropometry and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.
-However, these metrics are confounded with other determinants of health such as infectious disease and are not meaningful at the scale of individuals or households \citep{Perumal2018}.
-Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and food insecurity can certainly occur in the absence of food availability decline \citep{Sen1983}.
+Food security has traditionally been difficult to measure, and this has led to an incomplete or inaccurate assessment of global hunger.
+Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.
+However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.
+Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.
 Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.
@@
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
-Using a Rash model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing he responses given to the 8 questions asked in the survey.
-Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalences can be computed.
-A moderately food insecure adult is likely to have compromised on food quality and variety, as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
+Using a Rasch model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing the responses given to the 8 questions asked in the survey.
+Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalence of such food insecurity events can be computed.
+A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
 A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.
@@
 The FAO uses this data to estimate the prevalence of food insecurity, and reports national-level estimates of the percentage of the population over the thresholds for moderate and severe food insecurity.
-Drawing on individual-level data from these microdatasets in 75 countries, we used machine learning to estimate levels of food insecurity around the world at the subnational level.
-The model is then used to infer levels of food insecurity around the world at the subnational level and to forecast food insecurity to the year 2030.
+Drawing on individual-level data from these microdatasets in 75 countries, we use machine learning methods to estimate levels of food insecurity globally at the subnational level.
+The model is then used to forecast food insecurity to the year 2030 under a middle-of-the-road scenario.
 
@@
 \subsection{Food Insecurity over Time}
-Globally, in the year 2020, we estimate that 829 million people are above the threshold for severe food insecurity, and an additional 1.5 billion people are above the threshold for moderate food insecurity (See Fig.
+Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and an additional 2.5 billion people are above the threshold for moderate food insecurity (See Fig.
 \ref{fig:timeseries}).
-These numbers are higher than the estimates in the 2019 FAO State of Food Insecurity report, which estimated about 700 million severely food insecure people and an additional 1.31 billion people experiencing at least moderate food insecurity in 2018.
+These numbers are higher than the estimates in the 2019 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \ref{sofi2020}.
 
-While the total number of people experiencing moderate and severe food insecurity held roughly constant for the 2010s and even increased in latter part of that decade, our model predict that it will decline throughout the 2020s.
-However, this overall decline will not be uniform across the world.
-Our model expects that, in most of sub-Saharan Africa, moderate food insecurity will increase throughout the 2020s and severe food insecurity will plateau, although some countries, such as Kenya, Ethiopia, and Ghana, will see improvements.
-Food insecurity will fall in most other world regions, particularly South Asia and East Asia \& the Pacific.
+Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.
+However, for both indicators, these changes are very slight: from 2010 to 2030, the number of moderately food insecure people will increase 2\%, and the number of severely food insecure people will decrease 8\%.
+The increases in moderate food insecurity have been in lower and lower-middle income countries, while the increases in severe food insecurity are entirely from lower income countries.
+For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.
 
@@
   \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
-  \caption{Number of people over the thresholds for moderate and severe food insecurity, by UN SDG regions over time, with a 3-year smoothing.
+  \caption{Number of people over the thresholds for moderate and severe food insecurity, by World Bank income groups as of 2020.
 Panel (A) shows the number of food insecure people with region totals stacked, to show global trends and totals over time.
@@
 Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig.
-\ref{fig:rates}, our model finds the most progress has been made in East \& Southeast Asia, while Central \& Southern Asia is expected to see improvements in food insecurity throughout the 2020s.
-Other poor and middle income areas, including much of Africa, the Middle East and Latin America, have made only slight progress and will continue to make slight progress between 2010 and 2030.
-In high-income countries in Europe and North America, rates of people in at least moderate food insecurity move little, and rates of severe food insecurity are expected to increase slightly.
+\ref{fig:rates}, we find clear improvements in all except high income countries.
+Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.
+Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.
+In spite of this progress in the bottom three income quartiles, in high-income countries, rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.
 Overall, the world has made steady but slow progress on reducing the rate of food insecurity.
@@
   \includegraphics[width=\linewidth]{img/Rates.pdf}
-  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by UN SDG regions over time, with a 3-year smoothing.}
+  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020}
   \label{fig:rates}
@@
 
-\subsection{Food Insecurity over Space}
-We found substantial heterogeneity in the global distribution of severe food insecurity (See Fig.
+\subsection{Food Insecurity Across Space}
+We find substantial heterogeneity in the global distribution of severe food insecurity across countries and subnational regions (See Fig.
 \ref{fig:map}).
-For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with at least 15\% of people over the threshold for severe food insecurity in at last one subnational area in every country except Gabon and Equatorial Guinea.
+For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one subnational area in every country except Gabon, Djibouti and Equatorial Guinea.
 Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.
-In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10-15\% in areas like northern Brazil, many central Asian and middle-eastern countries, and India and Indonesia.
+In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in northern Brazil, many central Asian and middle-eastern countries, as well as India and Indonesia.
 
@@
 In 2020, much of Africa, south and southeast Asia, and parts of Latin America had over 50\% of the population living with moderate food insecurity.
-Even in countries widely considered to be developed, such as Australia, the United States, and parts of eastern and southern Europe, over 10\% of the population is above the threshold for moderate food insecurity.
+Even in highly developed economies, such as the United States and parts of eastern and southern Europe, over 10\% of the population is above the threshold for moderate food insecurity.
 
@@
   \includegraphics{img/FullMap.png}
-  \caption{Modeled spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
+  \caption{Spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
   \label{fig:map}
@@
 \subsection{Important Predictors of Food Insecurity}
-Overall, our model was able to predict food insecurity with high accuracy in the countries that had training data ($R^2 = 0.9981$ and $R^2 = 0.9976$ for moderate and severe food insecurity, respectively).
-Examining how permuting a variable affects the predictive accuracy of the model can give an idea of how important that variable is for predicting food insecurity.
+Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.944$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.799$ and $R^2 = 0.772$ for moderate and severe food insecurity, respectively).
+The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.
 
-For both moderate and severe food insecurity, the most important variable for predicting food security levels was the Poverty Headcount Index, followed by the rate of stunting.
-For predicting the rate of people with at least moderate food insecurity, GDP per capita was quite important.
-However, for the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.
+For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index, followed by the rate of stunting.
+For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.
+For the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.
+Across both indicators, economic variables were the most relevant predictors.
 
@@
 \section{Discussion}
-This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts out to the year 2030.
-We found that, while food insecurity has been increasing and has likely worsened in 2020 with economic downturn associated with the novel coronavirus, our models predict that number of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in south and east Asia.
-Nevertheless, many regions, most notably sub-Saharan Africa, are expected to see overall increases in the number of food insecure people.
+This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.
+We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the coronavirus pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.
+Nevertheless, lower and lower-middle income countries, are expected to see overall increases in the number of food insecure people.
 
-Our variable importance metrics show that economic conditions are the largest driver of global variation in rates of food insecurity.
-For both moderate and severe food insecurity, the poverty headcount index, the Gini coefficient, and GDP per capita were major predictors of the rate of food insecurity.
-The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger was also an important predictor of food insecurity.
-However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality were less important predictors of food insecurity.
-Geographic and climate factors, such as urban percentage, topographic ruggedness and mean annual temperature and precipitation were not important predictors of food insecurity at a global scale.
-Finally, while economic conditions are widely associated with food insecurity at a global scale, the fact that Europe and North America is predicted to see little improvement in food insecurity outcomes between 2010 and 2030, in spite of economic growth and continued reductions in poverty, suggests that there is an upper limit to the extent to which increasing economic development can improve food security outcomes.
+Our analysis of the relative importance of different predictors for future food insecurity dynamics shows that economic conditions are the largest driver of global variation in rates of food insecurity.
+For both moderate and severe food insecurity, the Poverty Headcount Index, the Gini coefficient, and GDP per capita appear as highly relevant predictors of the rate of food insecurity.
+The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger is also an important predictor of food insecurity.
+However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality are less important as explanatory factors of changes in food insecurity.
+Geographic and climate factors, such as the percentage of urban population, topographic ruggedness and mean annual temperature or precipitation do not appear to be robust predictors of food insecurity at a global scale.
+Finally, while economic conditions are widely associated with food insecurity at a global scale, the fact that high income countries have seen little improvement in food insecurity outcomes, in spite of marginal annual increases in GDP per capita, suggests that there is an upper limit to the extent to which increasing economic development can improve food security outcomes.
 
-These results contrast with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
-This is likely due in part to our differing modeling approaches.
+Our results anticipating of increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
+However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.
+In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 3\% decrease, from 788 to 767 million severely food insecure people over the same period.
+
+These discrepancies are in part due to our differing modeling approaches.
 While the SOFI report extrapolated recent trends in undernourishment and stunting, we modeled food insecurity based on interactions between projections in factors like GDP per capita, population, and the prevalence of stunting.
-Additionally, our diverging predictions for different SDG 2 indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while food insecurity, as measured by the FIES, decreases.
-The discrepancies between our projections illustrate both how hunger is a complex phenomenon that is resistant to simple quantification, as well as the challenges in forecasting the future, where different modeling approaches will yield different predictions.
+Additionally, our diverging predictions for different hunger indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while severe food insecurity, as measured by the FIES, decreases.
+The discrepancies between our projections and those of the FAO illustrate both the complexity of hunger as a social phenomenon, as well as the challenges in predicting future socioeconomic developments, where different modeling approaches may yield different forecasts.
 
-When making predictions in countries where we did not have primary data, we are assuming that the relationship between variables like poverty, stunting, inequality, and hunger are the same in all locations.
-We tested the accuracy of these cases by using ten-fold cross validation at the country level, and we found reasonable accuracy ($R^2 = 0.778$, MAE = 0.073 for moderate and $R^2 = 0.696$, MAE = 0.044 for severe food insecurity).
-This suggestions that our model explains between 70-78\% of the variation in food security outcomes and is on average, within 4-7\% in its estimates of rates of food insecurity, when applied to countries outside of the training set.
+When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.
+We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find reasonable accuracy ($R^2 = 0.799$, MAE = 0.068 for moderate food insecurity and $R^2 = 0.772$, MAE = 0.038 for severe food insecurity).
+This suggestions that our model explains between 77\% to 79\% of the variation in food security outcomes and is on average, within 3-7 percentage points in its estimates of rates of food insecurity, when applied to countries outside of the training set.
 
-Additionally, when making predictions into the future, we are assuming that long-term patterns of demographic change, urbanization, and development will maintain their long-held trajectories for the next decade.
+Additionally, when making predictions into the future, we assume that the long-term patterns of demographic change, urbanization, and development will maintain their trajectories for the next decade.
 At a global scale, this is almost certainly the case: it is highly unlikely that rates of fertility or economic growth will shift suddenly and globally.
 Nevertheless, at more local scales, sudden crises can lead to severe and unforeseeable increases in food insecurity, as crises in the previous decade in Yemen, Syria and Venezuela have shown.
-Thus, while our model's predictions at a global scale are very likely to bear out, there will probably be individual countries and areas where our forecasts end up being off.
 
-While we found that food insecurity will decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, only expects the number of people experiencing moderate food insecurity to fall by 12\% and the number of people experiencing severe food insecurity to fall by 20\%.
-This still leaves billions of people eating less than they should and nearly half a billion people going entire days without eating a decade from now, a number that falls far short of the SDG 2 goal of ending hunger.
-Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, or obesity.
-Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done in pursuit of SDG 2.
+While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate food insecurity to increase and the number of people experiencing severe food insecurity to fall only slightly between 2010 and 2030.
+This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, a falling far short of the SDG 2 goal of ending hunger.
+Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, sustainable diets, or obesity.
+Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done at the political level in pursuit of SDG 2.
 
 \section{Methods}
-
-%do we need to add a section where we state which fies data we use? like countries and years?
-
 \subsection{Disaggregation}
@@
 Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
-We used a similar methodology, except we calculated post-stratification weights at a subnational level.
-We used year-specific subnational estimates of population percentages by gender, age, urbanization, and, where available, wealth, and created a separate set of weights for each subnational area across all individuals.
-For age and gender, we used gridded data from WorldPop \citep{Tatem2017}, for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}, where available.
-This methodology rests on the mild assumption that subnational differences in rates of food insecurity are more due to factors related to demographics, urbanization, and wealth than to other unobservable factors.
+We use a similar methodology and calculate post-stratification weights at a subnational level.
+We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.
+For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}, where available.
+This methodology rests on the mild assumption that subnational differences in rates of food insecurity are driven by factors related to demographics, urbanization, and wealth rather than by other unobservable factors.
 
@@
 We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).
-We aimed for covariates that had subnational spatial resolution and peer-reviewed projections available through the year 2030.
-For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we used the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.
+We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.
+For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.
 
-Many of the variables required harmonizing subnational historic data with projections at the national level.
-This included mean years of schooling, GDP per capita, as well as for projections of population.
-We used observed trajectories in the historical distribution of variables among subnational areas within a country to estimate the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.
+Many of the variables require the harmonization of subnational historical data with projections at the national level.
+These include mean years of schooling, GDP per capita, as well as population.
+We use observed trajectories in the historical distribution of variables among subnational areas within a country to predict the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.
 
-In cases of health variables that have shown long-term trends and represented a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these variables for 2030.
-The involves taking the rate of change between each pair of years in the dataset, and then taking the mean rate of change over the period for which data is available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.
-This method has been used to estimate rates of stunting and wasting in 2025 \citep{Local2020}; we simply use the same method to make forecasts to the year 2030.
+For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions up to variables for 2030.
+We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.
+This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.
 
-For the climate variables temperature and precipitation, we combined historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
+For the climatic variables, temperature and precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
 
-Finally, we adjusted several variables to account for the effects of COVID-19, including GDP per capita using June 2020 estimates from the World Bank \citep{prospects2020}, stunting and wasting using August 2020 estimates published in \textit{The Lancet} \citep{headey2020impacts}, as well as the poverty headcount index \cite{Cuaresma2018}.
+Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.
+These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.
 For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.
@@
 	\end{tabular}
-	\caption{Covariates Included in Present-Day Analysis}
+	\caption{Covariates Included in the Analysis}
 	\label{tab:covars}
@@
 We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
-We used random forest regressions, as this approach performs well in the presence of non-linearities and important interactions among predictor variables \citep{hastie2009elements}.
-A random forest regression involves generating a large number of decision trees, each under slightly different conditions and using different sub-samples of the data, and then aggregating the predictions of the decision trees.
-
-Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.
-Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.
+We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.
+A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.
+Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.
 
@@
 %STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
+Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.
+Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.
 

### 2020-11-15 models/model_randomforest.R

@@ -36,9 +36,13 @@ prm2 <- expand.grid(list(node=c(1:10, 12, 14, 15, 16, 17, 18, 19, 20, 25, 100),
                         mtry=c(13:1),
                         depth=c(1:13, -1)))
 
+prm$sev.mae <- NA
+prm$mod.mae <- NA
+
 prm <- bind_rows(prm, prm2[!paste(prm2$node, prm2$depth, prm2$mtry, sep='-') %in% paste(prm$node, prm$depth, prm$mtry, sep='-'), ])
 
 prm$ix <- 1:nrow(prm)
+
 
 #############################################
 #Run model under 10-fold cross validation
@@ -78,9 +82,11 @@ for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
     moddat$fies.mod.pred[!ix] <- inv.logit(predict(rf.mod, moddat[!ix,])$predicted)
     moddat$fies.sev.pred[!ix] <- inv.logit(predict(rf.sev, moddat[!ix,])$predicted)
     
-    prm$sev.rsq[i] <- cor(moddat$fies.sev.pred, moddat$fies.sev)^2
-    prm$mod.rsq[i] <- cor(moddat$fies.mod.pred, moddat$fies.mod)^2
   }
+  prm$sev.rsq[i] <- cor(moddat$fies.sev.pred, moddat$fies.sev)^2
+  prm$mod.rsq[i] <- cor(moddat$fies.mod.pred, moddat$fies.mod)^2
+  prm$sev.mae[i] <- mean(abs(moddat$fies.sev.pred - moddat$fies.sev))
+  prm$mod.mae[i] <- mean(abs(moddat$fies.mod.pred - moddat$fies.mod))
 }
 
 write.csv(prm, 'data/prm.csv', row.names=F)

### 2020-11-15 viz/Error_and_VIP.R

@@ -14,7 +14,8 @@ vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
                                             'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
                                             'fies.mod.pred', 'fies.sev.pred',
                                             'urban', 'rural', 'fies.sev', 'fies.mod',
-                                            'population', 'YEAR', 'rural_perc', 'region')]
+                                            'population', 'YEAR', 'rural_perc', 'region',
+                                            'fies.mod.logit', 'fies.sev.logit')]
 
 # change varaibles names
 vars_full <- c("Stunting", "Wasting", "Mean Years of Schooling", "Topographic Ruggedness",
@@ -23,6 +24,10 @@ vars_full <- c("Stunting", "Wasting", "Mean Years of Schooling", "Topographic Ru
 
 v.mod <- vimp(rf.mod, importance='permute')
 v.sev <- vimp(rf.sev, importance='permute')
+
+#####################################
+# Plot Variable importance
+##################################
 
 df <- data.frame(var=c(names(v.mod$importance), names(v.sev$importance)),
                  val=c(v.mod$importance, v.sev$importance),
@@ -46,6 +51,22 @@ ggplot(df) +
   theme_bw() + 
   labs(x='', y='Increase in Model Error When Variable is Permutated')
 ggsave('VIMP.pdf', width=7, height=3.75)
+
+################################
+# Plot Error Over Time
+#################################
+
+erdf <- data.frame(var=c(v.mod$err.rate, v.sev$err.rate),
+                 trees=c(1:5000, 1:5000),
+                 mod=rep(c('Moderate', 'Severe'), each=5000)) %>%
+  na.omit
+
+ggplot(erdf) + 
+  geom_line(aes(x=trees, y=var)) + 
+  facet_wrap(. ~ mod, scales='free_y') + 
+  theme_bw() + 
+  labs(x='Number of Trees', y='Error Rate')
+ggsave('Error_Over_Training.pdf', width=7, height=3.75)
 
 ##################################
 # assess residuals
@@ -75,37 +96,9 @@ sev.res <- ggplot(moddat) +
[…]

### 2020-11-15 viz/Viz_Mod_Results.R

@@ -47,89 +47,33 @@ system('pdfcrop FullMap.pdf FullMap.pdf')
 ###########################
 # Setup Reference Map
 ############################
-region_cols <- c("High income"="#279B48",
-                 "Upper-middle income"="#02558B",
-                 "Lower-middle income"="#F36D25",
-                 "Low income"="#EB1C2D")
+region_cols <- c("High Income"="#279B48",
+                 "Upper-Middle Income"="#02558B",
+                 "Lower-Middle Income"="#F36D25",
+                 "Low Income"="#EB1C2D")
 
-regdat <- cty %>% select(ISO3 = iso_a3, region_wb = income_grp) #use world bank regions by income
-
-# rename and change order from high income to low income
-regdat$region_wb[regdat$region_wb == "1. High income: OECD"] <- "High income"
-regdat$region_wb[regdat$region_wb == "2. High income: nonOECD"] <- "High income"
-regdat$region_wb[regdat$region_wb == "3. Upper middle income"] <- "Upper-middle income"
-regdat$region_wb[regdat$region_wb == "4. Lower middle income"] <- "Lower-middle income"
-regdat$region_wb[regdat$region_wb == "5. Low income"] <- "Low income"
+regdat <- cty %>% 
+  select(ISO3 = iso_a3, region_wb = income_grp) %>% #use world bank regions by income
+  # rename and change order from high income to low income
+  mutate(region_wb = case_when(region_wb == "1. High income: OECD" ~ "High Income",
+                               region_wb == "2. High income: nonOECD" ~ "High Income",
+                               region_wb == "3. Upper middle income" ~ "Upper-Middle Income",
+                               region_wb == "4. Lower middle income" ~ "Lower-Middle Income",
+                               region_wb == "5. Low income" ~ "Low Income"))
 
 
 reg <- ggplot(regdat) + 
-  geom_sf(aes(fill=factor(regdat$region_wb, levels = c("High income", 
-                                                       "Upper-middle income", 
-                                                       "Lower-middle income", 
[…]

### 2020-11-17 docs/modeling_fies.tex
almost done with draft; improve validation

@@
 When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.
-We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find reasonable accuracy ($R^2 = 0.799$, MAE = 0.068 for moderate food insecurity and $R^2 = 0.772$, MAE = 0.038 for severe food insecurity).
-This suggestions that our model explains between 77\% to 79\% of the variation in food security outcomes and is on average, within 3-7 percentage points in its estimates of rates of food insecurity, when applied to countries outside of the training set.
+We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 3-7 percentage points on average when applied to countries outside of the training set.
 
@@
 A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.
+By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.
 Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.
+We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.
 
 To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.
-Key hyperparameters include the number of decision trees to generate, the number of variables randomly selected as candidates for splitting a node, and the average number of observations in a leaf node.
-We calculated the Out-Of-Bag (OOB) error for different combinations of hyperparameters for the moderate and severe models.
-By choosing the combinations of hyperparameters with the smallest OOB error we found the optimal combination for both models.
-We found that 5000 trees are more than sufficient for the error rates to converge.
+These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.
+We use the combination of hyperparameters that perform best under ten-fold cross validation based on $R^2$.
 
-After fitting the models, we examined the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
-We used a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable}.
-Similarly to model predictions, we transformed the error term with the inverse logit so that the errors would be on the scale of the outcome variable and therefore more interpretable.
+When performing cross validation, it is typical to select observations at random.
+However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.
+Thus, for each iteration of cross validation, we leave out data from 10\% of the countries in our dataset, rather than 10\% of the total observations.
+Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.
+By this validation metric, we find that our models perform well, with an $R^2$ of 0.799 and mean average error (MAE) of 0.068 for moderate food insecurity and an $R^2$ of 0.772 and a MAE of 0.038 for severe food insecurity.
+After fitting models on the entire dataset, we observe an $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.944 and a MAE of 0.019 for severe food insecurity.
 
-\subsection{Model Validation}
-We observed a MAE of 0.0029 for our model of severe food insecurity ($R^2$ of 0.9978) and an MAE 0.0054 for our model of moderate food insecurity ($R^2$ of 0.9981).
-This means that, on average, our model predictions were only of by 0.29\% and 0.54\%, respectively, for the subnational areas in countries where we had microdata.
+After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
+We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable, breiman2001random}.
+These metric can provide some interpretability to model predictions.
 
-To further validate the models, we ran a ten-fold cross validation at the country level.
-We randomly divided the countries into ten parts, and re-fit the model 10 times, with each part left out one.
-Then, we predicted the values for the ten percent that was left out using the model fit with 90\% of the countries.
-We randomly left out data at the country level because within each country we had values from different subnational areas and different years, and these values are likely to be very similar to each other.
-Because we are extending our models predictions to countries that were not found at all in our training data, a better test of our model is to assess its predictive accuracy on countries that were entirely absent from the training data.
-By this validation metric, we found that our model performed reasonably well, with an $R^2$ of 0.778 and an MAE of 0.073 for moderate food insecurity and an $R^2$ of 0.696 and a MAE of 0.044 for severe food insecurity.
-
-For a detailed description, implementation and validation of the random forest models see the Supplemental Info.
-
-\section{Conclusion}
+The Supplementary Materials provide a detailed description of the implementation and validation of the random forest models.
 
 \printbibliography
-
-\section*{Supplemental Info}
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\setcounter{section}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thesection}{S\arabic{section}}
-
-\section{Overview of the FIES}
-We used raw microdata released by the FAO from 75 countries that vary by world region and income level (See Fig. \ref{fig:fies_countries}.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/FIES_Countries.pdf}
-  \caption{Countries that had microdata on the FIES and were used in our model, shown in blue.}
-  \label{fig:fies_countries}
-\end{figure}
-
-This microdata comes with the probability that an individual is over the threshold for moderate and severe food insecurity already calculated by the FAO, based on their responses to the 8 FIES questions (See List \ref{itm:fies}).
-Thus, we did not perform those calculations ourselves.
-Nevertheless, we give here an overview of the procedure.
-For more detail see Cafiero (2018) in \textit{Measurement} \citep{Cafiero2018}.
-
-The FIES is calculated using a Rash model, which was developed by the psychometrics literature.
-The Rash model assumes that each individual and their responses to the FIES questions can be placed on a one-dimensional scale of food insecurity, and that the log odds of a respondent \textit{r} answering yes to one of the FIES questions \textit{i} is a linear function of the difference between the severity of the food insecurity experienced by \textit{r} and the severity of item {i}.
-
-The severity of each item and each respondents level of food insecurity can be estimated with Maximum Likelihood.
-After fitting the model to each country, the FAO found that the assumptions of the Rash model held up well, as evidenced by model fit diagnostics.
-
-After separately estimating the Rash model for each country, the FAO then developed a global reference scale for the severity of each of the 8 questions, by iteratively harmonizing the severity values in each country.
-Then, these global reference points were calibrated against other surveys that had asked questions similar to the FIES, such as the HFSSM and the ELCSA.
-
-Based on the global reference scale, two thresholds are set based on questions 1 and 8: whether the responded was worried they would not have enough food to eat because of a lack of money or other resources, and whether there was a time when they went without eating for a whole day because of a lack of money or other resources.
-
-%%% Discuss reference scale and calibration
-%%% Map of countries with Training Data
-\begin{enumerate}
-	\item During the last 12 MONTHS, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
-	\item Still thinking about the last 12 MONTHS, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
-	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
-	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
-	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
-	\item Was there a time when your household ran out of food because of a lack of money or other resources?
-	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
-	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
-  \label{itm:fies}
-\end{enumerate}
-
-\section{Predicting Covariates Into the Future}
-\subsection{Annualized Rate of Change}
-For many of our covariates, to extrapolate recent trends to the future, it is necessary to draw on annualized rates of change, which are then used to anticipate the rate of future change.
-This method can be used for any variable that is a fraction or a rate.
-We use this method to estimate future rates of stunting, wasting, and malaria mortality, as well as to model future subnational shares of GDP and population.
-Thus, we give an overview of the methodology here.
-
-First, for each subnational area, $a$, the rate of change (ROC) is calculated between each pair of adjacent years, $y$, based on a value, $p$, such that $0 < p < 1$, available for each year.
-
-\begin{equation}
-  \text{ROC}_{a,y} = \text{logit} \left( \frac{p_{a,y}}{p_{a,y-1}} \right)
-  \label{eqn:a}
-\end{equation}
-
-Then, each ROC is weighted to give more weight to recent years, such that weight $W$ at year $y$ is.  For a dataset with observations for 2010 to 2017, that would be:
-
-\begin{equation}
-  W_y = \frac{y - 2010}{\sum_{y=2010}^{2017} y - 2010}
-  \label{eqn:b}
-\end{equation}
-
-Then, the annualized rate of change (AROC) is calculated as:
-
-\begin{equation}
-  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2017} W_y \times \text{AROC}_{a, y} \right)
-  \label{eqn:c}
-\end{equation}
-
-Finally, the projections, Proj, for each year from 2018 to 2030 are given as:
-
-\begin{equation}
-  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2017} ) + AROC_{a} \times ( y - 2017 ) )
-  \label{eqn:d}
-\end{equation}
-
-For cases where we are modeling shares of a whole, for example when modeling subnational shares of population or GDP, we then re-scale the shares to ensure they total to 1.
-
-\subsection{Population}
-Although population is not a covariate in our random forest model, it is an important precursor to other covariates, such as GDP per capita, and future estimates of population are also necessary to convert modeled future rates of food insecurity into total headcounts of food insecurity.
-Thus, in addition to our other covariates, we also modeled population into the future.
-
-We do this by combining historic subnational data from the Subnational Development Database \citep{Smits2019} with national level population estimates for the 21st century \citep{KC2017}.
-To account for within-country trends and changes in population distribution, we use the AROC method outlined in equations \ref{eqn:a} - \ref{eqn:d} to project each subnational areas share of national population totals.
-We then disaggregate the national totals given by KC et al.
-to estimate future subnational populations.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/population.png}
-  \caption{Population}
-\end{figure}
-
-\subsection{Urban Percentage}
-
-We drew our historic and future estimates of urbanization entirely from Jones and O'Neill, published in \textit{Environmental Research Letters} \citep{Jones2016}, who modeled spatially explicit scenarios of urbanization consistent with the Shared Socioeconomic Pathways.
-As with our other covariates, we used estimates for SSP2, the middle-of-the-road scenario.
-Because definitions of ``urban" and ``rural" vary significantly across datasets, using one dataset that is entirely consistent within itself was an important consideration.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/urban_perc.png}
-  \caption{Percentage of people living in urban areas.}
-\end{figure}
-
-\subsection{Wasting}
-To estimate the prevalence of wasting for each administrative area in our dataset, we used data from the Local Burden of Disease project published in \textit{Nature} \citep{Local2020}.
-For the years 2010 to 2017, we simply took the mean rate of wasting in each administrative area from the dataset.
-Higher-income countries that were not included in the dataset were modeled as having a rate of 0 wasting.
-We then modeled wasting for the years 2018-2030 using the AROC method, which the Local Burden of Disease group similarly used to estimate wasting for the year 2025 \citep{Local2020}.
-
-To account for the effects of the novel coronavirus on global rates of wasting, we assumed that long-term trends and rates of wasting would hold steady, but that they increase globally by 14.3\% in 2020, based on estimates published in \textit{The Lancet} \citep{headey2020impacts}.
-We model this impact as uniform across all countries where wasting occurs.
-Then we model the rate of wasting in each administrative area in 2021 as being the mean of the rate in 2020 that accounts for the shock and the previously modeled rate for 2022.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/wasting.png}
-  \caption{Prevalence of wasting.}
-\end{figure}
-
-\subsection{Stunting}
-
-We model stunting using the same methodology we use to summarize and model wasting, because the Local Burden of Disease dataset that we draw on for wasting also includes stunting.
-We also included the 14.3\% increase in prevalence for the year 2020 that was estimated for wasting \citep{headey2020impacts}, even though stunting is a long-term consequence of malnutrition and will likely not be as readily observable in a population as wasting would be.
-However, we use stunting in our model not as a cause of food insecurity, but rather as a proxy for chronic conditions of hunger, which are exacerbated by the coronavirus pandemic.
-Thus, estimating an impact of the coronavirus shock on stunting is appropriate for our model.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/stunting.png}
-  \caption{Prevalence of stunting.}
-\end{figure}
-
-\subsection{GDP Per Capita}
-We estimate global subnational GDP per capita using historic subnational estimates of Gross National Income (GNI) from the Subnational Human Development Database \cite{Smits2019} and forecasts of national GDP within the SSP framework from Dellink et al.
-\cite{Dellink2017}.
-We first use data from the World Bank to estimate the mean country-level ratio of GNI to GDP and adjust the subnational estimates of GNI according to that country-specific ratio.
-We then calculate each subnational area's share of country GDP, and then use the AROC method to estimate each subnational area's share of total country GDP.
-Based on these trajectories in each subnational areas share of total country GDP, we estimate what each subnational area's share of GDP will be for each year from 2019 to 2030.
-We divide that share of GDP by the population of the administrative area, which we had previously estimated.
-Thus, national GDP totals are consistent with the projections of Dellink et al., but are disaggregated according to the historic patters of GDP share in the Subnational Human Development Database.
-
-Finally, we used estimates of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the coronavirus \citep{prospects2020}.
-We then model GDP as resuming its previous growth trajectory based on previously-estimated growth rates from 2022-2030, but growing from an estimate for 2021 that includes the coronavirus shock.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/gdp_percap.png}
-  \caption{GDP per capita.}
-\end{figure}
-
-\subsection{Mean Years of Schooling}
-We estimate global subnational mean years of schooling using historic subnational estimates from the Subnational Human Development Database \cite{Smits2019} and forecasts of national mean years of schooling from KC et al.
-\cite{KC2017}.
-Using the historic subnational data, we estimate each subnational administrative area's difference in years of schooling from the national mean, and use this value to disaggregate the future projections to a subnational level.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/school_mean.png}
-  \caption{Mean years of schooling.}
-\end{figure}
-
-\subsection{Gini Coefficient}
-We use estimates of national income inequality from Rao et al. \citep{Rao2019a}.  These estimates include all years in our analysis, and are not disaggregated to a subnational level.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/gini.png}
-  \caption{Gini Coefficient}
-\end{figure}
-
-\subsection{Poverty Headcount Index}
-We used estimates of the Poverty Headcount Index for the recent past and future from the World Poverty Clock by the World Data Lab \citep{Cuaresma2018}, updated to include the impact of the coronavirus.
-In our model we include national level poverty rates at a threshold of \$1.90 which is defined as \textit{extreme} poverty.
-By using the concept of parameterized Lorenz curves in combination with mean income/consumption and population projections future poverty rates are calculated.
-For more details on the methodology and information on global and country level poverty check the website \url{worldpoverty.io}.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/hci.png}
-  \caption{Poverty Headcount Index}
-\end{figure}
-
-\subsection{Water Scarcity Share}
-We used the water scarcity index (WSI) developed by \citep{greve2018global}.
-The index represents the ratio of the decadal averages in water demand to water supply and it assesses changes in water scarcity on a 0.5 by 0.5 degree global grid.
-Based on this index, we estimate the share of people living in areas with less than 1000m\textsuperscript{3} of water available per capita per year.
-The WSI has been used already to build the water scarcity clock.
-For more information visit \href{https://worldwater.io/}{worldwater.io}.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/ws_share.png}
-  \caption{Water Scarcity Share}
-\end{figure}
-
-\subsection{Mean Annual Precipitation}
-For mean annual precipitation, we use data for the years 2010 to 2019 from the TerraClimate dataset \citep{abatzoglou2018terraclimate}, summarized to each subnational area.
-For data for the years 2020-2030, we use models from the Inter-Sectoral Model Intercomparison Project \citep{warszawski2014inter}.
-Specifically, we use an ensemble of bias-corrected projections under representative concentration pathway (RCP) 6.0, taking the mean of the ensemble.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/precip.png}
-  \caption{Mean annual precipitation}
-\end{figure}
-
-\subsection{Mean Temperature}
-For mean temperature, we used the same datasets that we'd used for precipitation, as each of those datasets included both temperature in addition to precipitation.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/tave.png}
-  \caption{Mean temperature (Celsius)}
-\end{figure}
-
-\subsection{Topographic Ruggedness}
-As a measure of accessibility, we include the mean topographic ruggedness of each subnational area.
-We first used a gridded dataset of elevation from the USGS \cite{USGS1996} and calculated the index at each grid cell using the methodology from Riley et al.
-\citep{Riley1999}.
-We then summarized the values of all the grid cells within each administrative area.
-Because this variable does not change over time, it is constant across all years in the model.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/ruggedness.png}
-  \caption{Topographic ruggedness}
-\end{figure}
-
-\subsection{Malaria (\textit{P. falciparum}) Mortality Rate}
-We used data from \textit{The Lancet} to estimate the mortality (deaths per 100,000 population per year) due to Malaria \citep{Weiss2019}, taking the average pixel values within each subnational administrative area.
-We then estimated future malaria mortality based on AROC method.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/covars/mal_falciparum.png}
-  \caption{Rate of mortality due to \textit{P. falciparum} Malaria}
-\end{figure}
-
-\section{Description, Implementation and Validation of the RF regression model}
-
-The RF regression is a machine learning method based on the creation of a large number of decision trees.
-By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to single decision trees \citep{friedman2001elements}.
-For each of the decision trees in a random forest model, observations are selected at random with replacement, a method known as bootstrap aggregating, or bagging, and features are area also selected at random.
-
-For implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}.
-Specifically, we apply the functions \texttt{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \texttt{rfsrc()} to perform the RF regression models.
-After predicting rates of moderate-to-severe and severe food insecurity we plot the modeled vs.
-observed rates and calculated the MSE and ${R}^2$ (See Fig.
-\ref{fig:rf_in-sample} and Fig.
-\ref{fig:rf_out-sample}) for both models.
-Additionally, we use the function \texttt{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig.
-\ref{fig:rf_error}) and the importance of individual variables (See Fig.
-\ref{fig:rf_vimp}).
-
-%STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
-Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.
-Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.
-
-To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions. 
-%MAYBE STATE THIS SOMEWHERE ELSE, NOT SURE WHERE
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/model/error_rf.png}
-  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_error}
-\end{figure}
-
-%HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH
-
-In addition to the parameter that describes the number of trees to be created, two other important hyperparameters must be tuned.
-In the function \texttt{tune.rfsrc()} these parameters are called \texttt{mtry} and \texttt{nodesize}.
-\texttt{mtry} describes the number of variables randomly selected as candidates for splitting a node and \texttt{nodesize} describes the average number of observations in a leaf node.
-The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest out-of-bag (OOB) error.
-
-%EXPLAIN THE OOB ERROR HERE?
-
-The function \texttt{tune.rfsrc()} does exactly that and returns optimal values for \texttt{mtry} and \texttt{nodesize}.
-We applied this function on both, the moderate-to-severe and the severe model.
-For the moderate-to-severe model the optimal combination is \texttt{mtry} = 12 and \texttt{nodesize} = 1 and for the severe model we found \texttt{mtry} = 10 and \texttt{nodesize} = 2 to be optimal.
-
-%STATE FUNCTION ARGUMENTS HERE?
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
-  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_in-sample}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
-  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_out-sample}
-\end{figure}
-
-%HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
-%For modeling the rates of moderate-to-severe and severe food insecurity we use the function \texttt{rfsrc()}. We specify
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/VIMP.pdf}
-  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_vimp}
-\end{figure}
-
-%https://kogalur.github.io/randomForestSRC/theory.html#section4.2
-Lastly we asses the importance of individual variables.
-This is based on the Breiman-Culter permutation variable importance information \citep{breiman2001random}.
-This involved comparing the prediction error on the OOB data to the prediction error of OOB cases where the x-variable $x$ is randomly permutated.
-The importance of an individual variable is thus the mean difference between the perturbed and unperturbed error rate, across all trees.
-%MORE INFORMATION HERE
-\end{document}

### 2020-11-17 docs/modeling_fies_supplement.tex
almost done with draft; improve validation

\documentclass{article}

\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[a4paper, margin=1.25in]{geometry}
\usepackage{breakcites}
\usepackage{subcaption}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{textgreek}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{booktabs}
\usepackage{fancyhdr}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{lineno}
\usepackage[
  style=numeric,
  citestyle=numeric-comp,
  backend=biber,
  doi=true,
  natbib=true,
  sorting=none
]{biblatex}

\pagestyle{fancy}
\fancyhf{}
\lfoot{Supplemental Materials for \textit{A Global Picture of Hunger}}
\rfoot{\thepage}

\renewcommand{\footrulewidth}{0.4pt}
\renewcommand{\headrulewidth}{0pt}

\addbibresource{library.bib}

\begin{document}

\begin{center}
\section*{Supplemental Info for \textit{A Global Picture of Hunger}}
\end{center}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{section}{0}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}

\section{Overview of the FIES}
We use raw microdata released by the FAO from 75 countries from different world regions and income categories (See Fig. \ref{fig:fies_countries}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/FIES_Countries.pdf}
  \caption{Countries used in our model, shown in blue.}
  \label{fig:fies_countries}
\end{figure}

The microdata includes information about the probability that an individual is over the threshold for moderate and severe food insecurity, calculated by the FAO on the basis of the responses to the 8 FIES questions (See List \ref{itm:fies}).  The FIES is calculated using a Rasch model, which was developed by the psychometrics literature.  The Rasch model assumes that each individual and their responses to the FIES questions can be placed on a one-dimensional scale of food insecurity, and that the log odds of a respondent answering affirmatively to one of the FIES questions is a linear function of the difference between the severity of the food insecurity experienced by the individual and the severity of the item put forward in the corresponding question.

The severity of each item and the respondent’s level of food insecurity can be estimated making use of maximum likelihood methods. The assumptions of the Rasch model hold up well for the data at hand, as evidenced by model fit diagnostics \cite{Cafiero2018}.

After separately estimating the Rasch model for each country, FAO develops a global reference scale for the severity of each of the eight questions by iteratively harmonizing the severity values in each country. These global reference points are calibrated against other surveys that ask questions similar to those in the FIES, such as the HFSSM and the ELCSA.

Based on the global reference scale, two thresholds are set based on questions 1 and 8: whether the respondent is worried they would not have enough food to eat because of a lack of resources, and whether there was a time when they went without eating for a whole day because of a lack of money or other resources.

\begin{enumerate}
	\item During the last 12 MONTHS, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
	\item Still thinking about the last 12 MONTHS, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
	\item Was there a time when your household ran out of food because of a lack of money or other resources?
	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
  \label{itm:fies}
\end{enumerate}

\section{Predicting Covariates Into the Future}
\subsection{Annualized Rate of Change}
For many of our covariates, to extrapolate recent trends to the future, we draw on historically annualized rates of change, which are used to project the future change in the variable. We use this method to estimate future rates of stunting, wasting, and malaria mortality, as well as to model future subnational shares of GDP and population.

For each subnational area, $a$, the rate of change (ROC) is calculated between each pair of adjacent years, $y$, based on a value, $p$, such that $0 < p < 1$, available for each year.

\begin{equation}
  \text{ROC}_{a,y} = \text{logit} \left( \frac{p_{a,y}}{p_{a,y-1}} \right)
  \label{eqn:a}
\end{equation}

Each ROC is weighted so as to give proportionally more weight to recent years.  For a dataset with observations for 2010 to 2017, that would be:

\begin{equation}
  W_y = \frac{y - 2010}{\sum_{y=2010}^{2017} y - 2010}
  \label{eqn:b}
\end{equation}

The annualized rate of change (AROC) is calculated as:

\begin{equation}
  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2017} W_y \times \text{AROC}_{a, y} \right)
  \label{eqn:c}
\end{equation}

Finally, the projections, Proj, for each year up to 2030 are given by:

\begin{equation}
  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2017} ) + AROC_{a} \times ( y - 2017 ) )
  \label{eqn:d}
\end{equation}

For cases where we model shares with a summing constraint, such as the regional shares of a variable measured at the national level, we rescale the shares to ensure that the constraint holds.

\pagebreak
\subsection{Population}
Although population is not a covariate in our random forest model, it is an important determinant of other covariates such as GDP per capita. Future estimates of population are also necessary to convert modeled future rates of food insecurity into total headcounts of food insecurity.

We project population by combining historical subnational data from the Subnational Development Database \citep{Smits2019} with national level population estimates for the 21st century \citep{KC2017}.  To account for within-country trends and changes in population distribution, we use the AROC method outlined in equations \ref{eqn:a} - \ref{eqn:d} to project each subnational areas share of national population totals. We then disaggregate the national totals given by KC et al. to estimate future subnational populations.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/population.png}
  \caption{Population}
\end{figure}

\pagebreak
\subsection{Urbanization Rates}

We draw our historical and future estimates of urbanization entirely from the spatially explicit scenarios of urban population by Jones and O'Neill \citep{Jones2016}, which provide a series of urbanization rate projections which are consistent with the Shared Socioeconomic Pathways.  As with our other covariates, we use projections consistent with SSP2, the middle-of-the-road scenario.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/urban_perc.png}
  \caption{Percentage of people living in urban areas.}
\end{figure}

\pagebreak
\subsection{Wasting}
To estimate the prevalence of wasting for each administrative area in our dataset, we use data from the Local Burden of Disease project.  For the years 2010 to 2017, we simply use the mean rate of wasting in each administrative area from the dataset.  Higher-income countries that are not included in the dataset are assumed to have zero wasting.  We then project wasting for the years 2018-2030 using the AROC method, which the Local Burden of Disease group similarly use to estimate wasting for the year 2025 \citep{Local2020}.

To account for the effects of the coronavirus pandemic on global rates of wasting, we assume that long-term trends in rates of wasting would hold steady, after they increase globally by 14.3\% in 2020, based on estimates published in \textit{The Lancet} \citep{headey2020impacts}.  We model this impact as uniform across all countries where wasting occurs.  We project the rate of wasting in each administrative area in 2021 as being the average of (a) the rate of change in 2020 accounting for the pandemic shock and (b) the predicted rate of change for 2022 obtained ignoring the pandemic.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/wasting.png}
  \caption{Prevalence of wasting.}
\end{figure}

\pagebreak
\subsection{Stunting}
We model stunting, sourced from the Local Burden of Disease dataset, using the same methodology employed for wasting. We also assume a 14.3\% increase in prevalence for the year 2020 \citep{headey2020impacts}, as is done for wasting, although stunting is a long-term consequence of malnutrition and is not likely to be observable in a population with the same lags as wasting would be. We use stunting in our modeling framework as a proxy for chronic conditions of hunger, which are probably exacerbated by the coronavirus pandemic.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/stunting.png}
  \caption{Prevalence of stunting.}
\end{figure}

\pagebreak
\subsection{GDP Per Capita}
We estimate global subnational GDP per capita using historic subnational estimates of Gross National Income (GNI) from the Subnational Human Development Database \cite{Smits2019} and forecasts of national GDP within the SSP framework from Dellink et al. \cite{Dellink2017}.  We first use data from the World Bank to estimate the mean country-level ratio of GNI to GDP and adjust the subnational estimates of GNI according to that country-specific ratio.  We then calculate each subnational area's share of country GDP, and then use the AROC method to estimate each subnational area's share of total country GDP.  Based on these trajectories in each subnational areas share of total country GDP, we estimate what each subnational area's share of GDP will be for each year from 2019 to 2030.  We divide that share of GDP by the population of the administrative area, which we had previously estimated.  Thus, national GDP totals are consistent with the projections of Dellink et al., but are disaggregated according to the historic patters of GDP share in the Subnational Human Development Database.

Finally, we used estimates of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the coronavirus \citep{prospects2020}.  We then model GDP as resuming its previous growth trajectory based on previously-estimated growth rates from 2022-2030, but growing from an estimate for 2021 that includes the coronavirus shock.

We measure global subnational GDP per capita using historic subnational estimates of Gross National Income (GNI) from the Subnational Human Development Database \citep{Smits2019} and forecasts of national GDP within the SSP framework form Dellink et al. \citep{Dellink2017}. We source data on the mean country-level ratio of GNI to GDP from the World Bank and adjust the subnational estimates of GNI according to the ratio for each country in our sample. We calculate the share of national GDP for each subnational region, and use the AROC method to project each subnational area’s share of total country GDP up to 2030. National GDP totals are consistent with the projections of Dellink et al., but are disaggregated according to the historic patterns of subnational GDP shares sourced from the Subnational Human Development Database. We use nowcasts and forecasts of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the coronavirus \citep{prospects2020}. After the shock created by the COVID pandemic, GDP is assumed to resuming its previous growth trajectory for the period 2022-2030.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/gdp_percap.png}
  \caption{GDP per capita.}
\end{figure}

\pagebreak
\subsection{Mean Years of Schooling}
We use data on global subnational mean years of schooling from the Subnational Human Development Database \cite{Smits2019} and forecasts of national mean years of schooling from KC et al. \cite{KC2017}.  Using the subnational data, we calculate the difference in years of schooling form the nation mean each subnational administrative area, and use this value to disaggregate the future projections to a subnational level, keeping these differences constant over the projection horizon.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/school_mean.png}
  \caption{Mean years of schooling.}
\end{figure}

\pagebreak
\subsection{Gini Coefficient}
We use data on historical levels and projections of national income inequality from Rao et al. \citep{Rao2019a}.  These estimates are not disaggregated to a subnational level.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/gini.png}
  \caption{Gini Coefficient}
\end{figure}

\pagebreak
\subsection{Poverty Headcount Index}
We used historical and projected data on the Poverty Headcount Index (using the standard threshold of \$1.90 a day) from the World Poverty Clock by the World Data Lab, which in turn uses Povcal data sourced from the World Bank \citep{Cuaresma2018}. These data are the version corresponding to the update in summer 2020 and include the expected impact of the COVID pandemic on global poverty dynamics. Using parameterized Lorenz curves in combination with mean income/consumption and population projections, poverty rates are projected.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/hci.png}
  \caption{Poverty Headcount Index}
\end{figure}

\pagebreak
\subsection{Water Scarcity}
We use data for the water scarcity index (WSI) developed by \citep{greve2018global}. The index represents the ratio of the decadal averages in water demand to water supply, based on population projections consistent with SSP2 and water availability projections from global hydrological and general circulation models.  Based on this index, we estimate the share of people living in areas with less than 1000m\textsuperscript{3} of water available per capita per year. The WSI has been used already to build the water scarcity clock.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/ws_share.png}
  \caption{Water Scarcity Share}
\end{figure}

\pagebreak
\subsection{Mean Annual Precipitation and Mean Temperature}
For mean annual precipitation and temperature, we use data for the years 2010 to 2019 from the TerraClimate dataset \citep{abatzoglou2018terraclimate}, aggregated to each subnational area.  For data for the years 2020-2030, we use projections from models from the Inter-Sectoral Model Intercomparison Project \citep{warszawski2014inter}.  Specifically, we use the mean of an ensemble of bias-corrected projections under representative concentration pathway (RCP) 6.0.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/precip.png}
  \caption{Mean Annual Precipitation}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/tave.png}
  \caption{Mean Temperature (Celsius)}
\end{figure}

\pagebreak
\subsection{Topographic Ruggedness}
As a measure of accessibility, our models include the mean topographic ruggedness of each subnational area as a covariate. We obtain the variable using a gridded dataset of elevation from the USGS \citep{USGS1996} and calculate the index at each grid cell using the methodology from Riley et al. \citep{Riely1999}. We then aggregate the values for all the grid cells within each administrative area. The variable is constant in the estimation and projection period.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/ruggedness.png}
  \caption{Topographic ruggedness}
\end{figure}

\pagebreak
\subsection{Malaria (\textit{P. falciparum}) Mortality Rate}
We used data from Weiss et al. to estimate the mortality (deaths per 100,000 population per year) due to Malaria \citep{Weiss2019}, and aggregate the information at the level of subnational administrative area. Projections of malaria mortality are based on the AROC method.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/covars/mal_falciparum.png}
  \caption{Rate of mortality due to \textit{P. falciparum} Malaria}
\end{figure}

\pagebreak
\section{Description, Implementation and Validation of the Random Forest Regression Model}
The RF regression is a machine learning method based on the creation of a large number of decision trees. By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.  For each of the decision trees in a random forest model, observations are selected at random with replacement, a method known as bootstrap aggregating, or bagging, and features are also selected at random.

For the implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}. Specifically, we apply the functions \texttt{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \texttt{rfsrc()} to perform the RF regression models. After predicting rates of moderate-to-severe and severe food insecurity, we plot the modeled vs. observed rates and calculated the MSE and ${R}^2$ (See Fig. \ref{fig:rf_in-sample} and Fig. \ref{fig:rf_out-sample}) for both models. Additionally, we use the function \texttt{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig. \ref{fig:rf_error}) and the importance of individual variables (See Fig. \ref{fig:rf_vimp}).

%STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.  Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.

To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions. 
%MAYBE STATE THIS SOMEWHERE ELSE, NOT SURE WHERE

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/error_rf.png}
  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_error}
\end{figure}

%HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH

In addition to the parameter that describes the number of trees to be created, two other important hyperparameters must be tuned. In the function \texttt{tune.rfsrc()} these parameters are called \texttt{mtry} and \texttt{nodesize}. \texttt{mtry} describes the number of variables randomly selected as candidates for splitting a node and \texttt{nodesize} describes the average number of observations in a leaf node. The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest out-of-bag (OOB) error.

%EXPLAIN THE OOB ERROR HERE?

The function \texttt{tune.rfsrc()} does exactly that and returns optimal values for \texttt{mtry} and \texttt{nodesize}. We applied this function on both, the moderate-to-severe and the severe model. For the moderate-to-severe model the optimal combination is \texttt{mtry} = 12 and \texttt{nodesize} = 1 and for the severe model we found \texttt{mtry} = 10 and \texttt{nodesize} = 2 to be optimal.

%STATE FUNCTION ARGUMENTS HERE?

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_in-sample}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_out-sample}
\end{figure}

%HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
%For modeling the rates of moderate-to-severe and severe food insecurity we use the function \texttt{rfsrc()}. We specify

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/VIMP.pdf}
  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_vimp}
\end{figure}

%https://kogalur.github.io/randomForestSRC/theory.html#section4.2
Lastly we asses the importance of individual variables. This is based on the Breiman-Culter permutation variable importance information \citep{breiman2001random}.  This involved comparing the prediction error on the OOB data to the prediction error of OOB cases where the x-variable $x$ is randomly permutated.  The importance of an individual variable is thus the mean difference between the perturbed and unperturbed error rate, across all trees. 
%MORE INFORMATION HERE
\end{document}

\printbibliography

### 2020-11-17 models/model_randomforest.R

@@ -48,12 +48,10 @@ prm$ix <- 1:nrow(prm)
 #Run model under 10-fold cross validation
 #at the country level
 ##########################################
-iso3s <- unique(moddat$ISO3)
-samp <- sample(1:10, length(iso3s), replace=T)
 for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
   cat(round(sum(!is.na(prm$sev.rsq))/nrow(prm)*100, 2), 'percent done!\n')
-  for (s in 1:10){
-    ix <- moddat$ISO3 %in% iso3s[samp != s]
+  for (iso3 in unique(moddat$ISO3)){
+    ix <- moddat$ISO3 != iso3
     
     mtry <- prm$mtry[i]
     node <- prm$node[i]
@@ -79,14 +77,14 @@ for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
                     nodesize = node,
                     depth = depth)
       
-    moddat$fies.mod.pred[!ix] <- inv.logit(predict(rf.mod, moddat[!ix,])$predicted)
-    moddat$fies.sev.pred[!ix] <- inv.logit(predict(rf.sev, moddat[!ix,])$predicted)
+    moddat$fies.mod.pred.cv[!ix] <- inv.logit(predict(rf.mod, moddat[!ix,])$predicted)
+    moddat$fies.sev.pred.cv[!ix] <- inv.logit(predict(rf.sev, moddat[!ix,])$predicted)
     
   }
-  prm$sev.rsq[i] <- cor(moddat$fies.sev.pred, moddat$fies.sev)^2
-  prm$mod.rsq[i] <- cor(moddat$fies.mod.pred, moddat$fies.mod)^2
-  prm$sev.mae[i] <- mean(abs(moddat$fies.sev.pred - moddat$fies.sev))
-  prm$mod.mae[i] <- mean(abs(moddat$fies.mod.pred - moddat$fies.mod))
+  prm$sev.rsq[i] <- cor(moddat$fies.sev.pred.cv, moddat$fies.sev)^2
+  prm$mod.rsq[i] <- cor(moddat$fies.mod.pred.cv, moddat$fies.mod)^2
+  prm$sev.mae[i] <- mean(abs(moddat$fies.sev.pred.cv - moddat$fies.sev))
+  prm$mod.mae[i] <- mean(abs(moddat$fies.mod.pred.cv - moddat$fies.mod))
 }
[…]

### 2020-11-17 viz/Error_and_VIP.R

@@ -95,10 +95,36 @@ sev.res <- ggplot(moddat) +
        x='Observed Rates Of Severe Food Insecurity',
        y='Modeled Rates Of Severe Food Insecurity')	; sev.res
 
-plot_grid(mod.res, sev.res, align='h',labels='AUTO')
-ggsave('in-sample_rf.png', width=14, height=7)
+grd <- plot_grid(mod.res, sev.res, align='h',labels='AUTO')
+ggsave(grd, filename='docs/img/in-sample_rf.png', width=14, height=7)
 
-###############################
-# Error over training period
-##################################
+#########################################
+# Out of Sample Residuals
+#########################################
+mae.mod <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred.cv))
+mae.sev <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred.cv))
+
+r2.mod <- cor(moddat$fies.mod, moddat$fies.mod.pred.cv)^2
+r2.sev <- cor(moddat$fies.sev, moddat$fies.sev.pred.cv)^2
+
+mod.res <- ggplot(moddat) + 
+  geom_point(aes(x=fies.mod, y=fies.mod.pred.cv), alpha = 0.3) + 
+  geom_abline(intercept = 0, slope = 1, color = "red", size = 0.5) +
+  xlim(0, 1) + ylim(0, 1) +
+  labs(caption=paste0('Mean Average Error: ',  round(mae.mod, 4),
+                      '\nR-Squared: ', round(r2.mod, 4)),
+       x='Observed Rates Of Moderat-to-Severe Food Insecurity',
+       y='Modeled Rates Of Moderat-to-Severe Food Insecurity'); mod.res
+
+sev.res <- ggplot(moddat) + 
+  geom_point(aes(x=fies.sev, y=fies.sev.pred.cv), alpha = 0.3) +
+  geom_abline(intercept = 0, slope = 1, color = "red", size = 0.5) +
+  xlim(0, 1) + ylim(0, 1) +
+  labs(caption=paste0('Mean Average Error: ',  round(mae.sev, 4),
+                      '\nR-Squared: ', round(r2.sev, 4)),
+       x='Observed Rates Of Severe Food Insecurity',
+       y='Modeled Rates Of Severe Food Insecurity')	; sev.res
+
+grd <- plot_grid(mod.res, sev.res, align='h',labels='AUTO')
+ggsave(grd, filename='docs/img/in-sample_rf.png', width=14, height=7)
 

### 2020-11-18 models/PCA.R
final FINAL hyperpar gridsearc

@@ -1,47 +1,0 @@
-setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
-
-preddatsum <- preddat %>%
-  group_by(ISO3, YEAR) %>%
-  summarize(fies.mod.pred = weighted.mean(fies.mod.pred*population, w=population),
-            fies.sev.pred = weighted.mean(fies.sev.pred*population, w=population)) %>%
-  group_by(ISO3) %>%
-  mutate(fies.mod.pred = fies.mod.pred/lag(fies.mod.pred),
-         fies.sev.pred = fies.sev.pred/lag(fies.sev.pred))
-
-# Run PCA regressions
-pcadatmod <- preddatsum %>%
-  select(ISO3, YEAR, fies.mod.pred) %>%
-  spread(YEAR, fies.mod.pred) %>%
-  ungroup %>%
-  select(-`2010`)
-
-pcamod <- princomp(pcadatmod %>% select(-ISO3))
-
-pcadatsev <- preddatsum %>%
-  select(ISO3, YEAR, fies.sev.pred) %>%
-  spread(YEAR, fies.sev.pred) %>%
-  ungroup %>%
-  select(-`2010`)
-
-pcasev <- princomp(pcadatsev %>% select(-ISO3))
-
-# Make Plot data
-pcaplt <- cbind(pcadatsev %>% 
-                     select(iso_a3=ISO3),
-                   pcamod$scores[ , 1:4] %>%
-                     data.frame %>%
-                     rename(mod1=Comp.1, mod2=Comp.2, mod3=Comp.3, mod4=Comp.4),
-                   pcasev$scores[ , 1:4] %>%
-                     data.frame %>%
-                     rename(sev1=Comp.1, sev2=Comp.2, sev3=Comp.3, sev4=Comp.4))
-
-# Plot
-world <- ne_countries(returnclass='sf')
-
-m <- merge(world %>%
-             select(iso_a3),
-           pcaplt)
-
-ggplot() + 
-  geom_sf(data=m, aes(fill=mod1)) + 
-  scale_fill_gradientn(colors=ihme_cols)

### 2020-11-18 models/model_randomforest.R

@@ -4,8 +4,6 @@
 # modeled for 2020, 2025, and 2030
 #######################################################
 
-set.seed(100)
-
 setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
 
 # read in data
@@ -28,21 +26,15 @@ moddat <- moddat %>%
   mutate(fies.mod.logit = logit(fies.mod),
          fies.sev.logit = logit(fies.sev))
 
-prm <- expand.grid(list(node=c(1:10, 12, 14, 16, 20, 25, 100),
-                        mtry=c(6:1),
-                        depth=c(1:7, -1)))
-
-prm2 <- expand.grid(list(node=c(1:10, 12, 14, 15, 16, 17, 18, 19, 20, 25, 100),
-                        mtry=c(13:1),
-                        depth=c(1:13, -1)))
-
-prm$sev.mae <- NA
-prm$mod.mae <- NA
-
-prm <- bind_rows(prm, prm2[!paste(prm2$node, prm2$depth, prm2$mtry, sep='-') %in% paste(prm$node, prm$depth, prm$mtry, sep='-'), ])
-
+prm <- expand.grid(list(node=c(1:10, 12, 14, 16, 18, 20),
+                        mtry=c(12:1),
+                        depth=c(1:12, -1)))
 prm$ix <- 1:nrow(prm)
 
+prm$sev.rsq <- NA
+prm$mod.rsq <- NA
+prm$sev.mae <- NA
+prm$mod.mae <- NA
 
 #############################################
 #Run model under 10-fold cross validation
@@ -98,9 +90,9 @@ system('~/telegram.sh "Done Running Random Forests"')
 # Re-run cross validation for best hyperparameters
 #################################################
 i <- mod.prm$ix
-for (s in 1:10){
-  ix <- moddat$ISO3 %in% iso3s[samp != s]
-  
+for (iso3 in unique(moddat$ISO3)){
+  ix <- moddat$ISO3 != iso3
+    
   mtry <- prm$mtry[i]
   node <- prm$node[i]
   if(prm$depth[i] < 0){
@@ -121,8 +113,8 @@ for (s in 1:10){
 }
 
 i <- sev.prm$ix
-for (s in 1:10){
-  ix <- moddat$ISO3 %in% iso3s[samp != s]
+for (iso3 in unique(moddat$ISO3)){
+  ix <- moddat$ISO3 != iso3
   
   mtry <- prm$mtry[i]
[…]

### 2020-11-18 docs/modeling_fies.tex
new draft with country-level leave-one-out cross validation

@@
 This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
-We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the developing world where over half the population is severely food insecure.
+We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the parts of developing world where over half the population is severely food insecure.
 Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been steadily declining, with large observed and forecasted gains in lower and lower-middle income countries, and decreases in total number of severely food insecure people.
@@
 \section{Main}
-Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
+Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of at least moderate food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
 The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger.
@@
 Such a model can be employed to obtain forecasts to 2030 and assess likely future trends in food insecurity.
-We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond country level to a sub-national level.
+We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond the country level to a sub-national level.
 To our knowledge, this is the first effort to model the FIES outcomes at that scale.
@@
 However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.
-Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.
+Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that are food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.
 Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.
@@
 \subsection{Food Insecurity over Time}
-Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and an additional 2.5 billion people are above the threshold for moderate food insecurity (See Fig.
+Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and 2.5 billion people are above the threshold for moderate food insecurity (See Fig.
 \ref{fig:timeseries}).
-These numbers are higher than the estimates in the 2019 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \ref{sofi2020}.
+These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \ref{sofi2020}.
 
 Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.
-However, for both indicators, these changes are very slight: from 2010 to 2030, the number of moderately food insecure people will increase 2\%, and the number of severely food insecure people will decrease 8\%.
+However, for both indicators, these changes are very slight: in 2030, the number of moderately food insecure people will have increased by 2.2\% since 2010, and the number of severely food insecure people will have decreased by 7.0\%.
 The increases in moderate food insecurity have been in lower and lower-middle income countries, while the increases in severe food insecurity are entirely from lower income countries.
@@
 Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.
-In spite of this progress in the bottom three income quartiles, in high-income countries, rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.
+In spite of this progress in the bottom three income quartiles, in high-income countries, the rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.
 Overall, the world has made steady but slow progress on reducing the rate of food insecurity.
@@
   \includegraphics[width=\linewidth]{img/Rates.pdf}
-  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020}
+  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020.}
   \label{fig:rates}
@@
   \centering
-  \includegraphics{img/FullMap.png}
+  \includegraphics{img/FullMap.pdf}
   \caption{Spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
@@
 \subsection{Important Predictors of Food Insecurity}
-Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.944$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.799$ and $R^2 = 0.772$ for moderate and severe food insecurity, respectively).
+Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.940$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.792$ and $R^2 = 0.746$ for moderate and severe food insecurity, respectively).
 The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.
@@
 We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the coronavirus pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.
-Nevertheless, lower and lower-middle income countries, are expected to see overall increases in the number of food insecure people.
+Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of food insecure people.
 
@@
 For both moderate and severe food insecurity, the Poverty Headcount Index, the Gini coefficient, and GDP per capita appear as highly relevant predictors of the rate of food insecurity.
-The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger is also an important predictor of food insecurity.
-However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality are less important as explanatory factors of changes in food insecurity.
+The rate of stunting, an indicator of both sanitary conditions and chronic, long-term under-nutrition is also an important predictor of food insecurity.
+However, other indicators of nutrition and sanitary conditions, such as the rates of wasting and malaria mortality are less important as explanatory factors of changes in food insecurity.
 Geographic and climate factors, such as the percentage of urban population, topographic ruggedness and mean annual temperature or precipitation do not appear to be robust predictors of food insecurity at a global scale.
-Finally, while economic conditions are widely associated with food insecurity at a global scale, the fact that high income countries have seen little improvement in food insecurity outcomes, in spite of marginal annual increases in GDP per capita, suggests that there is an upper limit to the extent to which increasing economic development can improve food security outcomes.
 
-Our results anticipating of increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
+While economic conditions, and especially GDP per capita, are important predictors of food insecurity at a global scale, there is an upper limit to the extent to which increasing economic development alone can improve food security outcomes.
+High income countries have seen little improvement in food insecurity outcomes over twenty years, and upper-middle income countries are expected to see diminishing gains throughout the 2020s.
+Among the high-income countries in our dataset, those with the lowest rates of moderate and severe food insecurity also had lower inequality and poverty.
+This suggests that our model's forecasted marginal increases in food insecurity in high income countries is due to trends of increasing income inequality, and that economic development alone cannot ensure that everyone is well-fed.
+
+Our results anticipating increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
 However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.
-In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 3\% decrease, from 788 to 767 million severely food insecure people over the same period.
+In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 2.6\% decrease, from 787 to 767 million severely food insecure people over the same period.
 
@@
 When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.
-We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 3-7 percentage points on average when applied to countries outside of the training set.
+We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.
 
@@
 We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.
-For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}, where available.
-This methodology rests on the mild assumption that subnational differences in rates of food insecurity are driven by factors related to demographics, urbanization, and wealth rather than by other unobservable factors.
+For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
+This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.
 
 \subsection{Covariates}
-We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).
+We model subnational rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).
 We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.
@@
 These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.
-For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.
 
@@
 However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.
-Thus, for each iteration of cross validation, we leave out data from 10\% of the countries in our dataset, rather than 10\% of the total observations.
+Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.
 Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.
-By this validation metric, we find that our models perform well, with an $R^2$ of 0.799 and mean average error (MAE) of 0.068 for moderate food insecurity and an $R^2$ of 0.772 and a MAE of 0.038 for severe food insecurity.
-After fitting models on the entire dataset, we observe an $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.944 and a MAE of 0.019 for severe food insecurity.
+By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.
+After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.
 
@@
 We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable, breiman2001random}.
-These metric can provide some interpretability to model predictions.
+This metric can provide some interpretability to model predictions.
 
-The Supplementary Materials provide a detailed description of the implementation and validation of the random forest models.
+The Supplementary Materials provide a detailed description of the data processing for the covariates, as well as implementation and validation of the random forest models.
+
+\section
 
 \printbibliography
+
+\end{document}

### 2020-11-18 docs/modeling_fies_supplement.tex
new draft with country-level leave-one-out cross validation

@@
 \section{Overview of the FIES}
-We use raw microdata released by the FAO from 75 countries from different world regions and income categories (See Fig. \ref{fig:fies_countries}.
+We use raw microdata released by the FAO from 75 countries from different world regions and income categories (See Fig. \ref{fig:fies_countries}).
 
@@
 
-Based on the global reference scale, two thresholds are set based on questions 1 and 8: whether the respondent is worried they would not have enough food to eat because of a lack of resources, and whether there was a time when they went without eating for a whole day because of a lack of money or other resources.
+Based on the global reference scale, two thresholds are set based on questions 1 and 8, corresponding to moderate and severe food insecurity, respectively.
 
 \begin{enumerate}
-	\item During the last 12 MONTHS, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
-	\item Still thinking about the last 12 MONTHS, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
+	\item During the last 12 months, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
+	\item Still thinking about the last 12 months, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
 	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
 	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
-	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
+	\item Still thinking about the last 12 months, was there a time when you ate less than you thought you should because of a lack of money or other resources?
 	\item Was there a time when your household ran out of food because of a lack of money or other resources?
 	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
-	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
+	\item During the last 12 months, was there a time when you went without eating for a whole day because of a lack of money or other resources?
   \label{itm:fies}
@@
 
-Each ROC is weighted so as to give proportionally more weight to recent years.  For a dataset with observations for 2010 to 2017, that would be:
+Each ROC is weighted so as to give proportionally more weight to recent years.  For a dataset with observations for 2010 to 2019, that would be:
 
 \begin{equation}
-  W_y = \frac{y - 2010}{\sum_{y=2010}^{2017} y - 2010}
+  W_y = \frac{y - 2010}{\sum_{y=2010}^{2019} y - 2010}
   \label{eqn:b}
@@
 \begin{equation}
-  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2017} W_y \times \text{AROC}_{a, y} \right)
+  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2019} W_y \times \text{AROC}_{a, y} \right)
   \label{eqn:c}
@@
 \begin{equation}
-  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2017} ) + AROC_{a} \times ( y - 2017 ) )
+  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2019} ) + AROC_{a} \times ( y - 2019 ) )
   \label{eqn:d}
@@
 \section{Description, Implementation and Validation of the Random Forest Regression Model}
-The RF regression is a machine learning method based on the creation of a large number of decision trees.
-By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.
+The random forest regression is a machine learning method based on the creation of a large number of decision trees.
 For each of the decision trees in a random forest model, observations are selected at random with replacement, a method known as bootstrap aggregating, or bagging, and features are also selected at random.
+For the implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}.
 
-For the implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}.
-Specifically, we apply the functions \texttt{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \texttt{rfsrc()} to perform the RF regression models.
-After predicting rates of moderate-to-severe and severe food insecurity, we plot the modeled vs.
-observed rates and calculated the MSE and ${R}^2$ (See Fig.
-\ref{fig:rf_in-sample} and Fig.
-\ref{fig:rf_out-sample}) for both models.
-Additionally, we use the function \texttt{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig.
-\ref{fig:rf_error}) and the importance of individual variables (See Fig.
-\ref{fig:rf_vimp}).
-
-%STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
 Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.
@@
 
-To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions. 
-%MAYBE STATE THIS SOMEWHERE ELSE, NOT SURE WHERE
+To determine the best hyperparameters to model both moderate ands severe food insecurity, we use leave-one-out cross validation at the country level.
+This involves running a separate model for each country, where we train a model on the remaining data from all of the other countries, and use that model to predict values for the country that was left out.
+Then, across the predicted values for all countries, we evaluate the mean average error (MAE) and $R^2$ of the differences between observed rates of food insecurity and predictions from the cross validation models.
+
+We evaluate models across a grid of hyperparameters, with combinations of: between 1 to 13 variables randomly selected as candidates for splitting a node (\texttt{mtry}), between 1 and 20 terminal node sizes (\texttt{nodesize}), and with a maximum tree depth (\texttt{depth}) between 1 to 13 nodes, as well as without any maximum tree depth.
+For each model in cross validation, we used 1000 trees (\texttt{ntree}).
+Due to the large size of the hyperparameter space, we sampled sets of hyperparameters randomly, eventually sampling about 50\% of all combinations of hyperparamters.
+We found that the optimal model for moderate food insecurity had 16 average terminal nodes, 2 variables selected as candidates for splitting a node, and no maximum tree depth; the optimal model for severe food insecurity also had 2 variables selected as candidates for splitting a node and no maximum tree depth, but with only 14 average terminal.
+
+After fitting models with cross validation, we evaluated the models performance and graphed a comparison between the observed and predicted values based on instances where individual countries were left out (See Fig.
+\ref{fig:rf_out-sample}), as well as for a model trained on the entire dataset, which can only be evaluated on in-sample data (See Fig.
+\ref{fig:rf_in-sample}).
+Finally, Figure \ref{fig:mod-error} shows the model error converging and the number of trees increases.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{img/model/error_rf.png}
-  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_error}
+  \includegraphics[width=\linewidth]{img/out-sample_rf.pdf}
+  \caption{Out-of-sample fit of the random forest regression model, where predictions for each country are based on a model that did not include that country in the training data.
+Panel (A) shows the model for moderate food insecurity and panel (B) shows the model for severe food insecurity.}
+  \label{fig:rf_out-sample}
 \end{figure}
-
-%HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH
-
-In addition to the parameter that describes the number of trees to be created, two other important hyperparameters must be tuned.
-In the function \texttt{tune.rfsrc()} these parameters are called \texttt{mtry} and \texttt{nodesize}.
-\texttt{mtry} describes the number of variables randomly selected as candidates for splitting a node and \texttt{nodesize} describes the average number of observations in a leaf node.
-The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest out-of-bag (OOB) error.
-
-%EXPLAIN THE OOB ERROR HERE?
-
-The function \texttt{tune.rfsrc()} does exactly that and returns optimal values for \texttt{mtry} and \texttt{nodesize}.
-We applied this function on both, the moderate-to-severe and the severe model.
-For the moderate-to-severe model the optimal combination is \texttt{mtry} = 12 and \texttt{nodesize} = 1 and for the severe model we found \texttt{mtry} = 10 and \texttt{nodesize} = 2 to be optimal.
-
-%STATE FUNCTION ARGUMENTS HERE?
 
@@
   \centering
-  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
-  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
+  \includegraphics[width=\linewidth]{img/in-sample_rf.pdf}
+  \caption{In-sample fit of the random forest regression model, where one model was trained on the entire dataset.
+Panel (A) shows the model for moderate food insecurity and panel (B) shows the model for severe food insecurity.}
   \label{fig:rf_in-sample}
@@
   \centering
-  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
-  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_out-sample}
+  \includegraphics[width=\linewidth]{img/Error_Over_Training.pdf}
+  \caption{Error rate of the random forest regression model as the number of trees in the model increases.
+Panel (A) shows the model for moderate food insecurity, and panel (B) shows the model for severe food insecurity.}
+  \label{fig:mod-error}
 \end{figure}
 
-%HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
-%For modeling the rates of moderate-to-severe and severe food insecurity we use the function \texttt{rfsrc()}. We specify
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{img/VIMP.pdf}
-  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
-  \label{fig:rf_vimp}
-\end{figure}
-
-%https://kogalur.github.io/randomForestSRC/theory.html#section4.2
-Lastly we asses the importance of individual variables.
-This is based on the Breiman-Culter permutation variable importance information \citep{breiman2001random}.
-This involved comparing the prediction error on the OOB data to the prediction error of OOB cases where the x-variable $x$ is randomly permutated.
-The importance of an individual variable is thus the mean difference between the perturbed and unperturbed error rate, across all trees.
-%MORE INFORMATION HERE
 \end{document}

### 2020-11-18 viz/Error_and_VIP.R

@@ -1,4 +1,4 @@
-# setwd('~/wdl-fies');library(ProjectTemplate);load.project()
+setwd('~/wdl-fies');library(ProjectTemplate);load.project()
 setwd('~/wdl-fies/docs/img')
 #setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
 
@@ -83,8 +83,8 @@ mod.res <- ggplot(moddat) +
   xlim(0, 1) + ylim(0, 1) +
   labs(caption=paste0('Mean Average Error: ',  round(mae.mod, 4),
                       '\nR-Squared: ', round(r2.mod, 4)),
-       x='Observed Rates Of Moderat-to-Severe Food Insecurity',
-       y='Modeled Rates Of Moderat-to-Severe Food Insecurity'); mod.res
+       x='Observed Moderate Food Insecurity',
+       y='Modeled Moderate Food Insecurity'); mod.res
 
 sev.res <- ggplot(moddat) + 
   geom_point(aes(x=fies.sev, y=fies.sev.pred), alpha = 0.3) +
@@ -92,11 +92,11 @@ sev.res <- ggplot(moddat) +
   xlim(0, 1) + ylim(0, 1) +
   labs(caption=paste0('Mean Average Error: ',  round(mae.sev, 4),
                       '\nR-Squared: ', round(r2.sev, 4)),
-       x='Observed Rates Of Severe Food Insecurity',
-       y='Modeled Rates Of Severe Food Insecurity')	; sev.res
+       x='Observed Severe Food Insecurity',
+       y='Modeled Severe Food Insecurity')	; sev.res
 
 grd <- plot_grid(mod.res, sev.res, align='h',labels='AUTO')
-ggsave(grd, filename='docs/img/in-sample_rf.png', width=14, height=7)
+ggsave(grd, filename='in-sample_rf.pdf', width=7, height=3.75)
 
 #########################################
 # Out of Sample Residuals
@@ -113,8 +113,8 @@ mod.res <- ggplot(moddat) +
   xlim(0, 1) + ylim(0, 1) +
   labs(caption=paste0('Mean Average Error: ',  round(mae.mod, 4),
                       '\nR-Squared: ', round(r2.mod, 4)),
-       x='Observed Rates Of Moderat-to-Severe Food Insecurity',
-       y='Modeled Rates Of Moderat-to-Severe Food Insecurity'); mod.res
[…]

### 2020-11-18 viz/Hunger_Clock_Data.R

@@ -29,5 +29,5 @@ sel <- preddat %>%
 gdl <- gdl %>%
   filter(GDLCODE %in% sel$GDLCODE)
 
-write.csv(sel, 'figures/HC_2020-10-28_data.csv', row.names=F) #update date
+write.csv(sel, 'figures/HC_2020-11-18_data.csv', row.names=F) #update date
 write_sf(gdl, 'figures/HC_2020-10-28_polys.geojson', driver='GeoJSON')

### 2020-11-18 docs/modeling_fies.tex
Fix errors in overleaf and change title

@@
 
-\title{A Global Picture of Hunger}
+\title{Monitoring and Projecting Global Hunger}
 
@@
 \ref{fig:timeseries}).
-These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \ref{sofi2020}.
+These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.
 
@@
 
-For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index, followed by the rate of stunting.
+For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig.
+\ref{fig:vimp}, followed by the rate of stunting.
 For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.
@@
 This is determined by the increase in the error when the variable is permuted before fitting the model.}
-  \label{fig:map}
+  \label{fig:vimp}
 \end{figure}
@@
 
-\section
-
 \printbibliography

### 2020-11-18 docs/modeling_fies_supplement.tex
Fix errors in overleaf and change title

@@
 \fancyhf{}
-\lfoot{Supplemental Materials for \textit{A Global Picture of Hunger}}
+\lfoot{Supplemental Materials for \textit{Monitoring and Projecting Global Hunger}}
 \rfoot{\thepage}
@@
 \begin{center}
-\section*{Supplemental Info for \textit{A Global Picture of Hunger}}
+\section*{Supplemental Materials for \textit{Monitoring and Projecting Global Hunger}}
 \end{center}
@@
 We obtain the variable using a gridded dataset of elevation from the USGS \citep{USGS1996} and calculate the index at each grid cell using the methodology from Riley et al.
-\citep{Riely1999}.
+\citep{Riley1999}.
 We then aggregate the values for all the grid cells within each administrative area.

### 2020-12-10 docs/modeling_fies_supplement.tex
Changes from Overleaf

@@
 
-We draw our historical and future estimates of urbanization entirely from the spatially explicit scenarios of urban population by Jones and O'Neill \citep{Jones2016}, which provide a series of urbanization rate projections which are consistent with the Shared Socioeconomic Pathways.
+We draw our historical and future estimates of urbanization entirely from the spatially explicit scenarios of urban population by Jones and O'Neill \citep{Jones2016}, which provide a series of urbanization rate projections which are consistent with the Shared Socioeconomic Pathways (SSPs).
 As with our other covariates, we use projections consistent with SSP2, the middle-of-the-road scenario.
@@
 
-To account for the effects of the coronavirus pandemic on global rates of wasting, we assume that long-term trends in rates of wasting would hold steady, after they increase globally by 14.3\% in 2020, based on estimates published in \textit{The Lancet} \citep{headey2020impacts}.
+To account for the effects of the COVID-19 pandemic on global rates of wasting, we assume that long-term trends in rates of wasting would hold steady, after they increase globally by 14.3\% in 2020, based on estimates published in \textit{The Lancet} \citep{headey2020impacts}.
 We model this impact as uniform across all countries where wasting occurs.
@@
 We also assume a 14.3\% increase in prevalence for the year 2020 \citep{headey2020impacts}, as is done for wasting, although stunting is a long-term consequence of malnutrition and is not likely to be observable in a population with the same lags as wasting would be.
-We use stunting in our modeling framework as a proxy for chronic conditions of hunger, which are probably exacerbated by the coronavirus pandemic.
+We use stunting in our modeling framework as a proxy for chronic conditions of hunger, which are probably exacerbated by the COVID-19 pandemic.
 
@@
 
-Finally, we used estimates of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the coronavirus \citep{prospects2020}.
-We then model GDP as resuming its previous growth trajectory based on previously-estimated growth rates from 2022-2030, but growing from an estimate for 2021 that includes the coronavirus shock.
+Finally, we used estimates of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the COVID-19 pandemic \citep{prospects2020}.
+We then model GDP as resuming its previous growth trajectory based on previously-estimated growth rates from 2022-2030, but growing from an estimate for 2021 that includes the pandemic shock.
 
@@
 National GDP totals are consistent with the projections of Dellink et al., but are disaggregated according to the historic patterns of subnational GDP shares sourced from the Subnational Human Development Database.
-We use nowcasts and forecasts of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the coronavirus \citep{prospects2020}.
-After the shock created by the COVID pandemic, GDP is assumed to resuming its previous growth trajectory for the period 2022-2030.
+We use nowcasts and forecasts of country-level changes in GDP growth for the years 2020 and 2021 to account for the impact of the COVID-19 pandemic \citep{prospects2020}.
+After the shock created by the pandemic, GDP is assumed to resuming its previous growth trajectory for the period 2022-2030.
 
@@
 We used historical and projected data on the Poverty Headcount Index (using the standard threshold of \$1.90 a day) from the World Poverty Clock by the World Data Lab, which in turn uses Povcal data sourced from the World Bank \citep{Cuaresma2018}.
-These data are the version corresponding to the update in summer 2020 and include the expected impact of the COVID pandemic on global poverty dynamics.
+These data are the version corresponding to the update in summer 2020 and include the expected impact of the COVID-19 pandemic on global poverty dynamics.
 Using parameterized Lorenz curves in combination with mean income/consumption and population projections, poverty rates are projected.
@@
 \subsection{Water Scarcity}
-We use data for the water scarcity index (WSI) developed by \citep{greve2018global}.
+We use data for the Water Scarcity Index (WSI) developed by \citep{greve2018global}.
 The index represents the ratio of the decadal averages in water demand to water supply, based on population projections consistent with SSP2 and water availability projections from global hydrological and general circulation models.
 Based on this index, we estimate the share of people living in areas with less than 1000m\textsuperscript{3} of water available per capita per year.
-The WSI has been used already to build the water scarcity clock.
+The WSI has been used already to build the Water Scarcity Clock.
 
@@
 
-To determine the best hyperparameters to model both moderate ands severe food insecurity, we use leave-one-out cross validation at the country level.
+To determine the best hyperparameters to model both moderate and severe food insecurity, we use leave-one-out cross validation at the country level.
 This involves running a separate model for each country, where we train a model on the remaining data from all of the other countries, and use that model to predict values for the country that was left out.
@@
 
+\printbibliography
+
 \end{document}
-
-\printbibliography

### 2020-12-10 docs/modeling_fies.tex

@@
 \author[2,5]{Juan Carlos Laso Bayas}
-\author[5,6]{Jesús Crespo Cuaresma}
-\author[2,7]{Homi Kharas}
+\author[3,5,6,7]{Jesús Crespo Cuaresma}
+\author[2,8]{Homi Kharas}
 
@@
 \affil[3]{Department of Economics, Vienna University of Economics and Business}
-\affil[4]{The Food and Agricultural Organization of the United Nations}
-\affil[5]{The International Institute for Applied Systems Analysis}
-\affil[6]{Wittgenstein Centre For Demography and Global Human Capital}
-\affil[7]{The Brookings Institute}
+\affil[4]{Statistics Division, Food and Agricultural Organization of the United Nations}
+\affil[5]{International Institute for Applied Systems Analysis}
+\affil[6]{Wittgenstein Centre for Demography and Global Human Capital}
+\affil[7]{Austrian Institute of Economic Research}
+\affil[8]{The Brookings Institute}
 \affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}
@@
 \begin{abstract}
-Using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the \href{https://worldhunger.io}{World Hunger Clock}.
-This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.
-We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the parts of developing world where over half the population is severely food insecure.
-Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been steadily declining, with large observed and forecasted gains in lower and lower-middle income countries, and decreases in total number of severely food insecure people.
+Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock (https://worldhunger.io).
+This is the first global picture of food security at a subnational level, based on the Food Insecurity Experience Scale, a food security metric developed by FAO that is most indicative of the individual or household lived experience of food insecurity and hunger.
+We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population being at least moderately food insecure in some high-income countries to parts of the developing world where over half the population is severely food insecure.
+Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been declining steadily yet slowly, with larger observed and forecasted gains in lower and lower-middle income countries, and global decreases in the total number of severely food insecure people.
 However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.
@@
 
-Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the UN \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.
+Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
@@
 Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalence of such food insecurity events can be computed.
-A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of income, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
+A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
 A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.
@@
 \ref{fig:timeseries}).
-These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.
+These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity (SOFI) report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.
 
@@
 Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig.
-\ref{fig:rates}, we find clear improvements in all except high income countries.
+\ref{fig:rates}), we find clear improvements in all except high income countries.
 Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.
@@
 For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig.
-\ref{fig:vimp}, followed by the rate of stunting.
+\ref{fig:vimp}), followed by the rate of stunting and the Gini coefficient.
 For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.
@@
 
-The results of this analysis can be further explored in more detail on the \href{https://worldhunger.io}{World Hunger Clock}, including statistics for each subnational administrative area.
+The results of this analysis can be further explored in more detail on the World Hunger Clock \href{https://worldhunger.io}{(worldhunger.io)}, including statistics for each subnational administrative area.
 
@@
 This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.
-We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the coronavirus pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.
+We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.
 Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of food insecure people.
@@
 When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.
-We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.
+We test the accuracy of these cases by using leave-one-out cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.
 
@@
 These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.
-We use the combination of hyperparameters that perform best under ten-fold cross validation based on $R^2$.
+We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$.
 
@@
 After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
-We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable, breiman2001random}.
+We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.
 This metric can provide some interpretability to model predictions.

### 2020-12-10 docs/modeling_fies.tex
update figures; incorporate Homi changes

@@
 \affil[7]{Austrian Institute of Economic Research}
-\affil[8]{The Brookings Institute}
+\affil[8]{The Brookings Institution}
 \affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}
@@
 \section{Main}
-Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of at least moderate food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
+Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate and severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
 The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger.
@@
 However, for both indicators, these changes are very slight: in 2030, the number of moderately food insecure people will have increased by 2.2\% since 2010, and the number of severely food insecure people will have decreased by 7.0\%.
-The increases in moderate food insecurity have been in lower and lower-middle income countries, while the increases in severe food insecurity are entirely from lower income countries.
+The increases in moderate food insecurity have been in lower and lower-middle income countries, while the decreases in severe food insecurity are entirely from lower income countries.
 For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.
@@
 \ref{fig:rates}), we find clear improvements in all except high income countries.
-Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.
-Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.
+Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2020s.
+Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in the 2020s.
 In spite of this progress in the bottom three income quartiles, in high-income countries, the rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.
@@
 For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.
-For the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.
-Across both indicators, economic variables were the most relevant predictors.
+The country's Gini coefficient is also important, more so for predicting the rate of severe food insecurity compared to moderate food insecurity.
+Across both outcomes, economic variables were the most relevant predictors.
 
@@
 This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.
-We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.
-Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of food insecure people.
+We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of economic growth in developing countries.
+Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of moderately food insecure people.
 
@@
 While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate food insecurity to increase and the number of people experiencing severe food insecurity to fall only slightly between 2010 and 2030.
-This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, a falling far short of the SDG 2 goal of ending hunger.
+This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, falling far short of the SDG 2 goal of ending hunger.
 Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, sustainable diets, or obesity.
@@
 
-For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions up to variables for 2030.
+For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions for variables up to 2030.
 We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.
@@
 After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
-We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.
+We use a common method of permutating each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.
 This metric can provide some interpretability to model predictions.

### 2020-12-10 viz/Viz_Mod_Results.R

@@ -73,7 +73,9 @@ reg <- ggplot(regdat) +
 #also change regions in preddat
 preddat <- merge(preddat, st_drop_geometry(regdat)) %>%
   select(-region) %>%
-  rename(region = "region_wb")
+  rename(region = "region_wb") %>%
+  mutate(region = factor(region, levels=rev(c("High Income", "Upper-Middle Income",
+                                          "Lower-Middle Income", "Low Income"))))
 
 ###############################################
 # Rates over time, not just raw number

### 2020-12-10 viz/Viz_Mod_Results.R
tweak figure legends

@@ -146,18 +146,19 @@ regnoleg <- ggplot(regdat) +
   guides(fill=FALSE) + 
   labs(fill='')
 
-plot_grid(ratelines,
+plot_grid(ratelines, 
           plot_grid(regnoleg,
-                    plot_grid(pltleg, 
-                              regleg, 
+                    plot_grid(regleg,
+                              pltleg, 
+                              #regleg, 
                               ncol=1, 
                               rel_heights=c(1, 4)),
                     ncol=2,
-                    rel_widths=c(1, 1.5)),
+                    rel_widths=c(1, 1.25)),
           ncol=1,
-          rel_heights=c(2.5, 1))
+          rel_heights=c(2.75, 1))
 
-ggsave('Rates.pdf', width=7, height=5)
+ggsave('Rates.pdf', width=5, height=3.25)
 #####################
 # totals time series
 ###############
@@ -206,4 +207,4 @@ lines <- ggplot(totals) +
 
 plot_grid(plot_grid(stack, lines, align='v', nrow=2, labels='AUTO'), 
           reg, ncol=1, rel_heights=c(5, 1))
-ggsave('TimeSeries.pdf', width=7, height=7)
+ggsave('TimeSeries.pdf', width=5, height=5)

### 2020-12-22 docs/modeling_fies.tex
update title
Changes from Overleaf

@@
 
-\title{Monitoring and Projecting Global Hunger}
+\title{Monitoring and Projecting Global Hunger: Are We On Track?}
 
@@
 Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock (https://worldhunger.io).
-This is the first global picture of food security at a subnational level, based on the Food Insecurity Experience Scale, a food security metric developed by FAO that is most indicative of the individual or household lived experience of food insecurity and hunger.
+This is the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of "Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
 We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population being at least moderately food insecure in some high-income countries to parts of the developing world where over half the population is severely food insecure.
-Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been declining steadily yet slowly, with larger observed and forecasted gains in lower and lower-middle income countries, and global decreases in the total number of severely food insecure people.
-However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.
-Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.
+Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining steadily yet slowly, with larger observed and forecasted gains in lower and lower-middle income countries, resulting in global decreases in the total number of severely food insecure people.
+However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, we predict it will continue to increase through the end of the 2020s.
+Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.
 \end{abstract}
@@
 Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate and severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES.
-The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger.
-Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population.
+The FIES was developed by FAO in 2013 as a global extension of pioneering work started in the US in the 1990's to develop a metric of food insecurity based on data gathered by interviewing the people who experience hunger and conditions that produce hunger.
+Other indicators of food insecurity, such as FAO's more commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population, developed on the basis of infrequent and heterogeneous data on food production and consumption.
 These do not directly capture the actual lived exposure of individuals to food shortages.
 
-The FIES is based on responses to polls and the data are limited to 75 countries where these have been conducted and microdata has been vetted and released by the FAO.
-The current practice is to assign the average regional prevalence rate for countries without survey data in order to compile regional and global aggregates.
-In addition, since the FIES is a relatively new metric, it does not have a sufficiently long time series to assess time trends with a reasonable degree of precision.
-To understand whether SDG 2 is being met, estimates of food insecurity outcomes in 2030 are required.
+Currently, the data used to generate the estimates that are reported at the UN SDG indicators database come primarily from the 75 countries where surveys have been commissioned, vetted, and released by FAO.
+To compile global and regional aggregates, the current practice at FAO is to use the prevalence rate for the countries that do have FIES data as the regional rate.
+Moreover, since the FIES is a relatively new metric, a sufficiently long time series to assess time trends with a reasonable degree of precision is still lacking.
+Thus, to fill these spatial and temporal data gaps and understand whether SDG 2 is being met, modeled estimates of food insecurity outcomes are thus required.
 
-This paper contributes to shifting the research frontier by filling these two gaps in terms of data coverage and forecasts by developing a statistical model of FIES outcomes based on covariates that are available for territories where actual polling data are not available.
-Such a model can be employed to obtain forecasts to 2030 and assess likely future trends in food insecurity.
-We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond the country level to a sub-national level.
-To our knowledge, this is the first effort to model the FIES outcomes at that scale.
+This paper contributes to shifting the research frontier by filling these two gaps in terms of spatial coverage and forecasts by developing a machine learning model of FIES-based food insecurity rates based on covariates that are available for territories where actual survey data are not available.
+This model is then employed to obtain forecasts to 2030 and assess likely future trends in food insecurity.
+We take advantage of individual-level characteristics collected in FAO's FIES surveys to push the analysis beyond the country level to a sub-national level.
+To our knowledge, this is the first effort to model FIES-based food insecurity at that scale.
 
-Food security has traditionally been difficult to measure, and this has led to an incomplete or inaccurate assessment of global hunger.
+Food security has traditionally been difficult to measure, and this has arguably led to an incomplete or possibly inaccurate assessment of global hunger.
 Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.
@@
 
-As researchers began to focus on food insecurity at the individual and household level, household microdata collecting information on household finances and consumption became a common proxy for food security \citep{Haddad1994}.
+As researchers began to focus on food insecurity at the individual and household level, microdata from household surveys collecting information on household finances and consumption became a basis to compute a common proxy for food security \citep{Haddad1994}.
 However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \citep{Maxwell1996}.
@@
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
-Using a Rasch model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing the responses given to the 8 questions asked in the survey.
-Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalence of such food insecurity events can be computed.
-A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
-A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.
-Because the model estimates the population over a threshold on a univariate food insecurity scale, the percentage of people over the threshold for moderate food insecurity includes people that are also over the threshold for severe food insecurity.
+Processing the responses given to the FIES questions through the Rasch model, each individual in the survey is assigned a score that allows them to be ranked on a scale of food insecurity severity.
+Once the severity measures are calibrated against a common reference (the global FIES reference scale) \citep{Cafiero2018} and individuals are ranked in this fashion, conventional thresholds for ``moderate'' and ``severe'' food insecurity are set and the prevalence of such food insecurity conditions can be computed.
+Thresholds are set in such a way that a moderately food insecure respondent is one who likely has compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
+A severely food insecure one is likely to have run out of food and has gone an entire day without eating at times during the year.
+The percentage of people estimated to be over the threshold set for moderate food insecurity includes people that are also over the threshold for severe food insecurity and represents the SDG indicator 2.1.2.
 
-Since 2014, in partnership with Gallup World Poll, the FAO administered the FIES in surveys around the world.
-The FAO uses this data to estimate the prevalence of food insecurity, and reports national-level estimates of the percentage of the population over the thresholds for moderate and severe food insecurity.
-Drawing on individual-level data from these microdatasets in 75 countries, we use machine learning methods to estimate levels of food insecurity globally at the subnational level.
-The model is then used to forecast food insecurity to the year 2030 under a middle-of-the-road scenario.
+Since 2014, FAO contracted the Gallup World Poll to administer the FIES survey module in as many countries as possible around the world and uses this data to estimate the prevalence of food insecurity at region and global levels.
+For a number of those countries for which it has been authorized to release the national-level estimates, FAO also reports the percentage of the national population that is estimated to be over the thresholds for moderate and severe food insecurity and publicly releases the microdata.
+
+Drawing on individual-level data from these publicly available microdatasets from 75 countries, we use machine learning methods to model food insecurity based on covariates and estimate levels of food insecurity globally, at the subnational level.
+The model is then used to forecast food insecurity to the year 2030 under a reasonable scenario on the possible evolution of the covariates.
 
@@
 While economic conditions, and especially GDP per capita, are important predictors of food insecurity at a global scale, there is an upper limit to the extent to which increasing economic development alone can improve food security outcomes.
-High income countries have seen little improvement in food insecurity outcomes over twenty years, and upper-middle income countries are expected to see diminishing gains throughout the 2020s.
+High income countries have seen little improvement in food insecurity outcomes over twenty years, and upper-middle income countries are expected to see diminishing gains as they continue to develop economically throughout the 2020s.
 Among the high-income countries in our dataset, those with the lowest rates of moderate and severe food insecurity also had lower inequality and poverty.
@@
 
-When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.
+When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and food insecurity are the same across all locations.
 We test the accuracy of these cases by using leave-one-out cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.
@@
 \subsection{Disaggregation}
-The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
+The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
 Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
@@
 
+\section{Acknowledgements}
+This reserach was supported by The International Fund for Agricultural Development (IFAD) grant number 2000002044.  We thank Kristofer Hamel and Wolfgang Fengler for their support in this work.
+
+\section{Author Contributions}
+MC, BM, and HK conceptualized the analysis; CC, JCC, and HK provided methodological oversight; MC, BM, and JCL processed the data; MC and BM ran the models and conducted model validation; All authors contributed to writing and reviewing the manuscript.
+
+\section{Competing Interests}
+We have no competing interests to disclose.
+
 \printbibliography

### 2020-12-22 docs/modeling_fies.tex
final draft for submission

@@
 
+\linenumbers
+
 \addbibresource{library.bib}
@@
 \begin{abstract}
-Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock (https://worldhunger.io).
-This is the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of "Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
+Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock \href{https://worldhunger.io}{(worldhunger.io)}.
+This is the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
 We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population being at least moderately food insecure in some high-income countries to parts of the developing world where over half the population is severely food insecure.

### 2021-01-26 docs/modeling_fies.tex
Updates from Overleaf

--- 2020-12-22 docs/modeling_fies.tex
@@
 \begin{abstract}
-Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock (https://worldhunger.io).
-This is the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of "Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
-We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population being at least moderately food insecure in some high-income countries to parts of the developing world where over half the population is severely food insecure.
-Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining steadily yet slowly, with larger observed and forecasted gains in lower and lower-middle income countries, resulting in global decreases in the total number of severely food insecure people.
+This paper presents the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
+Using microdata from 75 countries and filling gaps using machine learning, we find significant heterogeneity in levels of food insecurity around the world.
+Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining, resulting in global decreases in the total number of severely food insecure people.
 However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, we predict it will continue to increase through the end of the 2020s.
-Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.
+Overall, we conclude that current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.
 \end{abstract}
+
+\textit{Keywords: Food Security, Sustainable Development Goal 2, Machine Learning, Food Insecurity Experience Scale, Projections}
 
@@
 The model is then used to forecast food insecurity to the year 2030 under a reasonable scenario on the possible evolution of the covariates.
+
+\section{Methods}
+\subsection{Disaggregation}
+The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
+Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
+We use a similar methodology and calculate post-stratification weights at a subnational level.
+We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.
+For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
+This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.
+
+\subsection{Covariates}
+We model subnational rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).
+We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.
+For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.
+
+Many of the variables require the harmonization of subnational historical data with projections at the national level.
+These include mean years of schooling, GDP per capita, as well as population.
+We use observed trajectories in the historical distribution of variables among subnational areas within a country to predict the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.
+
+For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions for variables up to 2030.
+We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.
+This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.
+
+For the climatic variables, temperature and precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
+
+Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.
+These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.
+
+\begin{table}[H]
+  \centering
+	\begin{tabular}{lll}
+		\toprule
+    Name & Source & Scale \\
+		\midrule
+    Urban Percentage & \citep{Jones2016} & Subnational \\
+    Stunting & \citep{Local2020} & Subnational \\
+    Wasting & \citep{Local2020} & Subnational \\
+    Mean Years of Schooling & \citep{Smits2019, KC2017} & Subnational\\
+    GDP Per Capita & \citep{Smits2019, Dellink2017} & Subnational \\
+    Gini Coefficient & \citep{Rao2019a} & National\\
+    Poverty Headcount Index & \citep{Cuaresma2018} & National \\
+    Water Scarcity & \citep{greve2018global} & Subnational \\
+    Mean Annual Precipitation &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
+    Topographic Ruggedness &  \cite{USGS1996, Riley1999} & Subnational \\
+    Mean Temperature &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
+    Malaria (\textit{P. falciparum}) Mortality Rate &  \cite{Weiss2019} & Subnational \\
+		\bottomrule
+	\end{tabular}
+	\caption{Covariates Included in the Analysis}
+	\label{tab:covars}
+\end{table}
+
+\subsection{Modeling}
+We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
+We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.
+A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.
+By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.
+Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.
+We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.
+
+To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.
+These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.
+We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$.
+
+When performing cross validation, it is typical to select observations at random.
+However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.
+Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.
+Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.
+By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.
+After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.
+
+After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
+We use a common method of permutating each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.
+This metric can provide some interpretability to model predictions.
+
+The Supplementary Materials provide a detailed description of the data processing for the covariates, as well as implementation and validation of the random forest models.
 
@@
 
-\section{Methods}
-\subsection{Disaggregation}
-The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
-Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
-We use a similar methodology and calculate post-stratification weights at a subnational level.
-We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.
-For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
-This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.
-
-\subsection{Covariates}
-We model subnational rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).
-We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.
-For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.
-
-Many of the variables require the harmonization of subnational historical data with projections at the national level.
-These include mean years of schooling, GDP per capita, as well as population.
-We use observed trajectories in the historical distribution of variables among subnational areas within a country to predict the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.
-
-For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions for variables up to 2030.
-We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.
-This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.
-
-For the climatic variables, temperature and precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
-
-Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.
-These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.
-
-\begin{table}[H]
-  \centering
-	\begin{tabular}{lll}
-		\toprule
-    Name & Source & Scale \\
-		\midrule
-    Urban Percentage & \citep{Jones2016} & Subnational \\
-    Stunting & \citep{Local2020} & Subnational \\
-    Wasting & \citep{Local2020} & Subnational \\
-    Mean Years of Schooling & \citep{Smits2019, KC2017} & Subnational\\
-    GDP Per Capita & \citep{Smits2019, Dellink2017} & Subnational \\
-    Gini Coefficient & \citep{Rao2019a} & National\\
-    Poverty Headcount Index & \citep{Cuaresma2018} & National \\
-    Water Scarcity & \citep{greve2018global} & Subnational \\
-    Mean Annual Precipitation &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
-    Topographic Ruggedness &  \cite{USGS1996, Riley1999} & Subnational \\
-    Mean Temperature &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
-    Malaria (\textit{P. falciparum}) Mortality Rate &  \cite{Weiss2019} & Subnational \\
-		\bottomrule
-	\end{tabular}
-	\caption{Covariates Included in the Analysis}
-	\label{tab:covars}
-\end{table}
-
-\subsection{Modeling}
-We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
-We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.
-A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.
-By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.
-Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.
-We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.
-
-To get the best possible model performance, it is necessary to tune the hyperparamters, which control the process in which the random forest algorithm is run.
-These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.
-We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$.
-
-When performing cross validation, it is typical to select observations at random.
-However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.
-Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.
-Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.
-By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.
-After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.
-
-After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
-We use a common method of permutating each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.
-This metric can provide some interpretability to model predictions.
-
-The Supplementary Materials provide a detailed description of the data processing for the covariates, as well as implementation and validation of the random forest models.
-
 \section{Acknowledgements}

### 2021-03-04 data/covars/scripts/urban-rural.R
new submission for GFS

@@ -180,7 +180,7 @@ rates <- ref_past %>%
   mutate(admin_shares_total = sum(admin_share)) %>%
   ungroup %>%
   mutate(admin_share = admin_share/admin_shares_total) %>%
-  dplyr::select(ISO3, GDLCODE, YEAR, admin_share)
+  dplyr::select(ISO3, GDLCODE, YEAR, admin_share, admin_shares_total)
 
 #Now extrapolate these rates to the future
 fut <-merge(rates, new_ref_fut) %>%

### 2021-03-04 docs/modeling_fies.tex
revert to results-first order
Updates from Overleaf

@@
-\documentclass{article}
+\documentclass[titlepage]{article}
 
@@
 \end{abstract}
-
-\textit{Keywords: Food Security, Sustainable Development Goal 2, Machine Learning, Food Insecurity Experience Scale, Projections}
 
@@
 
-\section{Competing Interests}
-We have no competing interests to disclose.
+\section{Additional Information}
+All code necessary to reproduce these results is on Github at \href{https://github.com/mcooper/wdl-fies}{github.com/mcooper/wdl-fies}. We have no competing interests to disclose.
 

### 2021-03-04 docs/modeling_fies_supplement.tex
Updates from Overleaf

@@
 \fancyhf{}
-\lfoot{Supplemental Materials for \textit{Monitoring and Projecting Global Hunger}}
+\lfoot{Supplemental Materials for \textit{Monitoring and Projecting Global Hunger: Are We On Track?}}
 \rfoot{\thepage}
@@
 \begin{center}
-\section*{Supplemental Materials for \textit{Monitoring and Projecting Global Hunger}}
+\section*{Supplemental Materials for \textit{Monitoring and Projecting Global Hunger: Are We On Track?}}
 \end{center}

### 2021-03-04 docs/modeling_fies.tex

@@
 
+\usepackage{float}
+\usepackage[
+  colorlinks=true,
+  citecolor=blue
+]{hyperref}
+\usepackage{pdflscape}
+\usepackage{rotating}
+\usepackage{authblk}
+\usepackage{booktabs}
 \usepackage{graphicx}
-\usepackage{hyperref}
-\usepackage[a4paper, margin=1.25in]{geometry}
-\usepackage{breakcites}
-\usepackage{subcaption}
-\usepackage{float}
-\usepackage{textcomp}
-\usepackage{amsmath}
-\usepackage{textgreek}
-\usepackage{authblk}
-\usepackage{rotating}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{pdflscape}
-\usepackage{lineno}
+\usepackage{makecell}
 \usepackage[
-  style=numeric,
-  citestyle=numeric-comp,
+  style=authoryear,
+  citestyle=authoryear,
   backend=biber,
@@
   natbib=true,
-  sorting=none
+  sorting=nyt %name, year, title
 ]{biblatex}
@@
 \addbibresource{library.bib}
+
+%set up alignment in \makecell
+%\renewcommand{\cellalign/theadalign}{cl}
 
@@
 
-\maketitle
+%\maketitle
 \begin{abstract}
+
 This paper presents the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
@@
 		\midrule
-    Urban Percentage & \citep{Jones2016} & Subnational \\
-    Stunting & \citep{Local2020} & Subnational \\
-    Wasting & \citep{Local2020} & Subnational \\
-    Mean Years of Schooling & \citep{Smits2019, KC2017} & Subnational\\
-    GDP Per Capita & \citep{Smits2019, Dellink2017} & Subnational \\
-    Gini Coefficient & \citep{Rao2019a} & National\\
-    Poverty Headcount Index & \citep{Cuaresma2018} & National \\
-    Water Scarcity & \citep{greve2018global} & Subnational \\
-    Mean Annual Precipitation &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
-    Topographic Ruggedness &  \cite{USGS1996, Riley1999} & Subnational \\
-    Mean Temperature &  \cite{abatzoglou2018terraclimate, warszawski2014inter} & Subnational \\
-    Malaria (\textit{P. falciparum}) Mortality Rate &  \cite{Weiss2019} & Subnational \\
+    Urban Percentage & \cite{Jones2016} & Subnational \\ [1cm]
+    Stunting & \cite{Local2020} & Subnational \\ [1cm]
+    Wasting & \cite{Local2020} & Subnational \\ [1cm]
+    Mean Years of Schooling & \makecell[tl]{\cite{Smits2019} \\ \cite{KC2017}} & Subnational\\ [1cm]
+    GDP Per Capita & \makecell[tl]{\cite{Smits2019} \\ \cite{Dellink2017}} & Subnational \\ [1cm]
+    Gini Coefficient & \cite{Rao2019a} & National\\ [1cm]
+    Poverty Headcount Index & \cite{Cuaresma2018} & National \\ [1cm]
+    Water Scarcity & \cite{greve2018global} & Subnational \\ [1cm]
+    Mean Annual Precipitation & \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
+    Topographic Ruggedness &  \makecell[tl]{\cite{USGS1996} \\ \cite{Riley1999}} & Subnational \\ [1cm]
+    Mean Temperature &  \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
+    \makecell[tl]{Malaria (\textit{P. falciparum}) \\ Mortality Rate} & \cite{Weiss2019} & Subnational \\
 		\bottomrule
@@
   \centering
-  \includegraphics[width=\linewidth]{img/TimeSeries.pdf}
+  \includegraphics[width=\linewidth]{TimeSeries.pdf}
   \caption{Number of people over the thresholds for moderate and severe food insecurity, by World Bank income groups as of 2020.
@@
   \centering
-  \includegraphics[width=\linewidth]{img/Rates.pdf}
+  \includegraphics[width=\linewidth]{Rates.pdf}
   \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020.}
@@
   \centering
-  \includegraphics{img/FullMap.pdf}
+  \includegraphics[width=\linewidth]{FullMap.pdf}
   \caption{Spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
@@
   \centering
-  \includegraphics[width=\linewidth]{img/VIMP.pdf}
+  \includegraphics[width=\linewidth]{VIMP.pdf}
   \caption{Importance of each variable in predicting the percent of a population over thresholds for moderate and severe food insecurity.

### 2021-06-11 docs/response_to_reviewers.md
being making edits for reviewer

---
created: 2021-06-10T10:04:08-0400
modified: 2021-06-11T12:16:25-0400
title: Response to Reviewers
header-includes:
 - \hypersetup{colorlinks=true}
 - \hypersetup{urlcolor=blue}
---

This paper presents a global assessment of food security outcomes and prediction for the years to come. It leverages a useful household-level food security dataset and gap fills with a random forest-model to predict food security outcomes in missing locations. While it has potential to add important findings to the literature, there are several methodological gaps that need to be addressed before it can be considered for publication. The General Comments below reflect broader specific locations in the text where I take issue. That said, I have not fully addressed line-by-line critiques yet because the broader concerns are too numerousR I highly encourage the authors to not be discouraged, as I see this work as immensely valuable, but I want them to produce the most robust science possible.

First, there is a growing body of literature that shows wide variation estimates of population across gridded gridded population products (e.g., Chen et al. 2020; Bustos et al. 2020; Tuholske et al. 2019) and recently published fitness-for-use guidelines (see Leyk et al. 2019). The authors do not clearly explain why they chose WorldPop (as opposed to Landscan, ESRI's World Population Estimates, the Global Human Settlement Layer, High Resolution - see www.popgrid.org for more). Furthermore, it is not clear how the Worldpop data is even used. It is mentioned in the Main Text, but I do not see it in your Supplemental Materials. Population estimates should be re-run using other gridded population datasets as I suspect that they will yield markedly different results.

**We regret that we did not make it more clear how the population data was used - we have now made that more clear in the text, and that may have also made it more apparent why we chose the WorldPop dataset.  We use gridded population data to estimate demographic characteristics for administrative areas, and chose to use WorldPop because it has data on age and sex.  Examining the other datasets you mention and searching through that very useful review by Leyk et al., it seems that GPWv4 also includes such demographic data, but more coarsely aggregated (age categories are across wide ranges, such as 15-49, whereas WorldPop bins ages every five years).  Moreover, WorldPop is an annual product that we can match to the specific year of our survey data, whereas GPWv4 is only available for 2015.  We have now made it more apparent in the _Disaggregation_ subsection of the _Methods_ section why we chose WorldPop, and have added some discussion and relevant citations from your review.**

**Additionally, we also used population data at the administrative area level to estimate the food insecure population totals after modeling rates of food insecurity.  This data came from the [Subnational Human Development Database (SHDI)](https://doi.org/10.1038/sdata.2019.38), which in turn is derived from national censuses.  Because this analysis is conducted at the admin-area level, we believe it is not necessary to use high-resolution gridded data to estimate population totals and the SHDI data is sufficient.**

Second, like the population data employed, I have strong concerns with the temperature and precipitation data and methodology employed here. Why use TerraClimate over the widely used and higher resolution CHIRPS (Funk et al. 2015) or other climate reanalysis (ERA5, MERRA-2, etc.) and the new CHIRTS-Max (Funk et al 2019) or CHIRTS-daily (Verdin et al. 2020) high-resolution temperature products to establish your baseline? What temporal resolution is 'mean temperature'? Surely it is widely established the annualized climate data over large distance data does not reflect the true nature of coupling between household food security outcomes and climate (see Shukla et al. 2020 for one recent example). I point to the example of Sternberg 2012 that clearly shows how climate shocks/variability across the planet can telescope to food security outcomes at great distances. How climate (across spatial and temporal scales) affect local price and labor is key here, yet you do not even
discuss price at all, much less employ it, in your modeling paradigm. Yet we know that specifically for urban areas, where half the planet lives, food security outcomes are highly dependent on labor and price (see Blekking et al. 2020 for an example). Also, what about seasonality (again see Shukla et al. 2020 for one example)? None of these crucial aspects of household-level food security outcomes are clearly addressed in the current version of the manuscript. I suspect that the reason your 'climate' variables do not predict your outcome is because you spatially and temporally smooth your data to the point where it will no matter what not capture the true nature of how climate shocks impact food security (e.g. huge spatiotemporal coupling between where food is grown, when it is grown, how much the price is dictated by local, national, and global market forces, etc.). See Niles et al. 2021 for a much more robust approach.

**We thank you for this detailed feedback, and we agree that the approach to modeling the impacts of climate on food security is weak.  However, we are in many ways hobbled by our survey data, which is only available at the annual temporal scale and only reported at the national spatial scale, which we use to infer subnational scales.  This makes it much harder to trace the impacts of climate on food security and necessitates the smoothing of our climate data.  The papers you cite by Niles, Blekking, and Shukla all seem to have indicators of food security available at fine spatial scales (Niles uses DHS clusters/community GPS points, Blekking uses household GPS points) and fine temporal scales (Niles uses monthly data, Shukla uses Dekads).**

**We also note that it is important to consider the scope of our analysis.  Weather shocks as well as labor and food prices certainly explain a lot of the variation in food security between households and across seasons.  However, the scope of this analysis is to explore drivers of food insecurity across the globe and over decadal timeframes.  We are especially interested in modeling future trends in food security.  We therefore use GCM projections, which are useful insofar as they are indicative of shifting baselines, but are unable to anticipate actual droughts or floods that would impact food insecurity in 2030.**

**Given these considerations of how the scale of our data (annual, admin-area scale) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value. (Also, the CHIRPS and CHIRTS datasets, while excellent, are not truly global and unavailable for certainly latitudes, so couldnt be applied to this analysis).  However, you raise very important considerations in your feedback and we therefore make two improvements to the paper: (1) in addition to the survey year's rainfall totals and precipitation averages, we add a feature for the previous year's total rainfall and mean temperature, and additionally (2) we add a paragraph to the discussion highlighting many of the points you raise and further outlining the considerations that led to the analytical approach we took.**

Third, Random Forrest models suffer from training data bias. If the underlying training data if flawed, the Random Forest Model has no way of identifying this. What quality control is used to determine if your survey data is correct? One way to address this is to compare the Gallup data with other sources like DHS or LSMS surveys from the same year. Do they both measure similar national-level food security outcomes? I am also curious to know how your models preform with simple linear or logistic regression? Do you get the same results? I recognize Random Forest are nonparametric, but I would be keen to see how linear models preform with controls. Also, what happens when you trade out wasting/stunting as your predicted outcome versus the FIES-based metric?

Fourth, the spatial nature of your modeling is confusing and not clear. Are metrics averaged at the sub-national scale (if so what level of analysis and for which country)? How is degree of urbanization quantified (e.g. are Worldpop pixels masked at some threshold?)? Is the Gallup representative at the national level? District-level? What controls are used to ensure that the Gallup data is a representative sample (Not even explained in the Supplemental Materials)? Your methodology (including how you address issues of spatial scale) needs to be more clearly described in the Main Text and the Supplemental Material before this paper can be considered for publication.

https://userforum.dhsprogram.com/index.php?t=msg&th=7832&start=0&

Fifth, I take issue with this being the first global food security paper at a sub-national level (Pg 12, ln 11). Niles et al. is another similar paper, that is near global, with a much more robust modeling framework. Additionally, FEWS-net and WFP provide these assessments regularly or real time. Specifically, you do not present any sub-national findings. What are they? Why is sub-national important and how do your results advance our understanding of food security over space and time? Surely there may be important nuances in the sub-national data (e.g. Does Nairobi look like Dhaka? Is your assumption that two households with similar characteristics, irrespective of geography, correct? Do your results show that this assumption is correct?).

Finally, and importantly, how are you defining 'urban'? (See Catteneo et al. 2021 for further details). This is not clear in your Supplementary Material or main text. You do not address the growing body of literature that clearly demonstrates important differences between 'rural' and 'urban' food security measurement and outcomes (see Blekking et al. 2020, Tuholske et al. 2020, Haysom & Tawodzera 2018, Cockx et al. 2018) that are extremely important for stakeholders and policy development (Battersby 2017) and poverty alleviation (Christiaensen et al. 2017).

A few specifics:

Pg. 3, Ln 35 - 38: This assumption needs strong justification. Recent urban food security studies show wide heterogeneity in food security outcomes among 'low-income' households depending on the food security metric used (see Blekking et al. 2020, Tuholske et al. 2020, etc.). 'Urbanization level' at an administrative-unit level (if that is what you are using…it's not clear), thus, may not be good predictor of food security outcomes.

Pg. 4, Ln 20 - 23: Please expand your justification as to why you selected these climate datasets. Why is spatial averaging over large length scales appropriate to predict food security outcomes? Why is annual average precipitation and (I assume annual) average temperature appropriate?

Pg. 7, Ln 19 - 30: This results is in contrast to FEWS NET outlook that the those at risk to famine rose by 85 million people from 2015 - 2019 (I can't find the brief, but here is an article with the numbers: https://www.sciencemag.org/news/2020/04/how-team-scientists-studying-drought-helped-build-world-s-leading-famine-prediction). This was before cofid. Why do your numbers differ?

Pg. 9, 3.2: You have sub-national data, and you make assumptions that the characteristics of two similar households have the same food security outcomes, yet you present your findings at national level. An increasing body of research (e.g. ref) is moving away from national-level estimates as they mask the true nature of food security outcomes and are not relevant for policy maker or targeted interventions. Please justify these findings given the high degree of spatiotemporal heterogeneity in food security outcomes.

References

Battersby, J. (2017). MDGs to SDGs-new goals, same gaps: the continued absence of urban food security in the post-2015 global development agenda. African Geographical Review, 36(1), 115-129.

Blekking, J., Waldman, K., Tuholske, C., & Evans, T. (2020). Formal/informal employment and urban food security in Sub-Saharan Africa. Applied Geography, 114, 102131.

Bustos, M. F. A., Hall, O., Niedomysl, T., & Ernstson, U. (2020). A pixel level evaluation of five multitemporal global gridded population datasets: a case study in sweden, 1990-2015. Population and environment, 1-23.

Cattaneo, A., Nelson, A., & McMenomy, T. (2021). Global mapping of urban-rural catchment areas reveals unequal access to services. Proceedings of the National Academy of Sciences, 118(2).

Chen R, Yan H, Liu F, Du W, Yang Y. Multiple Global Population Datasets: Differences and Spatial Distribution Characteristics. ISPRS International Journal of Geo-Information. 2020;9: 637.

Christiaensen, L., & Kanbur, R. (2017). Secondary towns and poverty reduction: refocusing the urbanization agenda. Annual Review of Resource Economics, 9, 405-419.

Cockx, L., Colen, L., & De Weerdt, J. (2018). From corn to popcorn? Urbanization and dietary change: evidence from rural-urban migrants in Tanzania. World Development, 110, 140-159.

Funk, C., Peterson, P., Landsfeld, M., Pedreros, D., Verdin, J., Shukla, S., ... & Michaelsen, J. (2015). The climate hazards infrared precipitation with stations—a new environmental record for monitoring extremes. Scientific data, 2(1), 1-21.

Funk, C., Peterson, P., Peterson, S., Shukla, S., Davenport, F., Michaelsen, J., ... & Mata, N. (2019). A high-resolution 1983-2016 T max climate data record based on infrared temperatures and stations by the climate Hazard center. Journal of Climate, 32(17), 5639-5658.
Haysom, G., & Tawodzera, G. (2018). "Measurement drives diagnosis and response": Gaps in transferring food security assessment to the urban scale. Food Policy, 74, 117-125.

Leyk, S., Gaughan, A. E., Adamo, S. B., Sherbinin, A. D., Balk, D., Freire, S., ... & Pesaresi, M. (2019). The spatial allocation of population: a review of large-scale gridded population data products and their fitness for use. Earth System Science Data, 11(3), 1385-1409.

Niles, M. T., Emery, B. F., Wiltshire, S., Brown, M. E., Fisher, B., & Ricketts, T. H. (2021). Climate impacts associated with reduced diet diversity in children across nineteen countries. Environmental Research Letters, 16(1), 015010.

Shukla, S., Husak, G., Turner, W., Davenport, F., Funk, C., Harrison, L., & Krell, N. (2021). A slow rainy season onset is a reliable harbinger of drought in most food insecure regions in Sub-Saharan Africa. Plos one, 16(1), e0242883.

Sternberg, T. (2012). Chinese drought, bread and the Arab Spring. Applied Geography, 34, 519-524.

Tuholske, C., Caylor, K., Evans, T., & Avery, R. (2019). Variability in urban population distributions across Africa. Environmental Research Letters, 14(8), 085009.

Tuholske, C., Andam, K., Blekking, J., Evans, T., & Caylor, K. (2020). Comparing measures of urban food security in Accra, Ghana. Food Security, 1-15.

Verdin, A., Funk, C., Peterson, P., Landsfeld, M., Tuholske, C., & Grace, K. (2020). Development and validation of the CHIRTS-daily quasi-global high-resolution daily temperature data set. Scientific Data, 7(1), 1-14.

### 2021-06-11 docs/modeling_fies_update.tex
being making edits for reviewer

--- 2021-03-04 docs/modeling_fies.tex
@@
 \begin{abstract}
-
-This paper presents the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
+This paper presents the first global picture of food security at an administrative area, subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
 Using microdata from 75 countries and filling gaps using machine learning, we find significant heterogeneity in levels of food insecurity around the world.
@@
 We take advantage of individual-level characteristics collected in FAO's FIES surveys to push the analysis beyond the country level to a sub-national level.
-To our knowledge, this is the first effort to model FIES-based food insecurity at that scale.
+To our knowledge, this is the first effort to model FIES-based food insecurity at that scale and make projections to 2030.
 
@@
 \subsection{Disaggregation}
-The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
+The data on the FIES collected by Gallup on behalf of FAO is collected annually within a country, and is therefore at the country-year level.
+This data also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
 Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
-We use a similar methodology and calculate post-stratification weights at a subnational level.
-We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.
-For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
+In order to disaggregate the data to the administrative area level, we use a similar methodology and calculate post-stratification weights.
+We use year-specific administrative area estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each administrative area across all individuals.
+For subnational distributions of age and gender, we rely on gridded population data.
+While there are many different gridded population datasets, they often take different approaches to disaggregating population data, are available at different temporal resolutions, and have varying availability of demographic data \cite{Ley2019Sep, friedman2001elements, Tuholske2019Aug}.
+We therefore chose to use the WorldPop dataset, which has demographic data on age and sex and is available annually \citep{Tatem2017}.
+For rates of urbanization by administrative area, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}, and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
 This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.
@@
 \subsection{Covariates}
-We model subnational rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).
-We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.
+We model rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment at the administrative area level (See Table \ref{tab:covars}).
+We employ covariates for which projections are available up to year 2030 and, in most cases, also have administrative area spatial resolution.
 For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.
@@
 These include mean years of schooling, GDP per capita, as well as population.
-We use observed trajectories in the historical distribution of variables among subnational areas within a country to predict the future distribution of GDP, population, and schooling among subnational areas and disaggregate national-level future projections.
+We use observed trajectories in the historical distribution of variables among administrative areas within a country to predict the future distribution of GDP, population, and schooling among these areas and disaggregate national-level future projections.
 
-For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions for variables up to 2030.
+For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each administrative area to model these and to obtain predictions for variables up to 2030.
 We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.
@@
 \subsection{Modeling}
-We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
+We fit two models for all administrative areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
 We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.
@@
 When performing cross validation, it is typical to select observations at random.
-However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.
+However, in our training data we have multiple highly similar observations from each country across years and administrative areas, and we extend our model predictions to countries that were not found at all in our training data.
 Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.
@@
 \subsection{Food Insecurity Across Space}
-We find substantial heterogeneity in the global distribution of severe food insecurity across countries and subnational regions (See Fig.
+We find substantial heterogeneity in the global distribution of severe food insecurity across countries and administrative areas (See Fig.
 \ref{fig:map}).
-For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one subnational area in every country except Gabon, Djibouti and Equatorial Guinea.
+For the year 2020, mainland sub-Saharan Africa is the continent with the highest rates of severe food insecurity, with 20\% of people over the corresponding threshold in at least one administrative area in every country except Gabon, Djibouti and Equatorial Guinea.
 Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.
@@
 Even in highly developed economies, such as the United States and parts of eastern and southern Europe, over 10\% of the population is above the threshold for moderate food insecurity.
+
+We also find substantial within-country variation in rates of moderate and severe food insecurity in many countries.
+In Brazil, for example, northern regions are much more food insecure than southern regions, while in China, coastal regions and administrative areas that contain large cities such as Beijing and Shanghai are more food secure.
+Finally, in many African countries, arid regions have higher rates of moderate and severe food insecurity, such as the Sahel in Mali and Niger, the Central and Western Zones of Tanzania, and the Somali Region of Ethiopia.
 
@@
 \section{Discussion}
-This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.
+This paper presents the first global estimates of the experience of food insecurity at a administrative area level using the FIES, with forecasts until the year 2030.
 We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of economic growth in developing countries.
@@
 
-\section{Additional Information}
-All code necessary to reproduce these results is on Github at \href{https://github.com/mcooper/wdl-fies}{github.com/mcooper/wdl-fies}. We have no competing interests to disclose.
+\section{Competing Interests}
+We have no competing interests to disclose.
 

### 2021-06-11 viz/compare_raw_disag_and_DHS.R

@@ -1,0 +1,73 @@
+setwd('~/wdl-fies');library(ProjectTemplate);load.project()
+
+dhs <- read_sf('~/wdl-fies/data/disag_vars/rawdata/dhs_statcompiler/shps', 
+               'sdr_subnational_data_dhs_2015')
+
+fies_subnat$iso2 <- countrycode(substr(fies_subnat$GDLCODE, 1, 3), 'iso3c', 'iso2c')
+
+int <- intersect(fies_subnat$iso2, dhs$ISO)
+
+fies_sel <- fies_subnat %>%
+  filter(iso2 %in% int) %>%
+  group_by(GDLCODE, iso2) %>%
+  summarize(fies.mod = mean(fies.mod),
+            fies.sev = mean(fies.sev))
+
+fies_sel <- merge(gdl, fies_sel)
+
+dhs_sel <- dhs %>%
+  filter(ISO %in% int) %>%
+  select(dhs.stnt=CNNUTSCHA2, dhs.wast=CNNUTSCWH2, ISO)
+
+for (i in int){
+  
+  ggplot(fies_sel %>% filter(iso2 == i)) + 
+    geom_sf(aes(fill=fies.sev)) + 
+    scale_fill_gradient(low="#fef0d9", high="#b30000") + 
+    theme_void()
+  ggsave(paste0('~/wdl-fies/figures/compare_dhs/', i, '_fies.sev.png'))
+
+  ggplot(fies_sel %>% filter(iso2 == i)) + 
+    geom_sf(aes(fill=fies.mod)) + 
+    scale_fill_gradient(low="#fef0d9", high="#b30000") + 
+    theme_void()
+  ggsave(paste0('~/wdl-fies/figures/compare_dhs/', i, '_fies.mod.png'))
+
+  ggplot(dhs_sel %>% filter(ISO == i)) + 
+    geom_sf(aes(fill=dhs.stnt)) + 
+    scale_fill_gradient(low="#fef0d9", high="#b30000") + 
+    theme_void()
+  ggsave(paste0('~/wdl-fies/figures/compare_dhs/', i, '_dhs.stnt.png'))
+
+  ggplot(dhs_sel %>% filter(ISO == i)) + 
+    geom_sf(aes(fill=dhs.wast)) + 
+    scale_fill_gradient(low="#fef0d9", high="#b30000") + 
+    theme_void()
[…]

### 2021-06-11 docs/response_to_reviewers.md
more changes to response

@@
 created: 2021-06-10T10:04:08-0400
-modified: 2021-06-11T12:16:25-0400
+modified: 2021-06-11T16:36:24-0400
 title: Response to Reviewers
@@
 
+**To add more clarity, we have changed the phrasing in most places from "subnational" to "administrative area", as this might improve some of the confusion around the spatial scale of our analysis.**
+
 https://userforum.dhsprogram.com/index.php?t=msg&th=7832&start=0&
@@
 Do your results show that this assumption is correct?).
+
+**When stating that this was the first global analysis, we meant that this was the first analysis using the FIES metric, and have ensured that anywhere we make that claim throught the paper, we add that important caveat.
+We also believe that our use of the word "global" indicates that our analysis has estimated food insecurity in all countries or nearly all countries, which was not done by Niles et al (who only used 19 countires) and is not done by the FEWSNET reports (which only targets certain developing regions).
+Thus, we believe that this claim is valid under our interpretation, and we believe future readers would interpret the word "global" the same way.
+However, if you disagree, please let us know in a future review and we can change the language.**
+
+**Our lacking in interpretation of subnational findings was a real oversight, and we have added interpretation around that to the results and discussion sections.**
 

### 2021-06-11 viz/compare_raw_disag_and_DHS.R

@@ -59,6 +59,9 @@ summary(lm(fies.mod ~ dhs.stnt + ISO, data=comb))
 summary(lm(fies.sev ~ dhs.wast + ISO, data=comb))
 summary(lm(fies.mod ~ dhs.wast + ISO, data=comb))
 
+#These are significantly associated
+summary(lm(dhs.stnt ~ dhs.wast + ISO, data=comb))
+
 for (i in int){
   sel <- comb %>%
     filter(ISO == i)

### 2021-07-08 docs/response_to_reviewers.md
lag temperature and precipitation

@@
 created: 2021-06-10T10:04:08-0400
-modified: 2021-06-11T16:36:24-0400
+modified: 2021-07-08T14:25:54-0400
 title: Response to Reviewers
@@
 I point to the example of Sternberg 2012 that clearly shows how climate shocks/variability across the planet can telescope to food security outcomes at great distances.
-How climate (across spatial and temporal scales) affect local price and labor is key here, yet you do not even
-discuss price at all, much less employ it, in your modeling paradigm.
+How climate (across spatial and temporal scales) affect local price and labor is key here, yet you do not even discuss price at all, much less employ it, in your modeling paradigm.
 Yet we know that specifically for urban areas, where half the planet lives, food security outcomes are highly dependent on labor and price (see Blekking et al.
@@
 **We thank you for this detailed feedback, and we agree that the approach to modeling the impacts of climate on food security is weak.
-However, we are in many ways hobbled by our survey data, which is only available at the annual temporal scale and only reported at the national spatial scale, which we use to infer subnational scales.
+However, we are in many ways constrained by our survey data, which is only available at the annual temporal scale and only reported at the national spatial scale, which we use to infer subnational scales.
 This makes it much harder to trace the impacts of climate on food security and necessitates the smoothing of our climate data.
@@
 However, the scope of this analysis is to explore drivers of food insecurity across the globe and over decadal timeframes.
-We are especially interested in modeling future trends in food security.
+We are especially interested in modeling future trends in food security at these broad spatial and temporal scales.
 We therefore use GCM projections, which are useful insofar as they are indicative of shifting baselines, but are unable to anticipate actual droughts or floods that would impact food insecurity in 2030.**
 
-**Given these considerations of how the scale of our data (annual, admin-area scale) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value.
-(Also, the CHIRPS and CHIRTS datasets, while excellent, are not truly global and unavailable for certainly latitudes, so couldnt be applied to this analysis).
+**Given these considerations of how the scale of our data (annual, national scale) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value.
+(Also, the CHIRPS and CHIRTS datasets, while excellent, are not truly global and unavailable for certainly latitudes, so couldnt be applied to this global analysis).
 However, you raise very important considerations in your feedback and we therefore make two improvements to the paper: (1) in addition to the survey year's rainfall totals and precipitation averages, we add a feature for the previous year's total rainfall and mean temperature, and additionally (2) we add a paragraph to the discussion highlighting many of the points you raise and further outlining the considerations that led to the analytical approach we took.**
@@
 
+**Unforuntately, the Gallup World Poll (GWP) is currently the only source of nationally-respresntative data on the FIES, as this metric is not collected by surveys like the DHS and LSMS (although as the FIES has been adopted as an SDG indicator, this is changing, see [here](https://userforum.dhsprogram.com/index.php?t=msg&th=7832&start=0&)).
+Thus, we cannot use those datasets to validate our Random Forest model.
+However, we note that we do take care to conduct cross validtion at the country level.
+We leave out all surveys from 10% of countries when evaluating our model, and find our model it still performs favorably.**
+
+**Additionally, we have updated the paper to now model the FIES using linear regression and lasso regression (in the supplement), and note that, compared to logisitic regression, the random forest model performs quite well.
+We also now add some discussion around the importance of accounting for interactions between covariates based on a comparison between random forest and linear regression.**
+
 Fourth, the spatial nature of your modeling is confusing and not clear.
@@
 
-**To add more clarity, we have changed the phrasing in most places from "subnational" to "administrative area", as this might improve some of the confusion around the spatial scale of our analysis.**
-
-https://userforum.dhsprogram.com/index.php?t=msg&th=7832&start=0&
+**To add more clarity, we have changed the phrasing in most places from "subnational" to "administrative area", as this might improve some of the confusion around the spatial scale of our analysis.
+The Gallup data is a purposive sample designed to be representative at the national level, and we dissaggregated it based on demographic characteristics to the administrative area level.
+Gallup records whether individual respondents are in urban or rural areas based on the individuals own reporting.
+We then use global gridded data that is available under various SSPs to estimate the percent of the population in urban areas for each administrative area.
+We tried to cover this in the _Disaggregation_ subsection of the _Methods_ section, and have endeavoured to make this section even more clear and explicit.**
 

### 2021-07-08 data/covars/scripts/tavg_precip.R

@@ -194,7 +194,7 @@ pas <- read.csv('~/wdl-fies/data/covars/rawdata/temp-precip-past.csv') %>%
 #Fill missing years, get slopes and predict values from linear trend
 all <- bind_rows(fut, pas) %>%
   arrange(GDLCODE, YEAR) %>%
-  complete(GDLCODE, YEAR=2010:2030) %>%
+  complete(GDLCODE, YEAR=2009:2030) %>%
   group_by(GDLCODE) %>%
   nest %>%
   mutate(pint=map(data, function(df) lm(precip~YEAR, data=df)$coefficients[1]),
@@ -206,9 +206,13 @@ all <- bind_rows(fut, pas) %>%
   mutate(ppred = pint + pcoef*YEAR,
          tpred = tint + tcoef*YEAR,
          precip = ifelse(is.na(precip), ppred, precip),
-         tave = ifelse(is.na(tave), tpred, tave))
+         tave = ifelse(is.na(tave), tpred, tave)) %>%
+  group_by(GDLCODE) %>%
+  mutate(tave_l=lag(tave),
+         precip_l=lag(precip)) %>%
+  filter(YEAR > 2009)
 
-write.csv(all %>% select(GDLCODE, YEAR, tave, precip),
+write.csv(all %>% select(GDLCODE, YEAR, tave, precip, tave_l, precip_l),
           '~/wdl-fies/data/covars/results/tave-precip.csv', row.names=F)
 
 # library(sf)

### 2021-07-08 munge/03-Covariates.R

@@ -11,7 +11,7 @@ for (f in fs){
 #Select ony vars that have future data
 covars <- covars %>%
   select(YEAR, ISO3, GDLCODE, stunting, wasting, school_mean, ruggedness,
-         gdp_percap, gini, mal_falciparum, precip, hci, tave, 
+         gdp_percap, gini, mal_falciparum, precip, hci, tave, precip_l, tave_l,
          population, rural_perc, urban_perc, ws_share) %>%
   merge(regions %>%
           select(region=region_sdg, ISO3))

### 2021-07-09 docs/response_to_reviewers.md
run linear model, re-run rf with lagged climate vars

@@
 created: 2021-06-10T10:04:08-0400
-modified: 2021-07-08T14:25:54-0400
+modified: 2021-07-08T21:28:46-0400
 title: Response to Reviewers
@@
 We also believe that our use of the word "global" indicates that our analysis has estimated food insecurity in all countries or nearly all countries, which was not done by Niles et al (who only used 19 countires) and is not done by the FEWSNET reports (which only targets certain developing regions).
-Thus, we believe that this claim is valid under our interpretation, and we believe future readers would interpret the word "global" the same way.
+Thus, we believe that this claim is valid, and we believe future readers would interpret the word "global" the same way.
 However, if you disagree, please let us know in a future review and we can change the language.**
@@
 2017).
+
+**Our data on urban and rural areas comes from Jones and ONeill 2016, which use as thier data source the Global Rural-Urban Mapping Project (GRUMP), v1.
+GRUMP in turn bases their definition of urban on night-time lights and buffered settlement centroids.
+We have not made that more explicit in the text.
+We note the importance of urbanization in determining food security outcomes - that is why we have included it in our analysis.
+We have also included some of the literature you've highlighted in the paper.**
 

### 2021-07-09 models/model_linear.R

@@ -1,0 +1,71 @@
+setwd('~/wdl-fies'); library(ProjectTemplate); load.project()
+
+# read in data
+moddat <- merge(fies_subnat,
+                covars,
+                all.x=T, all.y=F) %>%
+  na.omit %>%
+  data.frame
+
+preddat <- covars %>%
+  data.frame
+
+# set up model
+vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
+                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
+                                            'urban', 'rural', 'fies.sev', 'fies.mod',
+                                            'population', 'YEAR', 'rural_perc', 'region')]
+
+moddat <- moddat %>%
+  mutate(fies.mod.logit = logit(fies.mod),
+         fies.sev.logit = logit(fies.sev))
+
+mm <- model.matrix(as.formula(paste("fies.mod.logit", 
+                                           paste(vars, collapse = "+"), sep= "~")),
+                data = moddat)
+pm <- model.matrix(as.formula(paste("YEAR", 
+                                           paste(vars, collapse = "+"), sep= "~")),
+                data = preddat)
+
+#############################################
+#Run model under cross validation
+#at the country level
+##########################################
+rsq <- function(y, yhat){
+  1 - (sum((y - yhat)^2))/sum((y - mean(yhat))^2)
+}
+
+for (iso3 in unique(moddat$ISO3)){
+  print(iso3)
+  ix <- moddat$ISO3 != iso3
+  
+  lasso.mod <- cv.glmnet(mm[ix, ], moddat$fies.mod.logit[ix], alpha=1)
+  lasso.sev <- cv.glmnet(mm[ix, ], moddat$fies.sev.logit[ix], alpha=1)
+
+  moddat$fies.mod.pred.cv[!ix] <- inv.logit(predict(lasso.mod, mm[!ix,])[, 1])
+  moddat$fies.sev.pred.cv[!ix] <- inv.logit(predict(lasso.sev, mm[!ix,])[, 1])
+}
+
+sev.rsq <- cor(moddat$fies.sev, moddat$fies.sev.pred.cv)^2
+mod.rsq <- cor(moddat$fies.mod, moddat$fies.mod.pred.cv)^2
[…]

### 2021-07-09 models/model_randomforest.R

@@ -37,12 +37,17 @@ prm$sev.mae <- NA
 prm$mod.mae <- NA
 
 #############################################
-#Run model under 10-fold cross validation
+#Run model under cross validation
 #at the country level
 ##########################################
+rsq <- function(y, yhat){
+  1 - (sum((y - yhat)^2))/sum((y - mean(yhat))^2)
+}
+
 for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
   cat(round(sum(!is.na(prm$sev.rsq))/nrow(prm)*100, 2), 'percent done!\n')
   for (iso3 in unique(moddat$ISO3)){
+    print(iso3)
     ix <- moddat$ISO3 != iso3
     
     mtry <- prm$mtry[i]
@@ -73,8 +78,8 @@ for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
     moddat$fies.sev.pred.cv[!ix] <- inv.logit(predict(rf.sev, moddat[!ix,])$predicted)
     
   }
-  prm$sev.rsq[i] <- cor(moddat$fies.sev.pred.cv, moddat$fies.sev)^2
-  prm$mod.rsq[i] <- cor(moddat$fies.mod.pred.cv, moddat$fies.mod)^2
+  prm$sev.rsq[i] <- cor(moddat$fies.sev, moddat$fies.sev.pred.cv)^2
+  prm$mod.rsq[i] <- cor(moddat$fies.mod, moddat$fies.mod.pred.cv)^2
   prm$sev.mae[i] <- mean(abs(moddat$fies.sev.pred.cv - moddat$fies.sev))
   prm$mod.mae[i] <- mean(abs(moddat$fies.mod.pred.cv - moddat$fies.mod))
 }

### 2021-07-09 docs/response_to_reviewers.md
more edits

@@
 created: 2021-06-10T10:04:08-0400
-modified: 2021-07-08T21:28:46-0400
+modified: 2021-07-08T21:31:22-0400
 title: Response to Reviewers

### 2021-07-09 models/model_randomforest.R

@@ -78,8 +78,8 @@ for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
     moddat$fies.sev.pred.cv[!ix] <- inv.logit(predict(rf.sev, moddat[!ix,])$predicted)
     
   }
-  prm$sev.rsq[i] <- cor(moddat$fies.sev, moddat$fies.sev.pred.cv)^2
-  prm$mod.rsq[i] <- cor(moddat$fies.mod, moddat$fies.mod.pred.cv)^2
+  prm$sev.rsq[i] <- rsq(moddat$fies.sev, moddat$fies.sev.pred.cv)
+  prm$mod.rsq[i] <- rsq(moddat$fies.mod, moddat$fies.mod.pred.cv)
   prm$sev.mae[i] <- mean(abs(moddat$fies.sev.pred.cv - moddat$fies.sev))
   prm$mod.mae[i] <- mean(abs(moddat$fies.mod.pred.cv - moddat$fies.mod))
 }

### 2021-07-13 docs/response_to_reviewers.md
finish response to reviewers, add new LASSO analysis

--- 2021-07-08 docs/response_to_reviewers.md
@@
 created: 2021-06-10T10:04:08-0400
-modified: 2021-07-08T14:25:54-0400
+modified: 2021-07-12T20:10:12-0400
 title: Response to Reviewers
@@
 
+**We thank the reviewer for the detailed and thoughtful feedback, which forced us to engaged more with issues of spatial scale, household vs national food security, as well as a more broad body of literature thn we originally examined.
+We believe that the paper is now much improved and will be of interest to a wider body of researchers, with more clear explanations of the methodlogy and spatial scale, and more detailed discussions of issues like population, urbanization, and climate.**
+
 This paper presents a global assessment of food security outcomes and prediction for the years to come.
@@
 The General Comments below reflect broader specific locations in the text where I take issue.
-That said, I have not fully addressed line-by-line critiques yet because the broader concerns are too numerousR I highly encourage the authors to not be discouraged, as I see this work as immensely valuable, but I want them to produce the most robust science possible.
+That said, I have not fully addressed line-by-line critiques yet because the broader concerns are too numerous.
+I highly encourage the authors to not be discouraged, as I see this work as immensely valuable, but I want them to produce the most robust science possible.
 
@@
 
-**We thank you for this detailed feedback, and we agree that the approach to modeling the impacts of climate on food security is weak.
-However, we are in many ways constrained by our survey data, which is only available at the annual temporal scale and only reported at the national spatial scale, which we use to infer subnational scales.
+**We thank you for this detailed feedback, and we agree that the approach to modeling the impacts of climate on food security is too simple to capture all of the pathways by which the climate affects food security.
+However, we are in many ways constrained by our survey data, which is only available at the annual temporal scale and only reported at the national spatial scale.
 This makes it much harder to trace the impacts of climate on food security and necessitates the smoothing of our climate data.
@@
 **Given these considerations of how the scale of our data (annual, national scale) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value.
-(Also, the CHIRPS and CHIRTS datasets, while excellent, are not truly global and unavailable for certainly latitudes, so couldnt be applied to this global analysis).
+(Also, the CHIRPS and CHIRTS datasets, while excellent, are not truly global and unavailable for certain latitudes, so couldnt be applied to this global analysis).
 However, you raise very important considerations in your feedback and we therefore make two improvements to the paper: (1) in addition to the survey year's rainfall totals and precipitation averages, we add a feature for the previous year's total rainfall and mean temperature, and additionally (2) we add a paragraph to the discussion highlighting many of the points you raise and further outlining the considerations that led to the analytical approach we took.**
@@
 **To add more clarity, we have changed the phrasing in most places from "subnational" to "administrative area", as this might improve some of the confusion around the spatial scale of our analysis.
-The Gallup data is a purposive sample designed to be representative at the national level, and we dissaggregated it based on demographic characteristics to the administrative area level.
+We have also added a paragraph to the methods section giving a high-level overview of the analysis, clearly stating the scale of the study (year x admin area).
+The Gallup data is designed to be representative at the national level through a weighting schema based on demographic characterstics, and we used the same methodology to get estimates at the administrative area level.
 Gallup records whether individual respondents are in urban or rural areas based on the individuals own reporting.
-We then use global gridded data that is available under various SSPs to estimate the percent of the population in urban areas for each administrative area.
-We tried to cover this in the _Disaggregation_ subsection of the _Methods_ section, and have endeavoured to make this section even more clear and explicit.**
+We then use global gridded data from Jones and O'Neill (2016) in _Environmental Research Letters_ to estimate the percent of the population in urban areas for each administrative area through the year 2030.**
 
@@
 
-**When stating that this was the first global analysis, we meant that this was the first analysis using the FIES metric, and have ensured that anywhere we make that claim throught the paper, we add that important caveat.
+**When stating that this is the first global picture of food insecurity, we meant that this was the first analysis using the FIES metric, and have ensured that anywhere we make that claim throught the paper, we include that important caveat.
 We also believe that our use of the word "global" indicates that our analysis has estimated food insecurity in all countries or nearly all countries, which was not done by Niles et al (who only used 19 countires) and is not done by the FEWSNET reports (which only targets certain developing regions).
-Thus, we believe that this claim is valid under our interpretation, and we believe future readers would interpret the word "global" the same way.
-However, if you disagree, please let us know in a future review and we can change the language.**
-
-**Our lacking in interpretation of subnational findings was a real oversight, and we have added interpretation around that to the results and discussion sections.**
+Thus, we believe that this claim is valid, and we believe future readers would interpret the word "global" the same way.
+However, if you disagree, please let us know in a future review and we can change the language.
+We have also added more description of the subnational findings to the results section.**
 
@@
 
+**Our data on urban and rural areas comes from Jones and ONeill 2016, which use as their data source the Global Rural-Urban Mapping Project (GRUMP), v1.
+GRUMP in turn bases their definition of urban on night-time lights and buffered settlement centroids.
+We have now made that more explicit in the text.
+We note the importance of urbanization in determining food security outcomes and now discuss how an analysis at this the admin area scale cannot capture all of the impacts of finer processes like urbanization and climate.
+We have also added some of the literature you've highlighted to the paper.**
+
 A few specifics:
@@
 
+**We agree that there is a substantial variation in food security status between urban and rural areas, even low income ones.
+However, we also include data on other available demographic factors, including gender and age, to perform our disaggregation.
+Moreover, we are not analyzing individual households - we are just making population-level estimates of rates of food insecurity at the administrative area scale, based on samples that include thousands of households.
+Some assumptions are built into this process, but they are common to all survey weighting methods, as we now note in the text.**
+
 Pg.
@@
 Why is annual average precipitation and (I assume annual) average temperature appropriate?
+
+**In addressing your previous higher-level comment, we have added significantly more discussion around climate and spatial scale to the paper.
+We add a note to this section highlighting that, given the scale of our food security data and our analysis, we are mostly focusing on climatic baselines as they relate to agro-ecological conditions**
 
@@
 
+**This is an interesting point.
+We note that the paragraph you take issue with is specifically focusing on the period of 2010 to 2030, and our figures, which include the range 2015-2019, show increases in severe and moderate food insecurity over this period, jibing with CHG's findings.
+We also took some effort to compare our results to other reports of trends in food security, focusing mostly on the metric of undernourishment in the FAO's State of Food Insecurity report.
+We now have a reference to the science news article you mentioned.**
+
 Pg.
@@
 Please justify these findings given the high degree of spatiotemporal heterogeneity in food security outcomes.
+
+**It is true that there is a large body of important literature that focuses on food security at a household scale, and based on your comments throughout the review, it is clear that you are very familiar with this literature.
+However, we disagree that "national-level estimates...
+are not relevant for policy makers".
+While local practitioners focus on differences in food security between households, many mulitnational practitions pay close attention to country-level (or admin-level) estimates of food security.
+We note the FAO's State of Food Insecurity (SOFI) reports, which are widely read, as well as the fact that the SDG targets, which are key metrics for policymakers and which include the FIES for SDG2, are monitored at the national level (see: https://sdg-tracker.org/zero-hunger).
+We discuss both the SOFI estimates and the relevance to SDG2 throughtout the paper.
+Moreover, as we mention in the acknowledgments, this project began out of a specific demand from an international policymaking organization.**
 

### 2021-07-13 docs/modeling_fies.tex
finish response to reviewers, add new LASSO analysis

@@
 ]{biblatex}
+
+% Use datatool to read in reported values
+% csv must have columns `var` and `val`
+\usepackage{datatool}
+\DTLsetseparator{,}
+\DTLloaddb{data}{vars.csv}
+\newcommand{\myvar}[1]{\DTLfetch{data}{var}{#1}{val}}
 
@@
 \begin{abstract}
-
 This paper presents the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of ``Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.
@@
 
-Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.
+Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \citep{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
@@
 \section{Methods}
+This analysis had three steps.
+First, in countries where FIES microdata was available at the national level, we generate subnational weights to disaggregate the microdata and estimate food insecurity rates at the first administrative level for each survey year.
+Second, we gather data on covariates related to food insecurity from previously published datasets that include future predictions under the Shared Socio-Economic Pathways framework.
+These covariates are processed to be at the same yearly, administrative area level, for all years from 2010 to 2030.
+Finally, we fit a machine learning model using the covariates in administrative areas where microdata is available, and use that model to make predictions for all years and all administrative areas.
+
 \subsection{Disaggregation}
@@
 Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
-We use a similar methodology and calculate post-stratification weights at a subnational level.
+This ensures that the results are representative at a national level.
+We use a similar methodology, except that we calculate post-stratification weights at a subnational level.
 We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.
-For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
-This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.
+For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates of population in urban and rural pixels published by Jones and O'Neill and deriving from night-time lights and buffered settlement centroids \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
+This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status, an assumption that underlies much of survey weighting methodology \citep{bethlehem2009applied}.
 
@@
 
-For the climatic variables, temperature and precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
+For the climatic variables, mean annual temperature and total annual precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
+These variables cannot speak to sudden shocks but do capture the climatic baselines relevant to agro-ecological conditions that constrain food production.
 
 Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.
-These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.
+These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \citep{Cuaresma2018}.
 
@@
     Water Scarcity & \cite{greve2018global} & Subnational \\ [1cm]
-    Mean Annual Precipitation & \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
+    \makecell[tl]{Total Annual Precipitation \\ (Current and Previous Year)} & \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
     Topographic Ruggedness &  \makecell[tl]{\cite{USGS1996} \\ \cite{Riley1999}} & Subnational \\ [1cm]
-    Mean Temperature &  \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
+    \makecell[tl]{Mean Temperature \\ (Current and Previous Year)} &  \makecell[tl]{\cite{abatzoglou2018terraclimate} \\ \cite{warszawski2014inter}} & Subnational \\ [1cm]
     \makecell[tl]{Malaria (\textit{P. falciparum}) \\ Mortality Rate} & \cite{Weiss2019} & Subnational \\
@@
 We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
-We use random forest regressions, which perform well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.
+We use both random forest regressions and LASSO (Least Absolute Shrinkage and Selection Operator) regressions, with the results of the LASSO regressions presented in the supplement.
 A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.
+This approach performs well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.
 By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.
@@
 These include the average number of observations in a leaf node, the number of variables randomly selected as candidates for splitting a node, and the maximum depth of each tree.
-We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$.
+We use the combination of hyperparameters that perform best under leave-one-out validation based on $R^2$ and mean average error (MAE).
 
@@
 Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.
-By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.
-After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.
+By this validation metric, we find that our models perform well, with an $R^2$ of \myvar{rf_cv_mod_rsq} and mean average error (MAE) of \myvar{rf_cv_mod_mae} for moderate food insecurity and an $R^2$ of \myvar{rf_cv_sev_rsq} and a MAE of \myvar{rf_cv_sev_mae} for severe food insecurity.
+After fitting models on the entire dataset, we observe a within-sample $R^2$ of \myvar{rf_wthn_mod_rsq} and an MAE of \myvar{rf_wthn_mod_mae} for moderate food insecurity and an $R^2$ of \myvar{rf_wthn_sev_rsq} and a MAE of \myvar{rf_wthn_sev_mae} for severe food insecurity.
 
@@
 \subsection{Food Insecurity over Time}
-Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and 2.5 billion people are above the threshold for moderate food insecurity (See Fig.
+Globally, in the year 2020, we estimate that \myvar{glob_2020_sev} million people are above the threshold for severe food insecurity, and \myvar{glob_2020_mod} billion people are above the threshold for moderate food insecurity (See Fig.
 \ref{fig:timeseries}).
@@
 
-Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.
-However, for both indicators, these changes are very slight: in 2030, the number of moderately food insecure people will have increased by 2.2\% since 2010, and the number of severely food insecure people will have decreased by 7.0\%.
-The increases in moderate food insecurity have been in lower and lower-middle income countries, while the decreases in severe food insecurity are entirely from lower income countries.
-For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.
+Under realistic scenario assumptions, we find recent increases in the number of moderate and severely food insecure people around the world, yet expect overall decreases from 2010 to 2030.
+These changes are more slight for people who are at least moderately food insecure: in 2030, the number of moderately food insecure people will have decreased by \myvar{decprc_2010_2030_mod}\% since 2010, and the number of severely food insecure people will have decreased by \myvar{decprc_2010_2030_sev}\%.
+While the number of moderately and severely food insecure people are modeled to decrease overall, they are projected to increase in lower income countries through 2030.
 
@@
 Outside of Africa, serious pockets of severe food insecurity also occur in Venezuela, Syria, Papua New Guinea, Yemen, and Afghanistan.
-In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in northern Brazil, many central Asian and middle-eastern countries, as well as India and Indonesia.
+In many large middle-income countries, severe food insecurity is also quite prevalent, with rates between 10\% and 15\% in parts of Latin America, many central Asian and middle-eastern countries, as well as India and Indonesia.
 
@@
 Even in highly developed economies, such as the United States and parts of eastern and southern Europe, over 10\% of the population is above the threshold for moderate food insecurity.
+
+Some countries show persistent subnational gaps in rates of food insecurity, typically in areas that are more remote.
+For example, the Somali province of Ethiopia, Sahelo-Saharan Mali, and Amazonian Brazil are all generally more food insecure subnational areas than the more populous and accessible parts of their respective countries.
 
@@
 \subsection{Important Predictors of Food Insecurity}
-Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.940$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.792$ and $R^2 = 0.746$ for moderate and severe food insecurity, respectively).
+Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = \myvar{rf_wthn_mod_rsq}$ and $R^2 = \myvar{rf_wthn_sev_rsq}$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = \myvar{rf_cv_mod_rsq}$ and $R^2 = \myvar{rf_cv_sev_rsq}$ for moderate and severe food insecurity, respectively).
 The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.
@@
 For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig.
-\ref{fig:vimp}), followed by the rate of stunting and the Gini coefficient.
+\ref{fig:vimp}), followed by the Gini coefficient and the rate of stunting.
 For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.
@@
 
-Our results anticipating increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
+It should be noted that our model was using covariates and making predictions a coarser scale - the administrative area - than much of the food security literature, which, since the 1990s, has increasing focused at the scale of the household \citep{Cooper2020}.
+Thus, this analysis cannot speak to the localized impact of factors like climate, urbanization, or topography at the finer spatio-temporal scales at which these variables may operate \citep{Shukla2021, Blekking2020, Cockx2018, Tuholske2020}.
+For example, our predictor variables for temperature and precipitation, derived from future simulations of the climate, can only model the effects of shifting baselines as data is smoothed over the course of the year - our model cannot anticipate the effects of short-term and extreme climate events on food insecurity.
+Additionally, our model can only capture coarse, administrative area metrics of urbanization and cannot address household-level differences in food security between urban and rural households.
+Nevertheless, the fact that these variables explained some of the variation in food security outcomes, even at administrative area scale, underscores their importance for food security.
+
+Our results anticipating decreases in the number of people over the threshold for moderate food insecurity conflict with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
 However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.
-In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 2.6\% decrease, from 787 to 767 million severely food insecure people over the same period.
+In this case, our predictions still diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a \myvar{decrprc_2019_2030_sev}\% decrease, from \myvar{sev_tot_2019} to \myvar{sev_tot_2030} million severely food insecure people over the same period.
 
 These discrepancies are in part due to our differing modeling approaches.
-While the SOFI report extrapolated recent trends in undernourishment and stunting, we modeled food insecurity based on interactions between projections in factors like GDP per capita, population, and the prevalence of stunting.
-Additionally, our diverging predictions for different hunger indicators are not necessarily mutually exclusive: it is theoretically possible for undernourishment to increase while severe food insecurity, as measured by the FIES, decreases.
+Food insecurity was worsening slightly in the latter half of the 2010s, a trend noted elsewhere \cite{Voosen2020}, and reflected in our own analysis (Fig.
+\ref{fig:timeseries}), and the SOFI report extrapolated these recent trends in undernourishment and stunting.
+Our approach, on the other hand, modeled food insecurity based on interactions between projections in underlying drivers like GDP per capita and population change.
+Additionally, our diverging predictions for different hunger indicators are not necessarily mutually exclusive, as we are using different indicators: it is theoretically possible for undernourishment to increase while severe food insecurity, as measured by the FIES, decreases.
 The discrepancies between our projections and those of the FAO illustrate both the complexity of hunger as a social phenomenon, as well as the challenges in predicting future socioeconomic developments, where different modeling approaches may yield different forecasts.
@@
 
-While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate food insecurity to increase and the number of people experiencing severe food insecurity to fall only slightly between 2010 and 2030.
+While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate or severe food insecurity to increase in low income countries and the overall number of people experiencing moderate or severe food insecurity to fall only slightly between 2010 and 2030.
 This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, falling far short of the SDG 2 goal of ending hunger.
@@
 \section{Acknowledgements}
-This reserach was supported by The International Fund for Agricultural Development (IFAD) grant number 2000002044.  We thank Kristofer Hamel and Wolfgang Fengler for their support in this work.
+This research was supported by The International Fund for Agricultural Development (IFAD) grant number 2000002044.  We thank Kristofer Hamel and Wolfgang Fengler for their support in this work.
 

### 2021-07-13 docs/modeling_fies_supplement.tex
finish response to reviewers, add new LASSO analysis

--- 2020-12-10 docs/modeling_fies_supplement.tex
@@
 
+\section{Comparison Modeling Approach: LASSO Regression}
+
+In addition to a random forest model, we also fit a similar model using LASSO (Least Absolute Shrinkage and Selection Operator) regression.
+LASSO is a method of regression that penalizes the size of the coefficients to guard against overfitting, and is therefore more suitable for prediction applications \cite{hastie2009elements}.
+Additionally, the LASSO method can shrink some coefficients to zero, providing an indication of which covariates are not useful.
+The LASSO method recasts multiple regression as a convex optimization, specifically:
+
+{\small
+	\begin{align*}
+		\text{minimize}_{\beta{_0},\beta} \bigg[ \frac{1}{2} \sum\limits_{i=1}^N \Big( y_i - \beta_0 - \sum\limits_{j}x_{ij}\beta_j \Big)^2 + \lambda \sum\limits_{j=1}^p|\beta_j|\bigg] \numberthis \label{eqn:LASSO}
+	\end{align*}
+}%
+
+Where $\beta_j$ is the parameter vector from the regression model [1], including both the $\beta$ and $\gamma$ parameters, and the tuning parameter $\lambda$ moderates the level of penalization and regularization.
+
+Examining the model diagnostics for the LASSO regression and comparing them to those for the random forest models, LASSO regression generally performs worse, especially at higher rates of food insecurity and for severe food insecurity.
+This indicates that a model that can flexibly account for interactions between predictor variables, such as a random forest model, is necessary to more accurately predict food security outcomes.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/out-sample_lasso.pdf}
+  \caption{Out-of-sample fit of the LASSO regression model, where predictions for each country are based on a model that did not include that country in the training data.
+Panel (A) shows the model for moderate food insecurity and panel (B) shows the model for severe food insecurity.}
+  \label{fig:lasso_out-sample}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/in-sample_lasso.pdf}
+  \caption{In-sample fit of the LASSO regression model, where one model was trained on the entire dataset.
+Panel (A) shows the model for moderate food insecurity and panel (B) shows the model for severe food insecurity.}
+  \label{fig:lasso_in-sample}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{img/VIMP_Lasso.pdf}
+  \caption{Coefficients for LASSO regression.  Each coefficient shows the predicted change in the rate of Moderate/Severe food insecurity for a one standard deviation increase in the given variable.}
+  \label{fig:lasso_coefs}
+\end{figure}
+
 \printbibliography

### 2021-07-13 models/model_linear.R

@@ -46,8 +46,8 @@ for (iso3 in unique(moddat$ISO3)){
   moddat$fies.sev.pred.cv[!ix] <- inv.logit(predict(lasso.sev, mm[!ix,])[, 1])
 }
 
-sev.rsq <- cor(moddat$fies.sev, moddat$fies.sev.pred.cv)^2
-mod.rsq <- cor(moddat$fies.mod, moddat$fies.mod.pred.cv)^2
+sev.rsq <- rsq(moddat$fies.sev, moddat$fies.sev.pred.cv)
+mod.rsq <- rsq(moddat$fies.mod, moddat$fies.mod.pred.cv)
 sev.mae <- mean(abs(moddat$fies.sev.pred.cv - moddat$fies.sev))
 mod.mae <- mean(abs(moddat$fies.mod.pred.cv - moddat$fies.mod))
 

### 2021-07-13 models/model_randomforest.R

@@ -58,23 +58,24 @@ for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
       depth <- NULL
     }
     
-    rf.mod <- rfsrc(formula = as.formula(paste("fies.mod.logit", 
-                                               paste(vars, collapse = "+"), sep= "~")),
-                    data = moddat[ix, ],
-                    ntree = 1000, 
-                    mtry = mtry,
-                    nodesize = node,
-                    depth = depth)
-    
+    # rf.mod <- rfsrc(formula = as.formula(paste("fies.mod.logit", 
+    #                                            paste(vars, collapse = "+"), sep= "~")),
+    #                 data = moddat[ix, ],
+    #                 ntree = 1000, 
+    #                 mtry = mtry,
+    #                 nodesize = node,
+    #                 depth = depth)
+   
     rf.sev <- rfsrc(formula = as.formula(paste("fies.sev.logit", 
                                                paste(vars, collapse = "+"), sep= "~")),
                     data = moddat[ix, ],
                     ntree = 1000, 
                     mtry = mtry,
                     nodesize = node,
-                    depth = depth)
+                    depth = depth,
+                    seed=-10)
       
-    moddat$fies.mod.pred.cv[!ix] <- inv.logit(predict(rf.mod, moddat[!ix,])$predicted)
+    # moddat$fies.mod.pred.cv[!ix] <- inv.logit(predict(rf.mod, moddat[!ix,])$predicted)
     moddat$fies.sev.pred.cv[!ix] <- inv.logit(predict(rf.sev, moddat[!ix,])$predicted)
     
   }
@@ -86,8 +87,8 @@ for (i in sample(prm$ix[is.na(prm$sev.rsq)])){
 
 write.csv(prm, 'data/prm.csv', row.names=F)
 
-mod.prm <- prm[which.max(prm$mod.rsq), ]
-sev.prm <- prm[which.max(prm$sev.rsq), ]
+mod.prm <- prm[which.min(prm$mod.mae), ]
+sev.prm <- prm[which.min(prm$sev.mae), ]
 
 system('~/telegram.sh "Done Running Random Forests"')
 
@@ -129,16 +130,18 @@ for (iso3 in unique(moddat$ISO3)){
     depth <- NULL
   }
   
-  rf.sev <- rfsrc(formula = as.formula(paste("fies.mod.logit", 
+  rf.sev <- rfsrc(formula = as.formula(paste("fies.sev.logit", 
[…]

### 2021-07-13 viz/Error_and_VIP.R

@@ -6,6 +6,10 @@ library(cowplot)
 library(gridGraphics)
 options(scipen=100)
 
+rsq <- function(y, yhat){
+  1 - (sum((y - yhat)^2))/sum((y - mean(yhat))^2)
+}
+
 ####################################
 # Variable Importance
 ###################################
@@ -15,12 +19,16 @@ vars <- names(moddat)[!names(moddat) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
                                             'fies.mod.pred', 'fies.sev.pred',
                                             'urban', 'rural', 'fies.sev', 'fies.mod',
                                             'population', 'YEAR', 'rural_perc', 'region',
-                                            'fies.mod.logit', 'fies.sev.logit')]
+                                            'fies.mod.logit', 'fies.sev.logit', 
+                                            'fies.mod.pred.cv', 'fies.sev.pred.cv')]
 
 # change varaibles names
 vars_full <- c("Stunting", "Wasting", "Mean Years of Schooling", "Topographic Ruggedness",
-              "GDP Per Capita", "Gini Coefficient", "Malaria Mortality Rate", "Mean Annual Precipitation",
-              "Poverty Headcount Index", "Mean Temperature", "Urban Percentage", "Water Scarcity") #has to be in the same order as "vars"
+              "GDP Per Capita", "Gini Coefficient", "Malaria Mortality Rate",
+              "Total Annual Precipitation", 
+              "Poverty Headcount Index", "Mean Temperature", 
+              "Total Annual Precipitation\n(Previous Year)", "Mean Temperature\n(Previous Year)",
+              "Urban Percentage", "Water Scarcity") #has to be in the same order as "vars"
 
 v.mod <- vimp(rf.mod, importance='permute')
 v.sev <- vimp(rf.sev, importance='permute')
@@ -31,7 +39,7 @@ v.sev <- vimp(rf.sev, importance='permute')
 
 df <- data.frame(var=c(names(v.mod$importance), names(v.sev$importance)),
                  val=c(v.mod$importance, v.sev$importance),
[…]

### 2021-07-13 viz/Error_and_VIP_LASSO.R

@@ -1,0 +1,138 @@
+setwd('~/wdl-fies');library(ProjectTemplate);load.project()
+setwd('~/wdl-fies/docs/img')
+#setwd('C:/Users/bmuel/Desktop/GitHub/wdl-fies/docs/img')
+
+library(cowplot)
+library(gridGraphics)
+options(scipen=100)
+
+rsq <- function(y, yhat){
+  1 - (sum((y - yhat)^2))/sum((y - mean(yhat))^2)
+}
+
+####################################
+# Variable Coefficients
+###################################
+
+tmp_coeffs <- coef(lasso.mod, s = 'lambda.min')
+sevdf <- data.frame(term = tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1],
+                    estimate = tmp_coeffs@x) %>%
+  mutate(mod='Severe')
+
+tmp_coeffs <- coef(lasso.sev, s = 'lambda.min')
+moddf <- data.frame(term = tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1],
+                    estimate = tmp_coeffs@x) %>%
+  mutate(mod='Moderate')
+
+df <- bind_rows(sevdf, moddf) %>%
+  filter(term != '(Intercept)')
+
+for (t in unique(df$term)){
+  df$estimate[df$term == t] <- df$estimate[df$term == t]*sd(moddat.lasso[ , t])
+}
+
+vars <- names(moddat.lasso)[!names(moddat.lasso) %in% c('ISO3', 'GDLCODE', 'fies.mod.rur',
+                                            'fies.sev.rur', 'fies.mod.urb', 'fies.sev.urb',
+                                            'fies.mod.pred', 'fies.sev.pred',
+                                            'urban', 'rural', 'fies.sev', 'fies.mod',
+                                            'population', 'YEAR', 'rural_perc', 'region',
+                                            'fies.mod.logit', 'fies.sev.logit', 
+                                            'fies.mod.pred.cv', 'fies.sev.pred.cv')]
+
+# change varaibles names
+vars_full <- c("Stunting", "Wasting", "Mean Years of Schooling", "Topographic Ruggedness",
+              "GDP Per Capita", "Gini Coefficient", "Malaria Mortality Rate",
+              "Total Annual Precipitation", 
[…]

### 2021-07-13 viz/make_reference_values.R

@@ -1,0 +1,56 @@
+setwd('~/wdl-fies');library(ProjectTemplate);load.project()
+
+rsq <- function(y, yhat){
+  1 - (sum((y - yhat)^2))/sum((y - mean(yhat))^2)
+}
+
+rf_cv_mod_rsq <- rsq(moddat$fies.mod, moddat$fies.mod.pred.cv)
+rf_cv_mod_mae <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred.cv))
+rf_cv_sev_rsq <- rsq(moddat$fies.sev, moddat$fies.sev.pred.cv)
+rf_cv_sev_mae <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred.cv))
+
+rf_wthn_mod_rsq <- rsq(moddat$fies.mod, moddat$fies.mod.pred)
+rf_wthn_mod_mae <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred))
+rf_wthn_sev_rsq <- rsq(moddat$fies.sev, moddat$fies.sev.pred)
+rf_wthn_sev_mae <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred))
+
+ls_cv_mod_rsq <- rsq(moddat.lasso$fies.mod, moddat.lasso$fies.mod.pred.cv)
+ls_cv_mod_mae <- mean(abs(moddat.lasso$fies.mod - moddat.lasso$fies.mod.pred.cv))
+ls_cv_sev_rsq <- rsq(moddat.lasso$fies.sev, moddat.lasso$fies.sev.pred.cv)
+ls_cv_sev_mae <- mean(abs(moddat.lasso$fies.sev - moddat.lasso$fies.sev.pred.cv))
+
+ls_wthn_mod_rsq <- rsq(moddat.lasso$fies.mod, moddat.lasso$fies.mod.pred)
+ls_wthn_mod_mae <- mean(abs(moddat.lasso$fies.mod - moddat.lasso$fies.mod.pred))
+ls_wthn_sev_rsq <- rsq(moddat.lasso$fies.sev, moddat.lasso$fies.sev.pred)
+ls_wthn_sev_mae <- mean(abs(moddat.lasso$fies.sev - moddat.lasso$fies.sev.pred))
+
+
+glob_2020_sev
+glob_2020_mod
+
+incprc_2010_2030_mod
+incprc_2010_2030_sev
+
[…]

### 2021-07-16 docs/response_to_reviewers.md
resubmission

@@
 created: 2021-06-10T10:04:08-0400
-modified: 2021-07-12T20:10:12-0400
+modified: 2021-07-16T16:31:14-0400
 title: Response to Reviewers
@@
 
-**We thank the reviewer for the detailed and thoughtful feedback, which forced us to engaged more with issues of spatial scale, household vs national food security, as well as a more broad body of literature thn we originally examined.
+**We thank the reviewer for the detailed and thoughtful feedback, which forced us to engaged more with issues of spatial scale, household vs population food security, as well as a more broad body of literature thn we originally examined.
 We believe that the paper is now much improved and will be of interest to a wider body of researchers, with more clear explanations of the methodlogy and spatial scale, and more detailed discussions of issues like population, urbanization, and climate.**
@@
 
-**Given these considerations of how the scale of our data (annual, national scale) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value.
-(Also, the CHIRPS and CHIRTS datasets, while excellent, are not truly global and unavailable for certain latitudes, so couldnt be applied to this global analysis).
+**Given these considerations of how the scale of our data (annual, national sampling) and scope of our analysis (modeling food insecurity globally and out to 2030) makes it infeasible to explore the effect of climate on food insecurity beyond shifting baselines, we do not think that using other datasets like ERA-5 or MERRA-2 would add much value.
+(Also, the CHIRPS dataset, while excellent, is not truly global and unavailable for certain latitudes, so couldnt be applied to this global analysis).
 However, you raise very important considerations in your feedback and we therefore make two improvements to the paper: (1) in addition to the survey year's rainfall totals and precipitation averages, we add a feature for the previous year's total rainfall and mean temperature, and additionally (2) we add a paragraph to the discussion highlighting many of the points you raise and further outlining the considerations that led to the analytical approach we took.**
@@
 However, we note that we do take care to conduct cross validtion at the country level.
-We leave out all surveys from 10% of countries when evaluating our model, and find our model it still performs favorably.**
+We leave out all surveys a given country and then evaluate the model's performance on that country, and find our model it still performs favorably.**
 
-**Additionally, we have updated the paper to now model the FIES using linear regression and lasso regression (in the supplement), and note that, compared to logisitic regression, the random forest model performs quite well.
+**Additionally, we have updated the paper to now model the FIES using LASSO regression (in the supplement), and note that, compared to linear regression, the random forest model performs quite well.
 We also now add some discussion around the importance of accounting for interactions between covariates based on a comparison between random forest and linear regression.**
@@
 Gallup records whether individual respondents are in urban or rural areas based on the individuals own reporting.
-We then use global gridded data from Jones and O'Neill (2016) in _Environmental Research Letters_ to estimate the percent of the population in urban areas for each administrative area through the year 2030.**
+We then use global gridded data from Jones and O'Neill (2016) published in _Environmental Research Letters_ to estimate the percent of the population in urban areas for each administrative area through the year 2030.**
 
@@
 
-**It is true that there is a large body of important literature that focuses on food security at a household scale, and based on your comments throughout the review, it is clear that you are very familiar with this literature.
-However, we disagree that "national-level estimates...
+**First, we share some of your concerns about the inadequacy of national-level food security estimates, which is precisely why we chose to disaggregate our analysis to the admin area level.
+We further note that our data was originally available at the national scale, and we used demographic data to disaggregated it (See section 2.1).
+Disaggregating to the admin 1 area is possible with our data, but a household or individual-level analysis really is not.**
+
+**We also disagree that "national-level estimates...
 are not relevant for policy makers".
-While local practitioners focus on differences in food security between households, many mulitnational practitions pay close attention to country-level (or admin-level) estimates of food security.
+While local practitioners focus on differences in food security between inidivudals or households, many mulitnational practitions pay close attention to country-level (or admin-level) estimates of food security.
 We note the FAO's State of Food Insecurity (SOFI) reports, which are widely read, as well as the fact that the SDG targets, which are key metrics for policymakers and which include the FIES for SDG2, are monitored at the national level (see: https://sdg-tracker.org/zero-hunger).
 We discuss both the SOFI estimates and the relevance to SDG2 throughtout the paper.
-Moreover, as we mention in the acknowledgments, this project began out of a specific demand from an international policymaking organization.**
+Moreover, as we mention in the acknowledgments, this project began out of a specific demand from an international policymaking organization.
+So we feel that our analysis meets a clear stated need by policymakers, and is conducted at the appropriate spatial scale given the data that we have.**
 

### 2021-07-16 docs/test/modeling_fies.tex
resubmission

--- 2021-07-13 docs/modeling_fies.tex
@@
 % csv must have columns `var` and `val`
-\usepackage{datatool}
-\DTLsetseparator{,}
-\DTLloaddb{data}{vars.csv}
-\newcommand{\myvar}[1]{\DTLfetch{data}{var}{#1}{val}}
+% \usepackage{datatool}
+% \DTLsetseparator{,}
+% \DTLloaddb{data}{vars.csv}
+% \newcommand{\myvar}[1]{\DTLfetch{data}{var}{#1}{val}}
 
-\addbibresource{library.bib}
+%\addbibresource{library.bib}
+
+\bibliographystyle{splncs}
 
@@
 Food security has traditionally been difficult to measure, and this has arguably led to an incomplete or possibly inaccurate assessment of global hunger.
-Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.
-However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.
-Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that are food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.
+Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \cite{Puffer1973, Habicht1974}.
+However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \cite{Perumal2018}.
+Other proxies for food security, such as food availability estimated from crop yields \cite{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that are food insecure due to poor access, even when aggregate food availability is high \cite{Sen1983}.
 Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.
@@
 
-As researchers began to focus on food insecurity at the individual and household level, microdata from household surveys collecting information on household finances and consumption became a basis to compute a common proxy for food security \citep{Haddad1994}.
-However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \citep{Maxwell1996}.
-This led to the emergence of several indicators designed to be rapidly deployable, and based on the lived experience of food security \citep{Jones2013}.
-These metrics include the Household Food Security Survey Module (HFSSM) \citep{kennedy2005keynote}, originally developed for use in the US; the Latin American and Caribbean Food Security Scale (ELCSA); and the Household Food Insecurity and Access Scale (HFIAS) \citep{Coates2007}.
+As researchers began to focus on food insecurity at the individual and household level, microdata from household surveys collecting information on household finances and consumption became a basis to compute a common proxy for food security \cite{Haddad1994}.
+However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \cite{Maxwell1996}.
+This led to the emergence of several indicators designed to be rapidly deployable, and based on the lived experience of food security \cite{Jones2013}.
+These metrics include the Household Food Security Survey Module (HFSSM) \cite{kennedy2005keynote}, originally developed for use in the US; the Latin American and Caribbean Food Security Scale (ELCSA); and the Household Food Insecurity and Access Scale (HFIAS) \cite{Coates2007}.
 
-Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \citep{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.
+Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \cite{wambogo2018validity, smith2017world}.
 The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.
 Processing the responses given to the FIES questions through the Rasch model, each individual in the survey is assigned a score that allows them to be ranked on a scale of food insecurity severity.
-Once the severity measures are calibrated against a common reference (the global FIES reference scale) \citep{Cafiero2018} and individuals are ranked in this fashion, conventional thresholds for ``moderate'' and ``severe'' food insecurity are set and the prevalence of such food insecurity conditions can be computed.
+Once the severity measures are calibrated against a common reference (the global FIES reference scale) \cite{Cafiero2018} and individuals are ranked in this fashion, conventional thresholds for ``moderate'' and ``severe'' food insecurity are set and the prevalence of such food insecurity conditions can be computed.
 Thresholds are set in such a way that a moderately food insecure respondent is one who likely has compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally.
@@
 The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.
-Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.
+Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \cite{bethlehem2009applied}.
 This ensures that the results are representative at a national level.
@@
 We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.
-For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates of population in urban and rural pixels published by Jones and O'Neill and deriving from night-time lights and buffered settlement centroids \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.
-This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status, an assumption that underlies much of survey weighting methodology \citep{bethlehem2009applied}.
+For age and gender, we used gridded data from WorldPop \cite{Tatem2017}; for urbanization, we used spatially explicit estimates of population in urban and rural pixels published by Jones and O'Neill and deriving from night-time lights and buffered settlement centroids \cite{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \cite{dhsall}.
+This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status, an assumption that underlies much of survey weighting methodology \cite{bethlehem2009applied}.
 
@@
 We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.
-For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.
+For projections based on the Shared Socioeconomic Pathways (SSP) framework \cite{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \cite{van2011representative}.
 
@@
 
-For the climatic variables, mean annual temperature and total annual precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \citep{warszawski2014inter}.
+For the climatic variables, mean annual temperature and total annual precipitation, we combine historical observations with an ensemble of four bias-corrected simulations of the future climate from the Inter-Sectoral Impact Model Intercomparison Project (ISIMIP) \cite{warszawski2014inter}.
 These variables cannot speak to sudden shocks but do capture the climatic baselines relevant to agro-ecological conditions that constrain food production.
@@
 Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.
-These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \citep{Cuaresma2018}.
+These include June 2020 GDP predictions from the World Bank \cite{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \cite{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.
 
@@
 We fit two models for all subnational areas globally from 2010 to 2030: one for the percent of the population over the threshold for moderate food insecurity, and one for the percent of the population over the threshold for severe food insecurity.
-We use both random forest regressions and LASSO (Least Absolute Shrinkage and Selection Operator) regressions, with the results of the LASSO regressions presented in the supplement.
+We assessed several different modeling strategies and found that a random forest machine learning approach performed the best.
+A comparison of random forest and LASSO (Least Absolute Shrinkage and Selection Operator) regressions are presented in the supplement, and additional approaches are presented in a graduate thesis by one of the authors \cite{muller2021}.
 A random forest regression involves generating a large number of decision trees, each using different sub-samples of the data, and then aggregating the predictions of the decision trees.
-This approach performs well in the presence of non-linearities and interaction effects among predictor variables \citep{hastie2009elements}.
-By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.
+This approach performs well in the presence of non-linearities and interaction effects among predictor variables \cite{hastie2009elements}.
+By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \cite{friedman2001elements}.
 Given the bounded nature of our variables of interest, the random forest regression are applied to the logistically transformed rates.
-We create our models using the \texttt{randomForestSRC} package in R \citep{ishwaran2019randomforestsrc}.
+We create our models using the \texttt{randomForestSRC} package in R \cite{ishwaran2019randomforestsrc}.
 
@@
 Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.
-By this validation metric, we find that our models perform well, with an $R^2$ of \myvar{rf_cv_mod_rsq} and mean average error (MAE) of \myvar{rf_cv_mod_mae} for moderate food insecurity and an $R^2$ of \myvar{rf_cv_sev_rsq} and a MAE of \myvar{rf_cv_sev_mae} for severe food insecurity.
-After fitting models on the entire dataset, we observe a within-sample $R^2$ of \myvar{rf_wthn_mod_rsq} and an MAE of \myvar{rf_wthn_mod_mae} for moderate food insecurity and an $R^2$ of \myvar{rf_wthn_sev_rsq} and a MAE of \myvar{rf_wthn_sev_mae} for severe food insecurity.
+By this validation metric, we find that our models perform well, with an $R^2$ of 0.787 and mean average error (MAE) of 0.07 for moderate food insecurity and an $R^2$ of 0.756 and a MAE of 0.037 for severe food insecurity.
+After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.936 and an MAE of 0.038 for moderate food insecurity and an $R^2$ of 0.99 and a MAE of 0.007 for severe food insecurity.
 
 After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity.
-We use a common method of permutating each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.
+We use a common method of permutating each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \cite{ishwaran2007variable, breiman2001random}.
 This metric can provide some interpretability to model predictions.
@@
 \subsection{Food Insecurity over Time}
-Globally, in the year 2020, we estimate that \myvar{glob_2020_sev} million people are above the threshold for severe food insecurity, and \myvar{glob_2020_mod} billion people are above the threshold for moderate food insecurity (See Fig.
+Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and 2.5 billion people are above the threshold for moderate food insecurity (See Fig.
 \ref{fig:timeseries}).
-These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity (SOFI) report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \citep{sofi2020}.
+These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity (SOFI) report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \cite{sofi2020}.
 
 Under realistic scenario assumptions, we find recent increases in the number of moderate and severely food insecure people around the world, yet expect overall decreases from 2010 to 2030.
-These changes are more slight for people who are at least moderately food insecure: in 2030, the number of moderately food insecure people will have decreased by \myvar{decprc_2010_2030_mod}\% since 2010, and the number of severely food insecure people will have decreased by \myvar{decprc_2010_2030_sev}\%.
+These changes are more slight for people who are at least moderately food insecure: in 2030, the number of moderately food insecure people will have decreased by 2.6\% since 2010, and the number of severely food insecure people will have decreased by 12.4\%.
 While the number of moderately and severely food insecure people are modeled to decrease overall, they are projected to increase in lower income countries through 2030.
@@
   \centering
-  \includegraphics[width=\linewidth]{TimeSeries.pdf}
+  % \includegraphics[width=\linewidth]{TimeSeries.pdf}
   \caption{Number of people over the thresholds for moderate and severe food insecurity, by World Bank income groups as of 2020.
@@
   \centering
-  \includegraphics[width=\linewidth]{Rates.pdf}
+  % \includegraphics[width=\linewidth]{Rates.pdf}
   \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020.}
@@
   \centering
-  \includegraphics[width=\linewidth]{FullMap.pdf}
+  % \includegraphics[width=\linewidth]{FullMap.pdf}
   \caption{Spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
@@
 \subsection{Important Predictors of Food Insecurity}
-Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = \myvar{rf_wthn_mod_rsq}$ and $R^2 = \myvar{rf_wthn_sev_rsq}$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = \myvar{rf_cv_mod_rsq}$ and $R^2 = \myvar{rf_cv_sev_rsq}$ for moderate and severe food insecurity, respectively).
+Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.936$ and $R^2 = 0.99$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.787$ and $R^2 = 0.756$ for moderate and severe food insecurity, respectively).
 The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=\linewidth]{VIMP.pdf}
+  % \includegraphics[width=\linewidth]{VIMP.pdf}
   \caption{Importance of each variable in predicting the percent of a population over thresholds for moderate and severe food insecurity.
@@
 
-The results of this analysis can be further explored in more detail on the World Hunger Clock \href{https://worldhunger.io}{(worldhunger.io)}, including statistics for each subnational administrative area.
+The results of this analysis can be further explored in more detail on the World Hunger Clock (\href{https://worldhunger.io}{worldhunger.io}), including statistics for each subnational administrative area.
 
@@
 
-It should be noted that our model was using covariates and making predictions a coarser scale - the administrative area - than much of the food security literature, which, since the 1990s, has increasing focused at the scale of the household \citep{Cooper2020}.
-Thus, this analysis cannot speak to the localized impact of factors like climate, urbanization, or topography at the finer spatio-temporal scales at which these variables may operate \citep{Shukla2021, Blekking2020, Cockx2018, Tuholske2020}.
+It should be noted that our model was using covariates and making predictions a coarser scale - the administrative area - than much of the food security literature, which, since the 1990s, has increasing focused at the scale of the household \cite{Cooper2020}.
+Thus, this analysis cannot speak to the localized impact of factors like climate, urbanization, or topography at the finer spatio-temporal scales at which these variables may operate \cite{Shukla2021, Blekking2020, Cockx2018, Tuholske2020}.
 For example, our predictor variables for temperature and precipitation, derived from future simulations of the climate, can only model the effects of shifting baselines as data is smoothed over the course of the year - our model cannot anticipate the effects of short-term and extreme climate events on food insecurity.
@@
 
-Our results anticipating decreases in the number of people over the threshold for moderate food insecurity conflict with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.
-However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.
-In this case, our predictions still diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a \myvar{decrprc_2019_2030_sev}\% decrease, from \myvar{sev_tot_2019} to \myvar{sev_tot_2030} million severely food insecure people over the same period.
+Our results anticipating decreases in the number of people over the threshold for moderate food insecurity conflict with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \cite[Part 1, Page 11]{sofi2020}.
+However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \cite{sofi2020}.
+In this case, our predictions still diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 9.4\% decrease, from 790 to 715 million severely food insecure people over the same period.
 
@@
 
-\printbibliography
+%\printbibliography
+
+\bibliography{library}
 

### 2021-07-16 viz/make_reference_values.R

@@ -4,53 +4,98 @@ rsq <- function(y, yhat){
   1 - (sum((y - yhat)^2))/sum((y - mean(yhat))^2)
 }
 
-rf_cv_mod_rsq <- rsq(moddat$fies.mod, moddat$fies.mod.pred.cv)
-rf_cv_mod_mae <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred.cv))
-rf_cv_sev_rsq <- rsq(moddat$fies.sev, moddat$fies.sev.pred.cv)
-rf_cv_sev_mae <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred.cv))
+rf_cv_mod_rsq <- round(rsq(moddat$fies.mod, moddat$fies.mod.pred.cv), 3)
+rf_cv_mod_mae <- round(mean(abs(moddat$fies.mod - moddat$fies.mod.pred.cv)), 3)
+rf_cv_sev_rsq <- round(rsq(moddat$fies.sev, moddat$fies.sev.pred.cv), 3)
+rf_cv_sev_mae <- round(mean(abs(moddat$fies.sev - moddat$fies.sev.pred.cv)), 3)
 
-rf_wthn_mod_rsq <- rsq(moddat$fies.mod, moddat$fies.mod.pred)
-rf_wthn_mod_mae <- mean(abs(moddat$fies.mod - moddat$fies.mod.pred))
-rf_wthn_sev_rsq <- rsq(moddat$fies.sev, moddat$fies.sev.pred)
-rf_wthn_sev_mae <- mean(abs(moddat$fies.sev - moddat$fies.sev.pred))
+rf_wthn_mod_rsq <- round(rsq(moddat$fies.mod, moddat$fies.mod.pred), 3)
+rf_wthn_mod_mae <- round(mean(abs(moddat$fies.mod - moddat$fies.mod.pred)), 3)
+rf_wthn_sev_rsq <- round(rsq(moddat$fies.sev, moddat$fies.sev.pred), 3)
+rf_wthn_sev_mae <- round(mean(abs(moddat$fies.sev - moddat$fies.sev.pred)), 3)
 
-ls_cv_mod_rsq <- rsq(moddat.lasso$fies.mod, moddat.lasso$fies.mod.pred.cv)
[…]
